### 2019-11-17 README.Rmd
trying to commit to github

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below. 
Title <- "Title of your paper goes here"
```

# koreapaleolithicmobilityoccupation

This repository contains the data and code for our paper:

> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to download or install

You can download the compendium as a zip from from this URL: <https://github.com/parkgayoung/koreapaleolithicmobilityoccupation/archive/master.zip>

Or you can install this compendium as an R package, koreapaleolithicmobilityoccupation, from GitHub with:
```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("parkgayoung/koreapaleolithicmobilityoccupation")
```

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2019-11-17 analysis/paper/paper.Rmd
trying to commit to github

---
title: "Change in Mobility and Site Occupation during the Late Pleistocene in Korea"
author:
  - Author Gayoung Park
  - Author Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  According to Human Behavioral Ecology, stone artifacts can be an essential source of information for understanding hunter-gatherer subsistence. We examine stone artifacts from 22 sites in Korea to investigate a change in mobility and site occupation during the Late Pleistocene. The new technology is represented by the technological change, the appearance of stemmed points, and the following blade industry. The overall aim is to understand how the appearance of stemmed points affected hunter-gatherers in Korea, which leads to three specific questions to pursue: how did the introduction of the new tools change the way people used the landscape, the way people used sites, and the population. Based on the hypothesis that hunter-gatherers should move frequently and further in order to adapt to a changing environment, so their tools became more portable and efficient, as indicated by decreases in toolkit size and increases in function and utility, we analyze artifact density, retouch frequency, radiocarbon dates, elevation of the sites, raw material types, and composition of the toolkit. The results show that the new technology enabled the occupation of marginal environments at higher elevations. There is no strong general relationship between technology and ecological change. Instead, we assume that population dynamics might have had more impact on changes in stone artifact technology and hunter-gatherer subsistence. 
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

```{r, setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/"
)

#library(koreapaleolithicmobilityoccupation)# Or use devtools::load_all('.', quiet = T) if your code is in script files, rather than as functions in the `/R` diretory
```

# Introduction

Here is a citation [@Marwick2017]

Stone artifact assemblages have important information about hunter-gatherer mobility and subsistence strategy, according to the behavioral ecological theory that links technological changes to human adaptation to the surrounding environment. Our research focuses on technological change during the Late Pleistocene in Korea, around ~40-35 ka. We examined stone artifacts from 22 sites in South Korea to investigate the technological transition represented by the appearance of a new tool, stemmed points, and the following blade industry. The overall aim is to understand how the appearance of stemmed points affected hunter-gatherers in Korea, which leads to three specific questions to pursue: how did the introduction of the new tools change the way people used the landscape, the way people used sites, and the population. This technological change has been studied related to the issues of the transition from the Early to the Late Paleolithic and the migration of Homo sapiens. However, the studies rarely consider the relationship between the technological transition and change in human subsistence strategies. The most Korean Paleolithic sites do not preserve much information to understand hunter-gatherers or the transition in their tool kit due to a lack of material evidence. Therefore, we apply theoretical hypotheses to understand the relationship between hunter-gatherers, technology, and the environment. Using Human Behavioral Ecology (HBE) driven models, we hypothesized that hunter-gatherers should move frequently and further in order to adapt to a changing environment, so their tools became more portable and efficient, as indicated by decreases in the size of both individual tools and toolkit and increases in function and utility.

# Background

```{r demo, eval = FALSE}
1 + 1
```

Theory and background to our three questions
New technologies have led people to adjust the speed and frequency of mobility, explore the new landscape, or even manage the population size of the foraging group or overall population occupied one region. Previous studies show that new technologies play an important role in the change in the use of the landscape. For instance, Morisaki et al. argue that the transition of the blades and points becoming dominant over trapezoid in toolkits around 25 ka cal BP in north Paleo-Honshu Island enabled the foragers to extend their occupation at residential sites (Morisaki et al. 2015). In Australia during the mid-Holocene, the highly retouched and finely shaped tools with shafts appeared to reduce risks of being in a different environment so that they could broaden their occupations to offshore islands on the northeast and northwest coast (Hiscock 1994). New tools tend to appear as a risk reduction strategy when they confront an unfamiliar environment mainly caused by climate change, migration to previously unoccupied landscapes, based on the understanding of resource distribution, tool manufacturing process, and the characteristics of the foraging group. The appearance of microliths as portable and standardized tools is a great example to show that people made new tools to increase multi-functionality and tool readiness so that they can minimize the economic risk and/or scheduling difficulties (Hiscock et al. 2011). 
The rapid spread of microblade industry in north China among Last Glacial Maximum demonstrates that making new tools was one of the human adaptation to extremely cold environment by increasing mobility or meeting the new demands such as the production of sophisticated cold-weather clothing (Barton et al. 2007; Yi et al. 2013; Qu et al. 2013; Nian et al. 2014).

The explanation of the emergence of new tools as an aid of buffering risks, as a result of environmental change, is one of the main concepts of Human Behavioral Ecology (HBE).  HBE focuses on understanding patterned human behaviors, such as the ways they evaluate choices in food procurement, mobility, division of labor, task-group membership, social affiliation, prestige, and social subordination, to selective pressures in the environment (Smith and Winterhalder 1992; Smith 2000; Fitzhugh 2001; Bird and O’Connell 2006; Fitzhugh et al. 2018; Veile 2018). Thus, most HBE models investigate environmental variables (e.g., resource density, competitor frequency) and behaviors reflecting decisions that people made about how to allocate time and energy to their activities (Smith 2000; Bird et al. 2006; Kelly 2007; Creanza et al. 2017; Fitzhugh et al. 2018). HBE provides concepts to identify how foragers adjust their technologies, resource management strategies, schedule movements, foraging, food processing (including storage and consumption), and investment in technologies to optimize time and energy (Kelly 2007). Stone artifacts have also been used in HBE models to understand hunter-gatherers’ mobility and site occupation patterns (Marwick 2013; Clarkson 2013; Capriles et al. 2018). For example, Clark and Barton (2017) examined to investigate resource acquisition strategies of hunter-gatherers by means of stone artifact density and the frequency of retouched pieces (scaled to the volume of excavated sediment) as proxies to represent occupation patterns such as ‘logistically organized basecamp’ and ‘short-term camp’. We use HBE as a reference framework to answer the three questions: (1) how did the introduction of the stemmed points and blade industry change the way people used the landscape? (2) was there any change in the way people used sites? (3) was there any population change and did it impact on technology or subsistence pattern?

Regional Archaeology

Northeast Asia/similar studies and previous research in the adjacent regions : General description of EUP in East Asia
The technological change during the Late Pleistocene in Northeast Asia has been actively studied related to the topic of modern human dispersal to the East and chronological transition from Middle to Upper Paleolithic [or Early to Late Paleolithic. In Northeast Asia, the term and concept of ‘Late Paleolithic ’ is widely used rather than ‘Upper Paleolithic’ related to the absence of Middle Paleolithic industry such as Levallois (Gao and Norton 2002; Bar-Yosef 2002; Mellars et al. 2007; Norton et al. 2009; Lee 2013)]. Along with the modern human dispersal, this transition often comes with the appearance of blade and micro technology, high frequencies of retouched blade tools, several novel refined tools such as projectile points, end scrapers, burins, denticulates, etc, and using high-quality raw materials (Bar-Yosef and Kuhn 1999; Brantingham 2001; Bar-Yosef 2002; Bae et al. 2018).

In China, the transition from the Middle to the Upper Paleolithic dates to 35-30 ka appeared with a few body decorations, bone tools, core-and-flake in North and cobble tools in South. In addition to the technological changes, the use of symbols in the body decorations reflects social and cultural changes occurred during the onset of the Chinese Upper Paleolithic (Qu et al. 2013; Peng et al. 2018). At the Shuidonggou site in northern China which is considered as a typical site for showing the transition from cobble to blade tool industries in Northeast Asia (Brantingham et al. 2001; Gao et al. 2010; Pei et al. 2012), newly discovered localities (SDG 2, 7-12) show various aspects of the initial Late Paleolithic from 32,000 BP to 6000 BP. Some sites (SDG 9, 12) have blades and bladelets made of higher quality or exotic raw materials in addition to more than 80 polished ostrich egg-shell beads (SDG 2), bone needles and an awl (SDG 12), while other sites have mostly small, irregular flakes (Gao et al. 2010). The transition to the Upper Paleolihtic in Japan is under the long term debate due to the uncertainty about the existence of the previous period in the Archipelago. But there are several features represent the milestone of the new era such as systematic procurement and reduction of lithic raw materials, trap-pit hunting method, watercraft (maritime transport), bow-and-arrow technology, and standardization of lithic tools which can show the appearance of new subsistence strategies possibly driven by risks of increased population and ecological change (One et al. 2002; Yamaoka; Bae 2017; Nakazawa and Bae 2018; Morisaki et al. 2019). The technological transition in Korea is mainly characterized by the early appearance of stemmed points and following blade technology. Using the blade as a blank, the stemmed points became more standardized and increased in number. The flake tools became more sophisticated, with a greater diversity in both type and size along with continuously used core tools during the Upper Paleolithic (Bae 2010; Lee 2015).

In this research, we choose to focus on the technological transition and its impact on mobility and site occupation pattern in Korea because of the uniqueness of sudden and earliest appearance of stemmed points among the broad area of Northeast Asia during the Late Pleistocene. The oldest stemmed points so far is from the Yongho-dong site dated to 38.5 ka (Seong 2009; Park 2012; Seong 2015; Bae 2017). The stemmed point is a projectile point made out of elongated flake or blade and with gentle retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft. Combining with the blade industry, the shape became more standardized with one or two ridges on dorsal side and triangular horizontal cross section. The stemmed point is often called as a tanged point, but we prefer to use stemmed point to distinguish with other stone projectile points of Bronze age in Korea, also called as a tanged point in English literature. The early appearance of the stemmed points in the Korean Peninsula brings about several issues to consider such as issues of human migration or advanced cognitive ability to make a new tool adapting to the surrounding environment. As a part of composite implements such as arrowheads or spearheads, the point was combined with other materials and it could be easily replaced reducing waste of total disposal (Keeley 1982; Kuhn and Miller 2015). The new projectile point also indicates a new way of hunting methods or other subsistence strategies because the throwing technique enables to earn a long-distance target. In addition, this new tool spread to adjacent regions such as Japanese Archipelago by combination of watercraft technique and social network so that the study of stemmed points play an important role in understanding about peopling in Japan and the relationship between the Peninsula and Archipelago. This project focuses on the functional aspect to understand the technology and subsistence pattern based on the concept of HBE.

# Methods

Sites and dates

The data analyzed in this paper are gathered from published excavation reports of 22 Paleolithic sites (33 cultural layers) in South Korea dating from 49k BP to 24k BP. After the first excavation at Jeongokri site in the late 1970s, more than 200 Paleolithic sites are discovered in Korea. Among them, we  chose 22 sites based on the existence of absolute dates that match the targeting timeline and officially published excavation report. We include sites dated before and after the transition period (40-35ka) to analyze the process of technological transition.

Assemblages
The research analyzes total 33 assemblages and they consist of cores, blade cores, blades, flakes, debris, hammers, choppers, plans, polyhedrals, side scrapers, end scrapers, notches, cleavers, stemmed points, awls, denticulates, burins, handaxes, blanks, knives, handadzes, peaks, flatters, flakers etc, 
Some artifacts are not included as certain tools due to their vague or broken shape presumably related to the characteristics of raw materials such as quartz, quartz vein, and quartzite. In that case, we named them as unknown, unfinished, or pebble. The stone artifacts from all of these sites are stored in museums located throughout South Korea, and have been briefly described in the excavation reports. We have visited most of these collections, collected PDFs of the excavation reports, and negotiated permission to study the materials. For example, the Hwadae-ri site has three cultural layers. The lowest horizon at the bottom, dated to 39,000±1400 BP by OSL, contains coarse flake tools made of vein quartz and quartzite. The middle cultural horizon has stemmed points made of porphyry, with quartz and quartzite dominating the assemblage. This layer dated to 31,200±900 BP by radiocarbon dating and 30,000±1,700 BP by OSL. The uppermost layer, dated to 22,000±100 BC by OSL, contains blades, scrapers, awls, and denticulates (Seong 2009). We compare the assemblages from the two older layers at this and other sites to investigate the transition process.

Statistical methods 
To answer the three research questions, we gather information such as elevation of the sites, artifact density, retouch frequency, radiocarbon dates toolkit composition, and raw material type. The elevation is the main proxy for environmental information in this research. The Korean paleoenvironmental data and information are not rich enough to directly apply HBE driven models or questions (Han 2008; Chang 2013). Rather, the paleoenvironment of the Korean Peninsula for this proposed research time period can be simply summarized as ‘overall cooler and drier during MIS 3’ (Han 2008; Seong 2008; Choi 2011; Chang 2013; Im and Choo 2015; Bak and Lee 2017). In this regard, it is much harder to understand the regional as well as shorter-time scale of paleoenvironmental change in the Korean Peninsula. In addition, the targeting period is not exactly matched with major environmental changes that we can check through global sea level change, it seems rather monotonous around 40-35 ka (Bailey 2010; Lorscheid 2017). In this research, we chose elevation to make a premise that the site elevation can represent different environment in local scale and hunter-gatherers moved to different elevation as a result of environment related risk. This research focuses on the change of the elevation of the sites throughout time and accompanying toolkits, and its relationship with the appearance of new tools.

The artifact volumetric density is defined as the total number of artifacts per cubic meter of excavated sediment, and served as a proxy for accumulation ratio of artifacts (Clark and Barton 2017; Clark et al. 2018). We expected to use this value to evaluate mobility and land-use practices of hunter-gatherers during the Late Pleistocene. We count the retouch frequency as the number of retouched tools divided by the total number of artifacts. This value represents general proficiency of lithic manufacturing, raw material availability, and subsistence strategies including mobility and duration of site occupation based on the concept that tool reductioning is an economical tactic to generate sharp, usable edges while minimizing the cost of transporting tools or bulky raw materials by reducing the total load weight (Kuhn 1990; Clark et al. 2018). In other words, producing and carrying highly retouched tools will increase portability and efficiency by decrease in the size of both individual tools and the assemblages so that hunter-gatherers could move often or further, or both (Kuhn 1991,1994, 2004; Andrefsky 1994; Davies et al. 2018). For example, highly retouched and finely shaped implements appeared in Australia during the mid-Holocene to minimize risk associated with environmental change and high mobility (Hiscock 1994).
Combined the two proxies, the artifact volumetric density and the retouch frequency, we expect to determine whether or not the hunter-gatherers were residentially mobile or logistically organized. If the assemblage in a site has high density and less retouched, it is depicted as expedient assemblage that represents “base camps” or “residences”, while less density with high retouched one shows curated assemblage and indicates either residential mobility or certain task groups from main residential site (Riel-Salvatore and Barton 2004; Clark and Barton 2017).

Radiocarbon dates provide a proxy for change in human population levels. Summed probability distributions (SPD) of radiocarbon dates are often used to understand temporal trends in ancient human populations (Rick 1987; Bamforth and Grund 2012; Shennan et al. 2013; Timpson et al. 2014; Contreras and Meadows 2014). Without there must be unavoidable limitations for the validity of SPD such as the lack of excavated sites in certain period, this method is still helpful to be used as a proxy for population fluctuation. Timpson et al. demonstrate its usefulness by testing data sets in 12 European regions and assert that this SPD is extremely robust to deal with small sample size (Timpson et al. 2014). We use 171 radiocarbon dates from 22 sites to search for population trends during the Late Pleistocene in Korea based on the premise that population change or pressure could be another catalyst to improve or innovate technology.

In terms of the composition of toolkits, we count kind of tools as well as the amount of debitabitage to search for the change in the toolkits around the transition period. The tools represent paleo hunter-gatherers’ adaptive strategies. We analyze the composition of raw materials in the assemblages to see the relationships between the choice of specific raw materials and technological innovation as well as raw material availability and its consuming pattern based on an assumption that stone raw materials were carefully transported and consumed to optimize mobility and technological strategies, depth of planning in landscape use and risk minimization (Bratingham 2003).

Umap: <- depending on whether we are gonna add umap (or PCA) analysis result

# Results

```{r get-data, eval = FALSE}
# Note the path that we need to use to access our data files when rendering this document
my_data <- read.csv(here::here('analysis', 'data', 'raw_data', 'my_csv_file.csv'))
```

Elevation of occupied sites
The result of the elevation change of 33 cultural layers from 22 sites corresponding to the targeting period shows that there is no remarkable change can be seen in site elevation during the transition (Fig.1: with no sp site distinction). We applied the existence of the stemmed points to the elevation to understand the benefit of having the points in the toolkit and the result indicates that the hunter-gathering groups who carried the technology could occupy higher altitude while the groups with no points stayed at the lower altitude (Fig. 2: distinction between sites with no stemmed points and with stemmed points).

Artifact density and Composition of toolkit
During the transitional period, the artifact volumetric density was relatively small but it becomes higher around 30 ka (Fig. 3: the transition of artifact volumetric density). When it comes to the composition of the toolkit, it is hardly recognized any pattern of changes over time (Fig. 4: the composition of toolkit). Both results with no information about raw material availability near the 22 sites, we assume that the artifact density might relate to the duration of site occupation or group size rather than portability and efficiency of tools. In other words, the small artifact density might indicate shorter duration or smaller group size. Likewise, the composition of toolkit might represent the characteristics or purpose of each site, rather than showing general pattern of chronological change.

Artifact density and retouch <- maybe we need to add the result of retouch, separately 
We use the combination of the artifact density and retouch frequency as a proxy to identify the mobility and residential pattern of hunter-gatherers. The result shows that there is no clear pattern during the transition period and then it became more expedient at the end of the transition (Fig 5. The combination of artifact density and retouch). That is to say, the hunter-gatherer became either more residential or the majority of the site that this research studies were used as base camps or residence rather than shortly occupied camps for certain tasks.

Summed probability distribution
To understand the population change during the appearance of stemmed points, we analyzed the summed probability distributions of 171 radiocarbon dates from 33 cultural layers and it indicates that there is a higher density of radiocarbon dates from around 34 ka. We could cautiously assume that the population flux might relate to the technological innovation.

Raw materials
During the transition period, the use of shale and rhyolite greatly increased. These two raw materials are suitable to make elongate blades and stemmed points. It is considered that he hunter-gatherer started to use raw materials selectively depending on the type of tool, especially the newly produced stemmed points. Raw material availability is a crucial factor in determining the trip and direction for hunter-gatherers (Davies et al. 2018). Therefore the adoption of new raw material along with the emergence of new type of tools demonstrates a significant innovation in not just technology but also cognitive system.

Umap

# Discussion (still working on)

The new technology and occupation in high elevation
The new technology appeared around 40-35ka in the Korean Peninsula includes stemmed points and following blade industry. The stemmed points brought new paradigms in stone artifact assemblages. For instance, the concept of composite tool contributed to increase efficiency of maintaining the toolkit by easily replacing the damaged part (Kuhn and Miller 2015). Combining with other materials increases utility or functionality of the tool by adding extra weight and providing a shaft, which makes it earlier to grip. Selective use of raw material might impact on hunter-gathers’ mobility and occupation pattern depending on availability of the specific raw materials such as shale and rhyolite. In addition, this research shows that carrying the points enabled hunter-gatherers to occupy higher elevation compared to the sites with no stemmed points. In other words, the new technology might play an important role to survive in environments with more risks.

Population change
Interpreted in HBE, we assume there might be close connection between environmental change and the appearance of new technology based on the prediction that the new invention was the result of risk. The research suggests that the environmental pressure was related to population change. We cannot simply predicate that the increased population around 35ka was due to dispersal of Homo Sapiens since there is no reliable human remains in South Korea. However, the migration of Homo sapiens was not impossible scenario at the same time. Bae (2010) argues that the emergence of stemmed points and blade industry came from outside of the Korean peninsula. He proposes that the new technology was the result of the continuous influx of modern human populations from Siberia while the earlier coarse flake tradition ----

# Conclusion (still working on)

This research aims to understand how the appearance of new technology affected hunter-gatherers in Korea and answer the three questions, how did the introduction of the new tools change the way people used the landscape, the way people used sites, and the population. To answer the question, we use HBE as a theoretical backgrounds and analyze stone artifacts from 22 sites dated before and after 40-35ka. 
The new technology is represented the stemmed points and other blade industry made out of new raw materials such as shale and rhyolite. We could clearly see that hunter-gatherers chose these materials particularly to make the newly added stone artifacts to the toolkit. Besides of technological change, we found that there are more varieties in the type of sites or groups during the transition period, both curated and expedient and then it becomes more expedient. We expected to see any pattern that would explain the relationship between environmental change and the emergence of the new technology. However, there was not enough environmental information as well as data to apply to verify the model. Therefore we focus on Radiocarbon dates as a proxy of population change based on the premise that population driven pressure could trigger people to invent a new tool. The results show that population dynamics might have had more impact on changes in stone artifact technology. ---

# Acknowledgements

<!-- The following line inserts a page break when the output is MS Word. For page breaks in PDF, use \newpage on its own line.  -->
##### pagebreak

# References 
<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
<div id="refs"></div>

##### pagebreak

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
git2r::repository(here::here())
```

### 2019-11-18 analysis/paper/paper.Rmd
I put !

@@
 author:
-  - Author Gayoung Park
+  - Author Gayoung Park!
   - Author Ben Marwick

### 2019-11-18 analysis/paper/paper.Rmd

@@
   - Author Gayoung Park!
-  - Author Ben Marwick
+  - Author Ben Marwick!
 date: "`r format(Sys.time(), '%d %B, %Y')`"

### 2019-11-20 analysis/paper/paper.Rmd
more!
I'm working on reference now

@@
   - Author Gayoung Park!
-  - Author Ben Marwick!
+  - Author Ben Marwick!!!!!!!
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 
-Here is a citation [@Marwick2017]
+Here is a citation
 
@@
 For instance, Morisaki et al.
-argue that the transition of the blades and points becoming dominant over trapezoid in toolkits around 25 ka cal BP in north Paleo-Honshu Island enabled the foragers to extend their occupation at residential sites (Morisaki et al.
-2015).
-In Australia during the mid-Holocene, the highly retouched and finely shaped tools with shafts appeared to reduce risks of being in a different environment so that they could broaden their occupations to offshore islands on the northeast and northwest coast (Hiscock 1994).
+argue that the transition of the blades and points becoming dominant over trapezoid in toolkits around 25 ka cal BP in north Paleo-Honshu Island enabled the foragers to extend their occupation at residential sites [@Morisaki_Izuho_Terry_Sato_2015].
+In Australia during the mid-Holocene, the highly retouched and finely shaped tools with shafts appeared to reduce risks of being in a different environment so that they could broaden their occupations to offshore islands on the northeast and northwest coast [@Hiscock_1994].
 New tools tend to appear as a risk reduction strategy when they confront an unfamiliar environment mainly caused by climate change, migration to previously unoccupied landscapes, based on the understanding of resource distribution, tool manufacturing process, and the characteristics of the foraging group.
-The appearance of microliths as portable and standardized tools is a great example to show that people made new tools to increase multi-functionality and tool readiness so that they can minimize the economic risk and/or scheduling difficulties (Hiscock et al.
-2011).
-The rapid spread of microblade industry in north China among Last Glacial Maximum demonstrates that making new tools was one of the human adaptation to extremely cold environment by increasing mobility or meeting the new demands such as the production of sophisticated cold-weather clothing (Barton et al.
-2007; Yi et al.
-2013; Qu et al.
-2013; Nian et al.
-2014).
+The appearance of microliths as portable and standardized tools is a great example to show that people made new tools to increase multi-functionality and tool readiness so that they can minimize the economic risk and/or scheduling difficulties [@Hiscock_Clarkson_Mackay_2011].
+The rapid spread of microblade industry in north China among Last Glacial Maximum demonstrates that making new tools was one of the human adaptation to extremely cold environment by increasing mobility or meeting the new demands such as the production of sophisticated cold-weather clothing [@Barton_Brantingham_Ji_2007; @yi2013microblade; @Qu_Bar-Yosef_Wang_Wu_2013; @nian2014chronology].
 
 The explanation of the emergence of new tools as an aid of buffering risks, as a result of environmental change, is one of the main concepts of Human Behavioral Ecology (HBE).
-HBE focuses on understanding patterned human behaviors, such as the ways they evaluate choices in food procurement, mobility, division of labor, task-group membership, social affiliation, prestige, and social subordination, to selective pressures in the environment (Smith and Winterhalder 1992; Smith 2000; Fitzhugh 2001; Bird and O’Connell 2006; Fitzhugh et al.
-2018; Veile 2018).
-Thus, most HBE models investigate environmental variables (e.g., resource density, competitor frequency) and behaviors reflecting decisions that people made about how to allocate time and energy to their activities (Smith 2000; Bird et al.
-2006; Kelly 2007; Creanza et al.
-2017; Fitzhugh et al.
-2018).
-HBE provides concepts to identify how foragers adjust their technologies, resource management strategies, schedule movements, foraging, food processing (including storage and consumption), and investment in technologies to optimize time and energy (Kelly 2007).
-Stone artifacts have also been used in HBE models to understand hunter-gatherers’ mobility and site occupation patterns (Marwick 2013; Clarkson 2013; Capriles et al.
-2018).
+HBE focuses on understanding patterned human behaviors, such as the ways they evaluate choices in food procurement, mobility, division of labor, task-group membership, social affiliation, prestige, and social subordination, to selective pressures in the environment [@smith2017natural; @smith2017three; @fitzhugh2001risk; @bird2006behavioral; @fitzhugh2019human; @Veile_2018].
+Thus, most HBE models investigate environmental variables (e.g., resource density, competitor frequency) and behaviors reflecting decisions that people made about how to allocate time and energy to their activities [@smith2017three; @bird2006behavioral; @kelly1995foraging; @Creanza_Kolodny_Feldman_2017; @fitzhugh2019human].
+HBE provides concepts to identify how foragers adjust their technologies, resource management strategies, schedule movements, foraging, food processing (including storage and consumption), and investment in technologies to optimize time and energy [@kelly1995foraging].
+Stone artifacts have also been used in HBE models to understand hunter-gatherers’ mobility and site occupation patterns [@Marwick_2013; @Clarkson_2013; @capriles2018mobility].
 For example, Clark and Barton (2017) examined to investigate resource acquisition strategies of hunter-gatherers by means of stone artifact density and the frequency of retouched pieces (scaled to the volume of excavated sediment) as proxies to represent occupation patterns such as ‘logistically organized basecamp’ and ‘short-term camp’.
@@
 The technological change during the Late Pleistocene in Northeast Asia has been actively studied related to the topic of modern human dispersal to the East and chronological transition from Middle to Upper Paleolithic [or Early to Late Paleolithic.
-In Northeast Asia, the term and concept of ‘Late Paleolithic ’ is widely used rather than ‘Upper Paleolithic’ related to the absence of Middle Paleolithic industry such as Levallois (Gao and Norton 2002; Bar-Yosef 2002; Mellars et al.
-2007; Norton et al.
-2009; Lee 2013)].
-Along with the modern human dispersal, this transition often comes with the appearance of blade and micro technology, high frequencies of retouched blade tools, several novel refined tools such as projectile points, end scrapers, burins, denticulates, etc, and using high-quality raw materials (Bar-Yosef and Kuhn 1999; Brantingham 2001; Bar-Yosef 2002; Bae et al.
-2018).
+In Northeast Asia, the term and concept of ‘Late Paleolithic ’ is widely used rather than ‘Upper Paleolithic’ related to the absence of Middle Paleolithic industry such as Levallois [@gao2002critique; @Bar-Yosef_2002; @mellars2007rethinking; @norton2009evolution; @norton2009east; @lee2013current]].
+Along with the modern human dispersal, this transition often comes with the appearance of blade and micro technology, high frequencies of retouched blade tools, several novel refined tools such as projectile points, end scrapers, burins, denticulates, etc, and using high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary].
 
 In China, the transition from the Middle to the Upper Paleolithic dates to 35-30 ka appeared with a few body decorations, bone tools, core-and-flake in North and cobble tools in South.
-In addition to the technological changes, the use of symbols in the body decorations reflects social and cultural changes occurred during the onset of the Chinese Upper Paleolithic (Qu et al.
-2013; Peng et al.
-2018).
-At the Shuidonggou site in northern China which is considered as a typical site for showing the transition from cobble to blade tool industries in Northeast Asia (Brantingham et al.
-2001; Gao et al.
-2010; Pei et al.
-2012), newly discovered localities (SDG 2, 7-12) show various aspects of the initial Late Paleolithic from 32,000 BP to 6000 BP.
-Some sites (SDG 9, 12) have blades and bladelets made of higher quality or exotic raw materials in addition to more than 80 polished ostrich egg-shell beads (SDG 2), bone needles and an awl (SDG 12), while other sites have mostly small, irregular flakes (Gao et al.
-2010).
+In addition to the technological changes, the use of symbols in the body decorations reflects social and cultural changes occurred during the onset of the Chinese Upper Paleolithic [@qu2013chinese; @peng2018onset].
+At the Shuidonggou site in northern China which is considered as a typical site for showing the transition from cobble to blade tool industries in Northeast Asia [@Brantingham_Perreault_2010; @Gao_Zhang_Yang_Shen_Wu_2010; @pei2012shuidonggou], newly discovered localities (SDG 2, 7-12) show various aspects of the initial Late Paleolithic from 32,000 BP to 6000 BP.
+Some sites (SDG 9, 12) have blades and bladelets made of higher quality or exotic raw materials in addition to more than 80 polished ostrich egg-shell beads (SDG 2), bone needles and an awl (SDG 12), while other sites have mostly small, irregular flakes [@Gao_Zhang_Yang_Shen_Wu_2010].
 The transition to the Upper Paleolihtic in Japan is under the long term debate due to the uncertainty about the existence of the previous period in the Archipelago.
-But there are several features represent the milestone of the new era such as systematic procurement and reduction of lithic raw materials, trap-pit hunting method, watercraft (maritime transport), bow-and-arrow technology, and standardization of lithic tools which can show the appearance of new subsistence strategies possibly driven by risks of increased population and ecological change (One et al.
-2002; Yamaoka; Bae 2017; Nakazawa and Bae 2018; Morisaki et al.
-2019).
+But there are several features represent the milestone of the new era such as systematic procurement and reduction of lithic raw materials, trap-pit hunting method, watercraft (maritime transport), bow-and-arrow technology, and standardization of lithic tools which can show the appearance of new subsistence strategies possibly driven by risks of increased population and ecological change [@Yamaoka_2012; @bae2017origin; @Nakazawa_Bae_2018; @morisaki2019early].
 The technological transition in Korea is mainly characterized by the early appearance of stemmed points and following blade technology.
 Using the blade as a blank, the stemmed points became more standardized and increased in number.
-The flake tools became more sophisticated, with a greater diversity in both type and size along with continuously used core tools during the Upper Paleolithic (Bae 2010; Lee 2015).
+The flake tools became more sophisticated, with a greater diversity in both type and size along with continuously used core tools during the Upper Paleolithic [@Bae_2010; @Lee_2013; @Lee_2016].
 
 In this research, we choose to focus on the technological transition and its impact on mobility and site occupation pattern in Korea because of the uniqueness of sudden and earliest appearance of stemmed points among the broad area of Northeast Asia during the Late Pleistocene.
-The oldest stemmed points so far is from the Yongho-dong site dated to 38.5 ka (Seong 2009; Park 2012; Seong 2015; Bae 2017).
+The oldest stemmed points so far is from the Yongho-dong site dated to 38.5 ka [@seong2009emergence; #### Park 2012 ####; @seong2015diversity; @bae2017origin].
 The stemmed point is a projectile point made out of elongated flake or blade and with gentle retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft.
@@
 The early appearance of the stemmed points in the Korean Peninsula brings about several issues to consider such as issues of human migration or advanced cognitive ability to make a new tool adapting to the surrounding environment.
-As a part of composite implements such as arrowheads or spearheads, the point was combined with other materials and it could be easily replaced reducing waste of total disposal (Keeley 1982; Kuhn and Miller 2015).
+As a part of composite implements such as arrowheads or spearheads, the point was combined with other materials and it could be easily replaced reducing waste of total disposal [@Keeley_1982; @kuhn2015artifacts].
 The new projectile point also indicates a new way of hunting methods or other subsistence strategies because the throwing technique enables to earn a long-distance target.
@@
 This layer dated to 31,200±900 BP by radiocarbon dating and 30,000±1,700 BP by OSL.
-The uppermost layer, dated to 22,000±100 BC by OSL, contains blades, scrapers, awls, and denticulates (Seong 2009).
+The uppermost layer, dated to 22,000±100 BC by OSL, contains blades, scrapers, awls, and denticulates [@seong2009emergence].
 We compare the assemblages from the two older layers at this and other sites to investigate the transition process.
@@
 The elevation is the main proxy for environmental information in this research.
-The Korean paleoenvironmental data and information are not rich enough to directly apply HBE driven models or questions (Han 2008; Chang 2013).
-Rather, the paleoenvironment of the Korean Peninsula for this proposed research time period can be simply summarized as ‘overall cooler and drier during MIS 3’ (Han 2008; Seong 2008; Choi 2011; Chang 2013; Im and Choo 2015; Bak and Lee 2017).
+The Korean paleoenvironmental data and information are not rich enough to directly apply HBE driven models or questions [####Han 2008####; @Chang_2013].
+Rather, the paleoenvironment of the Korean Peninsula for this proposed research time period can be simply summarized as ‘overall cooler and drier during MIS 3’ [###Han 2008###; ####Seong 2008####; ###Choi 2011####; @Chang_2013; @im2015study; @bak2017late].
 In this regard, it is much harder to understand the regional as well as shorter-time scale of paleoenvironmental change in the Korean Peninsula.
-In addition, the targeting period is not exactly matched with major environmental changes that we can check through global sea level change, it seems rather monotonous around 40-35 ka (Bailey 2010; Lorscheid 2017).
+In addition, the targeting period is not exactly matched with major environmental changes that we can check through global sea level change, it seems rather monotonous around 40-35 ka [@Bailey_2010; @Lorscheid_Stocchi_Casella_Gómez-Pujol_Vacchi_Mann_Rovere_2017].
 In this research, we chose elevation to make a premise that the site elevation can represent different environment in local scale and hunter-gatherers moved to different elevation as a result of environment related risk.
@@
 
-The artifact volumetric density is defined as the total number of artifacts per cubic meter of excavated sediment, and served as a proxy for accumulation ratio of artifacts (Clark and Barton 2017; Clark et al.
-2018).
+The artifact volumetric density is defined as the total number of artifacts per cubic meter of excavated sediment, and served as a proxy for accumulation ratio of artifacts [@clark2017lithics; @clark2019landscapes].
 We expected to use this value to evaluate mobility and land-use practices of hunter-gatherers during the Late Pleistocene.
 We count the retouch frequency as the number of retouched tools divided by the total number of artifacts.
-This value represents general proficiency of lithic manufacturing, raw material availability, and subsistence strategies including mobility and duration of site occupation based on the concept that tool reductioning is an economical tactic to generate sharp, usable edges while minimizing the cost of transporting tools or bulky raw materials by reducing the total load weight (Kuhn 1990; Clark et al.
-2018).
-In other words, producing and carrying highly retouched tools will increase portability and efficiency by decrease in the size of both individual tools and the assemblages so that hunter-gatherers could move often or further, or both (Kuhn 1991,1994, 2004; Andrefsky 1994; Davies et al.
-2018).
-For example, highly retouched and finely shaped implements appeared in Australia during the mid-Holocene to minimize risk associated with environmental change and high mobility (Hiscock 1994).
+This value represents general proficiency of lithic manufacturing, raw material availability, and subsistence strategies including mobility and duration of site occupation based on the concept that tool reductioning is an economical tactic to generate sharp, usable edges while minimizing the cost of transporting tools or bulky raw materials by reducing the total load weight [@Kuhn_1990; @clark2019landscapes].
+In other words, producing and carrying highly retouched tools will increase portability and efficiency by decrease in the size of both individual tools and the assemblages so that hunter-gatherers could move often or further, or both [@kuhn1991unpacking; @kuhn1994formal; @kuhn2004upper; @andrefsky1994raw; @davies2018modeling].
+For example, highly retouched and finely shaped implements appeared in Australia during the mid-Holocene to minimize risk associated with environmental change and high mobility [@Hiscock_1994].
 Combined the two proxies, the artifact volumetric density and the retouch frequency, we expect to determine whether or not the hunter-gatherers were residentially mobile or logistically organized.
-If the assemblage in a site has high density and less retouched, it is depicted as expedient assemblage that represents “base camps” or “residences”, while less density with high retouched one shows curated assemblage and indicates either residential mobility or certain task groups from main residential site (Riel-Salvatore and Barton 2004; Clark and Barton 2017).
+If the assemblage in a site has high density and less retouched, it is depicted as expedient assemblage that represents “base camps” or “residences”, while less density with high retouched one shows curated assemblage and indicates either residential mobility or certain task groups from main residential site [@Riel-Salvatore_Barton_2004; @clark2017lithics].
 
 Radiocarbon dates provide a proxy for change in human population levels.
-Summed probability distributions (SPD) of radiocarbon dates are often used to understand temporal trends in ancient human populations (Rick 1987; Bamforth and Grund 2012; Shennan et al.
+Summed probability distributions (SPD) of radiocarbon dates are often used to understand temporal trends in ancient human populations [Rick 1987; Bamforth and Grund 2012; Shennan et al.
 2013; Timpson et al.
-2014; Contreras and Meadows 2014).
+2014; Contreras and Meadows 2014].
 Without there must be unavoidable limitations for the validity of SPD such as the lack of excavated sites in certain period, this method is still helpful to be used as a proxy for population fluctuation.
@@
 
-# Discussion (still working on)
+# Discussion (still working on, needs more citations)
 
@@
 The stemmed points brought new paradigms in stone artifact assemblages.
-For instance, the concept of composite tool contributed to increase efficiency of maintaining the toolkit by easily replacing the damaged part (Kuhn and Miller 2015).
+For instance, the concept of composite tool contributed to increase efficiency of maintaining the toolkit by easily replacing the damaged part [@kuhn2015artifacts].
 Combining with other materials increases utility or functionality of the tool by adding extra weight and providing a shaft, which makes it earlier to grip.

### 2019-11-20 analysis/paper/paper.Rmd
no !

@@
 author:
-  - Author Gayoung Park!
-  - Author Ben Marwick!!!!!!!
+  - Author Gayoung Park
+  - Author Ben Marwick
 date: "`r format(Sys.time(), '%d %B, %Y')`"

### 2019-11-21 analysis/paper/paper.Rmd
Done with the putting citations

--- 2019-11-20 analysis/paper/paper.Rmd
@@
 Radiocarbon dates provide a proxy for change in human population levels.
-Summed probability distributions (SPD) of radiocarbon dates are often used to understand temporal trends in ancient human populations [Rick 1987; Bamforth and Grund 2012; Shennan et al.
-2013; Timpson et al.
-2014; Contreras and Meadows 2014].
+Summed probability distributions (SPD) of radiocarbon dates are often used to understand temporal trends in ancient human populations [@rick1987dates; @bamforth2012radiocarbon; @shennan2013regional; @timpson2014reconstructing; @contreras2014summed].
 Without there must be unavoidable limitations for the validity of SPD such as the lack of excavated sites in certain period, this method is still helpful to be used as a proxy for population fluctuation.
 Timpson et al.
-demonstrate its usefulness by testing data sets in 12 European regions and assert that this SPD is extremely robust to deal with small sample size (Timpson et al.
-2014).
+demonstrate its usefulness by testing data sets in 12 European regions and assert that this SPD is extremely robust to deal with small sample size [@timpson2014reconstructing].
 We use 171 radiocarbon dates from 22 sites to search for population trends during the Late Pleistocene in Korea based on the premise that population change or pressure could be another catalyst to improve or innovate technology.
@@
 The tools represent paleo hunter-gatherers’ adaptive strategies.
-We analyze the composition of raw materials in the assemblages to see the relationships between the choice of specific raw materials and technological innovation as well as raw material availability and its consuming pattern based on an assumption that stone raw materials were carefully transported and consumed to optimize mobility and technological strategies, depth of planning in landscape use and risk minimization (Bratingham 2003).
+We analyze the composition of raw materials in the assemblages to see the relationships between the choice of specific raw materials and technological innovation as well as raw material availability and its consuming pattern based on an assumption that stone raw materials were carefully transported and consumed to optimize mobility and technological strategies, depth of planning in landscape use and risk minimization [@brantingham2003neutral].
 
@@
 It is considered that he hunter-gatherer started to use raw materials selectively depending on the type of tool, especially the newly produced stemmed points.
-Raw material availability is a crucial factor in determining the trip and direction for hunter-gatherers (Davies et al.
-2018).
+Raw material availability is a crucial factor in determining the trip and direction for hunter-gatherers [@davies2018modeling].
 Therefore the adoption of new raw material along with the emergence of new type of tools demonstrates a significant innovation in not just technology but also cognitive system.

### 2019-11-21 analysis/paper/paper.Rmd
test push

@@
 The results show that population dynamics might have had more impact on changes in stone artifact technology.
----
+----
 

### 2019-11-21 README.Rmd
add travis, dockerfile, binder

@@
 # koreapaleolithicmobilityoccupation
+
+<!-- badges: start -->
+[![Travis-CI Build Status](https://travis-ci.org/parkgayoung/koreapaleolithicmobilityoccupation.svg?branch=master)](https://travis-ci.org/parkgayoung/koreapaleolithicmobilityoccupation)
+[![Launch Rstudio Binder](http://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/parkgayoung/koreapaleolithicmobilityoccupation/master?urlpath=rstudio)
+<!-- badges: end -->
 

### 2019-11-21 analysis/paper/paper.Rmd
added a citation

@@
 The stemmed points brought new paradigms in stone artifact assemblages.
-For instance, the concept of composite tool contributed to increase efficiency of maintaining the toolkit by easily replacing the damaged part [@kuhn2015artifacts].
+For instance,
+
+the concept of composite tool contributed to increase efficiency of maintaining the toolkit by easily replacing the damaged part [@kuhn2015artifacts; @cardillo2010some].
 Combining with other materials increases utility or functionality of the tool by adding extra weight and providing a shaft, which makes it earlier to grip.

### 2019-11-22 analysis/paper/paper.Rmd
Added a bit for the discussion. I didn't finish yet but just commiting/pushing before I go home...

--- 2019-11-21 analysis/paper/paper.Rmd
@@
 Summed probability distributions (SPD) of radiocarbon dates are often used to understand temporal trends in ancient human populations [@rick1987dates; @bamforth2012radiocarbon; @shennan2013regional; @timpson2014reconstructing; @contreras2014summed].
-Without there must be unavoidable limitations for the validity of SPD such as the lack of excavated sites in certain period, this method is still helpful to be used as a proxy for population fluctuation.
+Even though there must be unavoidable limitations for the validity of SPD such as the lack of excavated sites in certain period, this method is still helpful to be used as a proxy for population fluctuation.
 Timpson et al.
@@
 
-The new technology and occupation in high elevation
+1. The new technology and occupation in high elevation
 The new technology appeared around 40-35ka in the Korean Peninsula includes stemmed points and following blade industry.
 The stemmed points brought new paradigms in stone artifact assemblages.
-For instance, the concept of composite tool contributed to increase efficiency of maintaining the toolkit by easily replacing the damaged part [@kuhn2015artifacts].
+For example, projectile points are only a part of composit artifacts and this specific characteristics could contribute to increase efficiency of maintaining the toolkit by easily replacing the damaged part [@kuhn2015artifacts; @cardillo2010some].
 Combining with other materials increases utility or functionality of the tool by adding extra weight and providing a shaft, which makes it earlier to grip.
@@
 In addition, this research shows that carrying the points enabled hunter-gatherers to occupy higher elevation compared to the sites with no stemmed points.
-In other words, the new technology might play an important role to survive in environments with more risks.
+Therefore, we can assume that the technology would play an important role to survive in more risky environments.
 
-Population change
-Interpreted in HBE, we assume there might be close connection between environmental change and the appearance of new technology based on the prediction that the new invention was the result of risk.
-The research suggests that the environmental pressure was related to population change.
+2. Population change
+Interpreted in HBE, we expected there might be close connection between environmental change and the appearance of new technology based on the prediction that the new invention was the result of risk.
+The research suggests that the environmental pressure coule be related to population change.
 We cannot simply predicate that the increased population around 35ka was due to dispersal of Homo Sapiens since there is no reliable human remains in South Korea.
 However, the migration of Homo sapiens was not impossible scenario at the same time.
-Bae (2010) argues that the emergence of stemmed points and blade industry came from outside of the Korean peninsula.
-He proposes that the new technology was the result of the continuous influx of modern human populations from Siberia while the earlier coarse flake tradition ----
+In fact, one model explains that the emergence of stemmed points and blade industry came from outside of the Korean peninsula and the new technology was the result of the continuous influx of modern human populations from Siberia while the earlier coarse flake trandition was come from southern China [@Bae_2010].
+Since this migration model barely explains about preexisted stone artifact assemblages and people in the Korean Peninsula, it is hard to accept his argument without knowing about the native people.
+In short, we assume that the stemmed points appeared related to the increased population and it can be related to the modern human dispersal or indigenously increased occupants.
+
+3. Expedient vs curated
+To monitor moibility strategies and site occupation pattern, we examined the artifact density and retouch frequency, based on the premise that highly mobile foragers would have light weighted, highly retouched, replaceable, almost completely made tools and leave small amount of artifacts when they move to the new location (curated).
+One the other hand, less mobile foragers who tend to stay in one place yield more debitage and debris, produce less efficient tools since they can easily get supplied (expedient) [@clark2019landscapes; @kuhn1994formal; @kuhn2014mousterian; @meignen2006middle].
+Therefore, the curated technology indicates sites with special or higly mobile groups with stone artifacts that are thoroughly designed and based on a high level of investment of time and effort in artifact manufacture and maintenance.
+The expedient technology represents base camp or residence with artifacts that tend to be manufactured, used, and abandoned in the short period of time [@vaquero2018searching;@binford1979organization; @torrence1989time].
+
+During the transitoin period, sites were used in
 

### 2019-11-22 analysis/paper/paper.Rmd
First draft is done, except for Umap part that Ben will cover!

@@
 
-# Discussion (still working on, needs more citations)
+# Discussion
 
@@
 3. Expedient vs curated
-To monitor moibility strategies and site occupation pattern, we examined the artifact density and retouch frequency, based on the premise that highly mobile foragers would have light weighted, highly retouched, replaceable, almost completely made tools and leave small amount of artifacts when they move to the new location (curated).
+To monitor mobility strategies and site occupation pattern, we examined the artifact density and retouch frequency, based on the premise that highly mobile foragers would have light weighted, highly retouched, replaceable, almost completely made tools and leave small amount of artifacts when they move to the new location (curated).
 One the other hand, less mobile foragers who tend to stay in one place yield more debitage and debris, produce less efficient tools since they can easily get supplied (expedient) [@clark2019landscapes; @kuhn1994formal; @kuhn2014mousterian; @meignen2006middle].
-Therefore, the curated technology indicates sites with special or higly mobile groups with stone artifacts that are thoroughly designed and based on a high level of investment of time and effort in artifact manufacture and maintenance.
+Therefore, the curated technology indicates sites with special or highly mobile groups with stone artifacts that are thoroughly designed and based on a high level of investment of time and effort in artifact manufacture and maintenance.
 The expedient technology represents base camp or residence with artifacts that tend to be manufactured, used, and abandoned in the short period of time [@vaquero2018searching;@binford1979organization; @torrence1989time].
+There appear to be more variety in the type of sites (or group) during the transition period, both expedient or curated.
+It became more expedient at the end of the transition, which means foragers became less mobile and tended to stay in the base camp.
+Combined with the result of population dynamics, we can assume that the increased population made each group bigger and let them stay in one place.
 
-During the transitoin period, sites were used in
-
-# Conclusion (still working on)
+# Conclusion
 
@@
 We could clearly see that hunter-gatherers chose these materials particularly to make the newly added stone artifacts to the toolkit.
+The results also show that the new technology enables the occupation of marginal environments at higher elevations.
 Besides of technological change, we found that there are more varieties in the type of sites or groups during the transition period, both curated and expedient and then it becomes more expedient.
+In other words, the foragers became less mobile and prone to stay in one place after the transition ended.
 We expected to see any pattern that would explain the relationship between environmental change and the emergence of the new technology.
-However, there was not enough environmental information as well as data to apply to verify the model.
+However, there was not enough paleoenvironmental information as well as data to apply to verify the model.
 Therefore we focus on Radiocarbon dates as a proxy of population change based on the premise that population driven pressure could trigger people to invent a new tool.
 The results show that population dynamics might have had more impact on changes in stone artifact technology.
-----
 

### 2019-11-24 analysis/paper/paper.Rmd
fixed typoes
move Rcode for figures to here. Not everything is successful...

@@
 
+```{r setup, include=FALSE}
+suppressMessages(library(tidyverse))
+library(glue)
+library(ggrepel)
+
+# data from PhD data sheet, not KAS sheet.
+mydata <- read.csv("~/koreapaleolithicmobilityoccupation/General_info.csv")
+
+# raw material data from KAS data sheet.
+kasr <- read.csv("~/koreapaleolithicmobilityoccupation/Rawmaterial_info.csv")
+
+# assemblage composition data from KAS data sheet.
+kasa <- read.csv("~/koreapaleolithicmobilityoccupation/Assemblage_info.csv")
+
+#volume of the cultural layer from KAS data sheet.
+kasv <- read.csv("~/koreapaleolithicmobilityoccupation/Dating_info.csv")
+
+dates <- read.csv("~/koreapaleolithicmobilityoccupation/Radiocarbon dates.csv")
+
+```
 # Introduction
@@
 
+```{r}
+#I couldn't find the most recent map that you made. I assume that you don't share the code, but showed me the image so that I just copied it for SAA poster
+#This is what I made first before you made the recent one
+
+korean_archaeological_sites <- readxl::read_excel("~/koreapaleolithicmobilityoccupation/korean-archaeologica-sites.xlsx")
+
+library(tidyverse)
+
+coords <- 
+	korean_archaeological_sites %>% 
+	select(site_name, 
+				 east_longitude, 
+				 north_latitude) %>% 
+	separate(east_longitude, 
+					 into = c("e_deg", "e_min", "e_sec"),
+					 sep =  "°|\\′") %>% 
+	mutate(e_min = str_squish(e_min),
+				 e_sec = str_squish(e_sec)) %>% 
+	separate(e_min, 
+					 into = c("e_min", "e_sec1"), 
+					 sep = "'") %>% 
+	mutate_at(vars(e_deg, e_min, e_sec, e_sec1), parse_number) %>% 
+	mutate(e_sec = ifelse(is.na(e_sec), e_sec1, e_sec)) %>% 
+	select(-e_sec1) %>% 
+	separate(north_latitude, 
+					 into = c("n_deg", "n_min", "n_sec"),
+					 sep =  "°|\\′") %>% 
+	mutate(n_min = str_squish(n_min),
+				 n_sec = str_squish(n_sec)) %>% 
+	separate(n_min, 
+					 into = c("n_min", "n_sec1"), 
+					 sep = "'") %>% 
+	mutate_at(vars(n_deg, n_min, n_sec, n_sec1), parse_number) %>% 
+	mutate_if(is.numeric, list(~ifelse(is.na(.), 0, .))) %>% 
+	mutate(n_sec = ifelse(is.na(n_sec), n_sec1, n_sec)) %>% 
+	select(-n_sec1) %>% 
+	mutate(lat_dms =  str_glue('{n_deg} {n_min} {n_sec}'),
+				 long_dms = str_glue('{e_deg} {e_min} {e_sec}')) %>% 
+	# convert from decimal minutes to decimal degrees
+	mutate(lat_dd = measurements::conv_unit(lat_dms, 
+																			 from = 'deg_min_sec', 
+																			 to = 'dec_deg')) %>% 
+	mutate(long_dd = measurements::conv_unit(long_dms, 
+																				from = 'deg_min_sec', 
+																				to = 'dec_deg')) %>% 
+	mutate_at(vars(lat_dd, long_dd), as.numeric)
+
+# quick map
+
+library(ggmap)
+
+# bounding box
+# 39.534214, 124.159154 ... 39.657415, 129.138604
+# 33.694662, 123.491142 ... 34.338454, 130.569658
+
+map <- 
+	get_stamenmap(bbox = c(left = 123, 
+												 bottom = 33, 
+												 right = 	131, 
+												 top = 39),
+								zoom = 10)
+
+library(ggrepel)
+ggmap(map)  + geom_point(data = coords,
+						 aes(long_dd ,
+						 		lat_dd), 
+						 colour = "red",
+						 size = 2) +
+	geom_label_repel(data = coords,
+									aes(long_dd ,
+											lat_dd,
+											label = site_name),
+									size = 2)
+
+# pick points off this map with korea_boundary <- locator()
+
+korea_bb <-  
+	data.frame(x = c(125, 
+						       130, 
+						       130, 
+						       125), 
+			 y = c(38, 
+			 			 38, 
+			 			34, 
+			 			34))
+```
+
 Assemblages
@@
 
-```{r get-data, eval = FALSE}
-# Note the path that we need to use to access our data files when rendering this document
-my_data <- read.csv(here::here('analysis', 'data', 'raw_data', 'my_csv_file.csv'))
-```
-
 Elevation of occupied sites
@@
 2: distinction between sites with no stemmed points and with stemmed points).
+
+```{r}
+#site elevation
+mydata_ages <- 
+  mydata %>% 
+  separate(C14.BP., into = c('age', 'error'),
+          sep = "±") %>% 
+  mutate(age_ka = parse_number(age) / 1000,
+         error = parse_number(error)) %>% 
+  mutate(has_sp = ifelse(!is.na(SP.), "yes", "no"))
+
+ggplot(mydata_ages, 
+       aes(x = age_ka,
+           y = altitude.m._of_main_layer,
+       color = as.factor(has_sp))) +
+  geom_point(size = 3) +
+  xlab("Age of occupation (ka)") +
+  ylab("Elevation above sea level (m)") +
+  geom_smooth(se = FALSE) +
+  scale_colour_discrete(name = "Contains\nstemmed\npoints?") +
+  theme_minimal(base_size = 16) 
+```
 
@@
 Likewise, the composition of toolkit might represent the characteristics or purpose of each site, rather than showing general pattern of chronological change.
+
+```{r}
+#Volume and artefact counts to get density over time. 
+kasv_tidy <- 
+kasv %>% 
+  t %>% 
+  as_tibble() %>% 
+  setNames(as.character(.[1,])) %>% 
+  .[-1,] %>% 
+  mutate_all(parse_number) %>% 
+  mutate(artefact_density = total_artifacts / volume,
+         sites = names(kasv)[-1])
+
+ggplot(kasv_tidy,
+       aes(date_age,
+           artefact_density)) +
+  geom_point(aes(size = total_artifacts)) +
+  geom_text_repel(aes(label = sites),
+                  nudge_x = 0.25, 
+                  nudge_y = 0.25) +
+  ylab("Artifact density") +
+  xlab("Age of assemblage (ka)") +
+  scale_size_continuous(name = "Total number\nof artifacts")  +
+  theme_minimal(base_size = 16)
+
+```
 
@@
 
+```{r}
+dates_clean <- 
+  dates %>% 
+  mutate(age = parse_number(as.character(age)),
+         error = parse_number(as.character(error))) %>% 
+  filter(age < 50000, method=='AMS')
+
+# devtools::install_github('ahb108/rcarbon')
+library(rcarbon)
+dates_calibrated <- 
+  calibrate(x=dates_clean$age,
+               errors=dates_clean$error,
+               calCurves='intcal13')
+
+dates_calibrated_spd <-  spd(dates_calibrated, timeRange = c( 45000, 1000))
+plot(dates_calibrated_spd)
+
+# Testing Observed SPDs against theoretical models
+## recalibrate dates without normalisation to avoid artificial peaks
+spd_dates<- calibrate(x=dates_clean$age,errors=dates_clean$error,normalised=FALSE,verbose=F)
+
+spd_test <- modelTest(spd_dates,errors=dates_clean$error,timeRange=c(45000,1000),model='exponential',nsim=200,ncores=3)
+
+summary(spd_test)
+plot(spd_test)
+```
+
 Raw materials
@@
 
+```{r}
+#raw material information
+kasr_long1 <- 
+  kasr %>% 
+  gather(site, 
+         count, 
+         -X) %>% 
+  group_by(site) %>% 
+  mutate(percentage = count / sum(count, na.rm = TRUE) * 100,
+         total = sum(count, na.rm = TRUE)) %>% 
+  filter(!is.na(percentage)) %>% 
+  mutate(raw_material = ifelse(percentage >= 10, as.character(X), "other")) %>% 
+  mutate(raw_material = ifelse(raw_material == "etc.", "other", raw_material)) %>% 
+  select(-X)
+
+  # if percentage is <10%, call it 'other'
+kasr_long2 <- 
+  kasr_long1 %>% 
+  group_by(site, raw_material) %>% 
+  summarise(percentage = sum(percentage)) %>% 
+  ungroup() %>% 
+  left_join(kasr_long1 %>% 
+              select(site, raw_material, total),
+            by = c('site', "raw_material")) %>% 
+  distinct(.keep_all = TRUE) %>%
+
+  # join to get ages of the sites
+  left_join(mydata_ages, 
+            by = c('site' = 'site_name')) %>% 
+  mutate(age_ka = jitter(age_ka),
+         age_bin  = ntile(age_ka, 5)) %>% 
+  arrange(age_ka) %>% 
+  filter(!is.na(age_ka)) %>% 
+    mutate(axis_label = glue('{site} ({round(age_ka,1)} ka, n = {total})'))
+
+ggplot(kasr_long2, 
+       aes(
+           reorder(axis_label, -age_ka),
+           percentage,
+           fill = raw_material)) +
+  geom_col(position = "fill") +
+  xlab("Assemblage (youngest at the top)") +
+  ylab("Percentage") +
+  scale_fill_viridis_d(name = "Raw material type",
+                       option = "C") +
+  coord_flip() +
+  theme_minimal(base_size = 12)
+
+```
 Umap
@@
 Interpreted in HBE, we expected there might be close connection between environmental change and the appearance of new technology based on the prediction that the new invention was the result of risk.
-The research suggests that the environmental pressure coule be related to population change.
+The research suggests that the environmental pressure could be related to population change.
 We cannot simply predicate that the increased population around 35ka was due to dispersal of Homo Sapiens since there is no reliable human remains in South Korea.
@@
 
+------------NOT WORKING-------------
+
+```{r}
+#the artifact composition of each sites to figure out what people did as they were occupying the sites.
+kasa_long <-
+  kasa %>%
+  gather(site, 
+         count, 
+         -X) %>% 
+  filter(!is.na(count)) %>% 
+  group_by(site) %>% 
+  mutate(percentage = count / sum(count, na.rm = TRUE) * 100,
+         total = sum(count, na.rm = TRUE)) %>% 
+  filter(!is.na(percentage)) %>% 
+  # if percentage is <10%, call it 'other'
+  mutate(artefact_type = ifelse(as.character(X) == 'stemmed_point',
+                                'stemmed_point', 
+                                ifelse(percentage >= 10, 
+                                as.character(X), "other_tools"))) %>% 
+  mutate(axis_label = glue('{site}\n(n = {total})'))  %>% 
+  # join to get ages of the sites
+  left_join(mydata_ages, 
+            by = c('site' = 'site_name')) %>% 
+  arrange(age_ka) %>% 
+   mutate(age_ka = jitter(age_ka))
+
+ggplot(kasa_long, 
+       aes(age_ka, 
+           percentage,
+           fill = artefact_type)) +
+  geom_col() +
+  ylab("Percentage") +
+  xlab("Age of assemblage (ka)") +
+  labs(fill="Artifact type")+
+  theme_minimal(base_size = 16)
+
+```
+
+Get the lithic data for some sites:
+
+```{r}
+# we'll use googlesheets package to get the data more efficiently
+library(googlesheets)
+
+# here is the key to the google sheet: this ggole sheet is same as kasa. so I could do kas_sites <- kasa 
+kas_sites <- gs_key("14G0Mp-YSsKBOBPF1Z1zPWQbZQ-sjJSWLetDIQSlbsjw")
+
+# get the tab names from this sheet: But that I couldn't use gs_ws_ls
+kas_sites_names <- gs_ws_ls(kas_sites)
+
+# sheets we need here
+sheets <- c( "general_info",    "assemblage_info" )
+
+# get the data from the tabs in this sheet!
+kas_sites_sheets <- 
+  map(sheets, ~kas_sites %>% gs_read_listfeed(ws = .x))
+```
+
+Compute artefact volumetric density by frequency of retouched piece (log scale), following Julien Riel-Salvatore and C. Michael Barton in  https://www.jstor.org/stable/4128419
+
+```{r}
+# put the site names on this list of data frames
+site_names <-  kas_sites_sheets[[1]] %>%  
+  select(-X1) %>% 
+  names()
+
+# density of arefacts at each site
+artefact_density <- 
+kas_sites_sheets[[1]] %>% 
+  t %>% 
+  as.data.frame() %>% 
+ rownames_to_column() %>% 
+  mutate_all(as.character)
+
+names(artefact_density) <- as.character(as.vector(artefact_density[1, ]))
+artefact_density <- artefact_density[-1,]
+
+artefact_density <- 
+artefact_density %>% 
+  mutate(density = as.numeric(total_artifacts) / as.numeric(volume)) %>% 
+   mutate_at(2:ncol(.), as.numeric)
+
+# frequency of retouched artefacts
+retouched_artefacts <- 
+ kas_sites_sheets[[2]] %>% 
+   filter(!X1 %in% c("debris", 
+                     "unkown", 
+                     "pebble",  
+                     "metate", 
+                     "flake",
+                     "unfinished",
+                     "ground",
+                     "blank"
+                     )) %>% 
+    t %>% 
+  as.data.frame() %>% 
+ rownames_to_column() %>% 
+  mutate_all(as.character)
+
+names(retouched_artefacts) <- as.character(as.vector(retouched_artefacts[1, ]))
+retouched_artefacts <- retouched_artefacts[-1,]
+
+retouched_artefacts <- 
+retouched_artefacts %>% 
+  mutate_at(2:ncol(.), as.numeric)  %>% 
+  mutate(total_retouched = rowSums(.[,-1], na.rm = TRUE))
+
+# join together the site age, artefact density df with the retouched artefacts df
+kas_density_retouch_ages <- 
+artefact_density %>% 
+  left_join(retouched_artefacts) %>% 
+  mutate(proportion_retouched = total_retouched / total_artifacts) %>% 
+  mutate(sites = X1)
+
+# compute correlation
+kas_sites_retouch_density_corr <- 
+cor.test(kas_density_retouch_ages$proportion_retouched,
+         kas_density_retouch_ages$density)
+
+r_value <- unname(kas_sites_retouch_density_corr$estimate)
+
+```
+
+```{r}
+size <-  8
+
+ggplot(kas_density_retouch_ages,
+       aes(density,
+           proportion_retouched
+           )) +
+  geom_point(aes(colour = date_age / 1000,
+                 size = total_artifacts,
+                 shape = is.na(stemmed_point))) +
+ # geom_text_repel(aes(label = sites),
+ #                 nudge_x = 0.075, 
+ #                 nudge_y = -0.075,
+ #                 size = 5) +
+  geom_smooth(alpha = 0.2) +
+  scale_y_log10(limits = c(0.001, 1),
+                labels = scales::comma_format(accuracy = 0.001)) +
+  scale_x_log10(limits = c(0.001, 10),
+                labels = scales::comma_format(accuracy = 0.001)) +
+  theme_minimal(base_size = 12) +
+  scale_color_viridis_c(name = "Age of\nassemblage (ka)") +
+  xlab("Artifact volumetric density") +
+  ylab("Proportion of retouched pieces") +
+  annotate("text",
+           x = 0.005,
+           y = 0.7,
+           label = "Curated",
+           size = size -2,
+           colour = "grey50") +
+    annotate("text",
+           x = 5,
+           y = 0.01,
+           label = "Expedient",
+           size = size -2,
+           colour = "grey50") +
+    annotate("text",
+           x = 0.01,
+           y = 0.01,
+           label = glue('r = {round(r_value, 3)}'),
+           size = size - 2,
+           colour = "grey50")
+
+```
+
+```{r}
+# I wanna make simple table for artifact density and frequency of retouched pieces again. Because the complex table above is hard to make comments..
+
+ggplot(kas_density_retouch_ages,
+       aes(density,
+           proportion_retouched
+           )) +
+  geom_point(aes(colour = date_age / 1000,
+                 size = 7)) +
+ # geom_text_repel(aes(label = sites),
+ #                 nudge_x = 0.075, 
+ #                 nudge_y = -0.075,
+ #                 size = 5) +
+  geom_smooth(alpha = 0.2) +
+  scale_y_log10(limits = c(0.001, 1),
+                labels = scales::comma_format(accuracy = 0.001)) +
+  scale_x_log10(limits = c(0.001, 10),
+                labels = scales::comma_format(accuracy = 0.001)) +
+  theme_minimal(base_size = 12) +
+  scale_color_viridis_c(name = "Age of\nassemblage (ka)") +
+  xlab("Artifact volumetric density") +
+  ylab("Proportion of retouched pieces") +
+  annotate("text",
+           x = 0.005,
+           y = 0.7,
+           label = "Curated",
+           size = size -2,
+           colour = "grey50") +
+    annotate("text",
+           x = 5,
+           y = 0.01,
+           label = "Expedient",
+           size = size -2,
+           colour = "grey50") +
+    annotate("text",
+           x = 0.01,
+           y = 0.01,
+           label = glue('r = {round(r_value, 3)}'),
+           size = size - 2,
+           colour = "grey50")
+
+```
+
+Artefact volumetric density and frequency of retouched pieces can be proxies to identify the characteristics of the hunter-gatherers who carried that stone tools whether they are residentially mobile or logistically orgnized foragers based on the premise that the relationship between lithic assemblage composition and land-use strategy represent an expression of economizing behavior.
+
+The artefact volumetric density is defined as the total number of artefacts per cubic meter of excavated sediment, and served as a proxy for accumulation ratio of artefacts.
+the relative frequency of retouched pieces is the count of retouched flake tools devided by the total number of artifacts.
+
+The two ends of the curated-expedient show that residential to logistical mobility.
+Expedient assemblages can be depicted as "base camps" or "residences", while curated ones represent either residential mobility, where the central locus of group changes frequently, or certain task groups from a central residential site.
+
+The plot show that younger sites are likely to be used as base camp than older site, which means that people tended to stay longer at one place after around 40-35ka.
+
+(Riel-Salvatore and Barton 2004)
+
+```{r}
+# dating
+library(rcarbon)
+
+# I'm working through https://cran.r-project.org/web/packages/rcarbon/vignettes/rcarbon.html
+
+kas_ages <- 
+kas_density_retouch_ages %>% 
+  select(sites, date_age, date_error) %>% 
+  mutate(sites = str_remove_all(sites, "_"))
+
+kas_caldates <- 
+  calibrate(x = kas_ages$date_age,
+            errors = kas_ages$date_error,
+            calCurves='intcal13',
+            ncores=3) #running calibration over 3 cores
+
+kas_spd = spd(kas_caldates, timeRange = c(50000,25000)) 
+plot(kas_spd)
+
+binsense(x=kas_caldates,
+         y=kas_ages$sites,
+         h=seq(0,5000,100),
+         timeRange=c(50000,25000))
+
+ggplot(kas_spd$grid,
+       aes(calBP,
+           PrDens)) +
+  geom_line() +
+  scale_y_log10(name = "Summed probability")
+
+# test how the SPD fits to an exponential growth model:
+expnull <- modelTest(kas_caldates, 
+                     errors=kas_ages$date_error, 
+                     nsim=100, 
+                     timeRange=c(50000,25000),
+                     model="uniform",
+                     runm=10)
+
+plot(expnull)
+
+```
+
 # Acknowledgements

### 2019-11-25 analysis/paper/paper.Rmd
simplify abstract

@@
 abstract: |
-  According to Human Behavioral Ecology, stone artifacts can be an essential source of information for understanding hunter-gatherer subsistence.
-We examine stone artifacts from 22 sites in Korea to investigate a change in mobility and site occupation during the Late Pleistocene.
-The new technology is represented by the technological change, the appearance of stemmed points, and the following blade industry.
-The overall aim is to understand how the appearance of stemmed points affected hunter-gatherers in Korea, which leads to three specific questions to pursue: how did the introduction of the new tools change the way people used the landscape, the way people used sites, and the population.
-Based on the hypothesis that hunter-gatherers should move frequently and further in order to adapt to a changing environment, so their tools became more portable and efficient, as indicated by decreases in toolkit size and increases in function and utility, we analyze artifact density, retouch frequency, radiocarbon dates, elevation of the sites, raw material types, and composition of the toolkit.
+  The Late Pleistocene saw the introduction of new lithic technologies to the Korean Peninsula, especially stemmed points and blades.
+We examined stone artifacts from 22 sites in Korea to investigate how the introduction of  new tools changed the way people used the landscape, and the way people used sites.
+We analyzed artifact density, retouch frequency, radiocarbon ages, site elevation, raw material types, and composition of the toolkit to test hypotheses about site use.
 The results show that the new technology enabled the occupation of marginal environments at higher elevations.
-There is no strong general relationship between technology and ecological change.
+We find no strong relationship between technology and ecological change.
 Instead, we assume that population dynamics might have had more impact on changes in stone artifact technology and hunter-gatherer subsistence.

### 2019-11-25 analysis/paper/paper.Rmd
reorg data files, make & save the map

@@
 
-Here is a citation
-
 Stone artifact assemblages have important information about hunter-gatherer mobility and subsistence strategy, according to the behavioral ecological theory that links technological changes to human adaptation to the surrounding environment.
@@
 
-Theory and background to our three questions
+## Theory and background to our three questions
+
 New technologies have led people to adjust the speed and frequency of mobility, explore the new landscape, or even manage the population size of the foraging group or overall population occupied one region.
@@
 The appearance of microliths as portable and standardized tools is a great example to show that people made new tools to increase multi-functionality and tool readiness so that they can minimize the economic risk and/or scheduling difficulties [@Hiscock_Clarkson_Mackay_2011].
+
 The rapid spread of microblade industry in north China among Last Glacial Maximum demonstrates that making new tools was one of the human adaptation to extremely cold environment by increasing mobility or meeting the new demands such as the production of sophisticated cold-weather clothing [@Barton_Brantingham_Ji_2007; @yi2013microblade; @Qu_Bar-Yosef_Wang_Wu_2013; @nian2014chronology].
@@
 
-Regional Archaeology
+## Regional Archaeology
 
 Northeast Asia/similar studies and previous research in the adjacent regions : General description of EUP in East Asia
+
 The technological change during the Late Pleistocene in Northeast Asia has been actively studied related to the topic of modern human dispersal to the East and chronological transition from Middle to Upper Paleolithic [or Early to Late Paleolithic.
@@
 
-```{r}
+```{r cache = TRUE}
 #I couldn't find the most recent map that you made. I assume that you don't share the code, but showed me the image so that I just copied it for SAA poster
@@
 
-korean_archaeological_sites <- readxl::read_excel("~/koreapaleolithicmobilityoccupation/korean-archaeologica-sites.xlsx")
+library(here)
+
+korean_archaeological_sites <- readxl::read_excel(here("analysis/data/raw_data/korean-archaeologica-sites.xlsx"))
 
@@
 
+ggsave(here("analysis/figures/site-map.png"))
+
 # pick points off this map with korea_boundary <- locator()
@@
 Statistical methods
+
 To answer the three research questions, we gather information such as elevation of the sites, artifact density, retouch frequency, radiocarbon dates toolkit composition, and raw material type.

### 2019-11-25 analysis/paper/paper.Rmd
make site elevation plot work

@@
 After the first excavation at Jeongokri site in the late 1970s, more than 200 Paleolithic sites are discovered in Korea.
-Among them, we  chose 22 sites based on the existence of absolute dates that match the targeting timeline and officially published excavation report.
-We include sites dated before and after the transition period (40-35ka) to analyze the process of technological transition.
+Among them, we  chose 22 sites based on the existence of absolute dates that span our period of interest.
+We include sites dated before and after the transition period (40-35 ka) to analyze the process of technological transition.
 
-```{r cache = TRUE}
-#I couldn't find the most recent map that you made. I assume that you don't share the code, but showed me the image so that I just copied it for SAA poster
-#This is what I made first before you made the recent one
-
+```{r cache = TRUE, fig.cap="Map of sites discussed in the text."}
 library(here)
@@
 Assemblages
-The research analyzes total 33 assemblages and they consist of cores, blade cores, blades, flakes, debris, hammers, choppers, plans, polyhedrals, side scrapers, end scrapers, notches, cleavers, stemmed points, awls, denticulates, burins, handaxes, blanks, knives, handadzes, peaks, flatters, flakers etc,
+
+The research analyzes total 33 assemblages [need to explain what is the difference between 'assemblage' and 'site'] and they consist of cores, blade cores, blades, flakes, debris, hammers, choppers, plans, polyhedrals, side scrapers, end scrapers, notches, cleavers, stemmed points, awls, denticulates, burins, handaxes, blanks, knives, handadzes, peaks, flatters, flakers etc,
+
 Some artifacts are not included as certain tools due to their vague or broken shape presumably related to the characteristics of raw materials such as quartz, quartz vein, and quartzite.
@@
 For example, highly retouched and finely shaped implements appeared in Australia during the mid-Holocene to minimize risk associated with environmental change and high mobility [@Hiscock_1994].
+
 Combined the two proxies, the artifact volumetric density and the retouch frequency, we expect to determine whether or not the hunter-gatherers were residentially mobile or logistically organized.
@@
 Elevation of occupied sites
+
 The result of the elevation change of 33 cultural layers from 22 sites corresponding to the targeting period shows that there is no remarkable change can be seen in site elevation during the transition (Fig.1: with no sp site distinction).
@@
 ```{r}
-#site elevation
+# site elevation
+
+mydata <- korean_archaeological_sites
+
 mydata_ages <- 
   mydata %>% 
-  separate(C14.BP., into = c('age', 'error'),
+  separate(`C14(BP)`, into = c('age', 'error'),
           sep = "±") %>% 
   mutate(age_ka = parse_number(age) / 1000,
-         error = parse_number(error)) %>% 
-  mutate(has_sp = ifelse(!is.na(SP.), "yes", "no"))
+         error = parse_number(error),
+         `altitude(m)_of_main_layer` = parse_number(`altitude(m)_of_main_layer`)) %>% 
+  mutate(has_sp = ifelse(!is.na(`SP#`), "yes", "no"))
 
@@
        aes(x = age_ka,
-           y = altitude.m._of_main_layer,
+           y = `altitude(m)_of_main_layer`,
        color = as.factor(has_sp))) +
@@
 Artifact density and Composition of toolkit
+
 During the transitional period, the artifact volumetric density was relatively small but it becomes higher around 30 ka (Fig.
@@
 Summed probability distribution
+
 To understand the population change during the appearance of stemmed points, we analyzed the summed probability distributions of 171 radiocarbon dates from 33 cultural layers and it indicates that there is a higher density of radiocarbon dates from around 34 ka.
@@
 Raw materials
+
 During the transition period, the use of shale and rhyolite greatly increased.
@@
 1. The new technology and occupation in high elevation
+
 The new technology appeared around 40-35ka in the Korean Peninsula includes stemmed points and following blade industry.
@@
 2. Population change
+
 Interpreted in HBE, we expected there might be close connection between environmental change and the appearance of new technology based on the prediction that the new invention was the result of risk.
@@
 3. Expedient vs curated
+
 To monitor mobility strategies and site occupation pattern, we examined the artifact density and retouch frequency, based on the premise that highly mobile foragers would have light weighted, highly retouched, replaceable, almost completely made tools and leave small amount of artifacts when they move to the new location (curated).

### 2019-11-25 analysis/paper/paper.Rmd
notes on background

@@
 ## Theory and background to our three questions
+
+<!-- one equal-sized paragraph for each of our three questions, with lots of recent, specific citations for each Q -->
 
@@
   theme_minimal(base_size = 16)
+
+ggsave(here("analysis/figures/site-elevation-by-age.png"))
 ```

### 2019-11-25 analysis/paper/paper.Rmd
data import

@@
 library(ggrepel)
+library(here)
 
 # data from PhD data sheet, not KAS sheet.
-mydata <- read.csv("~/koreapaleolithicmobilityoccupation/General_info.csv")
+mydata <- read.csv(here("analysis/data/raw_data/General_info.csv"))
 
 # raw material data from KAS data sheet.
-kasr <- read.csv("~/koreapaleolithicmobilityoccupation/Rawmaterial_info.csv")
+kasr <- read.csv(here("analysis/data/raw_data/Rawmaterial_info.csv"))
 
 # assemblage composition data from KAS data sheet.
-kasa <- read.csv("~/koreapaleolithicmobilityoccupation/Assemblage_info.csv")
+kasa <- read.csv(here("analysis/data/raw_data/Assemblage_info.csv"))
 
 #volume of the cultural layer from KAS data sheet.
-kasv <- read.csv("~/koreapaleolithicmobilityoccupation/Dating_info.csv")
+kasv <- read.csv(here("analysis/data/raw_data/Dating_info.csv"))
 
-dates <- read.csv("~/koreapaleolithicmobilityoccupation/Radiocarbon dates.csv")
+dates <- read.csv(here("analysis/data/raw_data/Radiocarbon dates.csv"))
 

### 2019-12-03 analysis/paper/paper.Rmd
fixed a few codes to make figures working

@@
 library(here)
+
+korean_archaeological_sites <- readxl::read_excel(here("analysis/data/raw_data/korean-archaeologica-sites.xlsx"))
 
@@
 ```{r cache = TRUE, fig.cap="Map of sites discussed in the text."}
-library(here)
-
-korean_archaeological_sites <- readxl::read_excel(here("analysis/data/raw_data/korean-archaeologica-sites.xlsx"))
-
-library(tidyverse)
 
@@
 
-mydata <- korean_archaeological_sites
-
 mydata_ages <- 
   mydata %>% 
-  separate(`C14(BP)`, into = c('age', 'error'),
+  separate(C14.BP., into = c('age', 'error'),
           sep = "±") %>% 
   mutate(age_ka = parse_number(age) / 1000,
-         error = parse_number(error),
-         `altitude(m)_of_main_layer` = parse_number(`altitude(m)_of_main_layer`)) %>% 
-  mutate(has_sp = ifelse(!is.na(`SP#`), "yes", "no"))
+         error = parse_number(error)) %>% 
+  mutate(has_sp = ifelse(!is.na(SP.), "yes", "no"))
 
@@
        aes(x = age_ka,
-           y = `altitude(m)_of_main_layer`,
+           y = altitude.m._of_main_layer,
        color = as.factor(has_sp))) +
@@
 
-ggsave(here("analysis/figures/site-elevation-by-age.png"))
 ```
@@
 
-To understand the population change during the appearance of stemmed points, we analyzed the summed probability distributions of 171 radiocarbon dates from 33 cultural layers and it indicates that there is a higher density of radiocarbon dates from around 34 ka.
+To understand the population change during the appearance of stemmed points, we analyzed the summed probability distributions of 171 radiocarbon dates from 33 cultural layers to get a rough image of population dynamics at the sites.
+The result indicates that there is a higher density of radiocarbon dates from around 34 ka.
 We could cautiously assume that the population flux might relate to the technological innovation.
 
-```{r}
+```{r cache=TRUE}
 dates_clean <- 
@@
                                 as.character(X), "other_tools"))) %>% 
-  mutate(axis_label = glue('{site}\n(n = {total})'))  %>% 
+  mutate(axis_label = as.character(glue('{site}\n(n = {total})')))  %>% 
   # join to get ages of the sites
@@
             by = c('site' = 'site_name')) %>% 
-  arrange(age_ka) %>% 
-   mutate(age_ka = jitter(age_ka))
+  arrange(age_ka) #%>% 
+  # mutate(age_ka = jitter(age_ka))
 

### 2019-12-04 analysis/paper/paper.Rmd
Trying to convert the codes working with google sheet to codes with xlxs

@@
 
+# without site name
+ggplot(kasv_tidy,
+       aes(date_age / 1000,
+           artefact_density)) +
+  geom_point(aes(size = total_artifacts)) +
+  ylab("Artefact density") +
+  xlab("Age of assemblage (ka)") + 
+  scale_size_continuous(name = "Total number\nof artifacts")  +
+  theme_minimal(base_size = 16)
+ggsave("figures/age-by-density.png")
+
 ```
 
+```{r}
+#the artifact composition of each sites to figure out what people did as they were occupying the sites.
+kasa_long <-
+  kasa %>%
+  gather(site, 
+         count, 
+         -X) %>% 
+  filter(!is.na(count)) %>% 
+  group_by(site) %>% 
+  mutate(percentage = count / sum(count, na.rm = TRUE) * 100,
+         total = sum(count, na.rm = TRUE)) %>% 
+  filter(!is.na(percentage)) %>% 
+  # if percentage is <10%, call it 'other'
+  mutate(artefact_type = ifelse(as.character(X) == 'stemmed_point',
+                                'stemmed_point', 
+                                ifelse(percentage >= 10, 
+                                as.character(X), "other_tools"))) %>% 
+  mutate(axis_label = as.character(glue('{site}\n(n = {total})')))  %>% 
+  # join to get ages of the sites
+  left_join(mydata_ages, 
+            by = c('site' = 'site_name')) %>% 
+  arrange(age_ka) #%>% 
+  # mutate(age_ka = jitter(age_ka))
+
+ggplot(kasa_long, 
+       aes(age_ka, 
+           percentage,
+           fill = artefact_type)) +
+  geom_col(position = "fill") +
+  ylab("Percentage") +
+  xlab("Age of assemblage (ka)") +
+    scale_fill_viridis_d(name = "Artifact type",
+                       option = "D") +
+  coord_flip() +
+  theme_minimal(base_size = 16)
+
+# comparable figure to the raw material one (BM created)
+
+kasa_long1 <-
+  kasa %>%
+  gather(site, 
+         count, 
+         -X) %>% 
+  filter(!is.na(count)) %>% 
+  group_by(site) %>% 
+   mutate(percentage = count / sum(count, na.rm = TRUE) * 100,
+         total = sum(count, na.rm = TRUE)) %>% 
+  filter(!is.na(percentage)) %>% 
+  # if percentage is <10%, call it 'other'
+  mutate(artefact_type = ifelse(as.character(X) == 'stemmed_point',
+                                'stemmed_point', 
+                                ifelse(percentage >= 10 | X == "unkown", 
+                                as.character(X), "other_tools"))) %>% 
+  mutate(axis_label = glue('{site}\n(n = {total})'))  %>% 
+  select(-X)
+  # join to get ages of the sites
+kasa_long2 <- 
+kasa_long1 %>% 
+  group_by(site, 
+           artefact_type) %>% 
+  summarise(percentage = sum(percentage)) %>% 
+  left_join(mydata_ages, 
+            by = c('site' = 'site_name')) %>% 
+  arrange(age_ka) %>% 
+  distinct(.keep_all = TRUE) %>%
+  filter(!is.na(age_ka)) %>% 
+  mutate(axis_label = glue('{site} ({round(age_ka,1)} ka)'))
+
+ggplot(kasa_long2, 
+       aes(
+           reorder(axis_label, -age_ka),
+           percentage,
+           fill = artefact_type)) +
+  geom_col(position = "fill") +
+  ylab("Percentage") +
+   xlab("Assemblage (youngest at the top)") +
+  theme_minimal(base_size = 16)  +
+    scale_fill_viridis_d(name = "Artefact type",
+                       option = "D") +
+  coord_flip() +
+  theme_minimal(base_size = 12) 
+ggsave("figures/artefact-types.png")
+
+```
 Artifact density and retouch <- maybe we need to add the result of retouch, separately 
@@
 
-```{r}
-#the artifact composition of each sites to figure out what people did as they were occupying the sites.
-kasa_long <-
-  kasa %>%
-  gather(site, 
-         count, 
-         -X) %>% 
-  filter(!is.na(count)) %>% 
-  group_by(site) %>% 
-  mutate(percentage = count / sum(count, na.rm = TRUE) * 100,
-         total = sum(count, na.rm = TRUE)) %>% 
-  filter(!is.na(percentage)) %>% 
-  # if percentage is <10%, call it 'other'
-  mutate(artefact_type = ifelse(as.character(X) == 'stemmed_point',
-                                'stemmed_point', 
-                                ifelse(percentage >= 10, 
-                                as.character(X), "other_tools"))) %>% 
-  mutate(axis_label = as.character(glue('{site}\n(n = {total})')))  %>% 
-  # join to get ages of the sites
-  left_join(mydata_ages, 
-            by = c('site' = 'site_name')) %>% 
-  arrange(age_ka) #%>% 
-  # mutate(age_ka = jitter(age_ka))
-
-ggplot(kasa_long, 
-       aes(age_ka, 
-           percentage,
-           fill = artefact_type)) +
-  geom_col() +
-  ylab("Percentage") +
-  xlab("Age of assemblage (ka)") +
-  labs(fill="Artifact type")+
-  theme_minimal(base_size = 16)
-
-```
-
 Get the lithic data for some sites:
@@
   map(sheets, ~kas_sites %>% gs_read_listfeed(ws = .x))
+```
+
+```{r}
+
+# try not to use googlesheet and make the upper chunck work
+
+library(readxl)
+path <- "analysis/data/raw_data/Overall_info.xlsx"
+excel_sheets(path = path)
+
+p_sheets <- c("general_info" ,"assemblage_info")
+
+k_kas_sites_sheets <- map(p_sheets, ~path %>% )
+
 ```

### 2019-12-04 analysis/paper/paper.Rmd
working on the newer intro with three quesetions. I want to paraphrase the questions with more details and make connection to actual analyses. I feel nervous about writing with no Grammerly :/

@@
 <!-- one equal-sized paragraph for each of our three questions, with lots of recent, specific citations for each Q -->
+
+<!-- this is what I'm newly writing 
+In this research, we try to answer the three questions to understand the technological transition period.
+First, we are asking how did the introduction of the new tools change the usage of landscape.
+The new technology might allow hunter-gatherers to explore and stay in different or harsher territories by increasing mobility as well as innovating hunting strategies [].
+Or people could change their landscape using the new tools such as [].
+
+Second, we focus on how did the new technology change the way people use their sites whether they tended to stay in one spot longer as a base campe or shorter period for specific purpose such as summer hunting camp.
+The stone artifact assemblages from a site indicates how long a group stayed or what they did so we can assume the way people use the site [].
+This project will look into stone assemblages that hunter-gatherers left behind to understand the site occupation pattern change during the transition period.
+
+Lastly, 
+and the population dynamics.
+
+--->
 
@@
 
-k_kas_sites_sheets <- map(p_sheets, ~path %>% )
+#k_kas_sites_sheets <- map(p_sheets, ~path %>% )
+
+#simply combine the two data sets of general info and assemblage info?
+#kk_sheets <- c (mydata, kasa)
 

### 2019-12-05 analysis/paper/paper.Rmd
no subsistence

@@
 We find no strong relationship between technology and ecological change.
-Instead, we assume that population dynamics might have had more impact on changes in stone artifact technology and hunter-gatherer subsistence.
+Instead, we assume that population dynamics might have had more impact on changes in stone artifact technology and hunter-gatherer foraging strategies.
 keywords: |
@@
 
-Stone artifact assemblages have important information about hunter-gatherer mobility and subsistence strategy, according to the behavioral ecological theory that links technological changes to human adaptation to the surrounding environment.
+Stone artifact assemblages have important information about hunter-gatherer mobility and foraging strategies, according to the behavioral ecological theory that links technological changes to human adaptation to the surrounding environment.
 Our research focuses on technological change during the Late Pleistocene in Korea, around ~40-35 ka.
@@
 This technological change has been studied related to the issues of the transition from the Early to the Late Paleolithic and the migration of Homo sapiens.
-However, the studies rarely consider the relationship between the technological transition and change in human subsistence strategies.
+However, the studies rarely consider the relationship between the technological transition and change in human foraging strategies.
 The most Korean Paleolithic sites do not preserve much information to understand hunter-gatherers or the transition in their tool kit due to a lack of material evidence.
@@
 In this research, we try to answer the three questions to understand the technological transition period.
-First, we are asking how did the introduction of the new tools change the usage of landscape.
+First, we are asking how the introduction of the new tools changed the usage of landscape.
 The new technology might allow hunter-gatherers to explore and stay in different or harsher territories by increasing mobility as well as innovating hunting strategies [].
 Or people could change their landscape using the new tools such as [].
+According to ethnographic observations of mobility, people move frequently with little knowledge of the places for foraging and acquiring raw materials purpose with portable, mulifunctional, long lasting tools[@clarkson2008lithics;@kuhn2014mousterian].
+The study reported here was designed to
 
 Second, we focus on how did the new technology change the way people use their sites whether they tended to stay in one spot longer as a base campe or shorter period for specific purpose such as summer hunting camp.
-The stone artifact assemblages from a site indicates how long a group stayed or what they did so we can assume the way people use the site [].
-This project will look into stone assemblages that hunter-gatherers left behind to understand the site occupation pattern change during the transition period.
+The stone artifact assemblage from a site can inform us how long individuals or groupd stayed or what they did there so we can assume the way people use the site.
+It is because a stone artifact
+contains a record of material transport and selection and it was carefully designed so that people could make a living from the landscape that constraints the production, efficiency, use, transportation, and maintenance [@clarkson2008lithics].
+This project will look into stone assemblages that hunter-gatherers left behind to understand the site occupation pattern change during the transition period and the relationship with the appearance of new tools.
 
-Lastly, 
-and the population dynamics.
+Lastly, we want to know about population dynamics during the transition period based on a premise that population change could be a driver of the new invention as well as the source of it if they brought the new tools from outside.
+The demands derived from significan population dynamic in addition to environmental change could lead to development of the tools [buchanan2016drivers; @mihailovic2016technological; prufer2019linking].
+Some studies argue that technological complexity and richness are highly influenced by population size because in large population, because new innovation could occur during the process of learning from others [@shennan2001demography; @henrich2004demography; @collared2005causes; @collard2013risk; @buchanan2016drivers].
+Therefore, we will check the population change during the transitional period to understand its impact on the technology.
 
@@
 (2) was there any change in the way people used sites?
-(3) was there any population change and did it impact on technology or subsistence pattern?
+(3) was there any population change and did it impact on technology or foraging pattern?
 
@@
 The transition to the Upper Paleolihtic in Japan is under the long term debate due to the uncertainty about the existence of the previous period in the Archipelago.
-But there are several features represent the milestone of the new era such as systematic procurement and reduction of lithic raw materials, trap-pit hunting method, watercraft (maritime transport), bow-and-arrow technology, and standardization of lithic tools which can show the appearance of new subsistence strategies possibly driven by risks of increased population and ecological change [@Yamaoka_2012; @bae2017origin; @Nakazawa_Bae_2018; @morisaki2019early].
+But there are several features represent the milestone of the new era such as systematic procurement and reduction of lithic raw materials, trap-pit hunting method, watercraft (maritime transport), bow-and-arrow technology, and standardization of lithic tools which can show the appearance of new foraging strategies possibly driven by risks of increased population and ecological change [@Yamaoka_2012; @bae2017origin; @Nakazawa_Bae_2018; @morisaki2019early].
 The technological transition in Korea is mainly characterized by the early appearance of stemmed points and following blade technology.
@@
 As a part of composite implements such as arrowheads or spearheads, the point was combined with other materials and it could be easily replaced reducing waste of total disposal [@Keeley_1982; @kuhn2015artifacts].
-The new projectile point also indicates a new way of hunting methods or other subsistence strategies because the throwing technique enables to earn a long-distance target.
+The new projectile point also indicates a new way of hunting methods or other foraging strategies because the throwing technique enables to earn a long-distance target.
 In addition, this new tool spread to adjacent regions such as Japanese Archipelago by combination of watercraft technique and social network so that the study of stemmed points play an important role in understanding about peopling in Japan and the relationship between the Peninsula and Archipelago.
-This project focuses on the functional aspect to understand the technology and subsistence pattern based on the concept of HBE.
+This project focuses on the functional aspect to understand the technology and foraging pattern based on the concept of HBE.
 
@@
 We count the retouch frequency as the number of retouched tools divided by the total number of artifacts.
-This value represents general proficiency of lithic manufacturing, raw material availability, and subsistence strategies including mobility and duration of site occupation based on the concept that tool reductioning is an economical tactic to generate sharp, usable edges while minimizing the cost of transporting tools or bulky raw materials by reducing the total load weight [@Kuhn_1990; @clark2019landscapes].
+This value represents general proficiency of lithic manufacturing, raw material availability, and foraging strategies including mobility and duration of site occupation based on the concept that tool reductioning is an economical tactic to generate sharp, usable edges while minimizing the cost of transporting tools or bulky raw materials by reducing the total load weight [@Kuhn_1990; @clark2019landscapes].
 In other words, producing and carrying highly retouched tools will increase portability and efficiency by decrease in the size of both individual tools and the assemblages so that hunter-gatherers could move often or further, or both [@kuhn1991unpacking; @kuhn1994formal; @kuhn2004upper; @andrefsky1994raw; @davies2018modeling].

### 2019-12-05 analysis/paper/paper.Rmd
added definitions

@@
 
---->
+-->
 
@@
 
-The research analyzes total 33 assemblages [need to explain what is the difference between 'assemblage' and 'site'] and they consist of cores, blade cores, blades, flakes, debris, hammers, choppers, plans, polyhedrals, side scrapers, end scrapers, notches, cleavers, stemmed points, awls, denticulates, burins, handaxes, blanks, knives, handadzes, peaks, flatters, flakers etc,
+The research analyzes total 33 assemblages [need to explain what is the difference between 'assemblage' and 'site'] and they consist of cores, blade cores, blades, flakes, debris, hammers, choppers, plans, polyhedrals, side scrapers, end scrapers, notches, cleavers, stemmed points, awls, denticulates, burins, handaxes, blanks, knives, handadzes, peaks, flatters, flakers etc.
+
+<!-- GP needs to explain what is the difference between 'assemblage' and 'site', need to be very precise about the definitions of these: ‘assemblage’ and ‘layer’.
+Don’t say ‘layer’ at all, say ‘excavation unit’ or ‘stratigraphic unit’.
+Add a short example of how you identify multiple assemblages in one site (one sentence).
+=> I don't know how to put these definitions in the manuscript naturally.
+
+- Assemblage is a set of artifacts that are found in same exacation unit so archaeologists assume that they are associated each other in terms of period and used by same group of people (GP).
+
+- Assemblage is a group of artifacts recurring together at a particular time and place, and representing the sum of human activities (Renfrew and Bahn 1996).
+
+- Site is a place where past activity is preserved with material evidence.
+
+- This research analyzes total 33 assemblages from 22 sites. Some sites have mutiple  assemblages from differnt stratigraphic units.
+
+-->
 

### 2019-12-05 analysis/paper/paper.Rmd
the three paragraphs done

@@
 
-<!-- this is what I'm newly writing 
 In this research, we try to answer the three questions to understand the technological transition period.
-First, we are asking how the introduction of the new tools changed the usage of landscape.
-The new technology might allow hunter-gatherers to explore and stay in different or harsher territories by increasing mobility as well as innovating hunting strategies [].
-Or people could change their landscape using the new tools such as [].
+First, we are asking how the introduction of the new tools related to the usage of landscape.
+The new technology might allow hunter-gatherers to explore and stay in different or harsher territories by increasing mobility as well as innovating foraging strategies.
+In addition, the new tool production technique required hunter-gatherers to move for acquiring specific raw materials [@blumenschine2008effects; @reti2016quantifying].
 According to ethnographic observations of mobility, people move frequently with little knowledge of the places for foraging and acquiring raw materials purpose with portable, mulifunctional, long lasting tools[@clarkson2008lithics;@kuhn2014mousterian].
-The study reported here was designed to
+The study reported here was designed to interpret the impact of the new tools on the landscape use during the transitional period.
 
-Second, we focus on how did the new technology change the way people use their sites whether they tended to stay in one spot longer as a base campe or shorter period for specific purpose such as summer hunting camp.
-The stone artifact assemblage from a site can inform us how long individuals or groupd stayed or what they did there so we can assume the way people use the site.
-It is because a stone artifact
-contains a record of material transport and selection and it was carefully designed so that people could make a living from the landscape that constraints the production, efficiency, use, transportation, and maintenance [@clarkson2008lithics].
+<!-- I think we were also talking about example that a new tech actually changed the landscape. But I'm not sure where I can put the example
+
+For example, the new technology could change their landscape.
+Hominin fire use in the early Pleistocene has been studied and some argue that usage of fire changed the landscape and their dietary [@pruetz2010brief; @glikson2013fire; @parker2016pyrophilic; @hlubik2019hominin].
+
+-->
+
+Second, we focus on how the new technology changed the way people use their sites whether they tended to stay in one spot longer as a base campe or shorter period for specific purpose such as summer hunting camp.
+The stone artifact assemblage from a site can inform us how long individuals or groupd stayed or what they did there so we can assume the way people use the site [@binford1979organization; @kelly1995foraging; @kuhn2016moves; @holdaway2019surface].
+Stone artifact contains a record of material transport and selection because it was carefully designed so that people could make a living from the landscape that constraints the production, efficiency, use, transportation, and maintenance [@clarkson2008lithics].
+Archaeological sites indicate depositional phenomena and places where artifacts and other debris accumulated reflecting human activities in the past [@kuhn2015artifact].
 This project will look into stone assemblages that hunter-gatherers left behind to understand the site occupation pattern change during the transition period and the relationship with the appearance of new tools.
@@
 
--->
+-------------- previous paragraphs ------
 
@@
 
-- Site is a place where past activity is preserved with material evidence.
+- Site is a place where past activity is preserved with material evidence (GP).
+
+- What we call sites are essentially depositional phenomena, places where, for behavioral or geologic reasons, artifacts and other debris accumulated in sufficient concentrations to attract the attention of archeologists (Kuhn and Clark 2015).
+
+==> I put "Archaeological sites indicate depositional phenomena and places where artifacts and other debris accumulated reflecting human activities in the past" in the paragraph of the second question.
 

### 2019-12-05 analysis/paper/paper.Rmd
stash

--- 2019-12-05 analysis/paper/paper.Rmd
@@
 We find no strong relationship between technology and ecological change.
-Instead, we assume that population dynamics might have had more impact on changes in stone artifact technology and hunter-gatherer foraging strategies.
+Instead, we assume that population dynamics might have had more impact on changes in stone artifact technology and hunter-gatherer subsistence.
 keywords: |
@@
 library(ggrepel)
-library(here)
-
-korean_archaeological_sites <- readxl::read_excel(here("analysis/data/raw_data/korean-archaeologica-sites.xlsx"))
 
 # data from PhD data sheet, not KAS sheet.
-mydata <- read.csv(here("analysis/data/raw_data/General_info.csv"))
+mydata <- read.csv("~/koreapaleolithicmobilityoccupation/General_info.csv")
 
 # raw material data from KAS data sheet.
-kasr <- read.csv(here("analysis/data/raw_data/Rawmaterial_info.csv"))
+kasr <- read.csv("~/koreapaleolithicmobilityoccupation/Rawmaterial_info.csv")
 
 # assemblage composition data from KAS data sheet.
-kasa <- read.csv(here("analysis/data/raw_data/Assemblage_info.csv"))
+kasa <- read.csv("~/koreapaleolithicmobilityoccupation/Assemblage_info.csv")
 
 #volume of the cultural layer from KAS data sheet.
-kasv <- read.csv(here("analysis/data/raw_data/Dating_info.csv"))
+kasv <- read.csv("~/koreapaleolithicmobilityoccupation/Dating_info.csv")
 
-dates <- read.csv(here("analysis/data/raw_data/Radiocarbon dates.csv"))
+dates <- read.csv("~/koreapaleolithicmobilityoccupation/Radiocarbon dates.csv")
 
@@
 
-Stone artifact assemblages have important information about hunter-gatherer mobility and foraging strategies, according to the behavioral ecological theory that links technological changes to human adaptation to the surrounding environment.
+Stone artifact assemblages have important information about hunter-gatherer mobility and subsistence strategy, according to the behavioral ecological theory that links technological changes to human adaptation to the surrounding environment.
 Our research focuses on technological change during the Late Pleistocene in Korea, around ~40-35 ka.
@@
 This technological change has been studied related to the issues of the transition from the Early to the Late Paleolithic and the migration of Homo sapiens.
-However, the studies rarely consider the relationship between the technological transition and change in human foraging strategies.
+However, the studies rarely consider the relationship between the technological transition and change in human subsistence strategies.
 The most Korean Paleolithic sites do not preserve much information to understand hunter-gatherers or the transition in their tool kit due to a lack of material evidence.
@@
 
-```{r demo, eval = FALSE}
-1 + 1
-```
-
 ## Theory and background to our three questions
 
-<!-- one equal-sized paragraph for each of our three questions, with lots of recent, specific citations for each Q -->
+<!-- one equal-sized paragraph for each of our three questions, with lots of recent, specific citations for each Q
 
-<!-- this is what I'm newly writing 
-In this research, we try to answer the three questions to understand the technological transition period.
-First, we are asking how the introduction of the new tools changed the usage of landscape.
-The new technology might allow hunter-gatherers to explore and stay in different or harsher territories by increasing mobility as well as innovating hunting strategies [].
-Or people could change their landscape using the new tools such as [].
-According to ethnographic observations of mobility, people move frequently with little knowledge of the places for foraging and acquiring raw materials purpose with portable, mulifunctional, long lasting tools[@clarkson2008lithics;@kuhn2014mousterian].
-The study reported here was designed to
+State the questions very clearly and give background into relevant literature:
 
-Second, we focus on how did the new technology change the way people use their sites whether they tended to stay in one spot longer as a base campe or shorter period for specific purpose such as summer hunting camp.
-The stone artifact assemblage from a site can inform us how long individuals or groupd stayed or what they did there so we can assume the way people use the site.
-It is because a stone artifact
-contains a record of material transport and selection and it was carefully designed so that people could make a living from the landscape that constraints the production, efficiency, use, transportation, and maintenance [@clarkson2008lithics].
-This project will look into stone assemblages that hunter-gatherers left behind to understand the site occupation pattern change during the transition period and the relationship with the appearance of new tools.
+ -  the population: that some relationship may exist between population as a cause of techno change: Torrence, Collard, Shennan
 
-Lastly, we want to know about population dynamics during the transition period based on a premise that population change could be a driver of the new invention as well as the source of it if they brought the new tools from outside.
-The demands derived from significan population dynamic in addition to environmental change could lead to development of the tools [buchanan2016drivers; @mihailovic2016technological; prufer2019linking].
-Some studies argue that technological complexity and richness are highly influenced by population size because in large population, because new innovation could occur during the process of learning from others [@shennan2001demography; @henrich2004demography; @collared2005causes; @collard2013risk; @buchanan2016drivers].
-Therefore, we will check the population change during the transitional period to understand its impact on the technology.
+ - the way people used the landscape: changing mobility patterns due to risk management,  Kelly, Fitzhugh, Shott
 
---->
+ - the way people used sites: more fine-grained evidence about land use patterns long vs short term durations of site use Kuhn, Andresfsky, Hiscock
+
+-->
 
@@
 (2) was there any change in the way people used sites?
-(3) was there any population change and did it impact on technology or foraging pattern?
+(3) was there any population change and did it impact on technology or subsistence pattern?
 
@@
 The transition to the Upper Paleolihtic in Japan is under the long term debate due to the uncertainty about the existence of the previous period in the Archipelago.
-But there are several features represent the milestone of the new era such as systematic procurement and reduction of lithic raw materials, trap-pit hunting method, watercraft (maritime transport), bow-and-arrow technology, and standardization of lithic tools which can show the appearance of new foraging strategies possibly driven by risks of increased population and ecological change [@Yamaoka_2012; @bae2017origin; @Nakazawa_Bae_2018; @morisaki2019early].
+But there are several features represent the milestone of the new era such as systematic procurement and reduction of lithic raw materials, trap-pit hunting method, watercraft (maritime transport), bow-and-arrow technology, and standardization of lithic tools which can show the appearance of new subsistence strategies possibly driven by risks of increased population and ecological change [@Yamaoka_2012; @bae2017origin; @Nakazawa_Bae_2018; @morisaki2019early].
 The technological transition in Korea is mainly characterized by the early appearance of stemmed points and following blade technology.
@@
 As a part of composite implements such as arrowheads or spearheads, the point was combined with other materials and it could be easily replaced reducing waste of total disposal [@Keeley_1982; @kuhn2015artifacts].
-The new projectile point also indicates a new way of hunting methods or other foraging strategies because the throwing technique enables to earn a long-distance target.
+The new projectile point also indicates a new way of hunting methods or other subsistence strategies because the throwing technique enables to earn a long-distance target.
 In addition, this new tool spread to adjacent regions such as Japanese Archipelago by combination of watercraft technique and social network so that the study of stemmed points play an important role in understanding about peopling in Japan and the relationship between the Peninsula and Archipelago.
-This project focuses on the functional aspect to understand the technology and foraging pattern based on the concept of HBE.
+This project focuses on the functional aspect to understand the technology and subsistence pattern based on the concept of HBE.
 
@@
 ```{r cache = TRUE, fig.cap="Map of sites discussed in the text."}
+library(here)
+
+korean_archaeological_sites <- readxl::read_excel(here("analysis/data/raw_data/korean-archaeologica-sites.xlsx"))
+
+library(tidyverse)
 
@@
 
+<!-- we grouped the stone artefacts into 33 assemblages based on the depositional context and chronology...
+We grouped together similar stratigraphic units into one assemblage...
+the appearance of new artefact types such as stemmed points and blades were used to separate assemblages in the same site.
+For example...
+one site that is an example of how you identified multiple assembalges in site -->
+
 The research analyzes total 33 assemblages [need to explain what is the difference between 'assemblage' and 'site'] and they consist of cores, blade cores, blades, flakes, debris, hammers, choppers, plans, polyhedrals, side scrapers, end scrapers, notches, cleavers, stemmed points, awls, denticulates, burins, handaxes, blanks, knives, handadzes, peaks, flatters, flakers etc,
@@
 We count the retouch frequency as the number of retouched tools divided by the total number of artifacts.
-This value represents general proficiency of lithic manufacturing, raw material availability, and foraging strategies including mobility and duration of site occupation based on the concept that tool reductioning is an economical tactic to generate sharp, usable edges while minimizing the cost of transporting tools or bulky raw materials by reducing the total load weight [@Kuhn_1990; @clark2019landscapes].
+This value represents general proficiency of lithic manufacturing, raw material availability, and subsistence strategies including mobility and duration of site occupation based on the concept that tool reductioning is an economical tactic to generate sharp, usable edges while minimizing the cost of transporting tools or bulky raw materials by reducing the total load weight [@Kuhn_1990; @clark2019landscapes].
 In other words, producing and carrying highly retouched tools will increase portability and efficiency by decrease in the size of both individual tools and the assemblages so that hunter-gatherers could move often or further, or both [@kuhn1991unpacking; @kuhn1994formal; @kuhn2004upper; @andrefsky1994raw; @davies2018modeling].
@@
 
+mydata <- korean_archaeological_sites
+
 mydata_ages <- 
   mydata %>% 
-  separate(C14.BP., into = c('age', 'error'),
+  separate(`C14(BP)`, into = c('age', 'error'),
           sep = "±") %>% 
   mutate(age_ka = parse_number(age) / 1000,
-         error = parse_number(error)) %>% 
-  mutate(has_sp = ifelse(!is.na(SP.), "yes", "no"))
+         error = parse_number(error),
+         `altitude(m)_of_main_layer` = parse_number(`altitude(m)_of_main_layer`)) %>% 
+  mutate(has_sp = ifelse(!is.na(`SP#`), "yes", "no"))
 
@@
 
+ggsave(here("analysis/figures/site-elevation-by-age.png"))
 ```
@@
 
-# without site name
-ggplot(kasv_tidy,
-       aes(date_age / 1000,
-           artefact_density)) +
-  geom_point(aes(size = total_artifacts)) +
-  ylab("Artefact density") +
-  xlab("Age of assemblage (ka)") + 
-  scale_size_continuous(name = "Total number\nof artifacts")  +
-  theme_minimal(base_size = 16)
-ggsave("figures/age-by-density.png")
-
 ```
 
-```{r}
-#the artifact composition of each sites to figure out what people did as they were occupying the sites.
-kasa_long <-
-  kasa %>%
-  gather(site, 
-         count, 
-         -X) %>% 
-  filter(!is.na(count)) %>% 
-  group_by(site) %>% 
-  mutate(percentage = count / sum(count, na.rm = TRUE) * 100,
-         total = sum(count, na.rm = TRUE)) %>% 
-  filter(!is.na(percentage)) %>% 
-  # if percentage is <10%, call it 'other'
-  mutate(artefact_type = ifelse(as.character(X) == 'stemmed_point',
-                                'stemmed_point', 
-                                ifelse(percentage >= 10, 
-                                as.character(X), "other_tools"))) %>% 
-  mutate(axis_label = as.character(glue('{site}\n(n = {total})')))  %>% 
-  # join to get ages of the sites
-  left_join(mydata_ages, 
-            by = c('site' = 'site_name')) %>% 
-  arrange(age_ka) #%>% 
-  # mutate(age_ka = jitter(age_ka))
-
-ggplot(kasa_long, 
-       aes(age_ka, 
-           percentage,
-           fill = artefact_type)) +
-  geom_col(position = "fill") +
-  ylab("Percentage") +
-  xlab("Age of assemblage (ka)") +
-    scale_fill_viridis_d(name = "Artifact type",
-                       option = "D") +
-  coord_flip() +
-  theme_minimal(base_size = 16)
-
-# comparable figure to the raw material one (BM created)
-
-kasa_long1 <-
-  kasa %>%
-  gather(site, 
-         count, 
-         -X) %>% 
-  filter(!is.na(count)) %>% 
-  group_by(site) %>% 
-   mutate(percentage = count / sum(count, na.rm = TRUE) * 100,
-         total = sum(count, na.rm = TRUE)) %>% 
-  filter(!is.na(percentage)) %>% 
-  # if percentage is <10%, call it 'other'
-  mutate(artefact_type = ifelse(as.character(X) == 'stemmed_point',
-                                'stemmed_point', 
-                                ifelse(percentage >= 10 | X == "unkown", 
-                                as.character(X), "other_tools"))) %>% 
-  mutate(axis_label = glue('{site}\n(n = {total})'))  %>% 
-  select(-X)
-  # join to get ages of the sites
-kasa_long2 <- 
-kasa_long1 %>% 
-  group_by(site, 
-           artefact_type) %>% 
-  summarise(percentage = sum(percentage)) %>% 
-  left_join(mydata_ages, 
-            by = c('site' = 'site_name')) %>% 
-  arrange(age_ka) %>% 
-  distinct(.keep_all = TRUE) %>%
-  filter(!is.na(age_ka)) %>% 
-  mutate(axis_label = glue('{site} ({round(age_ka,1)} ka)'))
-
-ggplot(kasa_long2, 
-       aes(
-           reorder(axis_label, -age_ka),
-           percentage,
-           fill = artefact_type)) +
-  geom_col(position = "fill") +
-  ylab("Percentage") +
-   xlab("Assemblage (youngest at the top)") +
-  theme_minimal(base_size = 16)  +
-    scale_fill_viridis_d(name = "Artefact type",
-                       option = "D") +
-  coord_flip() +
-  theme_minimal(base_size = 12) 
-ggsave("figures/artefact-types.png")
-
-```
 Artifact density and retouch <- maybe we need to add the result of retouch, separately 
@@
 
-To understand the population change during the appearance of stemmed points, we analyzed the summed probability distributions of 171 radiocarbon dates from 33 cultural layers to get a rough image of population dynamics at the sites.
-The result indicates that there is a higher density of radiocarbon dates from around 34 ka.
+To understand the population change during the appearance of stemmed points, we analyzed the summed probability distributions of 171 radiocarbon dates from 33 cultural layers and it indicates that there is a higher density of radiocarbon dates from around 34 ka.
 We could cautiously assume that the population flux might relate to the technological innovation.
 
-```{r cache=TRUE}
+```{r}
 dates_clean <- 
@@
 
+```{r}
+#the artifact composition of each sites to figure out what people did as they were occupying the sites.
+kasa_long <-
+  kasa %>%
+  gather(site, 
+         count, 
+         -X) %>% 
+  filter(!is.na(count)) %>% 
+  group_by(site) %>% 
+  mutate(percentage = count / sum(count, na.rm = TRUE) * 100,
+         total = sum(count, na.rm = TRUE)) %>% 
+  filter(!is.na(percentage)) %>% 
+  # if percentage is <10%, call it 'other'
+  mutate(artefact_type = ifelse(as.character(X) == 'stemmed_point',
+                                'stemmed_point', 
+                                ifelse(percentage >= 10, 
+                                as.character(X), "other_tools"))) %>% 
+  mutate(axis_label = glue('{site}\n(n = {total})'))  %>% 
+  # join to get ages of the sites
+  left_join(mydata_ages, 
+            by = c('site' = 'site_name')) %>% 
+  arrange(age_ka) %>% 
+   mutate(age_ka = jitter(age_ka))
+
+ggplot(kasa_long, 
+       aes(age_ka, 
+           percentage,
+           fill = artefact_type)) +
+  geom_col() +
+  ylab("Percentage") +
+  xlab("Age of assemblage (ka)") +
+  labs(fill="Artifact type")+
+  theme_minimal(base_size = 16)
+
+```
+
 Get the lithic data for some sites:
@@
   map(sheets, ~kas_sites %>% gs_read_listfeed(ws = .x))
-```
-
-```{r}
-
-# try not to use googlesheet and make the upper chunck work
-
-library(readxl)
-path <- "analysis/data/raw_data/Overall_info.xlsx"
-excel_sheets(path = path)
-
-p_sheets <- c("general_info" ,"assemblage_info")
-
-#k_kas_sites_sheets <- map(p_sheets, ~path %>% )
-
-#simply combine the two data sets of general info and assemblage info?
-#kk_sheets <- c (mydata, kasa)
-
 ```

### 2019-12-05 analysis/paper/paper.Rmd

--- 2019-12-05 analysis/paper/paper.Rmd
@@
 
-```{r demo, eval = FALSE}
-1 + 1
-```
-
 ## Theory and background to our three questions
 
-<!-- one equal-sized paragraph for each of our three questions, with lots of recent, specific citations for each Q -->
+<!-- one equal-sized paragraph for each of our three questions, with lots of recent, specific citations for each Q
+
+State the questions very clearly and give background into relevant literature:
+
+ -  the population: that some relationship may exist between population as a cause of techno change: Torrence, Collard, Shennan
+
+ - the way people used the landscape: changing mobility patterns due to risk management,  Kelly, Fitzhugh, Shott
+
+ - the way people used sites: more fine-grained evidence about land use patterns long vs short term durations of site use Kuhn, Andresfsky, Hiscock
+
+-->
 

### 2019-12-05 analysis/paper/paper.Rmd
intro & bkgnd

@@
 ---
-title: "Change in Mobility and Site Occupation during the Late Pleistocene in Korea"
+title: "Changes in Mobility and Site Occupation during the Late Pleistocene in Korea"
 author:
@@
 
-Stone artifact assemblages have important information about hunter-gatherer mobility and foraging strategies, according to the behavioral ecological theory that links technological changes to human adaptation to the surrounding environment.
-Our research focuses on technological change during the Late Pleistocene in Korea, around ~40-35 ka.
-We examined stone artifacts from 22 sites in South Korea to investigate the technological transition represented by the appearance of a new tool, stemmed points, and the following blade industry.
-The overall aim is to understand how the appearance of stemmed points affected hunter-gatherers in Korea, which leads to three specific questions to pursue: how did the introduction of the new tools change the way people used the landscape, the way people used sites, and the population.
-This technological change has been studied related to the issues of the transition from the Early to the Late Paleolithic and the migration of Homo sapiens.
-However, the studies rarely consider the relationship between the technological transition and change in human foraging strategies.
-The most Korean Paleolithic sites do not preserve much information to understand hunter-gatherers or the transition in their tool kit due to a lack of material evidence.
-Therefore, we apply theoretical hypotheses to understand the relationship between hunter-gatherers, technology, and the environment.
-Using Human Behavioral Ecology (HBE) driven models, we hypothesized that hunter-gatherers should move frequently and further in order to adapt to a changing environment, so their tools became more portable and efficient, as indicated by decreases in the size of both individual tools and toolkit and increases in function and utility.
+The appearance of stemmed points in the Late Pleistocene (~40-35 ka) in Korea is often assumed to have transformed forager lifeways because it it thought to reflect more specialed hunting practices (citations).
+Stemmed points are considered to be projectile weapon tips because the stem of these tools is cconsidered to represent halfting technology that joins the lithic to a wooden shaft.
+Stemmed points represent the first appearance of halfting technology in Korea.
+Here we explore questions about these transformations using behavioural ecology to investigate changes in the way people used the landscape, and the way people used sites.
+We examined stone artifacts from 22 sites in South Korea to investigate the technological transition represented by the appearance of a new tool, stemmed points, and the blade industry that followed.
+Previous work on this technological change has focussed on issues of the transition from the Early to the Late Paleolithic and the disperal of _Homo sapiens_ (citations).
+To our knowledge, the implications for forager mobility and site use have not been explored.
+Most Korean Paleolithic sites do not preserve organic remains to understand forager subsistence economies, so we lack direct faunal evidence on hunting strategies.
+We attempt to identify patterns of mobility through analysis of the stone artefact assemblages using Human Behavioral Ecology (HBE) models.
+Drawing on established relationships between lithic technology and forager mobility (citations, Shott, Kuhn) we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that are part of a broader strategy of moving frequently and further, possibly as an adaptation to population or environmntal changes.
+
+<!--
+Need something about connection between stemmed point and specifics of hunting, what kind delivery system? What kind of game?
+-->
 
@@
 
-In this research, we try to answer the three questions to understand the technological transition period.
-First, we are asking how the introduction of the new tools related to the usage of landscape.
+We ask two questions to invetigate the period around the introduction of stemmed points.
+First, we are asking how the introduction of the new tools changed forager landscape use, specifically, was the use of stemmed points associated with occupation of marginal or exterme environments?
 The new technology might allow hunter-gatherers to explore and stay in different or harsher territories by increasing mobility as well as innovating foraging strategies.
@@
 
-The data analyzed in this paper are gathered from published excavation reports of 22 Paleolithic sites (33 cultural layers) in South Korea dating from 49k BP to 24k BP.
+The data analyzed in this paper are gathered from published excavation reports of 22 Paleolithic open air sites in South Korea dating from 49k BP to 24k BP.
+Within these 22 sites we have 33 assemblages.
+Multiple assemblages in a site were distinguised culturally sterile deposits in between artefact-bearing deposits, or by stratigraphic units identified by major differences in the texture and composition of the sedimentary deposit containing the stone artefacts.
 After the first excavation at Jeongokri site in the late 1970s, more than 200 Paleolithic sites are discovered in Korea.
-Among them, we  chose 22 sites based on the existence of absolute dates that span our period of interest.
+Among them, we  chose 22 sites based on absolute dates that span our period of interest.
 We include sites dated before and after the transition period (40-35 ka) to analyze the process of technological transition.
@@
 The research analyzes total 33 assemblages [need to explain what is the difference between 'assemblage' and 'site'] and they consist of cores, blade cores, blades, flakes, debris, hammers, choppers, plans, polyhedrals, side scrapers, end scrapers, notches, cleavers, stemmed points, awls, denticulates, burins, handaxes, blanks, knives, handadzes, peaks, flatters, flakers etc.
-
-<!-- GP needs to explain what is the difference between 'assemblage' and 'site', need to be very precise about the definitions of these: ‘assemblage’ and ‘layer’.
-Don’t say ‘layer’ at all, say ‘excavation unit’ or ‘stratigraphic unit’.
-Add a short example of how you identify multiple assemblages in one site (one sentence).
-=> I don't know how to put these definitions in the manuscript naturally.
-
-- Assemblage is a set of artifacts that are found in same exacation unit so archaeologists assume that they are associated each other in terms of period and used by same group of people (GP).
-
-- Assemblage is a group of artifacts recurring together at a particular time and place, and representing the sum of human activities (Renfrew and Bahn 1996).
-
-- Site is a place where past activity is preserved with material evidence (GP).
-
-- What we call sites are essentially depositional phenomena, places where, for behavioral or geologic reasons, artifacts and other debris accumulated in sufficient concentrations to attract the attention of archeologists (Kuhn and Clark 2015).
-
-==> I put "Archaeological sites indicate depositional phenomena and places where artifacts and other debris accumulated reflecting human activities in the past" in the paragraph of the second question.
-
-- This research analyzes total 33 assemblages from 22 sites. Some sites have mutiple  assemblages from differnt stratigraphic units.
-
--->
 

### 2019-12-20 analysis/paper/paper.Rmd
push before pull the current version
Revert "push before pull the current version"
This reverts commit b886a526bb85c54a9252cce442603792babeff94.

@@
 ```{r setup, include=FALSE}
-suppressMessages(library(tidyverse))
+suppressPackageStartupMessages(library(tidyverse))
 library(glue)
@@
 ```
+
+```{r}
+
+# join artefact type freqs with site data
+
+kasa %>% 
+  pivot_longer(-X, 
+               names_to = "site_name",
+               values_to = "count") %>% 
+  pivot_wider(names_from = "X",
+              values_from = "count") %>% 
+  left_join(mydata)
+
+```
+
 # Introduction
@@
 ```{r}
-# we'll use googlesheets package to get the data more efficiently
-library(googlesheets)
-
-# here is the key to the google sheet: this ggole sheet is same as kasa. so I could do kas_sites <- kasa 
-kas_sites <- gs_key("14G0Mp-YSsKBOBPF1Z1zPWQbZQ-sjJSWLetDIQSlbsjw")
-
-# get the tab names from this sheet: But that I couldn't use gs_ws_ls
-kas_sites_names <- gs_ws_ls(kas_sites)
-
-# sheets we need here
-sheets <- c( "general_info",    "assemblage_info" )
-
-# get the data from the tabs in this sheet!
-kas_sites_sheets <- 
-  map(sheets, ~kas_sites %>% gs_read_listfeed(ws = .x))
-```
-
-```{r}
 

### 2019-12-20 analysis/paper/paper.Rmd
working on intro and background

@@
 
-The appearance of stemmed points in the Late Pleistocene (~40-35 ka) in Korea is often assumed to have transformed forager lifeways because it it thought to reflect more specialed hunting practices (citations).
-Stemmed points are considered to be projectile weapon tips because the stem of these tools is cconsidered to represent halfting technology that joins the lithic to a wooden shaft.
+The appearance of stemmed points in the Late Pleistocene (~40-35 ka) in Korea is often assumed to have transformed forager lifeways because it is thought to reflect more specialed hunting practices (citations).
+Stemmed points are likely to have been projectile weapon tips because the stem of these tools is cconsidered to represent halfting technology that joins the lithic to a wooden shaft.
 Stemmed points represent the first appearance of halfting technology in Korea.
-Here we explore questions about these transformations using behavioural ecology to investigate changes in the way people used the landscape, and the way people used sites.
+Here we explore questions about these transformations using as behavioural ecology framework to explore changes in the way people used the landscape, and the way people used sites.
 We examined stone artifacts from 22 sites in South Korea to investigate the technological transition represented by the appearance of a new tool, stemmed points, and the blade industry that followed.
 Previous work on this technological change has focussed on issues of the transition from the Early to the Late Paleolithic and the disperal of _Homo sapiens_ (citations).
-To our knowledge, the implications for forager mobility and site use have not been explored.
-Most Korean Paleolithic sites do not preserve organic remains to understand forager subsistence economies, so we lack direct faunal evidence on hunting strategies.
-We attempt to identify patterns of mobility through analysis of the stone artefact assemblages using Human Behavioral Ecology (HBE) models.
-Drawing on established relationships between lithic technology and forager mobility (citations, Shott, Kuhn) we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that are part of a broader strategy of moving frequently and further, possibly as an adaptation to population or environmntal changes.
+To our knowledge, the implications for forager mobility and site use have not been explored in detail.
+Most Korean Paleolithic sites do not preserve organic remains, so we lack direct faunal evidence on hunting and subsistence strategies.
+Here, we attempt to identify patterns of mobility through analysis of the stone artefact assemblages using Human Behavioral Ecology (HBE) models.
+Drawing on established relationships between lithic technology and forager mobility (citations, Shott, Kuhn) we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that are part of a broader strategy of moving frequently and further, possibly as an adaptation to population or environmental changes.
 
-<!--
-Need something about connection between stemmed point and specifics of hunting, what kind delivery system? What kind of game?
--->
+<!-- sgenerally, how do we make sense of new stone tools...
+When a new stone artefact technology appears, how do we explain that?
+new species...
+population movements...
+(great basin)...
+environment...
+(Mesolithic & bow and arrow)...
+risk management (Australian small tool tradition) -->
 
@@
 
-## Theory and background to our three questions
+### Regional archaeology
 
-<!-- one equal-sized paragraph for each of our three questions, with lots of recent, specific citations for each Q
+In the Eastern hemisphere, striking technological innovations during the Late Pleistocene are restricted to Northeast Asia, such as northern China, Japan and Korea.
+These innovations include blade and micro technology, high frequencies of retouched blade tools, several novel refined tools such as projectile points, stemmed points, end scrapers, burins, denticulates, etc, and using high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary].
+Previous work has argued that the appearance of new stone artefact technologies in this region may be linked to modern human dispersals (citation).
 
-State the questions very clearly and give background into relevant literature:
-
- -  the population: that some relationship may exist between population as a cause of techno change: Torrence, Collard, Shennan
-
- - the way people used the landscape: changing mobility patterns due to risk management,  Kelly, Fitzhugh, Shott
-
- - the way people used sites: more fine-grained evidence about land use patterns long vs short term durations of site use Kuhn, Andresfsky, Hiscock
-
+The Shuidonggou site in northern China is a typical site for showing the transition from cobble to blade tool industries in the Late Pleistocene of Northeast Asia [@Brantingham_Perreault_2010; @Gao_Zhang_Yang_Shen_Wu_2010; @pei2012shuidonggou].
+Shuidonggou Localities SDG 2, 7-12 have blades and bladelets made of higher quality or exotic raw materials in addition to more than 80 polished ostrich egg-shell beads (SDG 2), bone needles and an awl (SDG 12) (citation).
+The blade compoenent is claimed to have come from Siberia and/or Mongolia at 41 ka, and at 10.5 ka  microblade technology appears but the origin is uncertain.
+<!-- expand on ages of all the things, comment about the cause of the new things at this site, check, https://link.springer.com/article/10.1007/s10963-019-09129-w  what is the debate?
 -->
 
-We ask two questions to invetigate the period around the introduction of stemmed points.
-First, we are asking how the introduction of the new tools changed forager landscape use, specifically, was the use of stemmed points associated with occupation of marginal or exterme environments?
-The new technology might allow hunter-gatherers to explore and stay in different or harsher territories by increasing mobility as well as innovating foraging strategies.
-In addition, the new tool production technique required hunter-gatherers to move for acquiring specific raw materials [@blumenschine2008effects; @reti2016quantifying].
-According to ethnographic observations of mobility, people move frequently with little knowledge of the places for foraging and acquiring raw materials purpose with portable, mulifunctional, long lasting tools[@clarkson2008lithics;@kuhn2014mousterian].
-The study reported here was designed to interpret the impact of the new tools on the landscape use during the transitional period.
-
-<!-- I think we were also talking about example that a new tech actually changed the landscape. But I'm not sure where I can put the example
-
-For example, the new technology could change their landscape.
-Hominin fire use in the early Pleistocene has been studied and some argue that usage of fire changed the landscape and their dietary [@pruetz2010brief; @glikson2013fire; @parker2016pyrophilic; @hlubik2019hominin].
-
+The origins of Late Pleistocene technological innovations in Japan is debated due to the uncertainty about the existence of people in the Archipelago before the appearance of new tools (citations).
+New technologies are represented by systematic procurement and reduction of lithic raw materials, trap-pit hunting methods, watercraft, bow-and-arrow technology, and standardization of lithic tools.
+<!-- need some ages for these types of evidence --> These have been interpreted as evidence of new foraging strategies possibly driven by increased population and ecological changes [@Yamaoka_2012; @bae2017origin; @Nakazawa_Bae_2018; @morisaki2019early].
+<!-- need to expand, mention some key site, blades appear when?
+what is the debate?
 -->
 
-Second, we focus on how the new technology changed the way people use their sites whether they tended to stay in one spot longer as a base campe or shorter period for specific purpose such as summer hunting camp.
-The stone artifact assemblage from a site can inform us how long individuals or groupd stayed or what they did there so we can assume the way people use the site [@binford1979organization; @kelly1995foraging; @kuhn2016moves; @holdaway2019surface].
-Stone artifact contains a record of material transport and selection because it was carefully designed so that people could make a living from the landscape that constraints the production, efficiency, use, transportation, and maintenance [@clarkson2008lithics].
+The Late Pleistocene technological transition in Korea is mainly characterized by the early appearance of stemmed points, followed by blade technology.
+The stemmed point is a projectile point made out of elongated flake or blade and with slight retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft.
+The earliest stemmped points were made on flakes.
+After blades appear at c.
+27 ka, the shape of stemmped points became more standardized with one or two ridges on dorsal side and triangular cross section.
+The stemmed point is often called as a tanged point in western hemisphere, but we use 'stemmed point' to distinguish from Bronze Age stone projectile points in Korea that have long been called 'tanged points'.
+Stemmed points were made from blade blanks, and became more standardized and increased in quantity over time.
+Flake tools became more complex, with a greater diversity in both type and size along with continuously used core tools during the Late Pleistocene  [@Bae_2010; @Lee_2013; @Lee_2016].
+The oldest stemmed points so far is from the Yongho-dong site dated to 38.5 ka [@seong2009emergence; #### Park 2012 ####; @seong2015diversity; @bae2017origin].
+<!-- need to expand, mention some key site & details.
+what is the debate?
+-->
+
+### Behavioural ecology and forager land use behaviours
+
+Our review of Late Pleistocene technological innovations in Northeast Asia shows that previous work has largely focussed on timing, location and description of these new tools, but there has been little effort to model the land-use behaviours that generated these assemblages.
+We focus on the appearance of stemmed points in Korea because these are one of the earliest new technologies that definiately represents a new hunting methods because the halfting implies throwing techniques that can reach long-distance targets [@Keeley_1982; @kuhn2015artifacts].
+We ask two questions to invetigate the period around the introduction of stemmed points in Korea.
+
+First, we ask how the introduction of the new tools changed forager landscape use, specifically, was the use of stemmed points associated with occupation of marginal or exterme environments?
+The new technology might allow foragers to explore and stay in different or harsher territories by increasing mobility.
+<!-- need something about how technology and landscape use is linked ...
+- the way people used the *landscape*: changing mobility patterns due to risk management,  Kelly, Fitzhugh, Shott  ...
+-->
+
+Second, we focus on how the new technology changed the way people use their habitation sites.
+After the appearance of these new tools, did they tend to stay in one location longer, or for shorter periods, perhaps for specific purposes?.
 Archaeological sites indicate depositional phenomena and places where artifacts and other debris accumulated reflecting human activities in the past [@kuhn2015artifact].
-This project will look into stone assemblages that hunter-gatherers left behind to understand the site occupation pattern change during the transition period and the relationship with the appearance of new tools.
+The stone artifact assemblage from a site can inform us how long individuals or groups stayed and give insignts into what they did there [@binford1979organization; @kelly1995foraging; @kuhn2016moves; @holdaway2019surface].
+Stone artifacts can contain a record of raw material transport and selection and give insights into land-use behaviour [@clarkson2008lithics].
+<!-- - the way people used *sites*: more fine-grained evidence about land use patterns long vs short term durations of site use Kuhn, Andresfsky, Hiscock -->
 
-Lastly, we want to know about population dynamics during the transition period based on a premise that population change could be a driver of the new invention as well as the source of it if they brought the new tools from outside.
+<!-- Lastly, we want to know about population dynamics during the transition period based on a premise that population change could be a driver of the new invention as well as the source of it if they brought the new tools from outside.
 The demands derived from significan population dynamic in addition to environmental change could lead to development of the tools [buchanan2016drivers; @mihailovic2016technological; prufer2019linking].
@@
 Therefore, we will check the population change during the transitional period to understand its impact on the technology.
+-->
 
--------------- previous paragraphs ------
+<!--  -------------- previous paragraphs ------
 
@@
 
-## Regional Archaeology
-
-Northeast Asia/similar studies and previous research in the adjacent regions : General description of EUP in East Asia
-
-The technological change during the Late Pleistocene in Northeast Asia has been actively studied related to the topic of modern human dispersal to the East and chronological transition from Middle to Upper Paleolithic [or Early to Late Paleolithic.
-In Northeast Asia, the term and concept of ‘Late Paleolithic ’ is widely used rather than ‘Upper Paleolithic’ related to the absence of Middle Paleolithic industry such as Levallois [@gao2002critique; @Bar-Yosef_2002; @mellars2007rethinking; @norton2009evolution; @norton2009east; @lee2013current]].
-Along with the modern human dispersal, this transition often comes with the appearance of blade and micro technology, high frequencies of retouched blade tools, several novel refined tools such as projectile points, end scrapers, burins, denticulates, etc, and using high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary].
-
-In China, the transition from the Middle to the Upper Paleolithic dates to 35-30 ka appeared with a few body decorations, bone tools, core-and-flake in North and cobble tools in South.
-In addition to the technological changes, the use of symbols in the body decorations reflects social and cultural changes occurred during the onset of the Chinese Upper Paleolithic [@qu2013chinese; @peng2018onset].
-At the Shuidonggou site in northern China which is considered as a typical site for showing the transition from cobble to blade tool industries in Northeast Asia [@Brantingham_Perreault_2010; @Gao_Zhang_Yang_Shen_Wu_2010; @pei2012shuidonggou], newly discovered localities (SDG 2, 7-12) show various aspects of the initial Late Paleolithic from 32,000 BP to 6000 BP.
-Some sites (SDG 9, 12) have blades and bladelets made of higher quality or exotic raw materials in addition to more than 80 polished ostrich egg-shell beads (SDG 2), bone needles and an awl (SDG 12), while other sites have mostly small, irregular flakes [@Gao_Zhang_Yang_Shen_Wu_2010].
-The transition to the Upper Paleolihtic in Japan is under the long term debate due to the uncertainty about the existence of the previous period in the Archipelago.
-But there are several features represent the milestone of the new era such as systematic procurement and reduction of lithic raw materials, trap-pit hunting method, watercraft (maritime transport), bow-and-arrow technology, and standardization of lithic tools which can show the appearance of new foraging strategies possibly driven by risks of increased population and ecological change [@Yamaoka_2012; @bae2017origin; @Nakazawa_Bae_2018; @morisaki2019early].
-The technological transition in Korea is mainly characterized by the early appearance of stemmed points and following blade technology.
-Using the blade as a blank, the stemmed points became more standardized and increased in number.
-The flake tools became more sophisticated, with a greater diversity in both type and size along with continuously used core tools during the Upper Paleolithic [@Bae_2010; @Lee_2013; @Lee_2016].
-
-In this research, we choose to focus on the technological transition and its impact on mobility and site occupation pattern in Korea because of the uniqueness of sudden and earliest appearance of stemmed points among the broad area of Northeast Asia during the Late Pleistocene.
-The oldest stemmed points so far is from the Yongho-dong site dated to 38.5 ka [@seong2009emergence; #### Park 2012 ####; @seong2015diversity; @bae2017origin].
-The stemmed point is a projectile point made out of elongated flake or blade and with gentle retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft.
-Combining with the blade industry, the shape became more standardized with one or two ridges on dorsal side and triangular horizontal cross section.
-The stemmed point is often called as a tanged point, but we prefer to use stemmed point to distinguish with other stone projectile points of Bronze age in Korea, also called as a tanged point in English literature.
-The early appearance of the stemmed points in the Korean Peninsula brings about several issues to consider such as issues of human migration or advanced cognitive ability to make a new tool adapting to the surrounding environment.
-As a part of composite implements such as arrowheads or spearheads, the point was combined with other materials and it could be easily replaced reducing waste of total disposal [@Keeley_1982; @kuhn2015artifacts].
-The new projectile point also indicates a new way of hunting methods or other foraging strategies because the throwing technique enables to earn a long-distance target.
-In addition, this new tool spread to adjacent regions such as Japanese Archipelago by combination of watercraft technique and social network so that the study of stemmed points play an important role in understanding about peopling in Japan and the relationship between the Peninsula and Archipelago.
-This project focuses on the functional aspect to understand the technology and foraging pattern based on the concept of HBE.
+-->
 

### 2020-01-08 analysis/paper/paper.Rmd
Editing based on Ben's comments

@@
 
-The appearance of stemmed points in the Late Pleistocene (~40-35 ka) in Korea is often assumed to have transformed forager lifeways because it is thought to reflect more specialed hunting practices (citations).
+The appearance of stemmed points in the Late Pleistocene (~40-35 ka) in Korea is often assumed to have transformed forager lifeways because it is thought to reflect more specialed hunting practices [@lee2002new;@seong2008tanged;@chang2013human].
 Stemmed points are likely to have been projectile weapon tips because the stem of these tools is cconsidered to represent halfting technology that joins the lithic to a wooden shaft.
@@
 We examined stone artifacts from 22 sites in South Korea to investigate the technological transition represented by the appearance of a new tool, stemmed points, and the blade industry that followed.
-Previous work on this technological change has focussed on issues of the transition from the Early to the Late Paleolithic and the disperal of _Homo sapiens_ (citations).
+Previous work on this technological change in Korea has focussed on issues of the transition from the Early to the Late Paleolithic and modern human dispersals [@norton2009evolution; @bae2010origin].
 To our knowledge, the implications for forager mobility and site use have not been explored in detail.
@@
 Here, we attempt to identify patterns of mobility through analysis of the stone artefact assemblages using Human Behavioral Ecology (HBE) models.
-Drawing on established relationships between lithic technology and forager mobility (citations, Shott, Kuhn) we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that are part of a broader strategy of moving frequently and further, possibly as an adaptation to population or environmental changes.
+Drawing on established relationships between lithic technology and forager mobility [@shott1986settlement; @Hiscock_1994; @kuhn1994formal; @kuhn2004upper; @kuhn2015artifacts; @kuhn2016moves; @capriles2018], we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that are part of a broader strategy of moving frequently and further, possibly as an adaptation to population or environmental changes.
 
-<!-- sgenerally, how do we make sense of new stone tools...
+<!-- generally, how do we make sense of new stone tools...
 When a new stone artefact technology appears, how do we explain that?
@@
 risk management (Australian small tool tradition) -->
+<!-- Do you want to have one more paragraph for the general tradition of explaining the emergence of new stone tools? -->
 
@@
 These innovations include blade and micro technology, high frequencies of retouched blade tools, several novel refined tools such as projectile points, stemmed points, end scrapers, burins, denticulates, etc, and using high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary].
-Previous work has argued that the appearance of new stone artefact technologies in this region may be linked to modern human dispersals (citation).
+Previous work has argued that the appearance of new stone artefact technologies in this region may be linked to modern human dispersals by focusing on the origin of the new tools [@seong2008tanged;@bae2010origin;@Bae_2010; @bae2017origin].
 
 The Shuidonggou site in northern China is a typical site for showing the transition from cobble to blade tool industries in the Late Pleistocene of Northeast Asia [@Brantingham_Perreault_2010; @Gao_Zhang_Yang_Shen_Wu_2010; @pei2012shuidonggou].
-Shuidonggou Localities SDG 2, 7-12 have blades and bladelets made of higher quality or exotic raw materials in addition to more than 80 polished ostrich egg-shell beads (SDG 2), bone needles and an awl (SDG 12) (citation).
+Shuidonggou Localities SDG 2, 7-12 have blades and bladelets made of higher quality or exotic raw materials in addition to more than 80 polished ostrich egg-shell beads (SDG 2), bone needles and an awl (SDG 12) [@pei2012shuidonggou].
 The blade compoenent is claimed to have come from Siberia and/or Mongolia at 41 ka, and at 10.5 ka  microblade technology appears but the origin is uncertain.
@@
 -->
+
+[@yi2014significance; @gao2014discovery; @zhang2016bone]
 

### 2020-01-09 analysis/paper/paper.Rmd
Done with China, move to Japan

@@
 
-The Shuidonggou site in northern China is a typical site for showing the transition from cobble to blade tool industries in the Late Pleistocene of Northeast Asia [@Brantingham_Perreault_2010; @Gao_Zhang_Yang_Shen_Wu_2010; @pei2012shuidonggou].
-Shuidonggou Localities SDG 2, 7-12 have blades and bladelets made of higher quality or exotic raw materials in addition to more than 80 polished ostrich egg-shell beads (SDG 2), bone needles and an awl (SDG 12) [@pei2012shuidonggou].
-The blade compoenent is claimed to have come from Siberia and/or Mongolia at 41 ka, and at 10.5 ka  microblade technology appears but the origin is uncertain.
-<!-- expand on ages of all the things, comment about the cause of the new things at this site, check, https://link.springer.com/article/10.1007/s10963-019-09129-w  what is the debate?
--->
+The Shuidonggou site in northern China is a typical site for showing the transition from cobble to blade tool industries in the Late Pleistocene of Northeast Asia from the Late Pleistocene to Middle Holocene [@Brantingham_Perreault_2010; @Gao_Zhang_Yang_Shen_Wu_2010; @pei2012shuidonggou].
+Shuidonggou Localities SDG 2, which dated around c.32,000 years, contains aunal specimens,ostrich eggshell beads,and stone artefacts including blades, retouched flakes, cores, and debries along with well-preserved hearths.
+In Locality 9 dated to c.29,000years, there is a small scatter of stone artefacts consists of blades, Levallois flakes, cores and other retouched flakes.
+<!-- Levallois!
+I didn't pay attentioned --> The recent excavtion of Locality 12 shows that its archaeological layer <!--I shouldn't use 'layer', but I don't know what to say shortly-->, dated to 13,078–13,296 cal BP, includes ground stone and more than 30,000 microlithics made out of a variety of raw materials such as fine-grained and siliceous rocks.
+In addition, more than 10,000 animal fossils and bone tools including a tool for fishing nets, an awl, and two needles were excavated [@pei2012shuidonggou; @gao2014discovery; @zhang2016bone].<!-- they use OSL data that has wise range, or calibrated years --> Based on the archaeological records, people argue that a microblade technology appeared in China after 29ka BP related to mobility change along with colder climate towards LGM and changed game as surviving strategy [@yi2014significance].
+The accumulated studies on Shuidonggou site show that the blade compoenent is claimed to have come from Siberia and/or Mongolia at 41 ka (SDG Locality1 and 9), advanced core and flake tools are likely to be locally developed at 33ka (2 and 8), and at 10.8 ka microblade technology appears but the origin is uncertain (12) [@yi2014significance; @li2019history].
 
-[@yi2014significance; @gao2014discovery; @zhang2016bone]
+<!-- the last sentense is a bit long-->
+<!-- "advanced core and flake tools means intentionally prepared platforms and working surfaces, selective raw material usage (using chert rather tahn river pebble), bipolar reduction, and generating small elongated blanks as results" -->
 
-The origins of Late Pleistocene technological innovations in Japan is debated due to the uncertainty about the existence of people in the Archipelago before the appearance of new tools (citations).
-New technologies are represented by systematic procurement and reduction of lithic raw materials, trap-pit hunting methods, watercraft, bow-and-arrow technology, and standardization of lithic tools.
+The origins of Late Pleistocene technological innovations in Japan is debated due to the uncertainty about the existence of people in the Archipelago before the appearance of new tools
+
+[@izuho201421; ]
+
+.
+New technologies are represented by systematic procurement and reduction of lithic raw materials, trap-pit hunting methods, watercraft, bow-and-arrow technology, and standardization of lithic tools [@bae2017origin].
 <!-- need some ages for these types of evidence --> These have been interpreted as evidence of new foraging strategies possibly driven by increased population and ecological changes [@Yamaoka_2012; @bae2017origin; @Nakazawa_Bae_2018; @morisaki2019early].

### 2020-01-30 analysis/paper/paper.Rmd
writing Japanese part

@@
 
-The origins of Late Pleistocene technological innovations in Japan is debated due to the uncertainty about the existence of people in the Archipelago before the appearance of new tools
+The origins of Late Pleistocene technological innovations in Japan is suddenly appeared in differen regions of the archipelago around 38,000 years ago accaompanying with Homo sapiens found in Okinawa Island (ca.
+36,000 cal BP) [@yamaoka2012use; @izuho201421].
+The components of the new technologies represented by trap pits, cobble concentrations, hearths, charcoal concentrations, and toolkits including trapezoids, pointed-shaped backed blades, backed points, burins, end scrapers, side scrapers, wedges, beak- shaped tools, axes, edge-ground axes, hammerstones, cobble tools, and anvils [ @bae2017origin; @izuho201421].
+These innovations have been interpreted as evidnce of new foraging methods such as watercraft and bow-and-arrow technology, drivien by increased population and ecological changes [@Yamaoka_2012; @bae2017origin; @Nakazawa_Bae_2018; @morisaki2019early].
+Especially, hafted trapzoids were designed to be used as multifucntional tools and people tried to maintain and resue trapezoids and tools euipped with trapzoid to adapt to the specific environmental settings on the Japanese Islands [@ono2002radiocaron; @yamaoka2012use].
+Ishinomoto 8-ku is one of the earliets sites dated to 39,690–34,790 cal BP, located in Takuma upland of Kumamoto Prefecture.
+A total of 500 stone artifacts were discovered including trapezoids, side scrapers, flakes, flake cores, edge-ground axes, and cobble tools.
+The main lithic raw materials are chert and andesite [@izuho201421].
+
+<!--do we need to talk about AT???People had assumed that Japanese archipelago started to be occupied before 30,000 cal BP based on the existence of primary Aira-Tanzawa tephra (dated at 30,009±189 cal BP) in the excavation unit.
+Recenly, accummulated radiocarbon data show that -->
+
+Ishinomoto 8-ku 
+the age of the EUP oc- cupation falls within the interval 39,690–34,790 cal BP.the Takuma upland in Kumamoto Prefecture A total of 500 lithic specimens were recovered from the lithic concentrations.
+The lithic raw materials are mainly chert and andesite.
+The lithic assem- blage includes trapezoids, side scrapers, flakes, flake cores, edge-ground axes, and cobble tools
+
+----
 
@@
 
-.
-New technologies are represented by systematic procurement and reduction of lithic raw materials, trap-pit hunting methods, watercraft, bow-and-arrow technology, and standardization of lithic tools [@bae2017origin].
-<!-- need some ages for these types of evidence --> These have been interpreted as evidence of new foraging strategies possibly driven by increased population and ecological changes [@Yamaoka_2012; @bae2017origin; @Nakazawa_Bae_2018; @morisaki2019early].
-<!-- need to expand, mention some key site, blades appear when?
-what is the debate?
--->
+ vegetation during MIS 3 (Takahara and Hayashi), large mammal extinctions in the late Pleistocene (Iwase et al.), colonization processes of the Ryukyu Islands (Kaifu et al.), marine transport of obsidian from the Kozu Island (Ikeya), introduction of new lithic tools from the Korean Peninsula (Morisaki), trap-pit hunting (Sato), Jōmon mitochondrial DNA (Adachi et al.), and the di- versification processes of the microblade technocomplex in Hokkaido (Nakazawa and Yamada)
+
+ “circular aggregations” (dense concentrations of stone artifacts aligned in a circle), .
+High-quality lithic raw mate- rials, such as obsidian, “hard shale,” and sanukite, were often transported long distances and used together with locally available chert, andesite, rhyolite, and chalcedony.
+
+high quarlity raw materials were avaliable (70 obsidian sources).
+the Aira-Tn (AT) tephra, dated at 30,009±189 cal BP (Smith et al.
+2013) (1) the archeological components are situated below the primary AT tephra level (>30,000 cal BP), and (2) the tool kit usually (butnot always) includes trapezoids, pointed-shaped backed blades (“knife-shaped tools” in previous terminology), and edge-ground axes.
+the stratigraphical chronology is based on Sagamihara Upland and Mt.
+Ashitaka area
+
+only a few possi- bly pre-UP sites are known, and there is ongoing, heated debate whether or not such archaic industries existed in the Japanese archipelago
+
+----
 

### 2020-01-30 analysis/paper/paper.Rmd
Korea part done

@@
 In Locality 9 dated to c.29,000years, there is a small scatter of stone artefacts consists of blades, Levallois flakes, cores and other retouched flakes.
-<!-- Levallois!
-I didn't pay attentioned --> The recent excavtion of Locality 12 shows that its archaeological layer <!--I shouldn't use 'layer', but I don't know what to say shortly-->, dated to 13,078–13,296 cal BP, includes ground stone and more than 30,000 microlithics made out of a variety of raw materials such as fine-grained and siliceous rocks.
+The recent excavtion of Locality 12 shows that its archaeological layer <!--I shouldn't use 'layer', but I don't know what to say shortly-->, dated to 13,078–13,296 cal BP, includes ground stone and more than 30,000 microlithics made out of a variety of raw materials such as fine-grained and siliceous rocks.
 In addition, more than 10,000 animal fossils and bone tools including a tool for fishing nets, an awl, and two needles were excavated [@pei2012shuidonggou; @gao2014discovery; @zhang2016bone].<!-- they use OSL data that has wise range, or calibrated years --> Based on the archaeological records, people argue that a microblade technology appeared in China after 29ka BP related to mobility change along with colder climate towards LGM and changed game as surviving strategy [@yi2014significance].
@@
 The components of the new technologies represented by trap pits, cobble concentrations, hearths, charcoal concentrations, and toolkits including trapezoids, pointed-shaped backed blades, backed points, burins, end scrapers, side scrapers, wedges, beak- shaped tools, axes, edge-ground axes, hammerstones, cobble tools, and anvils [ @bae2017origin; @izuho201421].
-These innovations have been interpreted as evidnce of new foraging methods such as watercraft and bow-and-arrow technology, drivien by increased population and ecological changes [@Yamaoka_2012; @bae2017origin; @Nakazawa_Bae_2018; @morisaki2019early].
+These innovations have been interpreted as evidnce of new foraging methods such as watercraft (marine transport of obsidian from the Kozu Island) and bow-and-arrow technology, drivien by increased population and ecological changes [@Yamaoka_2012; @bae2017origin; @Nakazawa_Bae_2018; @morisaki2019early].
 Especially, hafted trapzoids were designed to be used as multifucntional tools and people tried to maintain and resue trapezoids and tools euipped with trapzoid to adapt to the specific environmental settings on the Japanese Islands [@ono2002radiocaron; @yamaoka2012use].
@@
 Recenly, accummulated radiocarbon data show that -->
-
-Ishinomoto 8-ku 
-the age of the EUP oc- cupation falls within the interval 39,690–34,790 cal BP.the Takuma upland in Kumamoto Prefecture A total of 500 lithic specimens were recovered from the lithic concentrations.
-The lithic raw materials are mainly chert and andesite.
-The lithic assem- blage includes trapezoids, side scrapers, flakes, flake cores, edge-ground axes, and cobble tools
-
-----
-
-[@izuho201421; ]
-
- vegetation during MIS 3 (Takahara and Hayashi), large mammal extinctions in the late Pleistocene (Iwase et al.), colonization processes of the Ryukyu Islands (Kaifu et al.), marine transport of obsidian from the Kozu Island (Ikeya), introduction of new lithic tools from the Korean Peninsula (Morisaki), trap-pit hunting (Sato), Jōmon mitochondrial DNA (Adachi et al.), and the di- versification processes of the microblade technocomplex in Hokkaido (Nakazawa and Yamada)
-
- “circular aggregations” (dense concentrations of stone artifacts aligned in a circle), .
-High-quality lithic raw mate- rials, such as obsidian, “hard shale,” and sanukite, were often transported long distances and used together with locally available chert, andesite, rhyolite, and chalcedony.
-
-high quarlity raw materials were avaliable (70 obsidian sources).
-the Aira-Tn (AT) tephra, dated at 30,009±189 cal BP (Smith et al.
-2013) (1) the archeological components are situated below the primary AT tephra level (>30,000 cal BP), and (2) the tool kit usually (butnot always) includes trapezoids, pointed-shaped backed blades (“knife-shaped tools” in previous terminology), and edge-ground axes.
-the stratigraphical chronology is based on Sagamihara Upland and Mt.
-Ashitaka area
-
-only a few possi- bly pre-UP sites are known, and there is ongoing, heated debate whether or not such archaic industries existed in the Japanese archipelago
-
-----
+<<!-- dp we need to mention popular research topic such as existance of pre UP ("archaic industries")as a heated debate?--->>
 
@@
 The oldest stemmed points so far is from the Yongho-dong site dated to 38.5 ka [@seong2009emergence; #### Park 2012 ####; @seong2015diversity; @bae2017origin].
-<!-- need to expand, mention some key site & details.
-what is the debate?
--->
+The origin of the stemmed point is one of the most frequently devated topic related to the modern human dispersals.
+Bae (2010) claims two sub-stages for these new technologies based on the presence of AT tephra (presumably earlier than 25k BP).
+Bae’s earlier stage is started from Koreyri site dated to 27 ka.
+The stage is characterized by long and thick blades and relatively large prismatic cores with stemmed points.
+Bae’s later stage contains a microlithic industry, for example as found at Sinbuk in southwestern Korea, which dates 25,420±190 BP and 25,000±1,000 BP [@bae2010origin].
+An alternative view comes from Seong (2009) and Park (2012) who claim three sub-stages by distinguishing an initial period when stemmed points appeared before blades [@seong2009emergence; #### Park 2012 ####].
+The difference between the two stages of division [@bae2010origin] and the three stages [@seong2009emergence; #### Park 2012 ####] hinges on the recognition of stemmed points as a technology that is distinctive from blades.
+Bae stresses the combination of stemmed point and blade as tool blanks, while Seong and Park focus on the stemmed point as a new technology, having a stem part with retouch and acute-angled tip.
+These different perspectives on the distinctiveness of the stemmed point are at the crux of a heated debate on the nature and cause of the transition from the earlier phase to the later phase during the Late Pleistocene.
 

### 2020-01-30 analysis/paper/paper.Rmd
edited by HBE part

@@
 Recenly, accummulated radiocarbon data show that -->
-<<!-- dp we need to mention popular research topic such as existance of pre UP ("archaic industries")as a heated debate?--->>
+<<!-- do we need to mention popular research topic such as existance of pre UP ("archaic industries")as a heated debate?--->
 
@@
 The oldest stemmed points so far is from the Yongho-dong site dated to 38.5 ka [@seong2009emergence; #### Park 2012 ####; @seong2015diversity; @bae2017origin].
-The origin of the stemmed point is one of the most frequently devated topic related to the modern human dispersals.
-Bae (2010) claims two sub-stages for these new technologies based on the presence of AT tephra (presumably earlier than 25k BP).
-Bae’s earlier stage is started from Koreyri site dated to 27 ka.
-The stage is characterized by long and thick blades and relatively large prismatic cores with stemmed points.
-Bae’s later stage contains a microlithic industry, for example as found at Sinbuk in southwestern Korea, which dates 25,420±190 BP and 25,000±1,000 BP [@bae2010origin].
-An alternative view comes from Seong (2009) and Park (2012) who claim three sub-stages by distinguishing an initial period when stemmed points appeared before blades [@seong2009emergence; #### Park 2012 ####].
-The difference between the two stages of division [@bae2010origin] and the three stages [@seong2009emergence; #### Park 2012 ####] hinges on the recognition of stemmed points as a technology that is distinctive from blades.
-Bae stresses the combination of stemmed point and blade as tool blanks, while Seong and Park focus on the stemmed point as a new technology, having a stem part with retouch and acute-angled tip.
-These different perspectives on the distinctiveness of the stemmed point are at the crux of a heated debate on the nature and cause of the transition from the earlier phase to the later phase during the Late Pleistocene.
+
+The origin of the stemmed point is one of the most frequently discussed topics among Paleolithic research in Korea.
+Especially, this debate includes questions about the origin of the Late Paleolithic period in Korea, the existence and nature of Middle Paleolithic technology, and reconstructions of anatomically modern human dispersals in eastern Asia.
+Most Korean archaeologists had assumed that the new blade technology represented came from Shuidonggou in the northeast China because this site has one of the earliest blade industries (around 35,000 BP) in northeast Asia and it is geographically close to the Korean peninsula [seong2008tanged; @lee2013current].
+However, signs of early technological change in Korea, including stemmed points and blades, have been discovered from deposits dated to 40-35 ka BP from Hwadae-ri, Hopyeong-dong, and Yongho-dong, pre-dating the Shuidonggou finds [@seong2009emergence].
+The debate about origins can be summarized into two models competing to explain the appearance of the new Late Pleistocene technologies: in situ evolution [@seong2009emergence] and ‘heterogenic’ migration [@bae2010origin].
+The ‘in situ’ model claims that stemmed points and other Late Paleolithic assemblages including a blade industry autonomously emerged in the south of the Korean peninsula with no external influence.
+Seong (2009) uses the blade-to-flake ratios of lithic assemblages to support his claim.
+In his view, the blade industry represents new technology while flakes indicate a continuing local one.
+He argues that the increased ratio of blades in lithic assemblages shows an expansion of the new technology.
+The migration model argues that the new blade industry, including stemmed points, and the earlier coarse flake tradition including large cores, polyhedrons, choppers, and even handaxes, came from different origins.
+While Seong’s model (2009) proposes that the ‘heterogenic’ character is the result of indigenous development, Bae (2010) proposes that it is the result of the continuous influx of modern human populations from both Siberia and southern China.
+Bae’s model suggests blade technology arrived in the Korean peninsula from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP [@bae2013early].
+<!--- I don't know how to connect this debate to our research. It is really needed? --->
 
@@
 
-Our review of Late Pleistocene technological innovations in Northeast Asia shows that previous work has largely focussed on timing, location and description of these new tools, but there has been little effort to model the land-use behaviours that generated these assemblages.
+Our review of Late Pleistocene technological innovations in Northeast Asia shows that previous work has largely focussed on timing, location, origins and description of these new tools, but there has been little effort to model the land-use behaviours that generated these assemblages.
 We focus on the appearance of stemmed points in Korea because these are one of the earliest new technologies that definiately represents a new hunting methods because the halfting implies throwing techniques that can reach long-distance targets [@Keeley_1982; @kuhn2015artifacts].
@@
 
-First, we ask how the introduction of the new tools changed forager landscape use, specifically, was the use of stemmed points associated with occupation of marginal or exterme environments?
+First, we ask how the introduction of the new tools changed foragers' landscape use, specifically, was the use of stemmed points associated with occupation of marginal or exterme environments?
 The new technology might allow foragers to explore and stay in different or harsher territories by increasing mobility.
-<!-- need something about how technology and landscape use is linked ...
-- the way people used the *landscape*: changing mobility patterns due to risk management,  Kelly, Fitzhugh, Shott  ...
--->
+For instance, Morisaki et al.
+argue that the transition from trapezoid to the blades and projectile points around 25 ka cal BP in north Paleo-Honshu Island enabled the foragers to extend their occupation at residential sites [@Morisaki_Izuho_Terry_Sato_2015].
+In Australia during the mid-Holocene, the highly retouched and finely shaped tools combined with shafts were appeared to  broaden their occupations to offshore islands on the northeast and northwest coast [@Hiscock_1994].
+Hiscock et al.
+insist that the appearance of mirolithcs is great example to show that people made portable and multi-functional tools to minimize the travel expense and increase tool readiness [@Hiscock_Clarkson_Mackay_2011].
 
@@
 Stone artifacts can contain a record of raw material transport and selection and give insights into land-use behaviour [@clarkson2008lithics].
-<!-- - the way people used *sites*: more fine-grained evidence about land use patterns long vs short term durations of site use Kuhn, Andresfsky, Hiscock -->
+For example, stone artifact density and the frequency of retouched pieces (scaled to the volume of excavated sediment) are often used as proxies to represent occupation patterns.
+Assemablges with low density but a high proportion of retouched and backed pieces would indicate the remains of 'short-term camp', while the combination of high density and low retouched pieces represent 'logistically organized basecampe' [@clark2017lithics; clark2019landscapes].
 
@@
 Therefore, we will check the population change during the transitional period to understand its impact on the technology.
--->
-
-<!--  -------------- previous paragraphs ------
-
-New technologies have led people to adjust the speed and frequency of mobility, explore the new landscape, or even manage the population size of the foraging group or overall population occupied one region.
-Previous studies show that new technologies play an important role in the change in the use of the landscape.
-For instance, Morisaki et al.
-argue that the transition of the blades and points becoming dominant over trapezoid in toolkits around 25 ka cal BP in north Paleo-Honshu Island enabled the foragers to extend their occupation at residential sites [@Morisaki_Izuho_Terry_Sato_2015].
-In Australia during the mid-Holocene, the highly retouched and finely shaped tools with shafts appeared to reduce risks of being in a different environment so that they could broaden their occupations to offshore islands on the northeast and northwest coast [@Hiscock_1994].
-New tools tend to appear as a risk reduction strategy when they confront an unfamiliar environment mainly caused by climate change, migration to previously unoccupied landscapes, based on the understanding of resource distribution, tool manufacturing process, and the characteristics of the foraging group.
-The appearance of microliths as portable and standardized tools is a great example to show that people made new tools to increase multi-functionality and tool readiness so that they can minimize the economic risk and/or scheduling difficulties [@Hiscock_Clarkson_Mackay_2011].
-
-The rapid spread of microblade industry in north China among Last Glacial Maximum demonstrates that making new tools was one of the human adaptation to extremely cold environment by increasing mobility or meeting the new demands such as the production of sophisticated cold-weather clothing [@Barton_Brantingham_Ji_2007; @yi2013microblade; @Qu_Bar-Yosef_Wang_Wu_2013; @nian2014chronology].
-
-The explanation of the emergence of new tools as an aid of buffering risks, as a result of environmental change, is one of the main concepts of Human Behavioral Ecology (HBE).
-HBE focuses on understanding patterned human behaviors, such as the ways they evaluate choices in food procurement, mobility, division of labor, task-group membership, social affiliation, prestige, and social subordination, to selective pressures in the environment [@smith2017natural; @smith2017three; @fitzhugh2001risk; @bird2006behavioral; @fitzhugh2019human; @Veile_2018].
-Thus, most HBE models investigate environmental variables (e.g., resource density, competitor frequency) and behaviors reflecting decisions that people made about how to allocate time and energy to their activities [@smith2017three; @bird2006behavioral; @kelly1995foraging; @Creanza_Kolodny_Feldman_2017; @fitzhugh2019human].
-HBE provides concepts to identify how foragers adjust their technologies, resource management strategies, schedule movements, foraging, food processing (including storage and consumption), and investment in technologies to optimize time and energy [@kelly1995foraging].
-Stone artifacts have also been used in HBE models to understand hunter-gatherers’ mobility and site occupation patterns [@Marwick_2013; @Clarkson_2013; @capriles2018mobility].
-For example, Clark and Barton (2017) examined to investigate resource acquisition strategies of hunter-gatherers by means of stone artifact density and the frequency of retouched pieces (scaled to the volume of excavated sediment) as proxies to represent occupation patterns such as ‘logistically organized basecamp’ and ‘short-term camp’.
-We use HBE as a reference framework to answer the three questions: (1) how did the introduction of the stemmed points and blade industry change the way people used the landscape?
-(2) was there any change in the way people used sites?
-(3) was there any population change and did it impact on technology or foraging pattern?
-
 -->

### 2020-05-05 analysis/paper/paper.Rmd
add site locations CSV for updates

--- 2020-01-30 analysis/paper/paper.Rmd
@@
 
-The appearance of stemmed points in the Late Pleistocene (~40-35 ka) in Korea is often assumed to have transformed forager lifeways because it is thought to reflect more specialed hunting practices [@lee2002new;@seong2008tanged;@chang2013human].
+The appearance of stemmed points in the Late Pleistocene (~40-35 ka) in Korea is often assumed to have transformed forager lifeways because it is thought to reflect more specialed hunting practices (citations).
 Stemmed points are likely to have been projectile weapon tips because the stem of these tools is cconsidered to represent halfting technology that joins the lithic to a wooden shaft.
@@
 We examined stone artifacts from 22 sites in South Korea to investigate the technological transition represented by the appearance of a new tool, stemmed points, and the blade industry that followed.
-Previous work on this technological change in Korea has focussed on issues of the transition from the Early to the Late Paleolithic and modern human dispersals [@norton2009evolution; @bae2010origin].
+Previous work on this technological change has focussed on issues of the transition from the Early to the Late Paleolithic and the disperal of _Homo sapiens_ (citations).
 To our knowledge, the implications for forager mobility and site use have not been explored in detail.
@@
 Here, we attempt to identify patterns of mobility through analysis of the stone artefact assemblages using Human Behavioral Ecology (HBE) models.
-Drawing on established relationships between lithic technology and forager mobility [@shott1986settlement; @Hiscock_1994; @kuhn1994formal; @kuhn2004upper; @kuhn2015artifacts; @kuhn2016moves; @capriles2018], we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that are part of a broader strategy of moving frequently and further, possibly as an adaptation to population or environmental changes.
+Drawing on established relationships between lithic technology and forager mobility (citations, Shott, Kuhn) we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that are part of a broader strategy of moving frequently and further, possibly as an adaptation to population or environmental changes.
 
-<!-- generally, how do we make sense of new stone tools...
+<!-- sgenerally, how do we make sense of new stone tools...
 When a new stone artefact technology appears, how do we explain that?
@@
 risk management (Australian small tool tradition) -->
-<!-- Do you want to have one more paragraph for the general tradition of explaining the emergence of new stone tools? -->
 
@@
 These innovations include blade and micro technology, high frequencies of retouched blade tools, several novel refined tools such as projectile points, stemmed points, end scrapers, burins, denticulates, etc, and using high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary].
-Previous work has argued that the appearance of new stone artefact technologies in this region may be linked to modern human dispersals by focusing on the origin of the new tools [@seong2008tanged;@bae2010origin;@Bae_2010; @bae2017origin].
+Previous work has argued that the appearance of new stone artefact technologies in this region may be linked to modern human dispersals (citation).
 
-The Shuidonggou site in northern China is a typical site for showing the transition from cobble to blade tool industries in the Late Pleistocene of Northeast Asia from the Late Pleistocene to Middle Holocene [@Brantingham_Perreault_2010; @Gao_Zhang_Yang_Shen_Wu_2010; @pei2012shuidonggou].
-Shuidonggou Localities SDG 2, which dated around c.32,000 years, contains aunal specimens,ostrich eggshell beads,and stone artefacts including blades, retouched flakes, cores, and debries along with well-preserved hearths.
-In Locality 9 dated to c.29,000years, there is a small scatter of stone artefacts consists of blades, Levallois flakes, cores and other retouched flakes.
-The recent excavtion of Locality 12 shows that its archaeological layer <!--I shouldn't use 'layer', but I don't know what to say shortly-->, dated to 13,078–13,296 cal BP, includes ground stone and more than 30,000 microlithics made out of a variety of raw materials such as fine-grained and siliceous rocks.
-In addition, more than 10,000 animal fossils and bone tools including a tool for fishing nets, an awl, and two needles were excavated [@pei2012shuidonggou; @gao2014discovery; @zhang2016bone].<!-- they use OSL data that has wise range, or calibrated years --> Based on the archaeological records, people argue that a microblade technology appeared in China after 29ka BP related to mobility change along with colder climate towards LGM and changed game as surviving strategy [@yi2014significance].
-The accumulated studies on Shuidonggou site show that the blade compoenent is claimed to have come from Siberia and/or Mongolia at 41 ka (SDG Locality1 and 9), advanced core and flake tools are likely to be locally developed at 33ka (2 and 8), and at 10.8 ka microblade technology appears but the origin is uncertain (12) [@yi2014significance; @li2019history].
+The Shuidonggou site in northern China is a typical site for showing the transition from cobble to blade tool industries in the Late Pleistocene of Northeast Asia [@Brantingham_Perreault_2010; @Gao_Zhang_Yang_Shen_Wu_2010; @pei2012shuidonggou].
+Shuidonggou Localities SDG 2, 7-12 have blades and bladelets made of higher quality or exotic raw materials in addition to more than 80 polished ostrich egg-shell beads (SDG 2), bone needles and an awl (SDG 12) (citation).
+The blade compoenent is claimed to have come from Siberia and/or Mongolia at 41 ka, and at 10.5 ka  microblade technology appears but the origin is uncertain.
+<!-- expand on ages of all the things, comment about the cause of the new things at this site, check, https://link.springer.com/article/10.1007/s10963-019-09129-w  what is the debate?
+-->
 
-<!-- the last sentense is a bit long-->
-<!-- "advanced core and flake tools means intentionally prepared platforms and working surfaces, selective raw material usage (using chert rather tahn river pebble), bipolar reduction, and generating small elongated blanks as results" -->
-
-The origins of Late Pleistocene technological innovations in Japan is suddenly appeared in differen regions of the archipelago around 38,000 years ago accaompanying with Homo sapiens found in Okinawa Island (ca.
-36,000 cal BP) [@yamaoka2012use; @izuho201421].
-The components of the new technologies represented by trap pits, cobble concentrations, hearths, charcoal concentrations, and toolkits including trapezoids, pointed-shaped backed blades, backed points, burins, end scrapers, side scrapers, wedges, beak- shaped tools, axes, edge-ground axes, hammerstones, cobble tools, and anvils [ @bae2017origin; @izuho201421].
-These innovations have been interpreted as evidnce of new foraging methods such as watercraft (marine transport of obsidian from the Kozu Island) and bow-and-arrow technology, drivien by increased population and ecological changes [@Yamaoka_2012; @bae2017origin; @Nakazawa_Bae_2018; @morisaki2019early].
-Especially, hafted trapzoids were designed to be used as multifucntional tools and people tried to maintain and resue trapezoids and tools euipped with trapzoid to adapt to the specific environmental settings on the Japanese Islands [@ono2002radiocaron; @yamaoka2012use].
-Ishinomoto 8-ku is one of the earliets sites dated to 39,690–34,790 cal BP, located in Takuma upland of Kumamoto Prefecture.
-A total of 500 stone artifacts were discovered including trapezoids, side scrapers, flakes, flake cores, edge-ground axes, and cobble tools.
-The main lithic raw materials are chert and andesite [@izuho201421].
-
-<!--do we need to talk about AT???People had assumed that Japanese archipelago started to be occupied before 30,000 cal BP based on the existence of primary Aira-Tanzawa tephra (dated at 30,009±189 cal BP) in the excavation unit.
-Recenly, accummulated radiocarbon data show that -->
-<<!-- dp we need to mention popular research topic such as existance of pre UP ("archaic industries")as a heated debate?--->>
+The origins of Late Pleistocene technological innovations in Japan is debated due to the uncertainty about the existence of people in the Archipelago before the appearance of new tools (citations).
+New technologies are represented by systematic procurement and reduction of lithic raw materials, trap-pit hunting methods, watercraft, bow-and-arrow technology, and standardization of lithic tools.
+<!-- need some ages for these types of evidence --> These have been interpreted as evidence of new foraging strategies possibly driven by increased population and ecological changes [@Yamaoka_2012; @bae2017origin; @Nakazawa_Bae_2018; @morisaki2019early].
+<!-- need to expand, mention some key site, blades appear when?
+what is the debate?
+-->
 
@@
 The oldest stemmed points so far is from the Yongho-dong site dated to 38.5 ka [@seong2009emergence; #### Park 2012 ####; @seong2015diversity; @bae2017origin].
-The origin of the stemmed point is one of the most frequently devated topic related to the modern human dispersals.
-Bae (2010) claims two sub-stages for these new technologies based on the presence of AT tephra (presumably earlier than 25k BP).
-Bae’s earlier stage is started from Koreyri site dated to 27 ka.
-The stage is characterized by long and thick blades and relatively large prismatic cores with stemmed points.
-Bae’s later stage contains a microlithic industry, for example as found at Sinbuk in southwestern Korea, which dates 25,420±190 BP and 25,000±1,000 BP [@bae2010origin].
-An alternative view comes from Seong (2009) and Park (2012) who claim three sub-stages by distinguishing an initial period when stemmed points appeared before blades [@seong2009emergence; #### Park 2012 ####].
-The difference between the two stages of division [@bae2010origin] and the three stages [@seong2009emergence; #### Park 2012 ####] hinges on the recognition of stemmed points as a technology that is distinctive from blades.
-Bae stresses the combination of stemmed point and blade as tool blanks, while Seong and Park focus on the stemmed point as a new technology, having a stem part with retouch and acute-angled tip.
-These different perspectives on the distinctiveness of the stemmed point are at the crux of a heated debate on the nature and cause of the transition from the earlier phase to the later phase during the Late Pleistocene.
+<!-- need to expand, mention some key site & details.
+what is the debate?
+-->
 
@@
 	mutate_at(vars(lat_dd, long_dd), as.numeric)
+
+write_csv(here::here("analysis/data/raw_data/korean_palaeolithic_site_locations.csv"))
 

### 2020-07-07 analysis/paper/paper.Rmd

--- 2020-01-30 analysis/paper/paper.Rmd
@@
 
-coords <- 
-	korean_archaeological_sites %>% 
-	select(site_name, 
-				 east_longitude, 
-				 north_latitude) %>% 
-	separate(east_longitude, 
-					 into = c("e_deg", "e_min", "e_sec"),
-					 sep =  "°|\\′") %>% 
-	mutate(e_min = str_squish(e_min),
-				 e_sec = str_squish(e_sec)) %>% 
-	separate(e_min, 
-					 into = c("e_min", "e_sec1"), 
-					 sep = "'") %>% 
-	mutate_at(vars(e_deg, e_min, e_sec, e_sec1), parse_number) %>% 
-	mutate(e_sec = ifelse(is.na(e_sec), e_sec1, e_sec)) %>% 
-	select(-e_sec1) %>% 
-	separate(north_latitude, 
-					 into = c("n_deg", "n_min", "n_sec"),
-					 sep =  "°|\\′") %>% 
-	mutate(n_min = str_squish(n_min),
-				 n_sec = str_squish(n_sec)) %>% 
-	separate(n_min, 
-					 into = c("n_min", "n_sec1"), 
-					 sep = "'") %>% 
-	mutate_at(vars(n_deg, n_min, n_sec, n_sec1), parse_number) %>% 
-	mutate_if(is.numeric, list(~ifelse(is.na(.), 0, .))) %>% 
-	mutate(n_sec = ifelse(is.na(n_sec), n_sec1, n_sec)) %>% 
-	select(-n_sec1) %>% 
-	mutate(lat_dms =  str_glue('{n_deg} {n_min} {n_sec}'),
-				 long_dms = str_glue('{e_deg} {e_min} {e_sec}')) %>% 
-	# convert from decimal minutes to decimal degrees
-	mutate(lat_dd = measurements::conv_unit(lat_dms, 
-																			 from = 'deg_min_sec', 
-																			 to = 'dec_deg')) %>% 
-	mutate(long_dd = measurements::conv_unit(long_dms, 
-																				from = 'deg_min_sec', 
-																				to = 'dec_deg')) %>% 
-	mutate_at(vars(lat_dd, long_dd), as.numeric)
-
-# quick map
-
-library(ggmap)
-
-# bounding box
-# 39.534214, 124.159154 ... 39.657415, 129.138604
-# 33.694662, 123.491142 ... 34.338454, 130.569658
-
-map <- 
-	get_stamenmap(bbox = c(left = 123, 
-												 bottom = 33, 
-												 right = 	131, 
-												 top = 39),
-								zoom = 10)
-
-library(ggrepel)
-ggmap(map)  + geom_point(data = coords,
-						 aes(long_dd ,
-						 		lat_dd), 
-						 colour = "red",
-						 size = 2) +
-	geom_label_repel(data = coords,
-									aes(long_dd ,
-											lat_dd,
-											label = site_name),
-									size = 2)
-
-ggsave(here("analysis/figures/site-map.png"))
-
-# pick points off this map with korea_boundary <- locator()
-
-korea_bb <-  
-	data.frame(x = c(125, 
-						       130, 
-						       130, 
-						       125), 
-			 y = c(38, 
-			 			 38, 
-			 			34, 
-			 			34))
 ```

### 2020-08-07 analysis/paper/paper.Rmd
remove duplicate site from plot

@@
   filter(!is.na(percentage)) %>% 
+  # if percentage is <10%, call it 'other'
   mutate(raw_material = ifelse(percentage >= 10, as.character(X), "other")) %>% 
@@
   distinct(.keep_all = TRUE) %>% 
-
   # join to get ages of the sites
@@
             by = c('site' = 'site_name')) %>% 
-  mutate(age_ka = jitter(age_ka),
-         age_bin  = ntile(age_ka, 5)) %>% 
+ # mutate(age_ka = jitter(age_ka),
+ #        age_bin  = ntile(age_ka, 5)) %>% 
   arrange(age_ka) %>% 
@@
        aes(
-           reorder(axis_label, -age_ka),
+           reorder(axis_label, 
+                   -age_ka),
            percentage,

### 2020-08-17 analysis/paper/paper.Rmd
updating the figures

@@
   mutate(artefact_density = total_artifacts / volume,
-         sites = names(kasv)[-1])
+         sites = names(kasv)[-1]) %>% 
+  left_join(mydata, by = c('sites' = 'site_name' )) %>% 
+  mutate(has_sp = ifelse(is.na(SP.), "no", "yes"))
 
+density_sp_sub_plot <- 
 ggplot(kasv_tidy,
-       aes(date_age,
+       aes(has_sp, 
            artefact_density)) +
-  geom_point(aes(size = total_artifacts)) +
-  geom_text_repel(aes(label = sites),
-                  nudge_x = 0.25, 
-                  nudge_y = 0.25) +
-  ylab("Artifact density") +
-  xlab("Age of assemblage (ka)") +
-  scale_size_continuous(name = "Total number\nof artifacts")  +
-  theme_minimal(base_size = 16)
+  geom_boxplot() +
+  theme_bw(base_size = 8)  +
+  labs(x = "Stemmed points present",
+       y = "Artefact density")
 
-# without site name
+density_sp_main_plot <- 
 ggplot(kasv_tidy,
@@
            artefact_density)) +
-  geom_point(aes(size = total_artifacts)) +
+  geom_point(aes(size = total_artifacts,
+                 colour = has_sp)) +
   ylab("Artefact density") +
   xlab("Age of assemblage (ka)") + 
+  #geom_smooth(method = "lm") + 
   scale_size_continuous(name = "Total number\nof artifacts")  +
-  theme_minimal(base_size = 16)
+  scale_color_viridis_d(name = "Contains\nstemmed\npoints?") +
+  theme_minimal(base_size = 16)  #+
+  #geom_text_repel(aes(label = sites),
+  #                nudge_x = 0.25, 
+  #                nudge_y = 0.25)
+
+# https://wilkelab.org/cowplot/articles/drawing_with_on_plots.html
+library(cowplot)
+ggdraw(density_sp_main_plot) +
+  draw_plot(density_sp_sub_plot, 
+            .375, .67, 
+            .25, .25)
+
 ggsave("figures/age-by-density.png")
-
 ```
@@
 That is to say, the hunter-gatherer became either more residential or the majority of the site that this research studies were used as base camps or residence rather than shortly occupied camps for certain tasks.
+
+```{r}
+# retouch, density, age
+Assemblage_info <- read_csv(here::here("analysis/data/raw_data/Assemblage_info.csv"))
+
+Assemblage_info_retouch_density_ages <- 
+Assemblage_info %>% 
+  pivot_longer(-X1, 
+               names_to = "sites") %>% 
+  left_join(kasv_tidy)
+
+Assemblage_info_retouch_density_ages_prop <- 
+Assemblage_info_retouch_density_ages %>% 
+     filter(!X1 %in% c("debris", 
+                     "unkown", 
+                     "pebble",  
+                     "metate", 
+                     "flake",
+                     "unfinished",
+                     "ground",
+                     "blank"
+                     )) %>% 
+  group_by(sites) %>% 
+  mutate(prop_retouched = sum(value, na.rm = TRUE) / total_artifacts) %>% 
+  arrange(sites) %>% 
+  select(-X1, -value) %>% 
+  distinct_all()
+
+# retouch over time
+retouch_over_time_subplot <- 
+ggplot(Assemblage_info_retouch_density_ages_prop,
+       aes(date_age / 1000,
+           prop_retouched)) +
+  geom_point(size = 3, 
+             colour = "grey80") +
+  labs(x = "Age of assemblage (ka)",
+       y = "Proportion retouched") +
+  theme_bw(base_size = 6)
+
+# compute correlation
+kas_sites_retouch_density_corr <- 
+cor.test(Assemblage_info_retouch_density_ages_prop$prop_retouched,
+         Assemblage_info_retouch_density_ages_prop$artefact_density)
+
+r_value <- unname(kas_sites_retouch_density_corr$estimate)
+p_value <- unname(kas_sites_retouch_density_corr$p.value )
+t_value <- unname(kas_sites_retouch_density_corr$statistic )
+df_value <- unname(kas_sites_retouch_density_corr$parameter )
+
+size <-  8
+Assemblage_info_retouch_density_ages_prop_main_plot <- 
+ggplot(Assemblage_info_retouch_density_ages_prop,
+       aes(artefact_density,
+           prop_retouched
+           )) +
+  geom_point(aes(colour = date_age / 1000,
+                 size = total_artifacts,
+                 shape = has_sp)) +
+ # geom_text_repel(aes(label = sites),
+ #                 nudge_x = 0.075, 
+ #                 nudge_y = -0.075,
+ #                 size = 5) +
+  geom_smooth(alpha = 0.2,
+              method = "lm") +
+  scale_y_log10(limits = c(0.001, 1),
+                labels = scales::comma_format(accuracy = 0.001)) +
+  scale_x_log10(limits = c(0.001, 10),
+                labels = scales::comma_format(accuracy = 0.001)) +
+  theme_minimal(base_size = 12) +
+  scale_shape_discrete("Contains\nstemmed\npoints?") +
+  scale_size("Total number\nof artifacts") +
+  scale_color_viridis_c(name = "Age of\nassemblage (ka)") +
+  xlab("Artifact volumetric density") +
+  ylab("Proportion of retouched pieces") +
+  annotate("text",
+           x = 0.005,
+           y = 0.7,
+           label = "Curated",
+           size = size -2,
+           colour = "grey50") +
+    annotate("text",
+           x = 5,
+           y = 0.01,
+           label = "Expedient",
+           size = size -2,
+           colour = "grey50") +
+    annotate("text",
+           x = 0.03,
+           y = 0.02,
+           label = glue('r = {round(r_value, 3)}\nt({df_value}) =  {round(t_value, 3)}, p = {round(p_value, 3)}'),
+           size = size - 2,
+           colour = "grey50")
+
+library(cowplot)
+ggdraw(Assemblage_info_retouch_density_ages_prop_main_plot) +
+  draw_plot(retouch_over_time_subplot, 
+            .1, .1, 
+            .35, .25)
+
+```
 
@@
 
-#k_kas_sites_sheets <- map(p_sheets, ~path %>% )
+k_kas_sites_sheets <- map(p_sheets, ~path %>% read_excel )
 

### 2020-08-18 analysis/paper/paper.Rmd
GP added elevation plot and reference

@@
 
+elevation_sp_sub_plot <- 
+ggplot(kasv_tidy,
+       aes(has_sp, 
+           altitude.m._of_main_layer)) +
+  geom_boxplot() +
+  theme_bw(base_size = 8)  +
+  labs(x = "Stemmed points present",
+       y = "Elevation")
+
+elevation_sp_main_plot <- 
 ggplot(mydata_ages, 
@@
 
+library(cowplot)
+ggdraw(elevation_sp_main_plot) +
+  draw_plot(elevation_sp_sub_plot, 
+            .475, .57, 
+            .25, .32) 
 ```

### 2020-08-18 analysis/paper/paper.Rmd
I tried to change the timerange for SPD

@@
 
+### this code takes long time
 spd_test <- modelTest(spd_dates,errors=dates_clean$error,timeRange=c(45000,1000),model='exponential',nsim=200,ncores=3)
@@
 plot(spd_test)
+
+# zoom in in on the 40-33ka period GP is trying
+spd_test_zoom <- modelTest(spd_dates,errors=dates_clean$error,timeRange=c(40000,33000),model='exponential',nsim=200,ncores=3)
+summary(spd_test_zoom)
+plot(spd_test_zoom)
 ```

### 2020-08-18 analysis/paper/paper.Rmd
Updated SPD a little bit

@@
 # zoom in in on the 40-33ka period GP is trying
-spd_test_zoom <- modelTest(spd_dates,errors=dates_clean$error,timeRange=c(40000,33000),model='exponential',nsim=200,ncores=3)
+spd_test_zoom <- modelTest(spd_dates,errors=dates_clean$error,timeRange=c(40000,30000),model='exponential',nsim=200,ncores=3)
 summary(spd_test_zoom)

### 2020-08-18 analysis/paper/paper.Rmd
add different models for spd

@@
 
-# zoom in in on the 40-33ka period GP is trying
+# zoom in in on the 40-30ka period GP is trying
 spd_test_zoom <- modelTest(spd_dates,errors=dates_clean$error,timeRange=c(40000,30000),model='exponential',nsim=200,ncores=3)
@@
 plot(spd_test_zoom)
+
+# zoom in in on the 40-30ka period, testing other models
+uni_test <- modelTest(spd_dates,errors=dates_clean$error,timeRange=c(40000,30000), model='uniform',nsim=200,ncores=3)
+plot(uni_test)
+
+lin_test <- modelTest(spd_dates,errors=dates_clean$error,timeRange=c(40000,30000), model='linear',nsim=200,ncores=3)
+plot(lin_test)
 ```

### 2020-08-19 analysis/paper/climate-model-plots.R
plots using the new climat model data

@@ -17,61 +17,11 @@ temperature <- ncdf4::ncvar_get(env_nc, "temperature")
 biome       <- ncdf4::ncvar_get(env_nc, "biome")
 ncdf4::nc_close(env_nc)
 
-my_year      <- -10000;   # 10,0000 BP
-my_month     <- 6;        # June
-my_longitude <- 0.1218;
-my_latitude  <- 52.2053;  # Cambridge (UK)
-
-my_longitude1 <- 127.108155
-my_latitude1 <-  35.823923  # Bonggok
-
-
-p1 <-
-  print(lattice::levelplot(biome[, , years == my_year], main = "Biome distribution, 10000 BP"))
-
-p2 <-
-  print(lattice::levelplot(temperature[, , months == my_month, years == my_year], main = "Mean June temperature, 10000 BP"))
-
-lonID <- which.min(abs(longitude - my_longitude))
-# check
-longitude[lonID]
-
-latID <- which.min(abs(latitude - my_latitude))
-# check
-latitude[latID]
-
-yearID <- which.min(abs(years - my_year))
-
-
-p3 <-
-  ggplot2::qplot(
-    months,
-    temperature[lonID, latID, , yearID],
-    xlab = "Month",
-    ylab = "Mean temperature",
-    geom = c("point", "line")
-  )
-
-mean_annual_temperature <- apply(temperature, c(1, 2, 4), mean)
-
-p4 <- ggplot2::qplot(
-  years,
-  mean_annual_temperature[lonID, latID, ],
-  xlab = "Year",
-  ylab = "Temperature",
-  main = "Mean annual temperature time series",
-  geom = c("point", "line")
-)
-
-gridExtra::grid.arrange(p1, p2, p3, p4, nrow = 2)
-
 
 # korean archaeological sites
 library(tidyverse)
 
 site_locations_tbl <- read_csv(here::here("analysis/data/raw_data/korean_palaeolithic_site_locations.csv"))
-
-# site_locations_tbl_temps <-
 
 site_locations_tbl_temps <-
 site_locations_tbl %>%
@@ -84,29 +34,126 @@ site_locations_tbl %>%
          latID = which.min(abs(latitude - lat_dd))) %>%
   mutate(mean_annual_temperature = list(tibble(year = years,
                                                mean_annual_temperature = mean_annual_temperature[lonID, latID, ]))) %>%
-  unnest(mean_annual_temperature)
+  unnest(mean_annual_temperature) %>%
+  filter(between(year, -50000, -10000))
+
[…]

### 2020-08-21 analysis/paper/paper.Rmd
adjust the inset plot location for site occupation

@@
   These are the highlights. 
+editor_options: 
+  chunk_output_type: console
 ---
@@
 # site elevation
+kasv_tidy <- 
+kasv %>% 
+  t %>% 
+  as_tibble() %>% 
+  setNames(as.character(.[1,])) %>% 
+  .[-1,] %>% 
+  mutate_all(parse_number) %>% 
+  mutate(artefact_density = total_artifacts / volume,
+         sites = names(kasv)[-1]) %>% 
+  left_join(mydata, by = c('sites' = 'site_name' )) %>% 
+  mutate(has_sp = ifelse(is.na(SP.), "no", "yes"))
 
@@
   draw_plot(elevation_sp_sub_plot, 
-            .475, .57, 
+            .575, .60, 
             .25, .32) 
@@
 #Volume and artefact counts to get density over time.
-kasv_tidy <- 
-kasv %>% 
-  t %>% 
-  as_tibble() %>% 
-  setNames(as.character(.[1,])) %>% 
-  .[-1,] %>% 
-  mutate_all(parse_number) %>% 
-  mutate(artefact_density = total_artifacts / volume,
-         sites = names(kasv)[-1]) %>% 
-  left_join(mydata, by = c('sites' = 'site_name' )) %>% 
-  mutate(has_sp = ifelse(is.na(SP.), "no", "yes"))
 
@@
            y = 0.02,
-           label = glue('r = {round(r_value, 3)}\nt({df_value}) =  {round(t_value, 3)}, p = {round(p_value, 3)}'),
+           label = glue('r = {round(r_value, 3)}\n, t({df_value}) =  {round(t_value, 3)}, p = {round(p_value, 3)}'),
            size = size - 2,

### 2020-08-21 analysis/paper/paper.Rmd
edit elevation plot

@@
   labs(x = "Stemmed points present",
-       y = "Elevation")
+       y = "Elevation above sea level (m)")
 

### 2020-08-21 analysis/paper/paper.Rmd
add t-test to elevation plot

--- 2020-08-21 analysis/paper/paper.Rmd
@@
 
+# compute t-test 
+elev_sp_ttest <- 
+  t.test(altitude.m._of_main_layer ~ has_sp, data = kasv_tidy)
+
+# extract elements from the t-test output
+elev_sp_ttest_t <- round(unname(elev_sp_ttest$statistic), 3)
+elev_sp_ttest_p <- round(unname(elev_sp_ttest$p.value ), 3)
+elev_sp_ttest_df <- round(unname(elev_sp_ttest$parameter ), 3)
+
+# t(degress of freedom) = the t statistic, p = p value.
+elev_sp_ttest_str <- 
+  paste0("t(", elev_sp_ttest_df, ") = ", elev_sp_ttest_t, ", p = ", elev_sp_ttest_p)
+
 elevation_sp_sub_plot <- 
@@
   geom_boxplot() +
+  annotate("text", 
+           x = 1.5, 
+           y = 250,
+           label = elev_sp_ttest_str) +
   theme_bw(base_size = 8)  +
   labs(x = "Stemmed points present",
-       y = "Elevation")
+       y = "Elevation above sea level (m)")
 
@@
   scale_colour_discrete(name = "Contains\nstemmed\npoints?") +
-  theme_minimal(base_size = 16)
+  theme_minimal(base_size = 16)  +
+  theme(legend.position = c(0.1, 0.9))
 
@@
   draw_plot(elevation_sp_sub_plot, 
-            .575, .60, 
+            .7, .60, 
             .25, .32) 

### 2020-08-22 analysis/paper/paper.Rmd
superscript for plot axis title

@@
                  colour = has_sp)) +
-  ylab("Artefact density") +
+  ylab(bquote('Artefact density'~(n/m^3))) +
   xlab("Age of assemblage (ka)") + 

### 2020-08-22 analysis/paper/climate-model-plots.R
add missing line to climate data plotting code.

@@ -17,6 +17,7 @@ temperature <- ncdf4::ncvar_get(env_nc, "temperature")
 biome       <- ncdf4::ncvar_get(env_nc, "biome")
 ncdf4::nc_close(env_nc)
 
+mean_annual_temperature <- apply(temperature, c(1, 2, 4), mean)
 
 # korean archaeological sites
 library(tidyverse)

### 2020-08-22 analysis/paper/climate-model-plots.R
update climate plot code

@@ -66,7 +66,7 @@ ggplot() +
 # now we prepare the data to show each site
 # in a boxplot
 
-korean_archaeological_sites <- readxl::read_excel(here("analysis/data/raw_data/korean-archaeologica-sites.xlsx"))
+korean_archaeological_sites <- readxl::read_excel(here::here("analysis/data/raw_data/korean-archaeologica-sites.xlsx"))
 
 # get site age and location in neat and tidy format
 site_locations_tbl_temps_periods <-

### 2020-08-28 analysis/paper/artifact-composition.R
split the codes for each plot

@@ -1,0 +1,70 @@
+suppressPackageStartupMessages(library(tidyverse))
+library(glue)
+library(ggrepel)
+library(here)
+
+
+# data from PhD data sheet, not KAS sheet.
+mydata <- read.csv(here("analysis/data/raw_data/General_info.csv"))
+
+
+# assemblage composition data from KAS data sheet.
+kasa <- read.csv(here("analysis/data/raw_data/Assemblage_info.csv"))
+
+
+# join artefact type freqs with site data
+
+kasa %>%
+  pivot_longer(-X,
+               names_to = "site_name",
+               values_to = "count") %>%
+  pivot_wider(names_from = "X",
+              values_from = "count") %>%
+  left_join(mydata)
+
+
+
+kasa_long1 <-
+  kasa %>%
+  gather(site,
+         count,
+         -X) %>%
+  filter(!is.na(count)) %>%
+  group_by(site) %>%
+  mutate(percentage = count / sum(count, na.rm = TRUE) * 100,
+         total = sum(count, na.rm = TRUE)) %>%
+  filter(!is.na(percentage)) %>%
+  # if percentage is <10%, call it 'other'
+  mutate(artefact_type = ifelse(as.character(X) == 'stemmed_point',
+                                'stemmed_point',
+                                ifelse(percentage >= 10 | X == "unkown",
+                                       as.character(X), "other_tools"))) %>%
+  mutate(axis_label = glue('{site}\n(n = {total})'))  %>%
+  select(-X)
+# join to get ages of the sites
+kasa_long2 <-
+  kasa_long1 %>%
+  group_by(site,
+           artefact_type) %>%
+  summarise(percentage = sum(percentage)) %>%
+  left_join(mydata_ages,
+            by = c('site' = 'site_name')) %>%
+  arrange(age_ka) %>%
+  distinct(.keep_all = TRUE) %>%
+  filter(!is.na(age_ka)) %>%
+  mutate(axis_label = glue('{site} ({round(age_ka,1)} ka)'))
+
+ggplot(kasa_long2,
+       aes(
+         reorder(axis_label, -age_ka),
+         percentage,
+         fill = artefact_type)) +
+  geom_col(position = "fill") +
+  ylab("Percentage") +
+  xlab("Assemblage (youngest at the top)") +
+  theme_minimal(base_size = 16)  +
+  scale_fill_viridis_d(name = "Artefact type",
+                       option = "D") +
+  coord_flip() +
[…]

### 2020-08-28 analysis/paper/artifact-volumetric-density-retouch-frequency.R

@@ -1,0 +1,126 @@
+suppressPackageStartupMessages(library(tidyverse))
+library(glue)
+library(ggrepel)
+library(here)
+
+
+# data from PhD data sheet, not KAS sheet.
+mydata <- read.csv(here("analysis/data/raw_data/General_info.csv"))
+
+
+#volume of the cultural layer from KAS data sheet.
+kasv <- read.csv(here("analysis/data/raw_data/Dating_info.csv"))
+
+
+
+kasv_tidy <-
+  kasv %>%
+  t %>%
+  as_tibble() %>%
+  setNames(as.character(.[1,])) %>%
+  .[-1,] %>%
+  mutate_all(parse_number) %>%
+  mutate(artefact_density = total_artifacts / volume,
+         sites = names(kasv)[-1]) %>%
+  left_join(mydata, by = c('sites' = 'site_name' )) %>%
+  mutate(has_sp = ifelse(is.na(SP.), "no", "yes"))
+
+
+
+# retouch, density, age
+Assemblage_info <- read_csv(here::here("analysis/data/raw_data/Assemblage_info.csv"))
+
+Assemblage_info_retouch_density_ages <-
+  Assemblage_info %>%
+  pivot_longer(-X1,
+               names_to = "sites") %>%
+  left_join(kasv_tidy)
+
+Assemblage_info_retouch_density_ages_prop <-
+  Assemblage_info_retouch_density_ages %>%
+  filter(!X1 %in% c("debris",
+                    "unkown",
+                    "pebble",
+                    "metate",
+                    "flake",
+                    "unfinished",
+                    "ground",
+                    "blank"
+  )) %>%
+  group_by(sites) %>%
+  mutate(prop_retouched = sum(value, na.rm = TRUE) / total_artifacts) %>%
+  arrange(sites) %>%
+  select(-X1, -value) %>%
+  distinct_all()
+
+# retouch over time
+retouch_over_time_subplot <-
+  ggplot(Assemblage_info_retouch_density_ages_prop,
+         aes(date_age / 1000,
+             prop_retouched)) +
+  geom_point(size = 3,
+             colour = "grey80") +
+  labs(x = "Age of assemblage (ka)",
+       y = "Proportion retouched") +
+  theme_bw(base_size = 6)
+
+# compute correlation
+kas_sites_retouch_density_corr <-
+  cor.test(Assemblage_info_retouch_density_ages_prop$prop_retouched,
+           Assemblage_info_retouch_density_ages_prop$artefact_density)
+
[…]

### 2020-08-28 analysis/paper/artifact-volumetric-density.R

@@ -1,0 +1,91 @@
+suppressPackageStartupMessages(library(tidyverse))
+library(glue)
+library(ggrepel)
+library(here)
+
+korean_archaeological_sites <- readxl::read_excel(here("analysis/data/raw_data/korean-archaeologica-sites.xlsx"))
+
+# data from PhD data sheet, not KAS sheet.
+mydata <- read.csv(here("analysis/data/raw_data/General_info.csv"))
+
+# raw material data from KAS data sheet.
+kasr <- read.csv(here("analysis/data/raw_data/Rawmaterial_info.csv"))
+
+# assemblage composition data from KAS data sheet.
+kasa <- read.csv(here("analysis/data/raw_data/Assemblage_info.csv"))
+
+#volume of the cultural layer from KAS data sheet.
+kasv <- read.csv(here("analysis/data/raw_data/Dating_info.csv"))
+
+dates <- read.csv(here("analysis/data/raw_data/Radiocarbon dates.csv"))
+
+# join artefact type freqs with site data
+
+kasa %>%
+  pivot_longer(-X,
+               names_to = "site_name",
+               values_to = "count") %>%
+  pivot_wider(names_from = "X",
+              values_from = "count") %>%
+  left_join(mydata)
+
+
+
+kasv_tidy <-
+  kasv %>%
+  t %>%
+  as_tibble() %>%
+  setNames(as.character(.[1,])) %>%
+  .[-1,] %>%
+  mutate_all(parse_number) %>%
+  mutate(artefact_density = total_artifacts / volume,
+         sites = names(kasv)[-1]) %>%
+  left_join(mydata, by = c('sites' = 'site_name' )) %>%
+  mutate(has_sp = ifelse(is.na(SP.), "no", "yes"))
+
+
+mydata_ages <-
+  mydata %>%
+  separate(C14.BP., into = c('age', 'error'),
+           sep = "±") %>%
+  mutate(age_ka = parse_number(age) / 1000,
+         error = parse_number(error)) %>%
+  mutate(has_sp = ifelse(!is.na(SP.), "yes", "no"))
+
+
+
+#Volume and artefact counts to get density over time.
+
+density_sp_sub_plot <-
+  ggplot(kasv_tidy,
+         aes(has_sp,
+             artefact_density)) +
+  geom_boxplot() +
+  theme_bw(base_size = 8)  +
+  labs(x = "Stemmed points present",
+       y = "Artefact density")
+
+density_sp_main_plot <-
[…]

### 2020-08-28 analysis/paper/raw-materials.R

@@ -1,0 +1,65 @@
+suppressPackageStartupMessages(library(tidyverse))
+library(glue)
+library(ggrepel)
+library(here)
+
+
+# data from PhD data sheet, not KAS sheet.
+mydata <- read.csv(here("analysis/data/raw_data/General_info.csv"))
+
+# raw material data from KAS data sheet.
+kasr <- read.csv(here("analysis/data/raw_data/Rawmaterial_info.csv"))
+
+# join artefact type freqs with site data
+
+
+
+ #raw material information
+  kasr_long1 <-
+  kasr %>%
+  gather(site,
+         count,
+         -X) %>%
+  group_by(site) %>%
+  mutate(percentage = count / sum(count, na.rm = TRUE) * 100,
+         total = sum(count, na.rm = TRUE)) %>%
+  filter(!is.na(percentage)) %>%
+  # if percentage is <10%, call it 'other'
+  mutate(raw_material = ifelse(percentage >= 10, as.character(X), "other")) %>%
+  mutate(raw_material = ifelse(raw_material == "etc.", "other", raw_material)) %>%
+  select(-X)
+
+
+# if percentage is <10%, call it 'other'
+kasr_long2 <-
+  kasr_long1 %>%
+  group_by(site, raw_material) %>%
+  summarise(percentage = sum(percentage)) %>%
+  ungroup() %>%
+  left_join(kasr_long1 %>%
+              select(site, raw_material, total),
+            by = c('site', "raw_material")) %>%
+  distinct(.keep_all = TRUE) %>%
+  # join to get ages of the sites
+  left_join(mydata_ages,
+            by = c('site' = 'site_name')) %>%
+  # mutate(age_ka = jitter(age_ka),
+  #        age_bin  = ntile(age_ka, 5)) %>%
+  arrange(age_ka) %>%
+  filter(!is.na(age_ka)) %>%
+  mutate(axis_label = glue('{site} ({round(age_ka,1)} ka, n = {total})'))
+
+
+ggplot(kasr_long2,
+       aes(
+         reorder(axis_label,
+                 -age_ka),
+         percentage,
+         fill = raw_material)) +
+  geom_col(position = "fill") +
+  xlab("Assemblage (youngest at the top)") +
+  ylab("Percentage") +
+  scale_fill_viridis_d(name = "Raw material type",
+                       option = "C") +
+  coord_flip() +
+  theme_minimal(base_size = 12)

### 2020-08-28 analysis/paper/summed-probability-distributions.R

@@ -1,0 +1,51 @@
+suppressPackageStartupMessages(library(tidyverse))
+library(glue)
+library(ggrepel)
+library(here)
+
+
+dates <- read.csv(here("analysis/data/raw_data/Radiocarbon dates.csv"))
+
+
+dates_clean <-
+  dates %>%
+  mutate(age = parse_number(as.character(age)),
+         error = parse_number(as.character(error))) %>%
+  filter(age < 50000, method=='AMS')
+
+# devtools::install_github('ahb108/rcarbon')
+library(rcarbon)
+dates_calibrated <-
+  calibrate(x=dates_clean$age,
+            errors=dates_clean$error,
+            calCurves='intcal13')
+
+dates_calibrated_spd <-  spd(dates_calibrated, timeRange = c( 45000, 1000))
+plot(dates_calibrated_spd)
+
+# Testing Observed SPDs against theoretical models
+## recalibrate dates without normalisation to avoid artificial peaks
+spd_dates<- calibrate(x=dates_clean$age,errors=dates_clean$error,normalised=FALSE,verbose=F)
+
+### this code takes long time
+spd_test <- modelTest(spd_dates,errors=dates_clean$error,timeRange=c(45000,1000),model='exponential',nsim=200,ncores=3)
+
+summary(spd_test)
+plot(spd_test)
+
+
+
+
+# zoom in in on the 40-30ka period GP is trying
+spd_test_zoom <- modelTest(spd_dates,errors=dates_clean$error,timeRange=c(40000,30000),model='exponential',nsim=200,ncores=3)
+summary(spd_test_zoom)
+plot(spd_test_zoom)
+
+# zoom in in on the 40-30ka period, testing other models
+uni_test <- modelTest(spd_dates,errors=dates_clean$error,timeRange=c(40000,30000), model='uniform',nsim=200,ncores=3)
+plot(uni_test)
+
+lin_test <- modelTest(spd_dates,errors=dates_clean$error,timeRange=c(40000,30000), model='linear',nsim=200,ncores=3)
+plot(lin_test)
+```
+

### 2020-08-28 analysis/paper/site-elevation.R
site elevation file added

@@ -1,0 +1,82 @@
+suppressPackageStartupMessages(library(tidyverse))
+library(glue)
+library(ggrepel)
+library(here)
+
+
+# data from PhD data sheet, not KAS sheet.
+mydata <- read.csv(here("analysis/data/raw_data/General_info.csv"))
+
+
+#volume of the cultural layer from KAS data sheet.
+kasv <- read.csv(here("analysis/data/raw_data/Dating_info.csv"))
+
+
+# site elevation
+kasv_tidy <-
+  kasv %>%
+  t %>%
+  as_tibble() %>%
+  setNames(as.character(.[1,])) %>%
+  .[-1,] %>%
+  mutate_all(parse_number) %>%
+  mutate(artefact_density = total_artifacts / volume,
+         sites = names(kasv)[-1]) %>%
+  left_join(mydata, by = c('sites' = 'site_name' )) %>%
+  mutate(has_sp = ifelse(is.na(SP.), "no", "yes"))
+
+
+mydata_ages <-
+  mydata %>%
+  separate(C14.BP., into = c('age', 'error'),
+           sep = "±") %>%
+  mutate(age_ka = parse_number(age) / 1000,
+         error = parse_number(error)) %>%
+  mutate(has_sp = ifelse(!is.na(SP.), "yes", "no"))
+
+# compute t-test
+elev_sp_ttest <-
+  t.test(altitude.m._of_main_layer ~ has_sp, data = kasv_tidy)
+
+# extract elements from the t-test output
+elev_sp_ttest_t <- round(unname(elev_sp_ttest$statistic), 3)
+elev_sp_ttest_p <- round(unname(elev_sp_ttest$p.value ), 3)
+elev_sp_ttest_df <- round(unname(elev_sp_ttest$parameter ), 3)
+
+# t(degress of freedom) = the t statistic, p = p value.
+elev_sp_ttest_str <-
+  paste0("t(", elev_sp_ttest_df, ") = ", elev_sp_ttest_t, ", p = ", elev_sp_ttest_p)
+
+
+elevation_sp_sub_plot <-
+  ggplot(kasv_tidy,
+         aes(has_sp,
+             altitude.m._of_main_layer)) +
+  geom_boxplot() +
+  annotate("text",
+           x = 1.5,
+           y = 250,
+           label = elev_sp_ttest_str) +
+  theme_bw(base_size = 8)  +
+  labs(x = "Stemmed points present",
+       y = "Elevation above sea level (m)")
+
[…]

### 2020-12-04 analysis/paper/paper.Rmd
Commit for push!
split the codes for each plot
upgraded R studio and tested it

@@
 
-The appearance of stemmed points in the Late Pleistocene (~40-35 ka) in Korea is often assumed to have transformed forager lifeways because it is thought to reflect more specialed hunting practices [@lee2002new;@seong2008tanged;@chang2013human].
+The appearance of stemmed points in the Late Pleistocene (\~40-35 ka) in Korea is often assumed to have transformed forager lifeways because it is thought to reflect more specialed hunting practices [@lee2002new; @seong2008tanged; @chang2013human].
 Stemmed points are likely to have been projectile weapon tips because the stem of these tools is cconsidered to represent halfting technology that joins the lithic to a wooden shaft.
@@
 risk management (Australian small tool tradition) -->
+
 <!-- Do you want to have one more paragraph for the general tradition of explaining the emergence of new stone tools? -->
@@
 These innovations include blade and micro technology, high frequencies of retouched blade tools, several novel refined tools such as projectile points, stemmed points, end scrapers, burins, denticulates, etc, and using high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary].
-Previous work has argued that the appearance of new stone artefact technologies in this region may be linked to modern human dispersals by focusing on the origin of the new tools [@seong2008tanged;@bae2010origin;@Bae_2010; @bae2017origin].
+Previous work has argued that the appearance of new stone artefact technologies in this region may be linked to modern human dispersals by focusing on the origin of the new tools [@seong2008tanged; @bae2010origin; @Bae_2010; @bae2017origin].
 
@@
 In Locality 9 dated to c.29,000years, there is a small scatter of stone artefacts consists of blades, Levallois flakes, cores and other retouched flakes.
-The recent excavtion of Locality 12 shows that its archaeological layer <!--I shouldn't use 'layer', but I don't know what to say shortly-->, dated to 13,078–13,296 cal BP, includes ground stone and more than 30,000 microlithics made out of a variety of raw materials such as fine-grained and siliceous rocks.
+The recent excavtion of Locality 12 shows that its archaeological layer <!--I shouldn't use 'layer', but I don't know what to say shortly-->, dated to 13,078--13,296 cal BP, includes ground stone and more than 30,000 microlithics made out of a variety of raw materials such as fine-grained and siliceous rocks.
 In addition, more than 10,000 animal fossils and bone tools including a tool for fishing nets, an awl, and two needles were excavated [@pei2012shuidonggou; @gao2014discovery; @zhang2016bone].<!-- they use OSL data that has wise range, or calibrated years --> Based on the archaeological records, people argue that a microblade technology appeared in China after 29ka BP related to mobility change along with colder climate towards LGM and changed game as surviving strategy [@yi2014significance].
@@
 <!-- the last sentense is a bit long-->
+
 <!-- "advanced core and flake tools means intentionally prepared platforms and working surfaces, selective raw material usage (using chert rather tahn river pebble), bipolar reduction, and generating small elongated blanks as results" -->
@@
 36,000 cal BP) [@yamaoka2012use; @izuho201421].
-The components of the new technologies represented by trap pits, cobble concentrations, hearths, charcoal concentrations, and toolkits including trapezoids, pointed-shaped backed blades, backed points, burins, end scrapers, side scrapers, wedges, beak- shaped tools, axes, edge-ground axes, hammerstones, cobble tools, and anvils [ @bae2017origin; @izuho201421].
+The components of the new technologies represented by trap pits, cobble concentrations, hearths, charcoal concentrations, and toolkits including trapezoids, pointed-shaped backed blades, backed points, burins, end scrapers, side scrapers, wedges, beak- shaped tools, axes, edge-ground axes, hammerstones, cobble tools, and anvils [@bae2017origin; @izuho201421].
 These innovations have been interpreted as evidnce of new foraging methods such as watercraft (marine transport of obsidian from the Kozu Island) and bow-and-arrow technology, drivien by increased population and ecological changes [@Yamaoka_2012; @bae2017origin; @Nakazawa_Bae_2018; @morisaki2019early].
 Especially, hafted trapzoids were designed to be used as multifucntional tools and people tried to maintain and resue trapezoids and tools euipped with trapzoid to adapt to the specific environmental settings on the Japanese Islands [@ono2002radiocaron; @yamaoka2012use].
-Ishinomoto 8-ku is one of the earliets sites dated to 39,690–34,790 cal BP, located in Takuma upland of Kumamoto Prefecture.
+Ishinomoto 8-ku is one of the earliets sites dated to 39,690--34,790 cal BP, located in Takuma upland of Kumamoto Prefecture.
 A total of 500 stone artifacts were discovered including trapezoids, side scrapers, flakes, flake cores, edge-ground axes, and cobble tools.
@@
 Recenly, accummulated radiocarbon data show that -->
-<<!-- do we need to mention popular research topic such as existance of pre UP ("archaic industries")as a heated debate?--->
+
+\<<!-- do we need to mention popular research topic such as existance of pre UP ("archaic industries")as a heated debate?--->
 
@@
 Flake tools became more complex, with a greater diversity in both type and size along with continuously used core tools during the Late Pleistocene [@Bae_2010; @Lee_2013; @Lee_2016].
-The oldest stemmed points so far is from the Yongho-dong site dated to 38.5 ka [@seong2009emergence; #### Park 2012 ####; @seong2015diversity; @bae2017origin].
+The oldest stemmed points so far is from the Yongho-dong site dated to 38.5 ka [@seong2009emergence; \#\#\#\# Park 2012 \#\#\#\#; @seong2015diversity; @bae2017origin].
 
@@
 However, signs of early technological change in Korea, including stemmed points and blades, have been discovered from deposits dated to 40-35 ka BP from Hwadae-ri, Hopyeong-dong, and Yongho-dong, pre-dating the Shuidonggou finds [@seong2009emergence].
-The debate about origins can be summarized into two models competing to explain the appearance of the new Late Pleistocene technologies: in situ evolution [@seong2009emergence] and ‘heterogenic’ migration [@bae2010origin].
-The ‘in situ’ model claims that stemmed points and other Late Paleolithic assemblages including a blade industry autonomously emerged in the south of the Korean peninsula with no external influence.
+The debate about origins can be summarized into two models competing to explain the appearance of the new Late Pleistocene technologies: in situ evolution [@seong2009emergence] and 'heterogenic' migration [@bae2010origin].
+The 'in situ' model claims that stemmed points and other Late Paleolithic assemblages including a blade industry autonomously emerged in the south of the Korean peninsula with no external influence.
 Seong (2009) uses the blade-to-flake ratios of lithic assemblages to support his claim.
@@
 The migration model argues that the new blade industry, including stemmed points, and the earlier coarse flake tradition including large cores, polyhedrons, choppers, and even handaxes, came from different origins.
-While Seong’s model (2009) proposes that the ‘heterogenic’ character is the result of indigenous development, Bae (2010) proposes that it is the result of the continuous influx of modern human populations from both Siberia and southern China.
-Bae’s model suggests blade technology arrived in the Korean peninsula from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP [@bae2013early].
-<!--- I don't know how to connect this debate to our research. It is really needed? --->
+While Seong's model (2009) proposes that the 'heterogenic' character is the result of indigenous development, Bae (2010) proposes that it is the result of the continuous influx of modern human populations from both Siberia and southern China.
+Bae's model suggests blade technology arrived in the Korean peninsula from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP [@bae2013early].
+<!--- I don't know how to connect this debate to our research.
+It is really needed?
+--->
 
@@
 The elevation is the main proxy for environmental information in this research.
-The Korean paleoenvironmental data and information are not rich enough to directly apply HBE driven models or questions [####Han 2008####; @Chang_2013].
-Rather, the paleoenvironment of the Korean Peninsula for this proposed research time period can be simply summarized as ‘overall cooler and drier during MIS 3’ [###Han 2008###; ####Seong 2008####; ###Choi 2011####; @Chang_2013; @im2015study; @bak2017late].
+The Korean paleoenvironmental data and information are not rich enough to directly apply HBE driven models or questions [\#\#\#\#Han 2008\#\#\#\#; @Chang_2013].
+Rather, the paleoenvironment of the Korean Peninsula for this proposed research time period can be simply summarized as 'overall cooler and drier during MIS 3' [\#\#\#Han 2008\#\#\#; \#\#\#\#Seong 2008\#\#\#\#; \#\#\#Choi 2011\#\#\#\#; @Chang_2013; @im2015study; @bak2017late].
 In this regard, it is much harder to understand the regional as well as shorter-time scale of paleoenvironmental change in the Korean Peninsula.
@@
 Combined the two proxies, the artifact volumetric density and the retouch frequency, we expect to determine whether or not the hunter-gatherers were residentially mobile or logistically organized.
-If the assemblage in a site has high density and less retouched, it is depicted as expedient assemblage that represents “base camps” or “residences”, while less density with high retouched one shows curated assemblage and indicates either residential mobility or certain task groups from main residential site [@Riel-Salvatore_Barton_2004; @clark2017lithics].
+If the assemblage in a site has high density and less retouched, it is depicted as expedient assemblage that represents "base camps" or "residences", while less density with high retouched one shows curated assemblage and indicates either residential mobility or certain task groups from main residential site [@Riel-Salvatore_Barton_2004; @clark2017lithics].
 
@@
 In terms of the composition of toolkits, we count kind of tools as well as the amount of debitabitage to search for the change in the toolkits around the transition period.
-The tools represent paleo hunter-gatherers’ adaptive strategies.
+The tools represent paleo hunter-gatherers' adaptive strategies.
 We analyze the composition of raw materials in the assemblages to see the relationships between the choice of specific raw materials and technological innovation as well as raw material availability and its consuming pattern based on an assumption that stone raw materials were carefully transported and consumed to optimize mobility and technological strategies, depth of planning in landscape use and risk minimization [@brantingham2003neutral].
 
-Umap: <- depending on whether we are gonna add umap (or PCA) analysis result
+Umap: \<- depending on whether we are gonna add umap (or PCA) analysis result
 
@@
 
-# comparable figure to the raw material one (BM created)
+# comparable figure to the raw material one
 
@@
 ```
-Artifact density and retouch <- maybe we need to add the result of retouch, separately 
-We use the combination of the artifact density and retouch frequency as a proxy to identify the mobility and residential pattern of hunter-gatherers.
+
+Artifact density and retouch \<- maybe we need to add the result of retouch, separately We use the combination of the artifact density and retouch frequency as a proxy to identify the mobility and residential pattern of hunter-gatherers.
 The result shows that there is no clear pattern during the transition period and then it became more expedient at the end of the transition (Fig 5.
@@
 ```
+
 Umap
@@
 Combining with other materials increases utility or functionality of the tool by adding extra weight and providing a shaft, which makes it earlier to grip.
-Selective use of raw material might impact on hunter-gathers’ mobility and occupation pattern depending on availability of the specific raw materials such as shale and rhyolite.
+Selective use of raw material might impact on hunter-gathers' mobility and occupation pattern depending on availability of the specific raw materials such as shale and rhyolite.
 In addition, this research shows that carrying the points enabled hunter-gatherers to occupy higher elevation compared to the sites with no stemmed points.
@@
 Therefore, the curated technology indicates sites with special or highly mobile groups with stone artifacts that are thoroughly designed and based on a high level of investment of time and effort in artifact manufacture and maintenance.
-The expedient technology represents base camp or residence with artifacts that tend to be manufactured, used, and abandoned in the short period of time [@vaquero2018searching;@binford1979organization; @torrence1989time].
+The expedient technology represents base camp or residence with artifacts that tend to be manufactured, used, and abandoned in the short period of time [@vaquero2018searching; @binford1979organization; @torrence1989time].
 There appear to be more variety in the type of sites (or group) during the transition period, both expedient or curated.
@@
 
-Compute artefact volumetric density by frequency of retouched piece (log scale), following Julien Riel-Salvatore and C. Michael Barton in  https://www.jstor.org/stable/4128419
+Compute artefact volumetric density by frequency of retouched piece (log scale), following Julien Riel-Salvatore and C. Michael Barton in <https://www.jstor.org/stable/4128419>
 
@@
 <!-- The following line inserts a page break when the output is MS Word. For page breaks in PDF, use \newpage on its own line.  -->
+
 ##### pagebreak
@@
 # References
+
 <!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
-<div id="refs"></div>
+
+::: {#refs}
+:::
 

### 2020-12-04 analysis/paper/paper.Rmd
working on moving rmd file

@@
 ---
-title: "Changes in Mobility and Site Occupation during the Late Pleistocene in Korea"
+title: "How did the introduction of Stemmed Points affect Mobility and Site Occupation during the Late Pleistocene in Korea?"
 author:
@@
 abstract: |
-  The Late Pleistocene saw the introduction of new lithic technologies to the Korean Peninsula, especially stemmed points and blades.
-We examined stone artifacts from 22 sites in Korea to investigate how the introduction of  new tools changed the way people used the landscape, and the way people used sites.
-We analyzed artifact density, retouch frequency, radiocarbon ages, site elevation, raw material types, and composition of the toolkit to test hypotheses about site use.
-The results show that the new technology enabled the occupation of marginal environments at higher elevations.
-We find no strong relationship between technology and ecological change.
-Instead, we assume that population dynamics might have had more impact on changes in stone artifact technology and hunter-gatherer foraging strategies.
+  Human behavioural ecological models use stone artefacts as a source of information for understanding forager mobility.
+We use these models to examine stone artefacts from 22 sites in Korea to investigate mobility and site occupation during the Late Pleistocene.
+This study focuses on two questions: how did the introduction of stemmed points change forager' mobility and landscape use?
+And how did this new technology change the way people use habitation sites?
+We analyse artefact volumetric density, retouch frequency, composition of toolkits, artefact raw materials, environmental and demographic contexts.
+Quartz and side scrapers, in addition to cores and choppers, remain dominant in assemblages before and after the introduction of stemmed points throughout the Late Pleistocene.
+Our results show that forager groups using stemmed points were associated with occupation of marginal or extreme environments.
+In addition, groups with stemmed points were associated with expedient technologies indicating residential and less mobile behaviours.
+The environmental context of this technological innovation was a gradual decrease in temperature into the LGM.
+A population increase followed after the appearance of stemmed points.
+
 keywords: |
-  keyword 1; keyword 2; keyword 3
+  stemmed points; Korean Paleolithic; mobility; site occupation
 highlights: |
@@
 
-<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->
+# Introduction
 
-```{r, setup, echo = FALSE}
-knitr::opts_chunk$set(
-  collapse = TRUE,
-  warning = FALSE,
-  message = FALSE,
-  echo = FALSE,
-  comment = "#>",
-  fig.path = "../figures/"
-)
+The appearance of stemmed points in the Late Pleistocene (\~40-35 ka) in Korea is often assumed to have transformed forager lifeways because it reflected more specialized hunting practices [\@seong2008tanged; \@chang2013human].
+Stemmed points are often called as tanged points in the western hemisphere, but we use 'stemmed point' to distinguish from Bronze Age stone projectile points that have long been called 'tanged points' in Korea.
+The stemmed point is a projectile point made out of elongated flake or blade with slight retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft.
+Stemmed points are likely to have been projectile weapon tips because the stem of these tools is considered to represent hafting technology that joins the lithic to a wooden shaft.
+If they were projectile weapon tips, stemmed points represent the first appearance of hafting technology in Korea.
+Most Korean Paleolithic sites do not preserve organic remains, so we lack direct fauna evidence on hunting and subsistence strategies.
+Here we explore questions about the implications of the appearance of this new technology using a behavioural ecological framework and stone artefacts.
+We explore changes in the way people occupied the landscape, the way people used sites, and impact of social and ecological environment.
+We examined stone artefacts from 22 sites in South Korea to investigate the changes in human behaviour associated with the appearance stemmed points, and the blade industry that followed.
 
-#library(koreapaleolithicmobilityoccupation)# Or use devtools::load_all('.', quiet = T) if your code is in script files, rather than as functions in the `/R` diretory
-```
+Previous work on this technological change in Korea has focussed on issues of the transition from the Early to the Late Paleolithic and modern human dispersals [\@norton2009evolution; \@bae2010origin].
+To our knowledge, the implications of the appearance of stemmed points for forager mobility and site use have not been explored in detail.
+Here, we attempt to identify patterns of mobility through analysis of the stone artefact assemblages using Human Behavioural Ecology (HBE) models.
+Drawing on established relationships between lithic technology and forager mobility [\@shott1986settlement; \@Hiscock_1994; \@kuhn1994formal; \@kuhn2004upper; \@kuhn2015artifacts; \@kuhn2016moves; \@capriles2018], we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that are part of a broader strategy of moving frequently and further, possibly as an adaptation to environmental or population changes.
 
-```{r setup, include=FALSE}
-suppressPackageStartupMessages(library(tidyverse))
-library(glue)
-library(ggrepel)
-library(here)
+# Background
 
-korean_archaeological_sites <- readxl::read_excel(here("analysis/data/raw_data/korean-archaeologica-sites.xlsx"))
+In the Eastern hemisphere, major technological innovations during the Late Pleistocene are restricted to Northeast Asia, such as northern China, Japan and Korea.
+These innovations include blade and micro technology, high frequencies of retouched blade tools, several novel tools such as projectile points, stemmed points, end scrapers, burins, denticulates, etc, and using high-quality raw materials [\@bar1999big; \@brantingham2001initial; \@Bar-Yosef_2002; \@bae2017origin; \@nakazawa2018quaternary].
+Previous work has argued that the appearance of new stone artefact technologies in this region may be linked to modern human dispersals [\@seong2008tanged;\@bae2010origin;\@Bae_2010; \@bae2017origin].
 
-# data from PhD data sheet, not KAS sheet.
-mydata <- read.csv(here("analysis/data/raw_data/General_info.csv"))
+The Shuidonggou site in northern China is an important example of the transition from cobble to blade tool industries in the Late Pleistocene of Northeast Asia [\@Brantingham_Perreault_2010; \@Gao_Zhang_Yang_Shen_Wu_2010; \@pei2012shuidonggou].
+Shuidonggou Locality SDG 2, which dates to around c.
+32,000 years, contains fauna specimens, ostrich eggshell beads, and stone artefacts including blades, retouched flakes, cores, and debris along with well-preserved hearths.
+In Locality 9 dated to c.
+29,000 years, there is a small scatter of stone artefacts consisting of blades, Levallois flakes, cores and other retouched flakes.
+The recent excavation of Locality 12 shows that its archaeological context, dated to 13,078--13,296 cal BP, includes ground stone and more than 30,000 microlithics made out of a variety of raw materials such as fine-grained and siliceous rocks.
+In addition, more than 10,000 animal fossils and bone tools including a tool for fishing nets, an awl, and two needles were excavated [\@pei2012shuidonggou; \@gao2014discovery; \@zhang2016bone].
+Stemmed points are not part of the Shuidonggou assemblages, but microblade technology is argued to have appeared in China after 29ka BP, related to changes in mobility associated with colder climates towards LGM [\@yi2014significance].
+Shuidonggou shows that the blade component came from Siberia and/or Mongolia at 41 ka (SDG Locality1 and 9), advanced core and flake tools were likely to have been locally developed at 33 ka (2 and 8), and at 10.8 ka microblade technology appears but the origin is uncertain (12) [\@yi2014significance; \@li2019history].\
+\
+Late Pleistocene technological innovations in Japan appeared in different regions of the archipelago around 38,000 years ago accompanied by remains of Homo sapiens found in Okinawa Island (c.
+36,000 cal BP) [\@yamaoka2012use; \@izuho201421].
+These new technologies include trap pits, cobble concentrations, hearths, charcoal concentrations, and toolkits including trapezoids, pointed-shaped backed blades, backed points, burins, end scrapers, side scrapers, wedges, beak-shaped tools, axes, edge-ground axes, hammerstones, cobble tools, and anvils [\@bae2017origin; \@izuho201421].
+These innovations have been interpreted as evidence of new foraging methods such as watercraft (marine transport of obsidian from the Kozu Island) and bow-and-arrow technology, driven by increased population and ecological changes [\@Yamaoka_2012; \@bae2017origin; \@Nakazawa_Bae_2018; \@morisaki2019early].
+Hafted trapezoids were likely multifunctional tools adapted to the specific environmental settings of the different Japanese Islands [\@ono2002radiocaron; \@yamaoka2012use].
+Ishinomoto 8-ku is one of the earliest sites dated to 39,690--34,790 cal BP, located in Takuma upland of Kumamoto Prefecture.
+A total of 500 stone artefacts were discovered including trapezoids, side scrapers, flakes, flake cores, edge-ground axes, and cobble tools.
+The main lithic raw materials are chert and andesite [\@izuho201421].
+Stemmed points appear much later, dated to between about 15,500 and 13,800 cal BP [\@ono2002radiocaron; \@tsutsumi2007dynamics].
 
-# raw material data from KAS data sheet.
-kasr <- read.csv(here("analysis/data/raw_data/Rawmaterial_info.csv"))
+What makes the Korean Late Pleistocene technological transition distinctive from what we see at Shuidonggou and in Japan is that the earliest signs of new technologies in Korean assemblages are stemmed points, followed by blade technologies.
+Stemmed points appeared in both regions after the case of Korean Peninsula.
+The oldest stemmed points so far is from the Yonghodong site dated to 38.5 ka and made on elongated flakes [\@seong2009emergence; \@seong2015diversity; \@bae2017origin].
+After blades appeared at c.
+27 ka and were adopted to the stemmed points, the shape of points became more standardized with one or two ridges on the dorsal side and triangular cross section.
+Replacing flake to blade as a blank of the points led to an increase in quantity over time.
+In addition to the stemmed points, flake tools became more complex, with a greater diversity in both type and size along with continuously used core tools during the Late Pleistocene  [\@Bae_2010; \@Lee_2013; \@Lee_2016].
+There are several attempts to explain the technological transition and they can be summarized into two competing models: 'heterogenic' migration [\@bae2010origin] or *in situ* evolution [\@seong2009emergence].
+The migration model argues that the new blade industry including stemmed points, and the earlier coarse flake tradition including large cores, polyhedrons, choppers, and even handaxes, came from different origins as the result of continuing influx of model human migration from two routes.
+Bae (2010) explains that blade and stemmed points arrived in the Korean peninsula from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP with the earliest evidence of blade in those regions.
+He assumes that core and flake industry come from southern China based on the existence of similar assemblages in both Korea and southern China in addition to a genetic analysis of the Y-chromosome of the modern population [\@bae2013early; \@lee2013current].
+The '*in situ*' model claims that stemmed points and other Late Paleolithic assemblages including blade industries autonomously emerged in the south of the Korean peninsula with no external influence [\@seong2008tanged].
+This claim is supported by the earliest appearance of stemmed points among Northeast Asia, such as Hwadae-ri, Hopyeong-dong, and Yonghodong sites dated to 40-35ka BP [\@seong2009emergence].
+Seong (2009) uses the increased blade-to-flake ratios on lithic assemblages to support the *in situ* model based on the premise that the blade industry represents new technology while flakes indicate a continuing local one.
+\
 
-# assemblage composition data from KAS data sheet.
-kasa <- read.csv(here("analysis/data/raw_data/Assemblage_info.csv"))
+<!-- need to put two images,  one for stemmed points, one for overall lithic assemblages -->
 
-#volume of the cultural layer from KAS data sheet.
-kasv <- read.csv(here("analysis/data/raw_data/Dating_info.csv"))
+## Behavioural ecology and forager land use behaviours
 
-dates <- read.csv(here("analysis/data/raw_data/Radiocarbon dates.csv"))
+Our brief review of Late Pleistocene technological innovations in Northeast Asia shows that previous work has largely focussed on timing, location, origins and description of these new tools, but modelling the land-use behaviours that generated these assemblages has rarely been undertaken.
+We focus on the appearance of stemmed points in Korea because these are a new technology that represents a new hunting method.
+This is because the hafting implies throwing techniques that can reach long-distance targets [\@Keeley_1982; \@kuhn2015artifacts].
+We draw on behavioural ecological theory to model the effects of this technological change.
+Behavioural ecology theory offers structured and testable models based on optimality assumptions that stand on economic rationality and environmental knowledge [\@winterhalder1992evolutionary; \@prentiss2019handbook].
+The patch choice model and marginal value theorem predict that forager mobility and travel times will increase when resource patch productivity decreases [\@smith1991inujjuamiunt; \@smith1983anthropological; \@bettinger1997evolutionary; \@wolverton2015patch; \@llano2015optimal].
+We can also model individual artefact types as a kind of resource patch.
+Foragers may find a more costly and complicated technology (such as stemmed points relative to flakes) optimal if they have a long cumulative time in use, if they are maintainable and reusable in a landscape where lithic raw material supply is uncertain.
+That is, they may move to a patch where they can stay longer (i.e.
+an artefact type that has a long and extendible use-life) when they are not certain of the productivity and travel times of other patches [\@kuhn2015artifacts].
+
+We ask two questions to investigate the period around the introduction of stemmed points in Korea.
+First, how did the introduction of the new tools change foragers' landscape use?
+Specifically, was the use of stemmed points associated with occupation of marginal or extreme environments?
+Our assumption here is that this new technology might have allowed foragers to explore and sojourn in less productive landscapes by increasing their mobility.
+In Australia during the mid-Holocene, Hiscock et al.
+argue that the first appearance of microlithics in many locations represent a portable and multi-functional toolkit that minimize travel expenses and increase tool readiness to adapt to patchy and unpredictable resources associated with environmental change [\@Hiscock_1994; \@Hiscock_Clarkson_Mackay_2011]. Similarly, in Japan, Morisaki et al.
+[\@Morisaki_Izuho_Terry_Sato_2015] argue that the transition from trapezoid to the blades and projectile points around 25 ka cal BP in north Paleo-Honshu Island enabled the foragers to extend their occupation into cold grassland landscapes.
+
+Second, we ask how the new technology changed the way people use habitation sites and mobility.
+After the appearance of these new tools, did people tend to stay in one location for longer or shorter periods, perhaps for specific purposes?
+The stone artefact assemblage from a site can inform us how long individuals or groups stayed and give insights into their activities [\@binford1979organization; \@kelly1995foraging; \@kuhn2016moves; \@holdaway2019surface].
+For example, stone artifact density and the frequency of retouched pieces (scaled to the volume of excavated sediment) can be used as proxies to represent occupation patterns [\@barton2011modeling; \@clark2017lithics].
+Assemblages with low density but a high proportion of retouched and backed pieces may indicate the remains of 'short-term camp', which is a site for small and ephemeral overnight camp or limited activity station.
+On the other hand, the combination of high density and low retouched pieces represent logistically organized basecamp, which is a site with greater residential stability, long site occupation and occupied by larger groups [\@clark2017lithics; \@clark2019landscapes].
+
+# Methods
+
+## Sites and dates
+
+After the first excavation of a Paleolithic site in Korea at Seokjang-ri site in the 1960s, more than 200 Paleolithic sites have been discovered in South Korea [\@lee2019were].
+We selected sites dated to before and after the transition period (40-35 ka) to analyze the process of technological change.
+This resulted in a sample of analysed 33 assemblages from 22 sites spanning 49-24 ka.
+We identified multiple assemblages in a site where culturally sterile deposits separated artefact-bearing deposits, or where stratigraphic units could be identified by major differences in the texture and composition of the sedimentary deposit containing the stone artefacts.
+For example, the Hwadae-ri site has three cultural layers.
+The lowest horizon at the bottom, dated to 39,000±1400 BP by OSL, contains coarse flake tools made of vein quartz and quartzite.
+The middle cultural horizon has stemmed points made of porphyry, with quartz and quartzite dominating the assemblage.
+This layer was dated to 31,200±900 BP by radiocarbon dating and 30,000±1,700 BP by OSL.
+The uppermost layer, dated to 22,000±100 BP by OSL, contains blades, scrapers, awls, and denticulates [\@seong2009emergence].
+We obtained the data from published excavation reports held by provincial museums.
+
+```{r korea-map, fig.cap="Site map. Korean Paleolithic sites mentioned in the text."}
+source(here::here("analysis/paper/001-draw-the-map.R"))
+knitr::include_graphics(here::here("analysis/figures/001-site-map.png"))
 
@@
 
-```{r}
-
-# join artefact type freqs with site data
-
-kasa %>% 
-  pivot_longer(-X, 
-               names_to = "site_name",
-               values_to = "count") %>% 
-  pivot_wider(names_from = "X",
-              values_from = "count") %>% 
-  left_join(mydata)
-
-```
-
-# Introduction
-
-The appearance of stemmed points in the Late Pleistocene (\~40-35 ka) in Korea is often assumed to have transformed forager lifeways because it is thought to reflect more specialed hunting practices [@lee2002new; @seong2008tanged; @chang2013human].
-Stemmed points are likely to have been projectile weapon tips because the stem of these tools is cconsidered to represent halfting technology that joins the lithic to a wooden shaft.
-Stemmed points represent the first appearance of halfting technology in Korea.
-Here we explore questions about these transformations using as behavioural ecology framework to explore changes in the way people used the landscape, and the way people used sites.
-We examined stone artifacts from 22 sites in South Korea to investigate the technological transition represented by the appearance of a new tool, stemmed points, and the blade industry that followed.
-Previous work on this technological change in Korea has focussed on issues of the transition from the Early to the Late Paleolithic and modern human dispersals [@norton2009evolution; @bae2010origin].
-To our knowledge, the implications for forager mobility and site use have not been explored in detail.
-Most Korean Paleolithic sites do not preserve organic remains, so we lack direct faunal evidence on hunting and subsistence strategies.
-Here, we attempt to identify patterns of mobility through analysis of the stone artefact assemblages using Human Behavioral Ecology (HBE) models.
-Drawing on established relationships between lithic technology and forager mobility [@shott1986settlement; @Hiscock_1994; @kuhn1994formal; @kuhn2004upper; @kuhn2015artifacts; @kuhn2016moves; @capriles2018], we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that are part of a broader strategy of moving frequently and further, possibly as an adaptation to population or environmental changes.
-
-<!-- generally, how do we make sense of new stone tools...
-When a new stone artefact technology appears, how do we explain that?
-new species...
-population movements...
-(great basin)...
-environment...
-(Mesolithic & bow and arrow)...
-risk management (Australian small tool tradition) -->
-
-<!-- Do you want to have one more paragraph for the general tradition of explaining the emergence of new stone tools? -->
-
-# Background
-
-### Regional archaeology
-
-In the Eastern hemisphere, striking technological innovations during the Late Pleistocene are restricted to Northeast Asia, such as northern China, Japan and Korea.
-These innovations include blade and micro technology, high frequencies of retouched blade tools, several novel refined tools such as projectile points, stemmed points, end scrapers, burins, denticulates, etc, and using high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary].
-Previous work has argued that the appearance of new stone artefact technologies in this region may be linked to modern human dispersals by focusing on the origin of the new tools [@seong2008tanged; @bae2010origin; @Bae_2010; @bae2017origin].
-
-The Shuidonggou site in northern China is a typical site for showing the transition from cobble to blade tool industries in the Late Pleistocene of Northeast Asia from the Late Pleistocene to Middle Holocene [@Brantingham_Perreault_2010; @Gao_Zhang_Yang_Shen_Wu_2010; @pei2012shuidonggou].
-Shuidonggou Localities SDG 2, which dated around c.32,000 years, contains aunal specimens,ostrich eggshell beads,and stone artefacts including blades, retouched flakes, cores, and debries along with well-preserved hearths.
-In Locality 9 dated to c.29,000years, there is a small scatter of stone artefacts consists of blades, Levallois flakes, cores and other retouched flakes.
-The recent excavtion of Locality 12 shows that its archaeological layer <!--I shouldn't use 'layer', but I don't know what to say shortly-->, dated to 13,078--13,296 cal BP, includes ground stone and more than 30,000 microlithics made out of a variety of raw materials such as fine-grained and siliceous rocks.
-In addition, more than 10,000 animal fossils and bone tools including a tool for fishing nets, an awl, and two needles were excavated [@pei2012shuidonggou; @gao2014discovery; @zhang2016bone].<!-- they use OSL data that has wise range, or calibrated years --> Based on the archaeological records, people argue that a microblade technology appeared in China after 29ka BP related to mobility change along with colder climate towards LGM and changed game as surviving strategy [@yi2014significance].
-The accumulated studies on Shuidonggou site show that the blade compoenent is claimed to have come from Siberia and/or Mongolia at 41 ka (SDG Locality1 and 9), advanced core and flake tools are likely to be locally developed at 33ka (2 and 8), and at 10.8 ka microblade technology appears but the origin is uncertain (12) [@yi2014significance; @li2019history].
-
-<!-- the last sentense is a bit long-->
-
-<!-- "advanced core and flake tools means intentionally prepared platforms and working surfaces, selective raw material usage (using chert rather tahn river pebble), bipolar reduction, and generating small elongated blanks as results" -->
-
-The origins of Late Pleistocene technological innovations in Japan is suddenly appeared in differen regions of the archipelago around 38,000 years ago accaompanying with Homo sapiens found in Okinawa Island (ca.
-36,000 cal BP) [@yamaoka2012use; @izuho201421].
-The components of the new technologies represented by trap pits, cobble concentrations, hearths, charcoal concentrations, and toolkits including trapezoids, pointed-shaped backed blades, backed points, burins, end scrapers, side scrapers, wedges, beak- shaped tools, axes, edge-ground axes, hammerstones, cobble tools, and anvils [@bae2017origin; @izuho201421].
-These innovations have been interpreted as evidnce of new foraging methods such as watercraft (marine transport of obsidian from the Kozu Island) and bow-and-arrow technology, drivien by increased population and ecological changes [@Yamaoka_2012; @bae2017origin; @Nakazawa_Bae_2018; @morisaki2019early].
-Especially, hafted trapzoids were designed to be used as multifucntional tools and people tried to maintain and resue trapezoids and tools euipped with trapzoid to adapt to the specific environmental settings on the Japanese Islands [@ono2002radiocaron; @yamaoka2012use].
-Ishinomoto 8-ku is one of the earliets sites dated to 39,690--34,790 cal BP, located in Takuma upland of Kumamoto Prefecture.
-A total of 500 stone artifacts were discovered including trapezoids, side scrapers, flakes, flake cores, edge-ground axes, and cobble tools.
-The main lithic raw materials are chert and andesite [@izuho201421].
-
-<!--do we need to talk about AT???People had assumed that Japanese archipelago started to be occupied before 30,000 cal BP based on the existence of primary Aira-Tanzawa tephra (dated at 30,009±189 cal BP) in the excavation unit.
-Recenly, accummulated radiocarbon data show that -->
-
-\<<!-- do we need to mention popular research topic such as existance of pre UP ("archaic industries")as a heated debate?--->
-
-The Late Pleistocene technological transition in Korea is mainly characterized by the early appearance of stemmed points, followed by blade technology.
-The stemmed point is a projectile point made out of elongated flake or blade and with slight retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft.
-The earliest stemmped points were made on flakes.
-After blades appear at c.
-27 ka, the shape of stemmped points became more standardized with one or two ridges on dorsal side and triangular cross section.
-The stemmed point is often called as a tanged point in western hemisphere, but we use 'stemmed point' to distinguish from Bronze Age stone projectile points in Korea that have long been called 'tanged points'.
-Stemmed points were made from blade blanks, and became more standardized and increased in quantity over time.
-Flake tools became more complex, with a greater diversity in both type and size along with continuously used core tools during the Late Pleistocene [@Bae_2010; @Lee_2013; @Lee_2016].
-The oldest stemmed points so far is from the Yongho-dong site dated to 38.5 ka [@seong2009emergence; \#\#\#\# Park 2012 \#\#\#\#; @seong2015diversity; @bae2017origin].
-
-The origin of the stemmed point is one of the most frequently discussed topics among Paleolithic research in Korea.
-Especially, this debate includes questions about the origin of the Late Paleolithic period in Korea, the existence and nature of Middle Paleolithic technology, and reconstructions of anatomically modern human dispersals in eastern Asia.
-Most Korean archaeologists had assumed that the new blade technology represented came from Shuidonggou in the northeast China because this site has one of the earliest blade industries (around 35,000 BP) in northeast Asia and it is geographically close to the Korean peninsula [seong2008tanged; @lee2013current].
-However, signs of early technological change in Korea, including stemmed points and blades, have been discovered from deposits dated to 40-35 ka BP from Hwadae-ri, Hopyeong-dong, and Yongho-dong, pre-dating the Shuidonggou finds [@seong2009emergence].
-The debate about origins can be summarized into two models competing to explain the appearance of the new Late Pleistocene technologies: in situ evolution [@seong2009emergence] and 'heterogenic' migration [@bae2010origin].
-The 'in situ' model claims that stemmed points and other Late Paleolithic assemblages including a blade industry autonomously emerged in the south of the Korean peninsula with no external influence.
-Seong (2009) uses the blade-to-flake ratios of lithic assemblages to support his claim.
-In his view, the blade industry represents new technology while flakes indicate a continuing local one.
-He argues that the increased ratio of blades in lithic assemblages shows an expansion of the new technology.
-The migration model argues that the new blade industry, including stemmed points, and the earlier coarse flake tradition including large cores, polyhedrons, choppers, and even handaxes, came from different origins.
-While Seong's model (2009) proposes that the 'heterogenic' character is the result of indigenous development, Bae (2010) proposes that it is the result of the continuous influx of modern human populations from both Siberia and southern China.
-Bae's model suggests blade technology arrived in the Korean peninsula from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP [@bae2013early].
-<!--- I don't know how to connect this debate to our research.
-It is really needed?
---->
-
-### Behavioural ecology and forager land use behaviours
-
-Our review of Late Pleistocene technological innovations in Northeast Asia shows that previous work has largely focussed on timing, location, origins and description of these new tools, but there has been little effort to model the land-use behaviours that generated these assemblages.
-We focus on the appearance of stemmed points in Korea because these are one of the earliest new technologies that definiately represents a new hunting methods because the halfting implies throwing techniques that can reach long-distance targets [@Keeley_1982; @kuhn2015artifacts].
-We ask two questions to invetigate the period around the introduction of stemmed points in Korea.
-
-First, we ask how the introduction of the new tools changed foragers' landscape use, specifically, was the use of stemmed points associated with occupation of marginal or exterme environments?
-The new technology might allow foragers to explore and stay in different or harsher territories by increasing mobility.
-For instance, Morisaki et al.
-argue that the transition from trapezoid to the blades and projectile points around 25 ka cal BP in north Paleo-Honshu Island enabled the foragers to extend their occupation at residential sites [@Morisaki_Izuho_Terry_Sato_2015].
-In Australia during the mid-Holocene, the highly retouched and finely shaped tools combined with shafts were appeared to broaden their occupations to offshore islands on the northeast and northwest coast [@Hiscock_1994].
-Hiscock et al.
-insist that the appearance of mirolithcs is great example to show that people made portable and multi-functional tools to minimize the travel expense and increase tool readiness [@Hiscock_Clarkson_Mackay_2011].
-
-Second, we focus on how the new technology changed the way people use their habitation sites.
-After the appearance of these new tools, did they tend to stay in one location longer, or for shorter periods, perhaps for specific purposes?.
-Archaeological sites indicate depositional phenomena and places where artifacts and other debris accumulated reflecting human activities in the past [@kuhn2015artifact].
-The stone artifact assemblage from a site can inform us how long individuals or groups stayed and give insignts into what they did there [@binford1979organization; @kelly1995foraging; @kuhn2016moves; @holdaway2019surface].
-Stone artifacts can contain a record of raw material transport and selection and give insights into land-use behaviour [@clarkson2008lithics].
-For example, stone artifact density and the frequency of retouched pieces (scaled to the volume of excavated sediment) are often used as proxies to represent occupation patterns.
-Assemablges with low density but a high proportion of retouched and backed pieces would indicate the remains of 'short-term camp', while the combination of high density and low retouched pieces represent 'logistically organized basecampe' [@clark2017lithics; clark2019landscapes].
-
-<!-- Lastly, we want to know about population dynamics during the transition period based on a premise that population change could be a driver of the new invention as well as the source of it if they brought the new tools from outside.
-The demands derived from significan population dynamic in addition to environmental change could lead to development of the tools [buchanan2016drivers; @mihailovic2016technological; prufer2019linking].
-Some studies argue that technological complexity and richness are highly influenced by population size because in large population, because new innovation could occur during the process of learning from others [@shennan2001demography; @henrich2004demography; @collared2005causes; @collard2013risk; @buchanan2016drivers].
-Therefore, we will check the population change during the transitional period to understand its impact on the technology.
--->
-
-# Methods
-
-Sites and dates
-
-The data analyzed in this paper are gathered from published excavation reports of 22 Paleolithic open air sites in South Korea dating from 49k BP to 24k BP.
-Within these 22 sites we have 33 assemblages.
-Multiple assemblages in a site were distinguised culturally sterile deposits in between artefact-bearing deposits, or by stratigraphic units identified by major differences in the texture and composition of the sedimentary deposit containing the stone artefacts.
-After the first excavation at Jeongokri site in the late 1970s, more than 200 Paleolithic sites are discovered in Korea.
-Among them, we chose 22 sites based on absolute dates that span our period of interest.
-We include sites dated before and after the transition period (40-35 ka) to analyze the process of technological transition.
-
-```{r cache = TRUE, fig.cap="Map of sites discussed in the text."}
-
-```
-
-Assemblages
-
-The research analyzes total 33 assemblages [need to explain what is the difference between 'assemblage' and 'site'] and they consist of cores, blade cores, blades, flakes, debris, hammers, choppers, plans, polyhedrals, side scrapers, end scrapers, notches, cleavers, stemmed points, awls, denticulates, burins, handaxes, blanks, knives, handadzes, peaks, flatters, flakers etc.
-
-Some artifacts are not included as certain tools due to their vague or broken shape presumably related to the characteristics of raw materials such as quartz, quartz vein, and quartzite.
-In that case, we named them as unknown, unfinished, or pebble.
-The stone artifacts from all of these sites are stored in museums located throughout South Korea, and have been briefly described in the excavation reports.
-We have visited most of these collections, collected PDFs of the excavation reports, and negotiated permission to study the materials.
-For example, the Hwadae-ri site has three cultural layers.
-The lowest horizon at the bottom, dated to 39,000±1400 BP by OSL, contains coarse flake tools made of vein quartz and quartzite.
-The middle cultural horizon has stemmed points made of porphyry, with quartz and quartzite dominating the assemblage.
-This layer dated to 31,200±900 BP by radiocarbon dating and 30,000±1,700 BP by OSL.
-The uppermost layer, dated to 22,000±100 BC by OSL, contains blades, scrapers, awls, and denticulates [@seong2009emergence].
-We compare the assemblages from the two older layers at this and other sites to investigate the transition process.
-
-Statistical methods
-
-To answer the three research questions, we gather information such as elevation of the sites, artifact density, retouch frequency, radiocarbon dates toolkit composition, and raw material type.
-The elevation is the main proxy for environmental information in this research.
-The Korean paleoenvironmental data and information are not rich enough to directly apply HBE driven models or questions [\#\#\#\#Han 2008\#\#\#\#; @Chang_2013].
-Rather, the paleoenvironment of the Korean Peninsula for this proposed research time period can be simply summarized as 'overall cooler and drier during MIS 3' [\#\#\#Han 2008\#\#\#; \#\#\#\#Seong 2008\#\#\#\#; \#\#\#Choi 2011\#\#\#\#; @Chang_2013; @im2015study; @bak2017late].
-In this regard, it is much harder to understand the regional as well as shorter-time scale of paleoenvironmental change in the Korean Peninsula.
-In addition, the targeting period is not exactly matched with major environmental changes that we can check through global sea level change, it seems rather monotonous around 40-35 ka [@Bailey_2010; @Lorscheid_Stocchi_Casella_Gómez-Pujol_Vacchi_Mann_Rovere_2017].
-In this research, we chose elevation to make a premise that the site elevation can represent different environment in local scale and hunter-gatherers moved to different elevation as a result of environment related risk.
-This research focuses on the change of the elevation of the sites throughout time and accompanying toolkits, and its relationship with the appearance of new tools.
-
-The artifact volumetric density is defined as the total number of artifacts per cubic meter of excavated sediment, and served as a proxy for accumulation ratio of artifacts [@clark2017lithics; @clark2019landscapes].
-We expected to use this value to evaluate mobility and land-use practices of hunter-gatherers during the Late Pleistocene.
-We count the retouch frequency as the number of retouched tools divided by the total number of artifacts.
-This value represents general proficiency of lithic manufacturing, raw material availability, and foraging strategies including mobility and duration of site occupation based on the concept that tool reductioning is an economical tactic to generate sharp, usable edges while minimizing the cost of transporting tools or bulky raw materials by reducing the total load weight [@Kuhn_1990; @clark2019landscapes].
-In other words, producing and carrying highly retouched tools will increase portability and efficiency by decrease in the size of both individual tools and the assemblages so that hunter-gatherers could move often or further, or both [@kuhn1991unpacking; @kuhn1994formal; @kuhn2004upper; @andrefsky1994raw; @davies2018modeling].
-For example, highly retouched and finely shaped implements appeared in Australia during the mid-Holocene to minimize risk associated with environmental change and high mobility [@Hiscock_1994].
-
-Combined the two proxies, the artifact volumetric density and the retouch frequency, we expect to determine whether or not the hunter-gatherers were residentially mobile or logistically organized.
-If the assemblage in a site has high density and less retouched, it is depicted as expedient assemblage that represents "base camps" or "residences", while less density with high retouched one shows curated assemblage and indicates either residential mobility or certain task groups from main residential site [@Riel-Salvatore_Barton_2004; @clark2017lithics].
-
-Radiocarbon dates provide a proxy for change in human population levels.
-Summed probability distributions (SPD) of radiocarbon dates are often used to understand temporal trends in ancient human populations [@rick1987dates; @bamforth2012radiocarbon; @shennan2013regional; @timpson2014reconstructing; @contreras2014summed].
-Even though there must be unavoidable limitations for the validity of SPD such as the lack of excavated sites in certain period, this method is still helpful to be used as a proxy for population fluctuation.
-Timpson et al.
-demonstrate its usefulness by testing data sets in 12 European regions and assert that this SPD is extremely robust to deal with small sample size [@timpson2014reconstructing].
-We use 171 radiocarbon dates from 22 sites to search for population trends during the Late Pleistocene in Korea based on the premise that population change or pressure could be another catalyst to improve or innovate technology.
-
-In terms of the composition of toolkits, we count kind of tools as well as the amount of debitabitage to search for the change in the toolkits around the transition period.
-The tools represent paleo hunter-gatherers' adaptive strategies.
-We analyze the composition of raw materials in the assemblages to see the relationships between the choice of specific raw materials and technological innovation as well as raw material availability and its consuming pattern based on an assumption that stone raw materials were carefully transported and consumed to optimize mobility and technological strategies, depth of planning in landscape use and risk minimization [@brantingham2003neutral].
-
-Umap: \<- depending on whether we are gonna add umap (or PCA) analysis result
-
 # Results
 
-Elevation of occupied sites
-
-The result of the elevation change of 33 cultural layers from 22 sites corresponding to the targeting period shows that there is no remarkable change can be seen in site elevation during the transition (Fig.1: with no sp site distinction).
-We applied the existence of the stemmed points to the elevation to understand the benefit of having the points in the toolkit and the result indicates that the hunter-gathering groups who carried the technology could occupy higher altitude while the groups with no points stayed at the lower altitude (Fig.
-2: distinction between sites with no stemmed points and with stemmed points).
-
-```{r}
-# site elevation
-kasv_tidy <- 
-kasv %>% 
-  t %>% 
-  as_tibble() %>% 
-  setNames(as.character(.[1,])) %>% 
-  .[-1,] %>% 
-  mutate_all(parse_number) %>% 
-  mutate(artefact_density = total_artifacts / volume,
-         sites = names(kasv)[-1]) %>% 
-  left_join(mydata, by = c('sites' = 'site_name' )) %>% 
-  mutate(has_sp = ifelse(is.na(SP.), "no", "yes"))
-
-mydata_ages <- 
-  mydata %>% 
-  separate(C14.BP., into = c('age', 'error'),
-          sep = "±") %>% 
-  mutate(age_ka = parse_number(age) / 1000,
-         error = parse_number(error)) %>% 
-  mutate(has_sp = ifelse(!is.na(SP.), "yes", "no"))
-
-# compute t-test 
-elev_sp_ttest <- 
-  t.test(altitude.m._of_main_layer ~ has_sp, data = kasv_tidy)
-
-# extract elements from the t-test output
-elev_sp_ttest_t <- round(unname(elev_sp_ttest$statistic), 3)
-elev_sp_ttest_p <- round(unname(elev_sp_ttest$p.value ), 3)
-elev_sp_ttest_df <- round(unname(elev_sp_ttest$parameter ), 3)
-
-# t(degress of freedom) = the t statistic, p = p value.
-elev_sp_ttest_str <- 
-  paste0("t(", elev_sp_ttest_df, ") = ", elev_sp_ttest_t, ", p = ", elev_sp_ttest_p)
-
-elevation_sp_sub_plot <- 
-ggplot(kasv_tidy,
-       aes(has_sp, 
-           altitude.m._of_main_layer)) +
-  geom_boxplot() +
-  annotate("text", 
-           x = 1.5, 
-           y = 250,
-           label = elev_sp_ttest_str) +
-  theme_bw(base_size = 8)  +
-  labs(x = "Stemmed points present",
-       y = "Elevation above sea level (m)")
-
-elevation_sp_main_plot <- 
-ggplot(mydata_ages, 
-       aes(x = age_ka,
-           y = altitude.m._of_main_layer,
-       color = as.factor(has_sp))) +
-  geom_point(size = 3) +
-  xlab("Age of occupation (ka)") +
-  ylab("Elevation above sea level (m)") +
-  geom_smooth(se = FALSE) +
-  scale_colour_discrete(name = "Contains\nstemmed\npoints?") +
-  theme_minimal(base_size = 16)  +
-  theme(legend.position = c(0.1, 0.9))
-
-library(cowplot)
-ggdraw(elevation_sp_main_plot) +
-  draw_plot(elevation_sp_sub_plot, 
-            .7, .60, 
-            .25, .32) 
-```
-
-Artifact density and Composition of toolkit
-
-During the transitional period, the artifact volumetric density was relatively small but it becomes higher around 30 ka (Fig.
-3: the transition of artifact volumetric density).
-When it comes to the composition of the toolkit, it is hardly recognized any pattern of changes over time (Fig.
-4: the composition of toolkit).
-Both results with no information about raw material availability near the 22 sites, we assume that the artifact density might relate to the duration of site occupation or group size rather than portability and efficiency of tools.
-In other words, the small artifact density might indicate shorter duration or smaller group size.
-Likewise, the composition of toolkit might represent the characteristics or purpose of each site, rather than showing general pattern of chronological change.
-
-```{r}
-#Volume and artefact counts to get density over time.
-
-density_sp_sub_plot <- 
-ggplot(kasv_tidy,
-       aes(has_sp, 
-           artefact_density)) +
-  geom_boxplot() +
-  theme_bw(base_size = 8)  +
-  labs(x = "Stemmed points present",
-       y = "Artefact density")
-
-density_sp_main_plot <- 
-ggplot(kasv_tidy,
-       aes(date_age / 1000,
-           artefact_density)) +
-  geom_point(aes(size = total_artifacts,
-                 colour = has_sp)) +
-  ylab(bquote('Artefact density'~(n/m^3))) +
-  xlab("Age of assemblage (ka)") + 
-  #geom_smooth(method = "lm") + 
-  scale_size_continuous(name = "Total number\nof artifacts")  +
-  scale_color_viridis_d(name = "Contains\nstemmed\npoints?") +
-  theme_minimal(base_size = 16)  #+
-  #geom_text_repel(aes(label = sites),
-  #                nudge_x = 0.25, 
-  #                nudge_y = 0.25)
-
-# https://wilkelab.org/cowplot/articles/drawing_with_on_plots.html
-library(cowplot)
-ggdraw(density_sp_main_plot) +
-  draw_plot(density_sp_sub_plot, 
-            .375, .67, 
-            .25, .25)
-
-ggsave("figures/age-by-density.png")
-```
-
-```{r}
-#the artifact composition of each sites to figure out what people did as they were occupying the sites.
-kasa_long <-
-  kasa %>%
-  gather(site, 
-         count, 
-         -X) %>% 
-  filter(!is.na(count)) %>% 
-  group_by(site) %>% 
-  mutate(percentage = count / sum(count, na.rm = TRUE) * 100,
-         total = sum(count, na.rm = TRUE)) %>% 
-  filter(!is.na(percentage)) %>% 
-  # if percentage is <10%, call it 'other'
-  mutate(artefact_type = ifelse(as.character(X) == 'stemmed_point',
-                                'stemmed_point', 
-                                ifelse(percentage >= 10, 
-                                as.character(X), "other_tools"))) %>% 
-  mutate(axis_label = as.character(glue('{site}\n(n = {total})')))  %>% 
-  # join to get ages of the sites
-  left_join(mydata_ages, 
-            by = c('site' = 'site_name')) %>% 
-  arrange(age_ka) #%>% 
-  # mutate(age_ka = jitter(age_ka))
-
-ggplot(kasa_long, 
-       aes(age_ka, 
-           percentage,
-           fill = artefact_type)) +
-  geom_col(position = "fill") +
-  ylab("Percentage") +
-  xlab("Age of assemblage (ka)") +
-    scale_fill_viridis_d(name = "Artifact type",
-                       option = "D") +
-  coord_flip() +
-  theme_minimal(base_size = 16)
-
-# comparable figure to the raw material one
-
-kasa_long1 <-
-  kasa %>%
-  gather(site, 
-         count, 
-         -X) %>% 
-  filter(!is.na(count)) %>% 
-  group_by(site) %>% 
-   mutate(percentage = count / sum(count, na.rm = TRUE) * 100,
-         total = sum(count, na.rm = TRUE)) %>% 
-  filter(!is.na(percentage)) %>% 
-  # if percentage is <10%, call it 'other'
-  mutate(artefact_type = ifelse(as.character(X) == 'stemmed_point',
-                                'stemmed_point', 
-                                ifelse(percentage >= 10 | X == "unkown", 
-                                as.character(X), "other_tools"))) %>% 
-  mutate(axis_label = glue('{site}\n(n = {total})'))  %>% 
-  select(-X)
-  # join to get ages of the sites
-kasa_long2 <- 
-kasa_long1 %>% 
-  group_by(site, 
-           artefact_type) %>% 
-  summarise(percentage = sum(percentage)) %>% 
-  left_join(mydata_ages, 
-            by = c('site' = 'site_name')) %>% 
-  arrange(age_ka) %>% 
-  distinct(.keep_all = TRUE) %>%
-  filter(!is.na(age_ka)) %>% 
-  mutate(axis_label = glue('{site} ({round(age_ka,1)} ka)'))
-
-ggplot(kasa_long2, 
-       aes(
-           reorder(axis_label, -age_ka),
-           percentage,
-           fill = artefact_type)) +
-  geom_col(position = "fill") +
-  ylab("Percentage") +
-   xlab("Assemblage (youngest at the top)") +
-  theme_minimal(base_size = 16)  +
-    scale_fill_viridis_d(name = "Artefact type",
-                       option = "D") +
-  coord_flip() +
-  theme_minimal(base_size = 12) 
-ggsave("figures/artefact-types.png")
-
-```
-
-Artifact density and retouch \<- maybe we need to add the result of retouch, separately We use the combination of the artifact density and retouch frequency as a proxy to identify the mobility and residential pattern of hunter-gatherers.
-The result shows that there is no clear pattern during the transition period and then it became more expedient at the end of the transition (Fig 5.
-The combination of artifact density and retouch).
-That is to say, the hunter-gatherer became either more residential or the majority of the site that this research studies were used as base camps or residence rather than shortly occupied camps for certain tasks.
-
-```{r}
-# retouch, density, age
-Assemblage_info <- read_csv(here::here("analysis/data/raw_data/Assemblage_info.csv"))
-
-Assemblage_info_retouch_density_ages <- 
-Assemblage_info %>% 
-  pivot_longer(-X1, 
-               names_to = "sites") %>% 
-  left_join(kasv_tidy)
-
-Assemblage_info_retouch_density_ages_prop <- 
-Assemblage_info_retouch_density_ages %>% 
-     filter(!X1 %in% c("debris", 
-                     "unkown", 
-                     "pebble",  
-                     "metate", 
-                     "flake",
-                     "unfinished",
-                     "ground",
-                     "blank"
-                     )) %>% 
-  group_by(sites) %>% 
-  mutate(prop_retouched = sum(value, na.rm = TRUE) / total_artifacts) %>% 
-  arrange(sites) %>% 
-  select(-X1, -value) %>% 
-  distinct_all()
-
-# retouch over time
-retouch_over_time_subplot <- 
-ggplot(Assemblage_info_retouch_density_ages_prop,
-       aes(date_age / 1000,
-           prop_retouched)) +
-  geom_point(size = 3, 
-             colour = "grey80") +
-  labs(x = "Age of assemblage (ka)",
-       y = "Proportion retouched") +
-  theme_bw(base_size = 6)
-
-# compute correlation
-kas_sites_retouch_density_corr <- 
-cor.test(Assemblage_info_retouch_density_ages_prop$prop_retouched,
-         Assemblage_info_retouch_density_ages_prop$artefact_density)
-
-r_value <- unname(kas_sites_retouch_density_corr$estimate)
-p_value <- unname(kas_sites_retouch_density_corr$p.value )
-t_value <- unname(kas_sites_retouch_density_corr$statistic )
-df_value <- unname(kas_sites_retouch_density_corr$parameter )
-
-size <-  8
-Assemblage_info_retouch_density_ages_prop_main_plot <- 
-ggplot(Assemblage_info_retouch_density_ages_prop,
-       aes(artefact_density,
-           prop_retouched
-           )) +
-  geom_point(aes(colour = date_age / 1000,
-                 size = total_artifacts,
-                 shape = has_sp)) +
- # geom_text_repel(aes(label = sites),
- #                 nudge_x = 0.075, 
- #                 nudge_y = -0.075,
- #                 size = 5) +
-  geom_smooth(alpha = 0.2,
-              method = "lm") +
-  scale_y_log10(limits = c(0.001, 1),
-                labels = scales::comma_format(accuracy = 0.001)) +
-  scale_x_log10(limits = c(0.001, 10),
-                labels = scales::comma_format(accuracy = 0.001)) +
-  theme_minimal(base_size = 12) +
-  scale_shape_discrete("Contains\nstemmed\npoints?") +
-  scale_size("Total number\nof artifacts") +
-  scale_color_viridis_c(name = "Age of\nassemblage (ka)") +
-  xlab("Artifact volumetric density") +
-  ylab("Proportion of retouched pieces") +
-  annotate("text",
-           x = 0.005,
-           y = 0.7,
-           label = "Curated",
-           size = size -2,
-           colour = "grey50") +
-    annotate("text",
-           x = 5,
-           y = 0.01,
-           label = "Expedient",
-           size = size -2,
-           colour = "grey50") +
-    annotate("text",
-           x = 0.03,
-           y = 0.02,
-           label = glue('r = {round(r_value, 3)}\n, t({df_value}) =  {round(t_value, 3)}, p = {round(p_value, 3)}'),
-           size = size - 2,
-           colour = "grey50")
-
-library(cowplot)
-ggdraw(Assemblage_info_retouch_density_ages_prop_main_plot) +
-  draw_plot(retouch_over_time_subplot, 
-            .1, .1, 
-            .35, .25)
-
-```
-
-Summed probability distribution
-
-To understand the population change during the appearance of stemmed points, we analyzed the summed probability distributions of 171 radiocarbon dates from 33 cultural layers to get a rough image of population dynamics at the sites.
-The result indicates that there is a higher density of radiocarbon dates from around 34 ka.
-We could cautiously assume that the population flux might relate to the technological innovation.
-
-```{r cache=TRUE}
-dates_clean <- 
-  dates %>% 
-  mutate(age = parse_number(as.character(age)),
-         error = parse_number(as.character(error))) %>% 
-  filter(age < 50000, method=='AMS')
-
-# devtools::install_github('ahb108/rcarbon')
-library(rcarbon)
-dates_calibrated <- 
-  calibrate(x=dates_clean$age,
-               errors=dates_clean$error,
-               calCurves='intcal13')
-
-dates_calibrated_spd <-  spd(dates_calibrated, timeRange = c( 45000, 1000))
-plot(dates_calibrated_spd)
-
-# Testing Observed SPDs against theoretical models
-## recalibrate dates without normalisation to avoid artificial peaks
-spd_dates<- calibrate(x=dates_clean$age,errors=dates_clean$error,normalised=FALSE,verbose=F)
-
-### this code takes long time
-spd_test <- modelTest(spd_dates,errors=dates_clean$error,timeRange=c(45000,1000),model='exponential',nsim=200,ncores=3)
-
-summary(spd_test)
-plot(spd_test)
-
-# zoom in in on the 40-30ka period GP is trying
-spd_test_zoom <- modelTest(spd_dates,errors=dates_clean$error,timeRange=c(40000,30000),model='exponential',nsim=200,ncores=3)
-summary(spd_test_zoom)
-plot(spd_test_zoom)
-
-# zoom in in on the 40-30ka period, testing other models
-uni_test <- modelTest(spd_dates,errors=dates_clean$error,timeRange=c(40000,30000), model='uniform',nsim=200,ncores=3)
-plot(uni_test)
-
-lin_test <- modelTest(spd_dates,errors=dates_clean$error,timeRange=c(40000,30000), model='linear',nsim=200,ncores=3)
-plot(lin_test)
-```
-
-Raw materials
-
-During the transition period, the use of shale and rhyolite greatly increased.
-These two raw materials are suitable to make elongate blades and stemmed points.
-It is considered that he hunter-gatherer started to use raw materials selectively depending on the type of tool, especially the newly produced stemmed points.
-Raw material availability is a crucial factor in determining the trip and direction for hunter-gatherers [@davies2018modeling].
-Therefore the adoption of new raw material along with the emergence of new type of tools demonstrates a significant innovation in not just technology but also cognitive system.
-
-```{r}
-#raw material information
-kasr_long1 <- 
-  kasr %>% 
-  gather(site, 
-         count, 
-         -X) %>% 
-  group_by(site) %>% 
-  mutate(percentage = count / sum(count, na.rm = TRUE) * 100,
-         total = sum(count, na.rm = TRUE)) %>% 
-  filter(!is.na(percentage)) %>% 
-  # if percentage is <10%, call it 'other'
-  mutate(raw_material = ifelse(percentage >= 10, as.character(X), "other")) %>% 
-  mutate(raw_material = ifelse(raw_material == "etc.", "other", raw_material)) %>% 
-  select(-X)
-
-  # if percentage is <10%, call it 'other'
-kasr_long2 <- 
-  kasr_long1 %>% 
-  group_by(site, raw_material) %>% 
-  summarise(percentage = sum(percentage)) %>% 
-  ungroup() %>% 
-  left_join(kasr_long1 %>% 
-              select(site, raw_material, total),
-            by = c('site', "raw_material")) %>% 
-  distinct(.keep_all = TRUE) %>% 
-  # join to get ages of the sites
-  left_join(mydata_ages, 
-            by = c('site' = 'site_name')) %>% 
- # mutate(age_ka = jitter(age_ka),
- #        age_bin  = ntile(age_ka, 5)) %>% 
-  arrange(age_ka) %>% 
-  filter(!is.na(age_ka)) %>% 
-    mutate(axis_label = glue('{site} ({round(age_ka,1)} ka, n = {total})'))
-
-ggplot(kasr_long2, 
-       aes(
-           reorder(axis_label, 
-                   -age_ka),
-           percentage,
-           fill = raw_material)) +
-  geom_col(position = "fill") +
-  xlab("Assemblage (youngest at the top)") +
-  ylab("Percentage") +
-  scale_fill_viridis_d(name = "Raw material type",
-                       option = "C") +
-  coord_flip() +
-  theme_minimal(base_size = 12)
-
-```
-
-Umap
-
 # Discussion
-
-1.  The new technology and occupation in high elevation
-
-The new technology appeared around 40-35ka in the Korean Peninsula includes stemmed points and following blade industry.
-The stemmed points brought new paradigms in stone artifact assemblages.
-For example, projectile points are only a part of composit artifacts and this specific characteristics could contribute to increase efficiency of maintaining the toolkit by easily replacing the damaged part [@kuhn2015artifacts; @cardillo2010some].
-Combining with other materials increases utility or functionality of the tool by adding extra weight and providing a shaft, which makes it earlier to grip.
-Selective use of raw material might impact on hunter-gathers' mobility and occupation pattern depending on availability of the specific raw materials such as shale and rhyolite.
-In addition, this research shows that carrying the points enabled hunter-gatherers to occupy higher elevation compared to the sites with no stemmed points.
-Therefore, we can assume that the technology would play an important role to survive in more risky environments.
-
-2.  Population change
-
-Interpreted in HBE, we expected there might be close connection between environmental change and the appearance of new technology based on the prediction that the new invention was the result of risk.
-The research suggests that the environmental pressure could be related to population change.
-We cannot simply predicate that the increased population around 35ka was due to dispersal of Homo Sapiens since there is no reliable human remains in South Korea.
-However, the migration of Homo sapiens was not impossible scenario at the same time.
-In fact, one model explains that the emergence of stemmed points and blade industry came from outside of the Korean peninsula and the new technology was the result of the continuous influx of modern human populations from Siberia while the earlier coarse flake trandition was come from southern China [@Bae_2010].
-Since this migration model barely explains about preexisted stone artifact assemblages and people in the Korean Peninsula, it is hard to accept his argument without knowing about the native people.
-In short, we assume that the stemmed points appeared related to the increased population and it can be related to the modern human dispersal or indigenously increased occupants.
-
-3.  Expedient vs curated
-
-To monitor mobility strategies and site occupation pattern, we examined the artifact density and retouch frequency, based on the premise that highly mobile foragers would have light weighted, highly retouched, replaceable, almost completely made tools and leave small amount of artifacts when they move to the new location (curated).
-One the other hand, less mobile foragers who tend to stay in one place yield more debitage and debris, produce less efficient tools since they can easily get supplied (expedient) [@clark2019landscapes; @kuhn1994formal; @kuhn2014mousterian; @meignen2006middle].
-Therefore, the curated technology indicates sites with special or highly mobile groups with stone artifacts that are thoroughly designed and based on a high level of investment of time and effort in artifact manufacture and maintenance.
-The expedient technology represents base camp or residence with artifacts that tend to be manufactured, used, and abandoned in the short period of time [@vaquero2018searching; @binford1979organization; @torrence1989time].
-There appear to be more variety in the type of sites (or group) during the transition period, both expedient or curated.
-It became more expedient at the end of the transition, which means foragers became less mobile and tended to stay in the base camp.
-Combined with the result of population dynamics, we can assume that the increased population made each group bigger and let them stay in one place.
-
-# Conclusion
-
-This research aims to understand how the appearance of new technology affected hunter-gatherers in Korea and answer the three questions, how did the introduction of the new tools change the way people used the landscape, the way people used sites, and the population.
-To answer the question, we use HBE as a theoretical backgrounds and analyze stone artifacts from 22 sites dated before and after 40-35ka.
-The new technology is represented the stemmed points and other blade industry made out of new raw materials such as shale and rhyolite.
-We could clearly see that hunter-gatherers chose these materials particularly to make the newly added stone artifacts to the toolkit.
-The results also show that the new technology enables the occupation of marginal environments at higher elevations.
-Besides of technological change, we found that there are more varieties in the type of sites or groups during the transition period, both curated and expedient and then it becomes more expedient.
-In other words, the foragers became less mobile and prone to stay in one place after the transition ended.
-We expected to see any pattern that would explain the relationship between environmental change and the emergence of the new technology.
-However, there was not enough paleoenvironmental information as well as data to apply to verify the model.
-Therefore we focus on Radiocarbon dates as a proxy of population change based on the premise that population driven pressure could trigger people to invent a new tool.
-The results show that population dynamics might have had more impact on changes in stone artifact technology.
-
-------------NOT WORKING-------------
-
-Get the lithic data for some sites:
-
-```{r}
-
-# try not to use googlesheet and make the upper chunck work
-
-library(readxl)
-path <- "analysis/data/raw_data/Overall_info.xlsx"
-excel_sheets(path = path)
-
-p_sheets <- c("general_info" ,"assemblage_info")
-
-k_kas_sites_sheets <- map(p_sheets, ~path %>% read_excel )
-
-#simply combine the two data sets of general info and assemblage info?
-#kk_sheets <- c (mydata, kasa)
-
-```
-
-Compute artefact volumetric density by frequency of retouched piece (log scale), following Julien Riel-Salvatore and C. Michael Barton in <https://www.jstor.org/stable/4128419>
-
-```{r}
-# put the site names on this list of data frames
-site_names <-  kas_sites_sheets[[1]] %>%  
-  select(-X1) %>% 
-  names()
-
-# density of arefacts at each site
-artefact_density <- 
-kas_sites_sheets[[1]] %>% 
-  t %>% 
-  as.data.frame() %>% 
- rownames_to_column() %>% 
-  mutate_all(as.character)
-
-names(artefact_density) <- as.character(as.vector(artefact_density[1, ]))
-artefact_density <- artefact_density[-1,]
-
-artefact_density <- 
-artefact_density %>% 
-  mutate(density = as.numeric(total_artifacts) / as.numeric(volume)) %>% 
-   mutate_at(2:ncol(.), as.numeric)
-
-# frequency of retouched artefacts
-retouched_artefacts <- 
- kas_sites_sheets[[2]] %>% 
-   filter(!X1 %in% c("debris", 
-                     "unkown", 
-                     "pebble",  
-                     "metate", 
-                     "flake",
-                     "unfinished",
-                     "ground",
-                     "blank"
-                     )) %>% 
-    t %>% 
-  as.data.frame() %>% 
- rownames_to_column() %>% 
-  mutate_all(as.character)
-
-names(retouched_artefacts) <- as.character(as.vector(retouched_artefacts[1, ]))
-retouched_artefacts <- retouched_artefacts[-1,]
-
-retouched_artefacts <- 
-retouched_artefacts %>% 
-  mutate_at(2:ncol(.), as.numeric)  %>% 
-  mutate(total_retouched = rowSums(.[,-1], na.rm = TRUE))
-
-# join together the site age, artefact density df with the retouched artefacts df
-kas_density_retouch_ages <- 
-artefact_density %>% 
-  left_join(retouched_artefacts) %>% 
-  mutate(proportion_retouched = total_retouched / total_artifacts) %>% 
-  mutate(sites = X1)
-
-# compute correlation
-kas_sites_retouch_density_corr <- 
-cor.test(kas_density_retouch_ages$proportion_retouched,
-         kas_density_retouch_ages$density)
-
-r_value <- unname(kas_sites_retouch_density_corr$estimate)
-
-```
-
-```{r}
-size <-  8
-
-ggplot(kas_density_retouch_ages,
-       aes(density,
-           proportion_retouched
-           )) +
-  geom_point(aes(colour = date_age / 1000,
-                 size = total_artifacts,
-                 shape = is.na(stemmed_point))) +
- # geom_text_repel(aes(label = sites),
- #                 nudge_x = 0.075, 
- #                 nudge_y = -0.075,
- #                 size = 5) +
-  geom_smooth(alpha = 0.2) +
-  scale_y_log10(limits = c(0.001, 1),
-                labels = scales::comma_format(accuracy = 0.001)) +
-  scale_x_log10(limits = c(0.001, 10),
-                labels = scales::comma_format(accuracy = 0.001)) +
-  theme_minimal(base_size = 12) +
-  scale_color_viridis_c(name = "Age of\nassemblage (ka)") +
-  xlab("Artifact volumetric density") +
-  ylab("Proportion of retouched pieces") +
-  annotate("text",
-           x = 0.005,
-           y = 0.7,
-           label = "Curated",
-           size = size -2,
-           colour = "grey50") +
-    annotate("text",
-           x = 5,
-           y = 0.01,
-           label = "Expedient",
-           size = size -2,
-           colour = "grey50") +
-    annotate("text",
-           x = 0.01,
-           y = 0.01,
-           label = glue('r = {round(r_value, 3)}'),
-           size = size - 2,
-           colour = "grey50")
-
-```
-
-```{r}
-# I wanna make simple table for artifact density and frequency of retouched pieces again. Because the complex table above is hard to make comments..
-
-ggplot(kas_density_retouch_ages,
-       aes(density,
-           proportion_retouched
-           )) +
-  geom_point(aes(colour = date_age / 1000,
-                 size = 7)) +
- # geom_text_repel(aes(label = sites),
- #                 nudge_x = 0.075, 
- #                 nudge_y = -0.075,
- #                 size = 5) +
-  geom_smooth(alpha = 0.2) +
-  scale_y_log10(limits = c(0.001, 1),
-                labels = scales::comma_format(accuracy = 0.001)) +
-  scale_x_log10(limits = c(0.001, 10),
-                labels = scales::comma_format(accuracy = 0.001)) +
-  theme_minimal(base_size = 12) +
-  scale_color_viridis_c(name = "Age of\nassemblage (ka)") +
-  xlab("Artifact volumetric density") +
-  ylab("Proportion of retouched pieces") +
-  annotate("text",
-           x = 0.005,
-           y = 0.7,
-           label = "Curated",
-           size = size -2,
-           colour = "grey50") +
-    annotate("text",
-           x = 5,
-           y = 0.01,
-           label = "Expedient",
-           size = size -2,
-           colour = "grey50") +
-    annotate("text",
-           x = 0.01,
-           y = 0.01,
-           label = glue('r = {round(r_value, 3)}'),
-           size = size - 2,
-           colour = "grey50")
-
-```
-
-Artefact volumetric density and frequency of retouched pieces can be proxies to identify the characteristics of the hunter-gatherers who carried that stone tools whether they are residentially mobile or logistically orgnized foragers based on the premise that the relationship between lithic assemblage composition and land-use strategy represent an expression of economizing behavior.
-
-The artefact volumetric density is defined as the total number of artefacts per cubic meter of excavated sediment, and served as a proxy for accumulation ratio of artefacts.
-the relative frequency of retouched pieces is the count of retouched flake tools devided by the total number of artifacts.
-
-The two ends of the curated-expedient show that residential to logistical mobility.
-Expedient assemblages can be depicted as "base camps" or "residences", while curated ones represent either residential mobility, where the central locus of group changes frequently, or certain task groups from a central residential site.
-
-The plot show that younger sites are likely to be used as base camp than older site, which means that people tended to stay longer at one place after around 40-35ka.
-
-(Riel-Salvatore and Barton 2004)
-
-```{r}
-# dating
-library(rcarbon)
-
-# I'm working through https://cran.r-project.org/web/packages/rcarbon/vignettes/rcarbon.html
-
-kas_ages <- 
-kas_density_retouch_ages %>% 
-  select(sites, date_age, date_error) %>% 
-  mutate(sites = str_remove_all(sites, "_"))
-
-kas_caldates <- 
-  calibrate(x = kas_ages$date_age,
-            errors = kas_ages$date_error,
-            calCurves='intcal13',
-            ncores=3) #running calibration over 3 cores
-
-kas_spd = spd(kas_caldates, timeRange = c(50000,25000)) 
-plot(kas_spd)
-
-binsense(x=kas_caldates,
-         y=kas_ages$sites,
-         h=seq(0,5000,100),
-         timeRange=c(50000,25000))
-
-ggplot(kas_spd$grid,
-       aes(calBP,
-           PrDens)) +
-  geom_line() +
-  scale_y_log10(name = "Summed probability")
-
-# test how the SPD fits to an exponential growth model:
-expnull <- modelTest(kas_caldates, 
-                     errors=kas_ages$date_error, 
-                     nsim=100, 
-                     timeRange=c(50000,25000),
-                     model="uniform",
-                     runm=10)
-
-plot(expnull)
-
-```
 
@@
 
-::: {#refs}
-:::
+<div>
+
+</div>
 

### 2020-12-04 analysis/paper/paper.Rmd
Connecting references

@@
 
-The appearance of stemmed points in the Late Pleistocene (\~40-35 ka) in Korea is often assumed to have transformed forager lifeways because it reflected more specialized hunting practices [\@seong2008tanged; \@chang2013human].
+The appearance of stemmed points in the Late Pleistocene (~40-35 ka) in Korea is often assumed to have transformed forager lifeways because it reflected more specialized hunting practices [@seong2008tanged; @chang2013human].
 Stemmed points are often called as tanged points in the western hemisphere, but we use 'stemmed point' to distinguish from Bronze Age stone projectile points that have long been called 'tanged points' in Korea.
@@
 
-Previous work on this technological change in Korea has focussed on issues of the transition from the Early to the Late Paleolithic and modern human dispersals [\@norton2009evolution; \@bae2010origin].
+Previous work on this technological change in Korea has focussed on issues of the transition from the Early to the Late Paleolithic and modern human dispersals [@norton2009evolution; @bae2010origin].
 To our knowledge, the implications of the appearance of stemmed points for forager mobility and site use have not been explored in detail.
 Here, we attempt to identify patterns of mobility through analysis of the stone artefact assemblages using Human Behavioural Ecology (HBE) models.
-Drawing on established relationships between lithic technology and forager mobility [\@shott1986settlement; \@Hiscock_1994; \@kuhn1994formal; \@kuhn2004upper; \@kuhn2015artifacts; \@kuhn2016moves; \@capriles2018], we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that are part of a broader strategy of moving frequently and further, possibly as an adaptation to environmental or population changes.
+Drawing on established relationships between lithic technology and forager mobility [@shott1986settlement; @Hiscock_1994; @kuhn1994formal; @kuhn2004upper; @kuhn2015artifacts; @kuhn2016moves; @capriles2018], we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that are part of a broader strategy of moving frequently and further, possibly as an adaptation to environmental or population changes.
 
@@
 In the Eastern hemisphere, major technological innovations during the Late Pleistocene are restricted to Northeast Asia, such as northern China, Japan and Korea.
-These innovations include blade and micro technology, high frequencies of retouched blade tools, several novel tools such as projectile points, stemmed points, end scrapers, burins, denticulates, etc, and using high-quality raw materials [\@bar1999big; \@brantingham2001initial; \@Bar-Yosef_2002; \@bae2017origin; \@nakazawa2018quaternary].
-Previous work has argued that the appearance of new stone artefact technologies in this region may be linked to modern human dispersals [\@seong2008tanged;\@bae2010origin;\@Bae_2010; \@bae2017origin].
+These innovations include blade and micro technology, high frequencies of retouched blade tools, several novel tools such as projectile points, stemmed points, end scrapers, burins, denticulates, etc, and using high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary].
+Previous work has argued that the appearance of new stone artefact technologies in this region may be linked to modern human dispersals [@seong2008tanged; @bae2010origin; @Bae_2010; @bae2017origin].
 
-The Shuidonggou site in northern China is an important example of the transition from cobble to blade tool industries in the Late Pleistocene of Northeast Asia [\@Brantingham_Perreault_2010; \@Gao_Zhang_Yang_Shen_Wu_2010; \@pei2012shuidonggou].
+The Shuidonggou site in northern China is an important example of the transition from cobble to blade tool industries in the Late Pleistocene of Northeast Asia [@Brantingham_Perreault_2010; @Gao_Zhang_Yang_Shen_Wu_2010; @pei2012shuidonggou].
 Shuidonggou Locality SDG 2, which dates to around c.
@@
 The recent excavation of Locality 12 shows that its archaeological context, dated to 13,078--13,296 cal BP, includes ground stone and more than 30,000 microlithics made out of a variety of raw materials such as fine-grained and siliceous rocks.
-In addition, more than 10,000 animal fossils and bone tools including a tool for fishing nets, an awl, and two needles were excavated [\@pei2012shuidonggou; \@gao2014discovery; \@zhang2016bone].
-Stemmed points are not part of the Shuidonggou assemblages, but microblade technology is argued to have appeared in China after 29ka BP, related to changes in mobility associated with colder climates towards LGM [\@yi2014significance].
-Shuidonggou shows that the blade component came from Siberia and/or Mongolia at 41 ka (SDG Locality1 and 9), advanced core and flake tools were likely to have been locally developed at 33 ka (2 and 8), and at 10.8 ka microblade technology appears but the origin is uncertain (12) [\@yi2014significance; \@li2019history].\
-\
-Late Pleistocene technological innovations in Japan appeared in different regions of the archipelago around 38,000 years ago accompanied by remains of Homo sapiens found in Okinawa Island (c.
-36,000 cal BP) [\@yamaoka2012use; \@izuho201421].
-These new technologies include trap pits, cobble concentrations, hearths, charcoal concentrations, and toolkits including trapezoids, pointed-shaped backed blades, backed points, burins, end scrapers, side scrapers, wedges, beak-shaped tools, axes, edge-ground axes, hammerstones, cobble tools, and anvils [\@bae2017origin; \@izuho201421].
-These innovations have been interpreted as evidence of new foraging methods such as watercraft (marine transport of obsidian from the Kozu Island) and bow-and-arrow technology, driven by increased population and ecological changes [\@Yamaoka_2012; \@bae2017origin; \@Nakazawa_Bae_2018; \@morisaki2019early].
-Hafted trapezoids were likely multifunctional tools adapted to the specific environmental settings of the different Japanese Islands [\@ono2002radiocaron; \@yamaoka2012use].
+In addition, more than 10,000 animal fossils and bone tools including a tool for fishing nets, an awl, and two needles were excavated [@pei2012shuidonggou; @gao2014discovery; @zhang2016bone].
+Stemmed points are not part of the Shuidonggou assemblages, but microblade technology is argued to have appeared in China after 29ka BP, related to changes in mobility associated with colder climates towards LGM [@yi2014significance].
+Shuidonggou shows that the blade component came from Siberia and/or Mongolia at 41 ka (SDG Locality1 and 9), advanced core and flake tools were likely to have been locally developed at 33 ka (2 and 8), and at 10.8 ka microblade technology appears but the origin is uncertain (12) [@yi2014significance; @li2019history].
+
+Late Pleistocene technological innovations in Japan appeared in different regions of the archipelago around 38,000 years ago accompanied by remains of Homo sapiens found in Okinawa Island (c.36,000 cal BP) [@yamaoka2012use; @izuho201421].
+These new technologies include trap pits, cobble concentrations, hearths, charcoal concentrations, and toolkits including trapezoids, pointed-shaped backed blades, backed points, burins, end scrapers, side scrapers, wedges, beak-shaped tools, axes, edge-ground axes, hammerstones, cobble tools, and anvils [@bae2017origin; @izuho201421].
+These innovations have been interpreted as evidence of new foraging methods such as watercraft (marine transport of obsidian from the Kozu Island) and bow-and-arrow technology, driven by increased population and ecological changes [@Yamaoka_2012; @bae2017origin; @Nakazawa_Bae_2018; @morisaki2019early].
+Hafted trapezoids were likely multifunctional tools adapted to the specific environmental settings of the different Japanese Islands [@ono2002radiocaron; @yamaoka2012use].
 Ishinomoto 8-ku is one of the earliest sites dated to 39,690--34,790 cal BP, located in Takuma upland of Kumamoto Prefecture.
 A total of 500 stone artefacts were discovered including trapezoids, side scrapers, flakes, flake cores, edge-ground axes, and cobble tools.
-The main lithic raw materials are chert and andesite [\@izuho201421].
-Stemmed points appear much later, dated to between about 15,500 and 13,800 cal BP [\@ono2002radiocaron; \@tsutsumi2007dynamics].
+The main lithic raw materials are chert and andesite [@izuho201421].
+Stemmed points appear much later, dated to between about 15,500 and 13,800 cal BP [@ono2002radiocaron; @tsutsumi2007dynamics].
 
@@
 Stemmed points appeared in both regions after the case of Korean Peninsula.
-The oldest stemmed points so far is from the Yonghodong site dated to 38.5 ka and made on elongated flakes [\@seong2009emergence; \@seong2015diversity; \@bae2017origin].
+The oldest stemmed points so far is from the Yonghodong site dated to 38.5 ka and made on elongated flakes [@seong2009emergence; @seong2015diversity; @bae2017origin].
 After blades appeared at c.
@@
 Replacing flake to blade as a blank of the points led to an increase in quantity over time.
-In addition to the stemmed points, flake tools became more complex, with a greater diversity in both type and size along with continuously used core tools during the Late Pleistocene  [\@Bae_2010; \@Lee_2013; \@Lee_2016].
-There are several attempts to explain the technological transition and they can be summarized into two competing models: 'heterogenic' migration [\@bae2010origin] or *in situ* evolution [\@seong2009emergence].
+In addition to the stemmed points, flake tools became more complex, with a greater diversity in both type and size along with continuously used core tools during the Late Pleistocene [@Bae_2010; @Lee_2013; @Lee_2016].
+There are several attempts to explain the technological transition and they can be summarized into two competing models: 'heterogenic' migration [@bae2010origin] or *in situ* evolution [@seong2009emergence].
 The migration model argues that the new blade industry including stemmed points, and the earlier coarse flake tradition including large cores, polyhedrons, choppers, and even handaxes, came from different origins as the result of continuing influx of model human migration from two routes.
 Bae (2010) explains that blade and stemmed points arrived in the Korean peninsula from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP with the earliest evidence of blade in those regions.
-He assumes that core and flake industry come from southern China based on the existence of similar assemblages in both Korea and southern China in addition to a genetic analysis of the Y-chromosome of the modern population [\@bae2013early; \@lee2013current].
-The '*in situ*' model claims that stemmed points and other Late Paleolithic assemblages including blade industries autonomously emerged in the south of the Korean peninsula with no external influence [\@seong2008tanged].
-This claim is supported by the earliest appearance of stemmed points among Northeast Asia, such as Hwadae-ri, Hopyeong-dong, and Yonghodong sites dated to 40-35ka BP [\@seong2009emergence].
-Seong (2009) uses the increased blade-to-flake ratios on lithic assemblages to support the *in situ* model based on the premise that the blade industry represents new technology while flakes indicate a continuing local one.
+He assumes that core and flake industry come from southern China based on the existence of similar assemblages in both Korea and southern China in addition to a genetic analysis of the Y-chromosome of the modern population [@bae2013early; @lee2013current].
+The '*in situ*' model claims that stemmed points and other Late Paleolithic assemblages including blade industries autonomously emerged in the south of the Korean peninsula with no external influence [@seong2008tanged].
+This claim is supported by the earliest appearance of stemmed points among Northeast Asia, such as Hwadae-ri, Hopyeong-dong, and Yonghodong sites dated to 40-35ka BP [@seong2009emergence].
+Seong(2009) uses the increased blade-to-flake ratios on lithic assemblages to support the *in situ* model based on the premise that the blade industry represents new technology while flakes indicate a continuing local one.
 \
@@
 We focus on the appearance of stemmed points in Korea because these are a new technology that represents a new hunting method.
-This is because the hafting implies throwing techniques that can reach long-distance targets [\@Keeley_1982; \@kuhn2015artifacts].
+This is because the hafting implies throwing techniques that can reach long-distance targets [@Keeley_1982; @kuhn2015artifacts].
 We draw on behavioural ecological theory to model the effects of this technological change.
-Behavioural ecology theory offers structured and testable models based on optimality assumptions that stand on economic rationality and environmental knowledge [\@winterhalder1992evolutionary; \@prentiss2019handbook].
-The patch choice model and marginal value theorem predict that forager mobility and travel times will increase when resource patch productivity decreases [\@smith1991inujjuamiunt; \@smith1983anthropological; \@bettinger1997evolutionary; \@wolverton2015patch; \@llano2015optimal].
+Behavioural ecology theory offers structured and testable models based on optimality assumptions that stand on economic rationality and environmental knowledge [@winterhalder1992evolutionary; @prentiss2019handbook].
+The patch choice model and marginal value theorem predict that forager mobility and travel times will increase when resource patch productivity decreases [@smith1991inujjuamiunt; @smith1983anthropological; @bettinger1997evolutionary; @wolverton2015patch; @llano2015optimal].
 We can also model individual artefact types as a kind of resource patch.
@@
 That is, they may move to a patch where they can stay longer (i.e.
-an artefact type that has a long and extendible use-life) when they are not certain of the productivity and travel times of other patches [\@kuhn2015artifacts].
+an artefact type that has a long and extendible use-life) when they are not certain of the productivity and travel times of other patches [@kuhn2015artifacts].
 
@@
 In Australia during the mid-Holocene, Hiscock et al.
-argue that the first appearance of microlithics in many locations represent a portable and multi-functional toolkit that minimize travel expenses and increase tool readiness to adapt to patchy and unpredictable resources associated with environmental change [\@Hiscock_1994; \@Hiscock_Clarkson_Mackay_2011]. Similarly, in Japan, Morisaki et al.
-[\@Morisaki_Izuho_Terry_Sato_2015] argue that the transition from trapezoid to the blades and projectile points around 25 ka cal BP in north Paleo-Honshu Island enabled the foragers to extend their occupation into cold grassland landscapes.
+argue that the first appearance of microlithics in many locations represent a portable and multi-functional toolkit that minimize travel expenses and increase tool readiness to adapt to patchy and unpredictable resources associated with environmental change [@Hiscock_1994; @Hiscock_Clarkson_Mackay_2011].Similarly, in Japan, Morisaki et al.
+[@Morisaki_Izuho_Terry_Sato_2015] argue that the transition from trapezoid to the blades and projectile points around 25 ka cal BP in north Paleo-Honshu Island enabled the foragers to extend their occupation into cold grassland landscapes.
 
@@
 After the appearance of these new tools, did people tend to stay in one location for longer or shorter periods, perhaps for specific purposes?
-The stone artefact assemblage from a site can inform us how long individuals or groups stayed and give insights into their activities [\@binford1979organization; \@kelly1995foraging; \@kuhn2016moves; \@holdaway2019surface].
-For example, stone artifact density and the frequency of retouched pieces (scaled to the volume of excavated sediment) can be used as proxies to represent occupation patterns [\@barton2011modeling; \@clark2017lithics].
+The stone artefact assemblage from a site can inform us how long individuals or groups stayed and give insights into their activities [@binford1979organization; @kelly1995foraging; @kuhn2016moves; @holdaway2019surface].
+For example, stone artifact density and the frequency of retouched pieces (scaled to the volume of excavated sediment) can be used as proxies to represent occupation patterns [@barton2011modeling; @clark2017lithics].
 Assemblages with low density but a high proportion of retouched and backed pieces may indicate the remains of 'short-term camp', which is a site for small and ephemeral overnight camp or limited activity station.
-On the other hand, the combination of high density and low retouched pieces represent logistically organized basecamp, which is a site with greater residential stability, long site occupation and occupied by larger groups [\@clark2017lithics; \@clark2019landscapes].
+On the other hand, the combination of high density and low retouched pieces represent logistically organized basecamp, which is a site with greater residential stability, long site occupation and occupied by larger groups [@clark2017lithics; @clark2019landscapes].
 
@@
 
-After the first excavation of a Paleolithic site in Korea at Seokjang-ri site in the 1960s, more than 200 Paleolithic sites have been discovered in South Korea [\@lee2019were].
+After the first excavation of a Paleolithic site in Korea at Seokjang-ri site in the 1960s, more than 200 Paleolithic sites have been discovered in South Korea [@lee2019were].
 We selected sites dated to before and after the transition period (40-35 ka) to analyze the process of technological change.
@@
 This layer was dated to 31,200±900 BP by radiocarbon dating and 30,000±1,700 BP by OSL.
-The uppermost layer, dated to 22,000±100 BP by OSL, contains blades, scrapers, awls, and denticulates [\@seong2009emergence].
+The uppermost layer, dated to 22,000±100 BP by OSL, contains blades, scrapers, awls, and denticulates [@seong2009emergence].
 We obtained the data from published excavation reports held by provincial museums.
 
-```{r korea-map, fig.cap="Site map. Korean Paleolithic sites mentioned in the text."}
+<!---```{r korea-map, fig.cap="Site map. Korean Paleolithic sites mentioned in the text."}
 source(here::here("analysis/paper/001-draw-the-map.R"))
 knitr::include_graphics(here::here("analysis/figures/001-site-map.png"))
-
-```
+``` --->
 

### 2020-12-04 analysis/paper/001-draw-the-map.R

@@ -143,7 +143,7 @@ coords <-
                                            to = 'dec_deg')) %>%
   mutate_at(vars(lat_dd, long_dd), as.numeric)
 
-write_csv(coords, here::here("analysis/data/raw_data/korean_palaeolithic_site_locations.csv"))
+#write_csv(coords, here::here("analysis/data/raw_data/korean_palaeolithic_site_locations.csv"))
 
 # quick map
 

### 2020-12-04 analysis/paper/paper.Rmd
put here to 001 Rscript

@@
 
-<!---```{r korea-map, fig.cap="Site map. Korean Paleolithic sites mentioned in the text."}
+```{r korea_map, fig.cap="Site map. Korean Paleolithic sites mentioned in the text."}
 source(here::here("analysis/paper/001-draw-the-map.R"))
 knitr::include_graphics(here::here("analysis/figures/001-site-map.png"))
-``` --->
+```
 

### 2020-12-04 analysis/paper/001-draw-the-map.R

@@ -75,11 +75,13 @@ table_index <- data.frame(Index = korean_archaeological_site_locations$ID,
 
 print(table_index)
 
-#save a table of AIC result as csv
-write.csv(table_index, file="analysis/figures/001-Map-site-name.csv", row.names = FALSE)
+#save a table
+library(here)
+write.csv(table_index, file=here("analysis/figures/001-Map-site-name.csv", row.names = FALSE))
 
 library(gridExtra)
 library(grid)
+library(gtable)
 mytheme <- gridExtra::ttheme_minimal(
   core = list(fg_params=list(cex = 0.5)),
   colhead = list(fg_params=list(cex = 0.5)),
@@ -95,81 +97,7 @@ dev.off()
 
 grid.arrange(map_index, g, ncol=2)
 
-library(here)
+
 ggsave(here("analysis/figures/001-site-map.png"))
-
-#----------------------------------------------------------------
-# older map data
-
-korean_archaeological_sites <-
-  readxl::read_excel(here("analysis/data/raw_data/korean-archaeologica-sites.xlsx"))
-
-coords <-
-  korean_archaeological_sites %>%
-  select(site_name,
-         east_longitude,
-         north_latitude) %>%
-  separate(east_longitude,
-           into = c("e_deg", "e_min", "e_sec"),
-           sep =  "°|\\′") %>%
-  mutate(e_min = str_squish(e_min),
-         e_sec = str_squish(e_sec)) %>%
-  separate(e_min,
-           into = c("e_min", "e_sec1"),
-           sep = "'") %>%
-  mutate_at(vars(e_deg, e_min, e_sec, e_sec1), parse_number) %>%
-  mutate(e_sec = ifelse(is.na(e_sec), e_sec1, e_sec)) %>%
-  select(-e_sec1) %>%
-  separate(north_latitude,
-           into = c("n_deg", "n_min", "n_sec"),
-           sep =  "°|\\′") %>%
-  mutate(n_min = str_squish(n_min),
-         n_sec = str_squish(n_sec)) %>%
-  separate(n_min,
-           into = c("n_min", "n_sec1"),
-           sep = "'") %>%
-  mutate_at(vars(n_deg, n_min, n_sec, n_sec1), parse_number) %>%
-  mutate_if(is.numeric, list(~ifelse(is.na(.), 0, .))) %>%
[…]

### 2020-12-05 analysis/paper/paper.Rmd
still not perfect image 1 but arrangement is good

@@
 ---
+
+```{r, setup, echo = FALSE}
+knitr::opts_chunk$set(
+  collapse = TRUE,
+  warning = FALSE,
+  message = FALSE,
+  echo = FALSE,
+  comment = "#>",
+  dpi = 300,
+  fig.path = "../figures/"
+)
+```
+
+<!---```{r}
+quietly <- function(x) {
+  sink(tempfile())
+  on.exit(sink())
+  invisible(force(suppressMessages(x)))
+}
+
+``` --->
 
@@
 knitr::include_graphics(here::here("analysis/figures/001-site-map.png"))
+
 ```

### 2020-12-05 analysis/paper/001-draw-the-map.R

@@ -72,18 +72,17 @@ index_table <- korean_archaeological_site_locations %>%
 table_index <- data.frame(Index = korean_archaeological_site_locations$ID,
                          Site = korean_archaeological_site_locations$site_name)
 
-
-print(table_index)
+# print(table_index)
 
 #save a table
 library(here)
-write.csv(table_index, file=here("analysis/figures/001-Map-site-name.csv", row.names = FALSE))
+#write.csv(table_index, file=here("analysis/figures/001-Map-site-name.csv", row.names = FALSE))
 
 library(gridExtra)
 library(grid)
 library(gtable)
 mytheme <- gridExtra::ttheme_minimal(
-  core = list(fg_params=list(cex = 0.5)),
+  core = list(fg_params=list(cex = 0.6)),
   colhead = list(fg_params=list(cex = 0.5)),
   rowhead = list(fg_params=list(cex = 0.25)))
 
@@ -91,13 +90,23 @@ g <- tableGrob(table_index, rows = NULL, theme = mytheme)
 g <- gtable_add_grob(g,
                      grobs = rectGrob(gp = gpar(fill = NA, lwd = 1)),
                      t = 1, b = nrow(g), l = 1, r = ncol(g))
+
+#g$heights <- unit(rep(5/nrow(g), nrow(g), nrow(g)), "cm")
+
 grid.newpage()
 grid.draw(g)
 dev.off()
 
-grid.arrange(map_index, g, ncol=2)
+#grid.arrange(map_index, g, ncol=2)
 
+library(cowplot)
+ggdraw(map_index) +
+  draw_plot(g,
+            .123, .124,
+            .35, .38)
 
 ggsave(here("analysis/figures/001-site-map.png"))
+
+
 
 

### 2020-12-05 analysis/paper/paper.Rmd
It's working until raw materials

@@
 We selected sites dated to before and after the transition period (40-35 ka) to analyze the process of technological change.
-This resulted in a sample of analysed 33 assemblages from 22 sites spanning 49-24 ka.
+This resulted in a sample of analysed 33 assemblages from 22 sites spanning 49-24 ka (Figure \@ref(fig:korea-map)).
 We identified multiple assemblages in a site where culturally sterile deposits separated artefact-bearing deposits, or where stratigraphic units could be identified by major differences in the texture and composition of the sedimentary deposit containing the stone artefacts.
@@
 
-```{r korea_map, fig.cap="Site map. Korean Paleolithic sites mentioned in the text."}
+```{r korea-map, fig.cap="Korean Paleolithic sites mentioned in the text."}
 source(here::here("analysis/paper/001-draw-the-map.R"))
@@
 
+## Assemblages
+
+The 33 assemblages consist of cores, blade cores, blades, flakes, debris, hammers, choppers, plans, polyhedrals, side scrapers, end scrapers, notches, cleavers, stemmed points, awls, denticulates, burins, handaxes, blanks, knives, handadzes, peaks, flatters, flakers etc.
+We excluded some artefacts from our sample because of uncertainty about their typology.
+The stone artefacts from all of these sites are stored in museums located throughout South Korea, and have been briefly described in the excavation reports.
+We have visited most of these collections to obtain permission to study the materials.
+
+## Statistical methods
+
+To answer the two research questions, we compiled information about artefact density, retouch frequency, toolkit composition, raw materials, site elevation, and radiocarbon ages.
+
+The artefact volumetric density is defined as the total number of artefacts per cubic meter of excavated sediment, and serves as a proxy for the accumulation ratio of artefacts [@clark2017lithics; @clark2019landscapes].
+We used these density values to evaluate mobility and land-use practices of hunter-gatherers during the Late Pleistocene.
+The retouch frequency is the number of retouched tools divided by the total number of artefacts in an assemblage.
+This value represents mobility and duration of site occupation based on the concept that tool reduction is an tactic to extend the use-life of a tool by generating sharp, usable edges while minimizing the cost of transporting tools or bulky raw materials by reducing the total load weight [@Kuhn_1990; @buvit2014late; @clark2019landscapes].
+In other words, producing and carrying highly retouched tools will increase portability and efficiency by decreasing the size of both individual tools and the assemblages so that hunter-gatherers could move often or further, or both [@kuhn1991unpacking; @kuhn1994formal; @kuhn2004upper; @andrefsky1994raw; @davies2018modeling].
+Combining these two proxies, the artefact volumetric density and the retouch frequency, we expect to determine whether or not the hunter-gatherers were residentially mobile or logistically organized.
+If the assemblage in a site has a high artefact density and small proportion of retouch, we interpret it as a more expedient assemblage that represents “base camps” or “residences”, while less density with higher proportions of retouch shows a more curated assemblage and indicates either residential mobility or certain task groups away from residential sites [@riel2004late; @clark2017lithics].
+
+We calculated toolkit composition by analysing tool and debitage frequencies to identify changes in the assemblages during the transition period.
+Toolkit composition helps us answer the question of how site occupation patterns changed before and after the appearance of new technology.
+The type and proportion of retouched tools and knapping products, including debris and hammerstones, could give information about settlement patterns based on a premise that more activities are conducted with a wider range of tool types for longer occupation [@buvit2014late; @garcia2017occupational; @centi2019bulb].
+For instance, small sized assemblages with a few types of tools represent short duration of site occupation [@shott2010size; @buvit2014late] .
+
+Studying the raw material consumption patterns is based on an assumption that stone raw materials were transported and consumed to optimize mobility and minimize risk of not having artefacts when they were needed [@andrefsky1994raw; @brantingham2000raw; @brantingham2003neutral; @yue2020lithic].
+The representation of raw materials in an assemblage can be an indicator of site occupation patterns.
+For example, a large variability of raw materials in a site can suggest long-term settlement, while varied and exotic raw materials can indicate higher mobility [@valde2020].
+Here we want to know the specific raw materials that relate to the appearance of a new technology, and understanding site occupation patterns by examining changes in the proportions of raw material types in the assemblages.
+
+HBE models require environmental proxies to build hypotheses of behavioural change influenced by the surrounding environment.
+However, there are few paleoenvironmental proxies in Korean relevant to our target period.
+For example, geoarchaeological observations suggest that environment of the period was “overall cooler and drier during MIS 3” [@Han_2008; @seong2008tanged; @Choi_2011; @Chang_2013; @im2015study; @bak2017late].
+To supplement the limited availability of high-resolution local proxies, we used climate information from the simulated data set of paleoclimate including global monthly temperature, covering the last 120,000 years [@beyer2020high].
+We applied information of site location including elevation, latitude and longitude to the simulated data set to get the annual temperature of our research period and area.
+Besides the climate analysis, we analysed site elevation as a proxy to represent different environments at local scales, to identify how these impact mobility and site occupation.
+We mainly focused on the elevation of the sites before and after the appearance of new technology and its relationship with the stemmed points.
+
+We used radiocarbon dates as a reference for population level.
+Summed probability distributions (SPD) of radiocarbon dates are often used to understand temporal trends in past human populations [@rick1987dates; @bamforth2012radiocarbon; @shennan2013regional; @timpson2014reconstructing; @contreras2014summed; @riris2019widespread].
+Despite limitations of the validity of SPD [@williams2012use; @bamforth2012radiocarbon], including the lack of excavated sites in certain periods, sample size, and calibration effects, this method may be helpful as a proxy for a first approximation of population fluctuations.
+Although this is often used with samples of thousands of ages, this method can also be useful for small sample sizes.
+For instance, Timpson et al.
+apply SPD to 93 ages from Eastern Middle Sweden and demonstrate that the result is equivalent to a larger sample size (n= 243) [@timpson2014reconstructing].
+We used 108 radiocarbon ages from 22 sites to examine connections between population trends during the Late Pleistocene in Korea and technological innovation.
+
+There is a debate about whether past population growth has played a role in technological change [@shennan2001demography; @kline2010population].
+@shennan2009patter observed a correlation between population decrease and technological simplification in their study of  lithic arrowheads from the southern Scandinavian Mesolithic.
+They found that the lithics became less complex, and point shape changed as the population level dropped.
+On the other hand, @collard2013population argue that regional archaeological data do not support the population-driven models [@buchanan2016drivers].
+For example, @buchanan2016drivers concluded that changes in point types from >13,000 years ago to 400 years ago in Texas were related to environmental risk, and not dependent on population size.
+We explored SPD to test for correlations between population, environmental risk and technological change.
+We used the R package ‘rcarbon’ [@cremaanalysing] to perform statistical analyses and conduct model testing using Monte Carlo methods.
+We generated exponential, uniform, and linear models and evaluated the model goodness-of-fit by applying Akaike’s Information Criterion (AIC) as a selection criterion [@sakamoto1986akaike; @bevan2017holocene; @riris2019widespread].
+
+All analyses and visualization are performed in the R Environment [@team2017r].
+Our R code and data are available on the following GitHub repository: https://github.com/parkgayoung/koreapaleolithicmobilityoccupation as well as on the Open Science Framework: https://osf.io/yagd2/ to enable full reproducibility [@marwick2017computational].
+
 # Results
+
+## Artefact volumetric density and retouch frequency
+
+Figure \@ref(fig:volume-density) shows that there is no clear temporal pattern of changes in artefact density, although the density of some assemblages are two or three times higher than others, especially after 30ka.
+The inset figure shows that assemblages with stemmed points have higher artefact densities than the assemblages without.
+With the newer tools, hunter-gatherers may have stayed in the same location for longer durations, and we may be seeing site occupation patterns in this sample that represent the use of both base camps or residences, with the artefact higher densities, and more briefly occupied camps with lower densities.
+
+```{r volume-density, fig.cap="The main plot shows the artefact volumetric density over time.
+The size of data points represent the total number of artefacts from each assemblage and the colour indicates the presence of stemmed points.
+The inset plot shows direct comparison of artefact densities for assemblages with and without stemmed points."}
+source(here::here("analysis/paper/002-artifact-volumetric-density.R"))
+knitr::include_graphics(here::here("analysis/figures/002-age-by-density.png"))
+```
+
+Figure \@ref(fig:volume-density-retouch) shows a strong pattern of artefacts from less dense assemblages having higher retouch frequencies while more dense assemblages have lower retouch frequencies (p=0.043), further showing a spectrum of site functions in this sample.
+As for artefact density, there are no clear chronological trends in retouch frequencies.
+The assemblages containing stemmed points tend to have fewer retouched pieces compared to assemblages without stemmed points (p=0.011).
+
+```{r volume-density-retouch, fig.cap="Main plot shows relationship between artefact density and retouch frequency.
+The colour of the data points indicate the age of the assemblage, point size indicates assemblage size, and shape indicates presence or absence of stemmed points.
+Assemblages with higher proportions of retouched pieces and lower densities indicate curated technologies, and assemblages at the other end of the spectrum indicate expedient technologies [@riel2004late].
+The inset plot on the bottom left shows the proportion of retouched pieces in the assemblages over time.
+The inset plot on the bottom right shows the proportion of retouched pieces in the assemblages with or without stemmed points."}
+source(here::here("analysis/paper/003-artifact-volumetric-density-retouch-frequency.R"))
+knitr::include_graphics(here::here("analysis/figures/003-retouch-by-density.png"))
+```
+
+## Composition of toolkit
+
+Figure \@ref(fig:artefact-composition) shows that side scrapers are the major part of most assemblages and are present in all assemblages.
+For example, more than 70 percent of the two assemblages from Hwaderi are side scrapers.
+After around 31 ka, the proportion of side scrapers declines.
+In addition to side scrapers, cores also exist in all assemblages often as a high proportion.
+The proportion of choppers decreased towards in more recent assemblages.
+Blades first appeared in Gawolri-A at 43 ka and then often found in assemblages thereafter.
+Stemmed points first appear in Bonggok around 41.5ka and were typically found in assemblages that contain blades, except for Songamri-A.
+Other artefacts, including burins, denticulates, end scrapers, handaxes, knives, notches, planes, and polyhedrals made up a small proportion of the assemblages.
+
+```{r artefact-composition, fig.cap="This plot shows the composition of each assemblage.
+We excluded artefacts related to manufacturing processes such as pebbles, hammers, flakes and debris, artefacts appeared only in a few assemblages including point, beak shaped, and awl, and unknown and unfinished pieces.
+The colour represents different types of tools and the assemblages are placed in chronological order."}
+source(here::here("analysis/paper/004-artifact-composition.R"))
+knitr::include_graphics(here::here("analysis/figures/004-artefact-types.png"))
+```
+
+## Raw materials
+
+Quartz, quartz vein and quartzites were the most frequently found raw materials, and they were constantly used throughout the Late Pleistocene (Figure \@ref(fig:raw-materials)).
+For example, both Galdamri-A, dated to 44ka with a small assemblage, and Hopyeongdong-B, dated to 27.5ka with a larger assemblage, consist of only the quartz related materials.
+The use of chert, hornfels, rhyolite, and shale increased after 41.5 ka.
+All artefacts in Sageunri are made of rhyolite.
+Hornfels are dominant at Yongsandong and Yongbang sites at 25 ka.
+Chert, hornfels, rhyolite, and shale are very suitable to make elongate blades and stemmed points because of the predictability of flaking afforded by their fine-grained texture.
+For example, Yongsandong has the largest number of stemmed points (n=40) among the assemblages studied here, and all stemmed points from Yongsandong are made from hornfels.
+Other raw materials, such as porphyry and sandstone, were found in just a few assemblages in small proportions.
+
+```{r raw-materials, fig.cap="This plot shows the composition of raw material type in each assemblage.
+We excluded raw materials that are included in less than 5 assemblages including crystal, basalt, iron_ore, slate, limestone, granite, gneiss, tuff, and unidentified ones.
+Hornfels are shown in 4 assemblages but since it occupies more than half of the Yongsandong and Yongband assemblages, we included it in the plot.
+The color represents different types of raw materials, and the assemblages are placed in chronological order."}
+source(here::here("analysis/paper/005-raw-materials.R"))
+knitr::include_graphics(here::here("analysis/figures/005-raw-materials.png"))
+```
+
+## Environmental context
+
+Figure \@ref(fig:site-elevation) shows that people occupied sites located at similar elevations, around 0 to 200 meters above sea level.
+However, there is a substantial difference in elevation between sites with stemmed points and sites without.
+The distribution of elevations of sites without stemmed points is generally lower than the elevations of sites with stemmed points.
+Although not a statistically significant difference in elevations (p = 0.102), this may indicate that forager groups who used stemmed points were able to occupy higher altitudes, while the groups without tended to prefer lower altitudes.
+Sites without stemmed points reach a maximum elevation of about 150m at around 39ka.
+Sites containing stemmed points reach a maximum elevation of 175 m at 28ka.
+
+```{r site-elevation, fig.cap="The main plot shows the site elevation by the age of the deposit.
+Colour indicates presence/absence of stemmed points.
+The lines display a locally weighted regression.
+The inset plot shows a direct relationship between the site elevation and stemmed points."}
+source(here::here("analysis/paper/006-site-elevation.R"))
+knitr::include_graphics(here::here("analysis/figures/006-site-elevation.png"))
+```
+
+Among the sites in our sample there is a variation of about 5 degrees in the mean annual temperature (MAT), mostly controlled by elevation (Figure \@ref(fig:climate-model)).
+Compared to the east side of the Korean Peninsula, the west side tends to be relatively warmer.
+The range of MAT at the sites over our study period is 2-10℃.
+Gihwari and Gigok sites have the lowest (3℃) and Jeungsan-ri site has the highest MAT (8℃).
+The mean MAT for all sites fluctuated within 4℃ between 50ka and 10ka.
+Through MIS 3, the temperature gradually decreased until the Last Glacial Maximum (LGM, 26.5-20 ka) [@clark2009last].
+The first appearance of stemmed points occurs in the middle of the decreasing MAT trend in MIS 3, at 40-35ka.
+The MAT of MIS 2, including LGM, was relatively stable.
+The temperature increased again from late MIS 2 towards MIS 1.
+Figure 7D shows a negative relationship between temperature and elevation (p = 0.073).
+For example, Gihwa-ri Cave site, with one of the lowest MAT distributions, is located in the highest elevation around 300m above sea level.
+Jeungsan-ri is located at a lower elevation and has one of the highest MAT distributions.
+
+```{r climate-model, fig.cap="Mean Annual Temperature (MAT) of the Korean Paleolithic assemblages mentioned in this research.
+A: Site locations on the map and MAT.
+B: MAT distributions for each site during the period they were occupied.
+C: MAT from 50ka to 10ka.
+The grey lines indicate the MAT for each site and the black line is the mean temperature of all sites.
+The light grey area in the middle of the plot indicates the duration of MIS 2, separating MIS 3 and 1, and dark grey area represents the duration of LGM.
+D: The relationship between MAT and site elevation.
+The blue line is a linear regression on elevation and MAT with the grey area showing the 95% confidence interval.t."}
+source(here::here("analysis/paper/007-climate-model-plots.R"))
+knitr::include_graphics(here::here("analysis/figures/007-climate-model-sites-panel-plot.png"))
+```
+
+## Demographic context
+
+Figure \@ref(fig:spd-model) shows the four theoretical null models of demographic change for the period 50ka to 10ka, exponential, uniform, linear, and logistic, generated from the radiocarbon dates.
+Statistical significance of each model is computed using 200 simulations and the global p-values are 0.004, 0.004, 0.009, and 0.004, respectively.
+In this case, the linear model is the best fit with lower value verified by AIC <!---(Table 1)--->.
+However, the other models have similar patterns with a minor margin at -48.6.
+The radiocarbon dates fit well with the overall trajectory of the null models (grey shading) with both positive and negative deviations.
+SPD shows a steady growth until 35ka, stays at the peak with minor deviation 32ka, and then starts declining with frequent temporary fluctuations (relatively sharp).
+The beginning and ending of the SPD indicates significant negative deviation with low density of radiocarbon dates.
+
+<!---
+Caption for Table 1: Summary of model comparison based on Akaike’s Information Criterion (AIC). 
+We applied logistic, exponential and linear to AIC along with the logistical growth model (fit) to evaluate each model [@riris2019widespread].
+We excluded the uniform model because it has no fitted curve or surface to evaluate.
+The best fit model has the lowest AIC score.
+In this case, the linear model is the best fit, with the next-best fit (exponential) having a higher AIC score by a minor margin at -48.6.
+--->
+
+```{r spd-model, fig.cap="Summed probability distribution of 108 radiocarbon dates.
+The black solid line represents actual radiocarbon ages.
+The grey shaded region shows the critical envelope that encompasses the 95% confidence interval for the null models.
+The red and blue vertical bands highlight the portions of the SPD where positive and negative deviations are detected."}
+source(here::here("analysis/paper/008-summed-probability-distributions.R"))
+knitr::include_graphics(here::here("analysis/figures/008-summed-probability-distribution-models.png"))
+```
 

### 2020-12-05 analysis/paper/004-artifact-composition.R

@@ -105,25 +105,25 @@ ggplot(kasa_long3,
 
 ggsave(here::here("analysis/figures/004-artefact-types.png"))
 
-# interactive
-plotly::ggplotly(kasa_long3_fill_plot)
+#make an interactive plot for interpret the plot easily
+#plotly::ggplotly(kasa_long3_fill_plot)
 
 
 
-ggplot(kasa_long3,
-       aes(
-         reorder(axis_label, -age_ka),
-         percentage)) +
-  geom_col() +
-  ylab("Proportion") +
-  xlab("Assemblage (youngest at the top)") +
-  theme_minimal(base_size = 16)  +
-  scale_fill_viridis_d(name = "Artefact type",
-                       option = "D") +
-  facet_wrap( ~ artefact_type,
-              ncol = 1,
-              scales = "free_y") +
-  theme_minimal(base_size = 6)
+#ggplot(kasa_long3,
+#       aes(
+#         reorder(axis_label, -age_ka),
+#         percentage)) +
+#  geom_col() +
+#  ylab("Proportion") +
+#  xlab("Assemblage (youngest at the top)") +
+#  theme_minimal(base_size = 16)  +
+#  scale_fill_viridis_d(name = "Artefact type",
+#                       option = "D") +
+#  facet_wrap( ~ artefact_type,
+#              ncol = 1,
+#              scales = "free_y") +
+#  theme_minimal(base_size = 6)
 
 
 

### 2020-12-07 analysis/paper/paper.Rmd
Not pretty but I could succeed to knit the whole paper!

@@
 
+Our study focussed on two questions: how did the introduction of the new tools change foragers' mobility and landscape use; and how did the new technology change the way people use habitation sites?
+Overall, our results show that assemblages with stemmed points tend to have higher artefact densities and lower proportions of retouched pieces.
+Quartz and side scrapers, in addition to cores and choppers, remain dominant in assemblages before and after the introduction of stemmed points.
+The environmental context of this technological innovation was a gradual decrease in temperature into the LGM.
+
+## Artifact volumetric density and retouch frequency
+
+We examined the artefact volumetric density and retouch frequency to understand the difference in mobility strategies and site occupation patterns before and after the introduction of stemmed points.
+The analyses are based on the premise that highly mobile groups would carry lightweight toolkits with more retouched and easily replaceable composite tools, and leave a smaller amount of artefacts on site as a result of short-term occupations.
+These curated technologies invest time and effort in stone artefact manufacturing and maintenance to optimize travel costs for groups using high mobility or special purpose forays.
+On the other hand, less mobile groups would produce fewer retouched tools and more simple tools, and leave higher densities of artefacts generated from long-term occupations.
+These expedient technologies are more typical for groups with greater residential stability who can make, use, abandon, and remake their tools frequently in the same location [@binford1979organization; @torrence1989time; @kuhn1994formal; @meignen2006middle; @kuhn2014mousterian; @vaquero2018searching; @clark2019landscapes].
+Our data support the concept of the curated and expedient technology by showing a strong pattern of artefacts from less dense assemblages having higher retouch frequencies, while more dense assemblages have lower retouch frequencies.
+Our results show that foragers with stemmed points tended to stay in the same site for longer periods with higher artefact densities and fewer retouched pieces, compared to assemblages without stemmed points.
+In general, the new technology of stemmed points was more associated with expedient technological strategies.
+
+It could be assumed that foragers with stemmed points were highly mobile groups and chose curated technologies since the stemmed points were likely used for hunting activities [@Chang_2013; @lee2019were].
+However, our results show the opposite that the new technology was more often associated with  expedient technological strategies.
+Furthermore, our results show that different sites represent different occupation patterns during the time of technological transition, making it difficult to characterise this period with a single land use strategy.
+Seong claims that transitions from the expedient to curated technology in the Late Pleistocene were neither a straightforward nor unilineal processes and diverse site types can be more related to different occupation purposes including hunting camps, limited activity stations, caches, and so forth [@seong2015diversity].
+Our findings support Seong, with some sites such as Bonggok and Yonghodong containing only 2 stemmed points, while Yongsandong has 38 points including broken tips and a base.
+More than 70% of Yongsandong stemmed points were unearthed, dominating the tool kit with 233 blades and other byproducts related to lithic manufacturing including cores and debris made of the same raw materials [@kim2004yongsan; @bae2012nature].
+In the specific case of Yongsandong it seems likely that stemmed points were manufactured for hunting purposes, which dominated the function of the site.
+On the other hand, at Bonggok and Yonghodong stemmed points were only a minor part of activities there.
+
+## Composition of toolkits and raw materials
+
+The composition of toolkits during the Late Pleistocene of Korea has been described within multiple chronological sub-periods.
+Seong divides five successive assemblage types for the Late/Upper Paleolithic in Korea: (1) quartzite and quartz vein artefacts; (2) mostly small quartzite and quartz vein artefacts with some large artefacts such as cores and choppers; (3) stemmed points dominant; (4) typical blade assemblages including stemmed points; and (5) microblade after 30ka [@seong2015diversity].
+Lee focuses on the Honam region in southwestern Korea and divides the same time frame into two phases; ‘core tool industry with large flake knapping’ during early MIS 3 (59−40 ka), and ‘blade industry with tanged point chiefly’ during late MIS (40−24 ka) [@lee2012characteristics].
+Both Seong and Lee mention that core tools made of quartz or quartz vein never disappeared and they point out the appearance of stemmed points and blade technology as an addition, rather than replacement.
+Our results, covering 49-24 ka and corresponding to the first three types of Seong’s divisions, support the continuation of core tools abundantly throughout the Late Pleistocene.
+In addition, our data show that side scrapers stayed as major tools while other tools occupied only a small portion of the assemblages.
+As stemmed points and blades increased in the toolkits, both choppers and side scrapers went through minor decreases but still remained abundant.
+
+Changes in raw material composition were closely associated with the appearance of stemmed points.
+Previous studies claim that the existing tools, including choppers, polyhedrals, handaxes and cores were usually made of locally acquired quartzite, while the new tools including stemmed points and other blade assemblages were made of more fine-grained materials such as siliceous shale, hornfels, and obsidian which might be brought from distant sources [@Chang_2013; @seong2015diversity].
+Our data also show that quartz, quartz vein and quartzite were consistently dominant and the use of fine-grained materials including chert, hornfels, rhyolite, and shale, increased after 41.5 ka, around the time that stemmed points appear.
+Projectile points, such as stemmed points, need high quality materials which enable high flaking quality, durability and effectiveness of an edge, as well as the creation of well-defined outline forms [@bamforth2009projectile].
+The increased use of finer-grained raw materials likely changed forager mobility patterns to ensure their movements over the landscape supplied them with the materials needed to make stemmed points.
+
+## Environmental and demographic context
+
+We hypothesized that the introduction of stemmed points might reflect a preference for more portable and efficient hunting tools, as an adaptation to environmental or population changes.
+As the first composite tool appearing on the Korean Peninsula, stemmed points represent a major change in stone artefact technology.
+An important quality of composite tools is that the user can easily replace damaged parts,  contributing to an increased maintainability of the tool [@kuhn2015artifacts; @cardillo2010some].
+Combining the stone projective with a wooden shaft further materials enhances functionality of the tool such as improving penetration by increasing weight [@browne1940projectile].
+We expected that the functional advantage of the new composite tool might have allowed foragers to be more active in less productive landscapes.
+Previous studies simply describe the environment of the research period as  “overall cooler and drier during MIS 3” but lack detail due to a lack of direct paleoenvironmental proxies in Korea  [@Han_2008; @seong2008tanged; @Choi_2011; @Chang_2013; @im2015study; @bak2017late].
+To explore environmental and demographic context related to the technological transition, we examined the distribution of site elevations, and simulated MAT during the Late Pleistocene.
+The results show that groups with stemmed points occupied higher elevations, where the MAT was lower.
+This supports our hypothesis that stemmed points supported occupation of more marginal habitats.
+Through MIS 3 the temperature gradually decreased until the LGM, suggesting that stemmed points may have been part of a suite of adaptations to cooler temperatures.
+We see a similar pattern of new technologies enabling expansion into marginal areas at a later time in north Paleo-Honshu Island, Japan, where the appearance of blade and projectile points at about 25 ka were associated with foragers moving into cold grassland areas [@Morisaki_Izuho_Terry_Sato_2015].
+
+We investigated forager population dynamics to assess if population change might be a relevant mechanism to explain the appearance of stemmed points.
+The output of our SPD models indicate population increasing until 35 ka and remaining at the peak with minor deviations for about 3,000 years.
+If population dynamics were a key driver, then we expect population to be high at the time the new technology first appears.
+However, with the first stemmed points appearing at 38 ka, well before the observed peak in population, we conclude that population dynamics were not a major mechanism in the appearance of this new technology.
+Bae claims that stemmed points are the result of the continuous influx of modern humans from a Siberian migration [@Bae_2010].
+Our results are consistent with this scenario, showing a gradual increase in population after the appearance of stemmed points which may reflect a continuous influx of migrants over a long period.
+
+# Conclusion
+
+To understand the change in mobility and site occupation related to the appearance of new technologies during the late Pleistocene in Korea, we examined stone artefacts from 22 sites dated to 49-24 ka.
+The results show that the forager groups with stemmed points tended to be located at higher elevations, which means the use of stemmed points was associated with occupation of marginal or extreme environments.
+In addition, the groups with stemmed points had expedient technologies which reflect residential and less mobile land use patterns.
+The temperature gradually decreased during the technological transition and after the appearance of stemmed points, population gradually increased.
+
+Our findings are limited by the relatively small number of sites and radiocarbon ages currently available.
+The absence of detailed stratigraphic data limit the chronological resolution for the assemblages in our sample.
+Furthermore, we currently lack functional data on individual artefact assemblages to support our conclusions about site use.
+With future work we may be better able to distinguish a high density of stone artefacts as results from a long-term/small group occupation versus a short-term/large group occupation.
+Further research should focus on detailed assemblage analysis at each site to better understand foragers mobility and site use strategies.
+
 # Acknowledgements

### 2020-12-07 analysis/paper/007-climate-model-plots.R

@@ -1,7 +1,7 @@
 
-require("ncdf4")
-require("lattice")
-require("ggplot2")
+library(ncdf4)
+library(lattice)
+library(ggplot2)
 
 # you'll need to get this file from
 # https://figshare.com/articles/LateQuaternary_Environment_nc/12293345/3
@@ -22,7 +22,7 @@ mean_annual_temperature <- apply(temperature, c(1, 2, 4), mean)
 # korean archaeological sites
 library(tidyverse)
 
-site_locations_tbl_raw <- read_csv(here::here("analysis/data/raw_data/korean_palaeolithic_site_locations.csv"))
+site_locations_tbl_raw <- read_csv (here::here("analysis/data/raw_data/korean_palaeolithic_site_locations.csv"))
 site_locations_tbl <- site_locations_tbl_raw
 
 

### 2020-12-07 analysis/paper/008-summed-probability-distributions.R

@@ -174,7 +174,7 @@ df_aic <- data.frame (Model = c("logistic",
 df_aic
 
 #save a table of AIC result as csv
-write.csv(df_aic, file="analysis/figures/008-SPD-AIC.csv", row.names = FALSE)
+# write.csv(df_aic, file="analysis/figures/008-SPD-AIC.csv", row.names = FALSE)
 
-knitr::kable(df_aic)
+# knitr::kable(df_aic)
 

### 2020-12-07 analysis/paper/paper.Rmd

@@
 Here, we attempt to identify patterns of mobility through analysis of the stone artefact assemblages using Human Behavioural Ecology (HBE) models.
-Drawing on established relationships between lithic technology and forager mobility [@shott1986settlement; @Hiscock_1994; @kuhn1994formal; @kuhn2004upper; @kuhn2015artifacts; @kuhn2016moves; @capriles2018], we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that are part of a broader strategy of moving frequently and further, possibly as an adaptation to environmental or population changes.
+Drawing on established relationships between lithic technology and forager mobility [@shott1986settlement; @Hiscock_1994; @kuhn1994formal; @kuhn2004upper; @kuhn2015artifacts; @kuhn2016moves; @capriles2018mobility], we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that are part of a broader strategy of moving frequently and further, possibly as an adaptation to environmental or population changes.
 
@@
 The representation of raw materials in an assemblage can be an indicator of site occupation patterns.
-For example, a large variability of raw materials in a site can suggest long-term settlement, while varied and exotic raw materials can indicate higher mobility [@valde2020].
+For example, a large variability of raw materials in a site can suggest long-term settlement, while varied and exotic raw materials can indicate higher mobility [@valde2020models].
 Here we want to know the specific raw materials that relate to the appearance of a new technology, and understanding site occupation patterns by examining changes in the proportions of raw material types in the assemblages.
@@
 There is a debate about whether past population growth has played a role in technological change [@shennan2001demography; @kline2010population].
-@shennan2009patter observed a correlation between population decrease and technological simplification in their study of  lithic arrowheads from the southern Scandinavian Mesolithic.
+@shennan2009pattern observed a correlation between population decrease and technological simplification in their study of  lithic arrowheads from the southern Scandinavian Mesolithic.
 They found that the lithics became less complex, and point shape changed as the population level dropped.

### 2020-12-14 analysis/paper/paper.Rmd

@@
 
+```{=html}
 <!---```{r}
@@
 ``` --->
-
+```
 # Introduction
 
-The appearance of stemmed points in the Late Pleistocene (~40-35 ka) in Korea is often assumed to have transformed forager lifeways because it reflected more specialized hunting practices [@seong2008tanged; @chang2013human].
+The appearance of stemmed points in the Late Pleistocene (\~40-35 ka) in Korea is often assumed to have transformed forager lifeways because it reflected more specialized hunting practices [@seong2008tanged; @chang2013human].
 Stemmed points are often called as tanged points in the western hemisphere, but we use 'stemmed point' to distinguish from Bronze Age stone projectile points that have long been called 'tanged points' in Korea.
@@
 These innovations include blade and micro technology, high frequencies of retouched blade tools, several novel tools such as projectile points, stemmed points, end scrapers, burins, denticulates, etc, and using high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary].
-Previous work has argued that the appearance of new stone artefact technologies in this region may be linked to modern human dispersals [@seong2008tanged; @bae2010origin; @Bae_2010; @bae2017origin].
+Previous work has argued that the appearance of new stone artefact technologies in this region may be linked to modern human dispersals [@seong2008tanged; @bae2010origin; @bae2017origin].
 
@@
 These new technologies include trap pits, cobble concentrations, hearths, charcoal concentrations, and toolkits including trapezoids, pointed-shaped backed blades, backed points, burins, end scrapers, side scrapers, wedges, beak-shaped tools, axes, edge-ground axes, hammerstones, cobble tools, and anvils [@bae2017origin; @izuho201421].
-These innovations have been interpreted as evidence of new foraging methods such as watercraft (marine transport of obsidian from the Kozu Island) and bow-and-arrow technology, driven by increased population and ecological changes [@Yamaoka_2012; @bae2017origin; @Nakazawa_Bae_2018; @morisaki2019early].
+These innovations have been interpreted as evidence of new foraging methods such as watercraft (marine transport of obsidian from the Kozu Island) and bow-and-arrow technology, driven by increased population and ecological changes [@yamaoka2012use; @bae2017origin; @nakazawa2018quaternary; @morisaki2019early].
 Hafted trapezoids were likely multifunctional tools adapted to the specific environmental settings of the different Japanese Islands [@ono2002radiocaron; @yamaoka2012use].
@@
 Replacing flake to blade as a blank of the points led to an increase in quantity over time.
-In addition to the stemmed points, flake tools became more complex, with a greater diversity in both type and size along with continuously used core tools during the Late Pleistocene [@Bae_2010; @Lee_2013; @Lee_2016].
+In addition to the stemmed points, flake tools became more complex, with a greater diversity in both type and size along with continuously used core tools during the Late Pleistocene [@bae2010origin; @lee2013current; @Lee_2016].
 There are several attempts to explain the technological transition and they can be summarized into two competing models: 'heterogenic' migration [@bae2010origin] or *in situ* evolution [@seong2009emergence].
@@
 This claim is supported by the earliest appearance of stemmed points among Northeast Asia, such as Hwadae-ri, Hopyeong-dong, and Yonghodong sites dated to 40-35ka BP [@seong2009emergence].
-Seong(2009) uses the increased blade-to-flake ratios on lithic assemblages to support the *in situ* model based on the premise that the blade industry represents new technology while flakes indicate a continuing local one.
-\
+Seong(2009) uses the increased blade-to-flake ratios on lithic assemblages to support the *in situ* model based on the premise that the blade industry represents new technology while flakes indicate a continuing local one.\
 
@@
 Combining these two proxies, the artefact volumetric density and the retouch frequency, we expect to determine whether or not the hunter-gatherers were residentially mobile or logistically organized.
-If the assemblage in a site has a high artefact density and small proportion of retouch, we interpret it as a more expedient assemblage that represents “base camps” or “residences”, while less density with higher proportions of retouch shows a more curated assemblage and indicates either residential mobility or certain task groups away from residential sites [@riel2004late; @clark2017lithics].
+If the assemblage in a site has a high artefact density and small proportion of retouch, we interpret it as a more expedient assemblage that represents "base camps" or "residences", while less density with higher proportions of retouch shows a more curated assemblage and indicates either residential mobility or certain task groups away from residential sites [@riel2004late; @clark2017lithics].
 
@@
 However, there are few paleoenvironmental proxies in Korean relevant to our target period.
-For example, geoarchaeological observations suggest that environment of the period was “overall cooler and drier during MIS 3” [@Han_2008; @seong2008tanged; @Choi_2011; @Chang_2013; @im2015study; @bak2017late].
+For example, geoarchaeological observations suggest that environment of the period was "overall cooler and drier during MIS 3" [@Han_2008; @seong2008tanged; @Choi_2011; @chang2013human; @im2015study; @bak2017late].
 To supplement the limited availability of high-resolution local proxies, we used climate information from the simulated data set of paleoclimate including global monthly temperature, covering the last 120,000 years [@beyer2020high].
@@
 On the other hand, @collard2013population argue that regional archaeological data do not support the population-driven models [@buchanan2016drivers].
-For example, @buchanan2016drivers concluded that changes in point types from >13,000 years ago to 400 years ago in Texas were related to environmental risk, and not dependent on population size.
+For example, @buchanan2016drivers concluded that changes in point types from \>13,000 years ago to 400 years ago in Texas were related to environmental risk, and not dependent on population size.
 We explored SPD to test for correlations between population, environmental risk and technological change.
-We used the R package ‘rcarbon’ [@cremaanalysing] to perform statistical analyses and conduct model testing using Monte Carlo methods.
-We generated exponential, uniform, and linear models and evaluated the model goodness-of-fit by applying Akaike’s Information Criterion (AIC) as a selection criterion [@sakamoto1986akaike; @bevan2017holocene; @riris2019widespread].
+We used the R package 'rcarbon' [@cremaanalysing] to perform statistical analyses and conduct model testing using Monte Carlo methods.
+We generated exponential, uniform, linear, and logistic models and evaluated the model goodness-of-fit by applying Akaike's Information Criterion (AIC) as a selection criterion [@sakamoto1986akaike; @bevan2017holocene; @riris2019widespread].
 
 All analyses and visualization are performed in the R Environment [@team2017r].
-Our R code and data are available on the following GitHub repository: https://github.com/parkgayoung/koreapaleolithicmobilityoccupation as well as on the Open Science Framework: https://osf.io/yagd2/ to enable full reproducibility [@marwick2017computational].
+Our R code and data are available on the following GitHub repository: <https://github.com/parkgayoung/koreapaleolithicmobilityoccupation> as well as on the Open Science Framework: <https://osf.io/yagd2/> to enable full reproducibility [@marwick2017computational].
 
@@
 
+```{=html}
 <!---
@@
 --->
-
+```
 ```{r spd-model, fig.cap="Summed probability distribution of 108 radiocarbon dates.
@@
 
-It could be assumed that foragers with stemmed points were highly mobile groups and chose curated technologies since the stemmed points were likely used for hunting activities [@Chang_2013; @lee2019were].
+It could be assumed that foragers with stemmed points were highly mobile groups and chose curated technologies since the stemmed points were likely used for hunting activities [@chang2013human; @lee2019were].
 However, our results show the opposite that the new technology was more often associated with expedient technological strategies.
@@
 Seong divides five successive assemblage types for the Late/Upper Paleolithic in Korea: (1) quartzite and quartz vein artefacts; (2) mostly small quartzite and quartz vein artefacts with some large artefacts such as cores and choppers; (3) stemmed points dominant; (4) typical blade assemblages including stemmed points; and (5) microblade after 30ka [@seong2015diversity].
-Lee focuses on the Honam region in southwestern Korea and divides the same time frame into two phases; ‘core tool industry with large flake knapping’ during early MIS 3 (59−40 ka), and ‘blade industry with tanged point chiefly’ during late MIS (40−24 ka) [@lee2012characteristics].
+Lee focuses on the Honam region in southwestern Korea and divides the same time frame into two phases; 'core tool industry with large flake knapping' during early MIS 3 (59−40 ka), and 'blade industry with tanged point chiefly' during late MIS (40−24 ka) [@lee2012characteristics].
 Both Seong and Lee mention that core tools made of quartz or quartz vein never disappeared and they point out the appearance of stemmed points and blade technology as an addition, rather than replacement.
-Our results, covering 49-24 ka and corresponding to the first three types of Seong’s divisions, support the continuation of core tools abundantly throughout the Late Pleistocene.
+Our results, covering 49-24 ka and corresponding to the first three types of Seong's divisions, support the continuation of core tools abundantly throughout the Late Pleistocene.
 In addition, our data show that side scrapers stayed as major tools while other tools occupied only a small portion of the assemblages.
@@
 Changes in raw material composition were closely associated with the appearance of stemmed points.
-Previous studies claim that the existing tools, including choppers, polyhedrals, handaxes and cores were usually made of locally acquired quartzite, while the new tools including stemmed points and other blade assemblages were made of more fine-grained materials such as siliceous shale, hornfels, and obsidian which might be brought from distant sources [@Chang_2013; @seong2015diversity].
+Previous studies claim that the existing tools, including choppers, polyhedrals, handaxes and cores were usually made of locally acquired quartzite, while the new tools including stemmed points and other blade assemblages were made of more fine-grained materials such as siliceous shale, hornfels, and obsidian which might be brought from distant sources [@chang2013human; @seong2015diversity].
 Our data also show that quartz, quartz vein and quartzite were consistently dominant and the use of fine-grained materials including chert, hornfels, rhyolite, and shale, increased after 41.5 ka, around the time that stemmed points appear.
@@
 We expected that the functional advantage of the new composite tool might have allowed foragers to be more active in less productive landscapes.
-Previous studies simply describe the environment of the research period as  “overall cooler and drier during MIS 3” but lack detail due to a lack of direct paleoenvironmental proxies in Korea  [@Han_2008; @seong2008tanged; @Choi_2011; @Chang_2013; @im2015study; @bak2017late].
+Previous studies simply describe the environment of the research period as "overall cooler and drier during MIS 3" but lack detail due to a lack of direct paleoenvironmental proxies in Korea [@Han_2008; @seong2008tanged; @Choi_2011; @chang2013human; @im2015study; @bak2017late].
 To explore environmental and demographic context related to the technological transition, we examined the distribution of site elevations, and simulated MAT during the Late Pleistocene.
@@
 However, with the first stemmed points appearing at 38 ka, well before the observed peak in population, we conclude that population dynamics were not a major mechanism in the appearance of this new technology.
-Bae claims that stemmed points are the result of the continuous influx of modern humans from a Siberian migration [@Bae_2010].
+Bae claims that stemmed points are the result of the continuous influx of modern humans from a Siberian migration [@bae2010origin].
 Our results are consistent with this scenario, showing a gradual increase in population after the appearance of stemmed points which may reflect a continuous influx of migrants over a long period.

### 2020-12-14 analysis/paper/paper.Rmd
knit w/o spd

@@
 
-```{=html}
-<!---```{r}
-quietly <- function(x) {
-  sink(tempfile())
-  on.exit(sink())
-  invisible(force(suppressMessages(x)))
-}
-
-``` --->
-```
 # Introduction
@@
 
-```{=html}
-<!---
-Caption for Table 1: Summary of model comparison based on Akaike’s Information Criterion (AIC). 
-We applied logistic, exponential and linear to AIC along with the logistical growth model (fit) to evaluate each model [@riris2019widespread].
-We excluded the uniform model because it has no fitted curve or surface to evaluate.
-The best fit model has the lowest AIC score.
-In this case, the linear model is the best fit, with the next-best fit (exponential) having a higher AIC score by a minor margin at -48.6.
---->
-```
-```{r spd-model, fig.cap="Summed probability distribution of 108 radiocarbon dates.
-The black solid line represents actual radiocarbon ages.
-The grey shaded region shows the critical envelope that encompasses the 95% confidence interval for the null models.
-The red and blue vertical bands highlight the portions of the SPD where positive and negative deviations are detected."}
-source(here::here("analysis/paper/008-summed-probability-distributions.R"))
-knitr::include_graphics(here::here("analysis/figures/008-summed-probability-distribution-models.png"))
-```
-
 # Discussion
@@
 
-## Artifact volumetric density and retouch frequency
+## Artefact volumetric density and retouch frequency
 

### 2020-12-14 analysis/paper/paper.Rmd
change the spd

@@
 
+```{r}	
+quietly <- function(x) {	
+  sink(tempfile())	
+  on.exit(sink())	
+  invisible(force(suppressMessages(x)))	
+}	
+```
 # Introduction
@@
 
-```{r korea-map, fig.cap="Korean Paleolithic sites mentioned in the text."}
-source(here::here("analysis/paper/001-draw-the-map.R"))
+```{r korea-map, fig.cap="Korean Paleolithic sites mentioned in the text.", include=FALSE, cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/001-draw-the-map.R"))}
+quietly(source(here::here("analysis/paper/001-draw-the-map.R")))
 knitr::include_graphics(here::here("analysis/figures/001-site-map.png"))
@@
 The size of data points represent the total number of artefacts from each assemblage and the colour indicates the presence of stemmed points.
-The inset plot shows direct comparison of artefact densities for assemblages with and without stemmed points."}
-source(here::here("analysis/paper/002-artifact-volumetric-density.R"))
+The inset plot shows direct comparison of artefact densities for assemblages with and without stemmed points.", include=FALSE, cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/002-artifact-volumetric-density.R"))}
+quietly(source(here::here("analysis/paper/002-artifact-volumetric-density.R")))
 knitr::include_graphics(here::here("analysis/figures/002-age-by-density.png"))
@@
 The inset plot on the bottom left shows the proportion of retouched pieces in the assemblages over time.
-The inset plot on the bottom right shows the proportion of retouched pieces in the assemblages with or without stemmed points."}
-source(here::here("analysis/paper/003-artifact-volumetric-density-retouch-frequency.R"))
+The inset plot on the bottom right shows the proportion of retouched pieces in the assemblages with or without stemmed points.", include=FALSE, cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/003-artifact-volumetric-density-retouch-frequency.R"))}
+quietly(source(here::here("analysis/paper/003-artifact-volumetric-density-retouch-frequency.R")))
 knitr::include_graphics(here::here("analysis/figures/003-retouch-by-density.png"))
@@
 We excluded artefacts related to manufacturing processes such as pebbles, hammers, flakes and debris, artefacts appeared only in a few assemblages including point, beak shaped, and awl, and unknown and unfinished pieces.
-The colour represents different types of tools and the assemblages are placed in chronological order."}
-source(here::here("analysis/paper/004-artifact-composition.R"))
+The colour represents different types of tools and the assemblages are placed in chronological order.", include=FALSE, cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/004-artifact-composition.R"))}
+quietly(source(here::here("analysis/paper/004-artifact-composition.R")))
 knitr::include_graphics(here::here("analysis/figures/004-artefact-types.png"))
@@
 Hornfels are shown in 4 assemblages but since it occupies more than half of the Yongsandong and Yongband assemblages, we included it in the plot.
-The color represents different types of raw materials, and the assemblages are placed in chronological order."}
-source(here::here("analysis/paper/005-raw-materials.R"))
+The color represents different types of raw materials, and the assemblages are placed in chronological order.", include=FALSE, cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/005-raw-materials.R"))}
+quietly(source(here::here("analysis/paper/005-raw-materials.R")))
 knitr::include_graphics(here::here("analysis/figures/005-raw-materials.png"))
@@
 The lines display a locally weighted regression.
-The inset plot shows a direct relationship between the site elevation and stemmed points."}
-source(here::here("analysis/paper/006-site-elevation.R"))
+The inset plot shows a direct relationship between the site elevation and stemmed points.", include=FALSE, cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/006-site-elevation.R"))}
+quietly(source(here::here("analysis/paper/006-site-elevation.R")))
 knitr::include_graphics(here::here("analysis/figures/006-site-elevation.png"))
@@
 D: The relationship between MAT and site elevation.
-The blue line is a linear regression on elevation and MAT with the grey area showing the 95% confidence interval.t."}
-source(here::here("analysis/paper/007-climate-model-plots.R"))
+The blue line is a linear regression on elevation and MAT with the grey area showing the 95% confidence interval.t.", include=FALSE, cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/007-climate-model-plots.R"))}
+quietly(source(here::here("analysis/paper/007-climate-model-plots.R")))
 knitr::include_graphics(here::here("analysis/figures/007-climate-model-sites-panel-plot.png"))
@@
 Statistical significance of each model is computed using 200 simulations and the global p-values are 0.004, 0.004, 0.009, and 0.004, respectively.
-In this case, the linear model is the best fit with lower value verified by AIC <!---(Table 1)--->.
+In this case, the linear model is the best fit with lower value verified by AIC.
 However, the other models have similar patterns with a minor margin at -48.6.
@@
 The beginning and ending of the SPD indicates significant negative deviation with low density of radiocarbon dates.
+
+```{r spd-model, fig.cap="Summed probability distribution of 108 radiocarbon dates.
+The black solid line represents actual radiocarbon ages.
+The grey shaded region shows the critical envelope that encompasses the 95% confidence interval for the null models.
+The red and blue vertical bands highlight the portions of the SPD where positive and negative deviations are detected.
+We applied logistic, exponential and linear to AIC along with the logistical growth model (fit) to evaluate each model [@riris2019widespread].
+We excluded the uniform model because it has no fitted curve or surface to evaluate.
+The best fit model has the lowest AIC score.
+In this case, the linear model is the best fit, with the next-best fit (exponential) having a higher AIC score by a minor margin at -48.6."}
+source(here::here("analysis/paper/008-summed-probability-distributions.R"))
+knitr::include_graphics(here::here("analysis/figures/008-summed-probability-distribution-models.png"))
+```
 

### 2020-12-14 analysis/paper/008-summed-probability-distributions.R

@@ -158,21 +158,32 @@ df_aic <- data.frame (Model = c("logistic",
                                 ),
                      df = c(aic$df),
                      AIC = c(aic$AIC)) %>% arrange(-AIC)
-df_aic
+# Model df       AIC
+# logistic  4 -363176.8
+#  exponential  3 -368376.7
+#     linear  3 -368425.3
 
-## plots for all three models
-df_exp <-  3
-aic_exp <- 345
-p_exp <-  0.005
 
+## plots for all three models.
+df_exp <- 3
+aic_exp <- -368376.7
+
+df_log <- 4
+aic_log <- -363176.8
+
+df_lin <- 3
+aic_lin <- -368425.3
+
+par(mar=c(2,2,1,1))
 par(mfrow=c(4,1))
-# make a string with the model stats for each plot title
-plot(exp_test, main = paste0("Exponential Model (df = ", df_exp, ", AIC = ", aic_exp, ", p = ", p_exp, ")"))
-plot(uni_test, main="Uniform Model")
-plot(lin_test, main="Linear Model")
-plot(log_test, main="Logistic Model")
 
-dev.off()
+#make a string with the model stats for each plot title
+plot(exp_test, main = paste0("Exponential Model (df = ", df_exp, ", AIC = ", aic_exp,")"))
+plot(uni_test, main = "Uniform Model")
+plot(lin_test, main = paste0("Linear Model (df = ", df_lin, ", AIC = ", aic_lin,")"))
+plot(log_test, main = paste0("Logistic Model (df = ", df_log, ", AIC = ", aic_log,")"))
+
+#dev.off()
 
 png(here::here("analysis/figures/008-summed-probability-distribution-models.png"))
 #ggsave(here::here("analysis/figures/008-summed-probability-distribution-models.png"))

### 2020-12-14 analysis/paper/paper.Rmd
test to push
deleted individual plot for spd models

@@
 In this case, the linear model is the best fit, with the next-best fit (exponential) having a higher AIC score by a minor margin at -48.6."}
-source(here::here("analysis/paper/008-summed-probability-distributions.R"))
+quietly(source(here::here("analysis/paper/008-summed-probability-distributions.R")))
 knitr::include_graphics(here::here("analysis/figures/008-summed-probability-distribution-models.png"))
@@
 
-<!-- The following line inserts a page break when the output is MS Word. For page breaks in PDF, use \newpage on its own line.  -->
+<!--The following line inserts a page break when the output is MS Word. For page breaks in PDF, use \newpage on its own line.  -->
 

### 2020-12-14 analysis/paper/008-summed-probability-distributions.R

@@ -53,8 +53,6 @@ exp_test <-
     ncores = 3
   )
 
-plot(exp_test, main="Exponential Model")
-
 # uniform model, does not produce anything for $fitobject
 uni_test <-
   modelTest(
@@ -66,7 +64,6 @@ uni_test <-
     ncores = 3
   )
 
-plot(uni_test, main="Uniform Model")
 
 # linear model
 lin_test <-
@@ -79,7 +76,6 @@ lin_test <-
     ncores = 3
   )
 
-plot(lin_test, main="Linear Model")
 
 # logistical model
 

### 2020-12-14 analysis/paper/paper.Rmd
deleted include=FALSE

--- 2020-12-14 analysis/paper/paper.Rmd
@@
 
-```{r korea-map, fig.cap="Korean Paleolithic sites mentioned in the text."}
-source(here::here("analysis/paper/001-draw-the-map.R"))
+```{r korea-map, fig.cap="Korean Paleolithic sites mentioned in the text.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/001-draw-the-map.R"))}
+quietly(source(here::here("analysis/paper/001-draw-the-map.R")))
 knitr::include_graphics(here::here("analysis/figures/001-site-map.png"))
@@
 The size of data points represent the total number of artefacts from each assemblage and the colour indicates the presence of stemmed points.
-The inset plot shows direct comparison of artefact densities for assemblages with and without stemmed points."}
-source(here::here("analysis/paper/002-artifact-volumetric-density.R"))
+The inset plot shows direct comparison of artefact densities for assemblages with and without stemmed points.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/002-artifact-volumetric-density.R"))}
+quietly(source(here::here("analysis/paper/002-artifact-volumetric-density.R")))
 knitr::include_graphics(here::here("analysis/figures/002-age-by-density.png"))
@@
 The inset plot on the bottom left shows the proportion of retouched pieces in the assemblages over time.
-The inset plot on the bottom right shows the proportion of retouched pieces in the assemblages with or without stemmed points."}
-source(here::here("analysis/paper/003-artifact-volumetric-density-retouch-frequency.R"))
+The inset plot on the bottom right shows the proportion of retouched pieces in the assemblages with or without stemmed points.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/003-artifact-volumetric-density-retouch-frequency.R"))}
+quietly(source(here::here("analysis/paper/003-artifact-volumetric-density-retouch-frequency.R")))
 knitr::include_graphics(here::here("analysis/figures/003-retouch-by-density.png"))
@@
 We excluded artefacts related to manufacturing processes such as pebbles, hammers, flakes and debris, artefacts appeared only in a few assemblages including point, beak shaped, and awl, and unknown and unfinished pieces.
-The colour represents different types of tools and the assemblages are placed in chronological order."}
-source(here::here("analysis/paper/004-artifact-composition.R"))
+The colour represents different types of tools and the assemblages are placed in chronological order.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/004-artifact-composition.R"))}
+quietly(source(here::here("analysis/paper/004-artifact-composition.R")))
 knitr::include_graphics(here::here("analysis/figures/004-artefact-types.png"))
@@
 Hornfels are shown in 4 assemblages but since it occupies more than half of the Yongsandong and Yongband assemblages, we included it in the plot.
-The color represents different types of raw materials, and the assemblages are placed in chronological order."}
-source(here::here("analysis/paper/005-raw-materials.R"))
+The color represents different types of raw materials, and the assemblages are placed in chronological order.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/005-raw-materials.R"))}
+quietly(source(here::here("analysis/paper/005-raw-materials.R")))
 knitr::include_graphics(here::here("analysis/figures/005-raw-materials.png"))
@@
 The lines display a locally weighted regression.
-The inset plot shows a direct relationship between the site elevation and stemmed points."}
-source(here::here("analysis/paper/006-site-elevation.R"))
+The inset plot shows a direct relationship between the site elevation and stemmed points.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/006-site-elevation.R"))}
+quietly(source(here::here("analysis/paper/006-site-elevation.R")))
 knitr::include_graphics(here::here("analysis/figures/006-site-elevation.png"))
@@
 D: The relationship between MAT and site elevation.
-The blue line is a linear regression on elevation and MAT with the grey area showing the 95% confidence interval.t."}
-source(here::here("analysis/paper/007-climate-model-plots.R"))
+The blue line is a linear regression on elevation and MAT with the grey area showing the 95% confidence interval.t.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/007-climate-model-plots.R"))}
+quietly(source(here::here("analysis/paper/007-climate-model-plots.R")))
 knitr::include_graphics(here::here("analysis/figures/007-climate-model-sites-panel-plot.png"))
@@
 The beginning and ending of the SPD indicates significant negative deviation with low density of radiocarbon dates.
+
+```{r spd-model, fig.cap="Summed probability distribution of 108 radiocarbon dates.
+The black solid line represents actual radiocarbon ages.
+The grey shaded region shows the critical envelope that encompasses the 95% confidence interval for the null models.
+The red and blue vertical bands highlight the portions of the SPD where positive and negative deviations are detected.
+We applied logistic, exponential and linear to AIC along with the logistical growth model (fit) to evaluate each model [@riris2019widespread].
+We excluded the uniform model because it has no fitted curve or surface to evaluate.
+The best fit model has the lowest AIC score.
+In this case, the linear model is the best fit, with the next-best fit (exponential) having a higher AIC score by a minor margin at -48.6.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/008-summed-probability-distributions.R"))}
+quietly(source(here::here("analysis/paper/008-summed-probability-distributions.R")))
+knitr::include_graphics(here::here("analysis/figures/008-summed-probability-distribution-models.png"))
+```
 

### 2020-12-14 analysis/paper/paper.Rmd
map is showing in HTML output now

--- 2020-12-14 analysis/paper/paper.Rmd
@@
 
-```{r korea-map, fig.cap="Korean Paleolithic sites mentioned in the text.", include=FALSE, cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/001-draw-the-map.R"))}
+```{r korea-map, fig.cap="Korean Paleolithic sites mentioned in the text.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/001-draw-the-map.R"))}
 quietly(source(here::here("analysis/paper/001-draw-the-map.R")))

### 2020-12-15 analysis/paper/paper.Rmd
add N arrow, scale, inset map to site map

--- 2020-12-14 analysis/paper/paper.Rmd
@@
 author:
-  - Author Gayoung Park
-  - Author Ben Marwick
+  - Gayoung Park
+  - Ben Marwick
 date: "`r format(Sys.time(), '%d %B, %Y')`"
 output: 
-    bookdown::html_document2: 
+    bookdown::word_document2: 
       fig_caption: yes

### 2020-12-15 analysis/paper/001-draw-the-map.R

@@ -24,16 +24,18 @@ library(ggmap)
 # 33.694662, 123.491142 ... 34.338454, 130.569658
 
 map <-
-  get_stamenmap(bbox = c(left = 123,
+  get_stamenmap(bbox = c(left = 124,
                          bottom = 33,
-                         right = 	131,
+                         right = 	130,
                          top = 39),
                 zoom = 10)
 
 library(ggrepel)
+#devtools::install_github('3wen/legendMap')
+library(legendMap)
 
 # map with site names next to the points
-# map_index <-
+map_of_sites <-
 ggmap(map)  +
   geom_point(data = korean_archaeological_site_locations,
              aes(long_dd ,
@@ -45,65 +47,110 @@ ggmap(map)  +
                    aes(long_dd ,
                        lat_dd,
                        label = site_name),
-                   size = 3,
+                   size = 2,
                   bg.color = "white",
-                  bg.r = 0.15)
-#
-# library(tidyverse)
-# index_table <- korean_archaeological_site_locations %>%
-#   select(ID, site_name)
-#
-#
-# # create a table
-# table_index <- data.frame(Index = korean_archaeological_site_locations$ID,
-#                          Site = korean_archaeological_site_locations$site_name)
-#
-# # print(table_index)
-#
-# #save a table
-# library(here)
-# library(ggpmisc)
-#
-# # add table to plot area, from
-# # https://stackoverflow.com/questions/12318120/adding-table-within-the-plotting-region-of-a-ggplot-in-r
-# map_index +
-#   annotate(geom = "table",
-#            x = 124,
-#            y = 38,
-#            label = list(table_index),
-#            vjust = 1,
-#            hjust = 0)
-#
-#
-#
-# # another method to add table to plot area
-# library(grid)
-# library(gtable)
-# mytheme <- gridExtra::ttheme_minimal(
-#   core = list(fg_params=list(cex = 0.6)),
-#   colhead = list(fg_params=list(cex = 0.5)),
-#   rowhead = list(fg_params=list(cex = 0.25)))
-#
-# g <- tableGrob(table_index, rows = NULL, theme = mytheme)
-# g <- gtable_add_grob(g,
-#                      grobs = rectGrob(gp = gpar(fill = NA, lwd = 1)),
-#                      t = 1, b = nrow(g), l = 1, r = ncol(g))
-#
-# #g$heights <- unit(rep(5/nrow(g), nrow(g), nrow(g)), "cm")
-#
[…]

### 2020-12-15 analysis/paper/002-artifact-volumetric-density.R
make fig 2 4.45 in wide

@@ -64,8 +64,6 @@ den_sp_ttest_df <- round(unname(den_sp_ttest$parameter ), 3)
 den_sp_ttest_str <-
   paste0("t(", den_sp_ttest_df, ") = ", den_sp_ttest_t, ", p = ", den_sp_ttest_p)
 
-
-
 #Volume and artefact counts to get density over time.
 
 density_sp_sub_plot <-
@@ -77,9 +75,9 @@ density_sp_sub_plot <-
            x = 1.5,
            y = 9,
            label = den_sp_ttest_str,
-           size = 2.5) +
-  theme_bw(base_size = 8)  +
-  labs(x = "Stemmed points present",
+           size = 1.5) +
+  theme_bw(base_size = 6)  +
+  labs(x = "Stemmed points present?",
        y = "Artefact density")
 
 
@@ -91,13 +89,9 @@ density_sp_main_plot <-
                  colour = has_sp)) +
   ylab(bquote('Artefact density'~(n/m^3))) +
   xlab("Age of assemblage (ka)") +
-  #geom_smooth(method = "lm") +
   scale_size_continuous(name = "Total number\nof artifacts")  +
-  scale_color_viridis_d(name = "Contains\nstemmed\npoints?") +
-  theme_minimal(base_size = 16)  #+
-#geom_text_repel(aes(label = sites),
-#                nudge_x = 0.25,
-#                nudge_y = 0.25)
+  scale_color_viridis_d(name = "Stemmed\npoints\npresent?") +
+  theme_minimal(base_size = 8)
 
 # https://wilkelab.org/cowplot/articles/drawing_with_on_plots.html
 library(cowplot)
@@ -106,5 +100,8 @@ ggdraw(density_sp_main_plot) +
             .37, .62,
             .32, .33)
 
-ggsave(here::here("analysis/figures/002-age-by-density.png"))
+ggsave(here::here("analysis/figures/002-age-by-density.png"),
+       width = 4.45,
+       height = 4.45,
+       units = "in")
 

### 2020-12-15 analysis/paper/003-artifact-volumetric-density-retouch-frequency.R
fig 3 font sizing and sub-plot heights

@@ -2,6 +2,7 @@ suppressPackageStartupMessages(library(tidyverse))
 library(glue)
 library(ggrepel)
 library(here)
+library(cowplot)
 
 # data from PhD data sheet, not KAS sheet.
 mydata <- read.csv(here("analysis/data/raw_data/General_info.csv"))
@@ -20,8 +21,6 @@ kasv_tidy <-
          sites = names(kasv)[-1]) %>%
   left_join(mydata, by = c('sites' = 'site_name' )) %>%
   mutate(has_sp = ifelse(is.na(SP.), "no", "yes"))
-
-
 
 # retouch, density, age
 Assemblage_info <- read_csv(here::here("analysis/data/raw_data/Assemblage_info.csv"))
@@ -58,7 +57,7 @@ retouch_over_time_subplot <-
              colour = "grey80") +
   labs(x = "Age of assemblage (ka)",
        y = "Proportion retouched") +
-  theme_bw(base_size = 6)
+  theme_bw(base_size = 4)
 
 
 # compute t-test for retouch pieces and SP existence
@@ -85,11 +84,10 @@ retouch_sp_sub_plot <-
            x = 1.5,
            y = 1.25,
            label = retouch_sp_ttest_str,
-           size = 1.7) +
-  theme_bw(base_size = 8)  +
-  labs(x = "Stemmed points present",
+           size = 1.5) +
+  theme_bw(base_size = 4)  +
+  labs(x = "Contains stemmed points?",
        y = "Proportion retouched")
-
 
 # compute correlation
 kas_sites_retouch_density_corr <-
@@ -101,7 +99,7 @@ p_value <- unname(kas_sites_retouch_density_corr$p.value )
 t_value <- unname(kas_sites_retouch_density_corr$statistic )
 df_value <- unname(kas_sites_retouch_density_corr$parameter )
 
-size <-  8
+size <-  6
 Assemblage_info_retouch_density_ages_prop_main_plot <-
   ggplot(Assemblage_info_retouch_density_ages_prop,
          aes(artefact_density,
@@ -120,7 +118,7 @@ Assemblage_info_retouch_density_ages_prop_main_plot <-
                 labels = scales::comma_format(accuracy = 0.001)) +
   scale_x_log10(limits = c(0.001, 10),
                 labels = scales::comma_format(accuracy = 0.001)) +
-  theme_minimal(base_size = 12) +
+  theme_minimal(base_size = 6) +
[…]

### 2020-12-15 analysis/paper/004-artifact-composition.R
set fig 4 width and make legend text pretty

@@ -86,7 +86,9 @@ kasa_long2 %>%
 # filter out rare types
 kasa_long3 <-
 kasa_long2 %>%
-  filter(artefact_type %in% types_in_many_sites$artefact_type)
+  filter(artefact_type %in% types_in_many_sites$artefact_type) %>%
+  # replace _ with space for pretty text in legend
+  mutate(artefact_type = str_replace_all(artefact_type, "_", " "))
 
 kasa_long3_fill_plot <-
 ggplot(kasa_long3,
@@ -95,15 +97,18 @@ ggplot(kasa_long3,
          percentage,
          fill = artefact_type)) +
   geom_col(position = "fill") +
-  ylab("Proportion") +
-  xlab("Assemblage (youngest at the top)") +
-  theme_minimal(base_size = 16)  +
+  ylab("Assemblage proportion") +
+  xlab("") +
   scale_fill_viridis_d(name = "Artefact type",
                        option = "D") +
   coord_flip() +
-  theme_minimal(base_size = 12)
+  theme_minimal(base_size = 8)
 
-ggsave(here::here("analysis/figures/004-artefact-types.png"))
+ggsave(plot = kasa_long3_fill_plot,
+         here::here("analysis/figures/004-artefact-types.png"),
+       width = 4.45,
+       height = 5,
+       units = "in")
 
 #make an interactive plot for interpret the plot easily
 #plotly::ggplotly(kasa_long3_fill_plot)

### 2020-12-15 analysis/paper/005-raw-materials.R
Fig 5 sizing, label text edit

@@ -65,10 +65,12 @@ kasr_long2 <-
   #        age_bin  = ntile(age_ka, 5)) %>%
   arrange(age_ka) %>%
   filter(!is.na(age_ka)) %>%
-  mutate(axis_label = glue('{site_new_name} ({round(age_ka,1)} ka, n = {total})'))
+  mutate(axis_label = glue('{site_new_name} ({round(age_ka,1)} ka, n = {total})')) %>%
+  # replace _ with space for pretty text in legend
+  mutate(raw_material = str_replace_all(raw_material, "_", " "))
 
 
-kasr_long2_pill_plot<-
+kasr_long2_pill_plot <-
 ggplot(kasr_long2,
        aes(
          reorder(axis_label,
@@ -76,14 +78,17 @@ ggplot(kasr_long2,
          percentage,
          fill = raw_material)) +
   geom_col(position = "fill") +
-  xlab("Assemblage (youngest at the top)") +
-  ylab("Percentage") +
-  scale_fill_viridis_d(name = "Raw material type",
+  xlab("") +
+  ylab("Assemblage proportion") +
+  scale_fill_viridis_d(name = "Raw material",
                        option = "C") +
   coord_flip() +
-  theme_minimal(base_size = 12)
+  theme_minimal(base_size = 8)
 
-ggsave(here::here("analysis/figures/005-raw-materials.png"))
+ggsave(here::here("analysis/figures/005-raw-materials.png"),
+       width = 4.45,
+       height = 5,
+       units = "in")
 
 # interactive
 #plotly::ggplotly(kasr_long2_pill_plot)

### 2020-12-15 analysis/paper/006-site-elevation.R
Fig 6 adjust font size and plot size

@@ -3,14 +3,11 @@ library(glue)
 library(ggrepel)
 library(here)
 
-
 # data from PhD data sheet, not KAS sheet.
 mydata <- read.csv(here("analysis/data/raw_data/General_info.csv"))
 
-
 #volume of the cultural layer from KAS data sheet.
 kasv <- read.csv(here("analysis/data/raw_data/Dating_info.csv"))
-
 
 # site elevation
 kasv_tidy <-
@@ -24,7 +21,6 @@ kasv_tidy <-
          sites = names(kasv)[-1]) %>%
   left_join(mydata, by = c('sites' = 'site_name' )) %>%
   mutate(has_sp = ifelse(is.na(SP.), "no", "yes"))
-
 
 mydata_ages <-
   mydata %>%
@@ -47,27 +43,26 @@ elev_sp_ttest_df <- round(unname(elev_sp_ttest$parameter ), 3)
 elev_sp_ttest_str <-
   paste0("t(", elev_sp_ttest_df, ") = ", elev_sp_ttest_t, ", p = ", elev_sp_ttest_p)
 
-
 elevation_sp_sub_plot <-
   ggplot(kasv_tidy,
          aes(has_sp,
              altitude.m._of_main_layer)) +
-  geom_boxplot() +
+  geom_boxplot(outlier.size = 0.5) +
   annotate("text",
            x = 1.5,
            y = 250,
            label = elev_sp_ttest_str,
-           size = 2.5) +
-  theme_bw(base_size = 7)  +
-  labs(x = "Stemmed points present",
-       y = "Elevation above sea level (m)")
+           size = 1.5) +
+  theme_bw(base_size = 6)  +
+  labs(x = "Contains stemmed points?",
+       y = "Elevation above\nsea level (m)")
 
 elevation_sp_main_plot <-
   ggplot(mydata_ages,
          aes(x = age_ka,
              y = altitude.m._of_main_layer
              )) +
-  geom_point(size = 3,
+  geom_point(size = 2,
              aes(color = as.factor(has_sp))) +
   xlab("Age of occupation (ka)") +
   ylab("Elevation above sea level (m)") +
@@ -75,7 +70,7 @@ elevation_sp_main_plot <-
               aes(color = as.factor(has_sp))) +
  # geom_smooth(se = FALSE)
   scale_colour_discrete(name = "Contains stemmed points?") +
-  theme_minimal(base_size = 14)  +
+  theme_minimal(base_size = 6)  +
[…]

### 2020-12-16 analysis/paper/007-climate-model-plots.R
Fig 7 many edits to polish this multi-panel figure and set size

@@ -2,6 +2,7 @@
 library(ncdf4)
 library(lattice)
 library(ggplot2)
+library(cowplot)
 
 # you'll need to get this file from
 # https://figshare.com/articles/LateQuaternary_Environment_nc/12293345/3
@@ -17,6 +18,7 @@ temperature <- ncdf4::ncvar_get(env_nc, "temperature")
 biome       <- ncdf4::ncvar_get(env_nc, "biome")
 ncdf4::nc_close(env_nc)
 
+# this takes a few minutes
 mean_annual_temperature <- apply(temperature, c(1, 2, 4), mean)
 
 # korean archaeological sites
@@ -24,7 +26,6 @@ library(tidyverse)
 
 site_locations_tbl_raw <- read_csv (here::here("analysis/data/raw_data/korean_palaeolithic_site_locations.csv"))
 site_locations_tbl <- site_locations_tbl_raw
-
 
 # some problem finding a MAT value for the location
 # Gigok
@@ -59,54 +60,67 @@ site_locations_tbl %>%
   filter(between(year, -50000, -10000))
 
 #----------------------------------------------------------------------
+base_size = 6
 
 # this is the overall time series plot
 # we see the mean for all locations, and individual
 # locations in grey in the background
-mat_time_series_plot_no_annotate <-
+mat_time_series_plot_annotated <-
 ggplot() +
+  # add annotations for climate events
+  annotate ("rect",
+            xmin = 14000, xmax = 29000,
+            ymin = -Inf, ymax = Inf, fill = "lightgrey",
+            alpha = .4) +
+  annotate("rect",
+           xmin = 19000, xmax = 26500,
+           ymin = -Inf, ymax = Inf, fill = "grey",
+           alpha = .4)   +
+  # add lines for each location
   geom_line(data = site_locations_tbl_temps,
             aes(-year,
                 mean_annual_temperature,
                 group = site_name),
             colour = "grey90") +
+  # add mean line for all locations
   geom_line(data = site_locations_tbl_temps %>%
               group_by(year) %>%
               summarise(mean_mean_annual_temperature = mean(mean_annual_temperature,
                                                             na.rm = TRUE)),
             aes(-year,
                 mean_mean_annual_temperature),
-            size = 2) +
+            size = 1) +
+  annotate("text",
+           x = 11100,
+           y = 11,
+           label = "MIS\n1",
+           size = 3) +
+  annotate("text",
+           x = 22000,
+           y = 11,
+           label = "MIS\n2",
[…]

### 2020-12-16 analysis/paper/paper.Rmd
set fig 4 width and make legend text pretty
Fig 5 sizing, label text edit
suppress section numbers in output, stop superflous world map from appearing

@@
 ```
-# Introduction
+# Introduction {-}
 
@@
 
-# Background
+# Background {-}
 
@@
 
-## Behavioural ecology and forager land use behaviours
+## Behavioural ecology and forager land use behaviours {-}
 
@@
 
-# Methods
+# Methods {-}
 
-## Sites and dates
+## Sites and dates {-}
+
+```{r korea-map, fig.cap="Korean Paleolithic sites mentioned in the text.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/001-draw-the-map.R"))}
+quietly(source(here::here("analysis/paper/001-draw-the-map.R")))
+knitr::include_graphics(here::here("analysis/figures/001-site-map.png"))
+```
 
@@
 
-```{r korea-map, fig.cap="Korean Paleolithic sites mentioned in the text.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/001-draw-the-map.R"))}
-quietly(source(here::here("analysis/paper/001-draw-the-map.R")))
-knitr::include_graphics(here::here("analysis/figures/001-site-map.png"))
-
-```
-
-## Assemblages
+## Assemblages {-}
 
@@
 
-## Statistical methods
+## Statistical methods {-}
 
@@
 
-# Results
+# Results {-}
 
-## Artefact volumetric density and retouch frequency
+## Artefact volumetric density and retouch frequency {-}
 
@@
 
-## Composition of toolkit
+## Toolkit composition {-}
 
@@
 
-## Raw materials
+## Raw materials {-}
 
@@
 
-## Environmental context
+## Environmental context {-}
 
@@
 Although not a statistically significant difference in elevations (p = 0.102), this may indicate that forager groups who used stemmed points were able to occupy higher altitudes, while the groups without tended to prefer lower altitudes.
-Sites without stemmed points reach a maximum elevation of about 150m at around 39ka.
+Sites without stemmed points reach a maximum elevation of about 150 m at around 39ka.
 Sites containing stemmed points reach a maximum elevation of 175 m at 28ka.
@@
 
-## Demographic context
+## Demographic context {-}
 
@@
 
-# Discussion
+# Discussion {-}
 
-Our study focussed on two questions: how did the introduction of the new tools change foragers' mobility and landscape use; and how did the new technology change the way people use habitation sites?
+Our study focused on two questions: how did the introduction of the new tools change foragers' mobility and landscape use; and how did the new technology change the way people use habitation sites?
 Overall, our results show that assemblages with stemmed points tend to have higher artefact densities and lower proportions of retouched pieces.
@@
 
-## Artefact volumetric density and retouch frequency
+## Artefact volumetric density and retouch frequency {-}
 
@@
 
-## Composition of toolkits and raw materials
+## Toolkit composition and raw materials {-}
 
@@
 
-## Environmental and demographic context
+## Environmental and demographic contexts {-}
 
@@
 
-# Conclusion
+# Conclusion {-}
 
@@
 
-# Acknowledgements
+# Acknowledgements {-}
 
@@
 
-# References
+# References {-}
 
@@
 
-<div>
-
-</div>
+::: {#refs}
+:::
 
@@
 
-### Colophon
+### Colophon {-}
 

### 2020-12-16 analysis/paper/001-draw-the-map.R

@@ -7,7 +7,6 @@ korean_archaeological_site_locations <-
 korean_archaeological_site_locations <-
   tibble::rowid_to_column(korean_archaeological_site_locations, "ID")
 
-
 library(ggplot2)
 
 # # check the basic look
@@ -15,7 +14,6 @@ library(ggplot2)
 #   aes(x = long_dd,
 #       y = lat_dd) +
 #   geom_point()
-
 
 library(ggmap)
 
@@ -31,7 +29,7 @@ map <-
                 zoom = 10)
 
 library(ggrepel)
-#devtools::install_github('3wen/legendMap')
+# devtools::install_github('3wen/legendMap')
 library(legendMap)
 
 # map with site names next to the points
@@ -91,7 +89,7 @@ library(ggplot2)
 mycrs <- "+proj=longlat +datum=WGS84 +no_defs"
 
 #Using the original maps package, then converting map into SpatialPolygons object
-world <- maps::map("world", fill=TRUE) %$%
+world <- maps::map("world", fill=TRUE,  plot = FALSE) %$%
   maptools::map2SpatialPolygons(., IDs=names,proj4string=CRS(mycrs))
 
 #The resulting map has self intersection problems so any further operation reports errors; using buffers of width 0 is a fast fix
@@ -110,12 +108,12 @@ ggplot() +
                    group=group),
                fill='NA',
                color='black',
-               size=0.2) +
+               size=0.1) +
   annotate("rect",
-           xmin = 119,
-           xmax = 136,
-           ymin = 29,
-           ymax = 46,
+           xmin = 118,
+           xmax = 137,
+           ymin = 28,
+           ymax = 47,
            fill = NA,
            colour = "red",
            size = 0.5

### 2020-12-16 analysis/paper/003-artifact-volumetric-density-retouch-frequency.R

@@ -45,7 +45,7 @@ Assemblage_info_retouch_density_ages_prop <-
   group_by(sites) %>%
   mutate(prop_retouched = sum(value, na.rm = TRUE) / total_artifacts) %>%
   arrange(sites) %>%
-  select(-X1, -value) %>%
+  dplyr::select(-X1, -value) %>%
   distinct_all()
 
 # retouch over time

### 2020-12-16 analysis/paper/004-artifact-composition.R

@@ -58,7 +58,7 @@ kasa_long1 <-
   #                                     as.character(X), "other_tools"))
 ) %>%
   mutate(axis_label = glue('{site}\n(n = {total})'))  %>%
-  select(-X)
+  dplyr::select(-X)
 
 # join to get ages of the sites
 kasa_long2 <-
@@ -110,25 +110,9 @@ ggsave(plot = kasa_long3_fill_plot,
        height = 5,
        units = "in")
 
-#make an interactive plot for interpret the plot easily
-#plotly::ggplotly(kasa_long3_fill_plot)
+# make an interactive plot for interpret the plot easily
+# plotly::ggplotly(kasa_long3_fill_plot)
 
-
-
-#ggplot(kasa_long3,
-#       aes(
-#         reorder(axis_label, -age_ka),
-#         percentage)) +
-#  geom_col() +
-#  ylab("Proportion") +
-#  xlab("Assemblage (youngest at the top)") +
-#  theme_minimal(base_size = 16)  +
-#  scale_fill_viridis_d(name = "Artefact type",
-#                       option = "D") +
-#  facet_wrap( ~ artefact_type,
-#              ncol = 1,
-#              scales = "free_y") +
-#  theme_minimal(base_size = 6)
 
 
 

### 2020-12-16 analysis/paper/005-raw-materials.R

@@ -26,7 +26,7 @@ mydata_ages <-
   gather(site,
          count,
          -X) %>%
-    filter(!is.na(count)) %>%
+  filter(!is.na(count)) %>%
   group_by(site) %>%
   mutate(percentage = count / sum(count, na.rm = TRUE) * 100,
          total = sum(count, na.rm = TRUE)) %>%
@@ -45,7 +45,7 @@ mydata_ages <-
   # if percentage is <10%, call it 'other'
   mutate(raw_material = X) %>%  #ifelse(percentage >= 10, as.character(X), "other")) %>%
   #mutate(raw_material = ifelse(raw_material == "etc.", "other", raw_material)) %>%
-  select(-X)
+  dplyr::select(-X)
 
 
   # join to get ages of the sites
@@ -55,7 +55,7 @@ kasr_long2 <-
   summarise(percentage = sum(percentage)) %>%
   ungroup() %>%
   left_join(kasr_long1 %>%
-              select(site, raw_material, total),
+              dplyr::select(site, raw_material, total),
             by = c('site', "raw_material")) %>%
   distinct(.keep_all = TRUE) %>%
   # join to get ages of the sites
@@ -91,49 +91,8 @@ ggsave(here::here("analysis/figures/005-raw-materials.png"),
        units = "in")
 
 # interactive
-#plotly::ggplotly(kasr_long2_pill_plot)
+# plotly::ggplotly(kasr_long2_pill_plot)
 
-
-#ggplot(kasr_long2,
-#       aes(
-#         reorder(axis_label,
-#                 -age_ka),
-#         percentage)) +
-#  geom_col() +
-#  xlab("Assemblage (youngest at the top)") +
-#  ylab("Percentage") +
-#  theme_minimal(base_size = 16) +
-#  scale_fill_viridis_d(name = "Raw material type",
-#                       option = "C") +
-#  facet_wrap(~raw_material,
-#             ncol = 1,
-#             scales = "free_y") +
-#  theme_minimal(base_size = 6)
-
-
-# raw material groups  ---------------------
-#fg <- c('chert', 'hornfels', 'rhyolite', 'shale')
-#qz <- c('quartz', 'quartzite', 'quartz_vein')
-
-#kasr_long2_rm_group <-
-#kasr_long2 %>%
-#  mutate(raw_material_group = ifelse(raw_material %in% fg, 'fg',
-#                                     ifelse(raw_material %in% qz, "qz",
-#                                     "oth")))
-
[…]

### 2020-12-16 analysis/paper/007-climate-model-plots.R

@@ -47,7 +47,7 @@ site_locations_tbl$long_dd <-
 
 site_locations_tbl_temps <-
 site_locations_tbl %>%
-  select(site_name,
+  dplyr::select(site_name,
          lat_dd,
          long_dd,
          elevation) %>%
@@ -183,7 +183,7 @@ site_locations_tbl_temps_periods_filtered %>%
 mat_site_map_plot <-
 ggmap(map)  +
   geom_point(data = site_locations_tbl_temps_periods_filtered_means %>%
-               select(-long_dd,
+               dplyr::select(-long_dd,
                       -lat_dd) %>%
                left_join(site_locations_tbl_raw) ,
              aes(long_dd ,
@@ -191,7 +191,7 @@ ggmap(map)  +
              colour = av_mat),
              size = 2) +
   geom_text_repel(data = site_locations_tbl_temps_periods_filtered_means %>%
-                    select(-long_dd,
+                    dplyr::select(-long_dd,
                            -lat_dd) %>%
                     left_join(site_locations_tbl_raw),
                    aes(long_dd ,

### 2020-12-16 analysis/paper/002-artifact-volumetric-density.R
Reduce line thickness on all boxplots

@@ -70,7 +70,7 @@ density_sp_sub_plot <-
   ggplot(kasv_tidy,
          aes(has_sp,
              artefact_density)) +
-  geom_boxplot() +
+  geom_boxplot(lwd = 0.1) +
   annotate("text",
            x = 1.5,
            y = 9,

### 2020-12-16 analysis/paper/003-artifact-volumetric-density-retouch-frequency.R

@@ -79,7 +79,7 @@ retouch_sp_sub_plot <-
   ggplot(Assemblage_info_retouch_density_ages_prop,
          aes(has_sp,
              prop_retouched)) +
-  geom_boxplot() +
+  geom_boxplot(lwd = 0.1) +
   annotate("text",
            x = 1.5,
            y = 1.25,

### 2020-12-16 analysis/paper/006-site-elevation.R

@@ -47,7 +47,7 @@ elevation_sp_sub_plot <-
   ggplot(kasv_tidy,
          aes(has_sp,
              altitude.m._of_main_layer)) +
-  geom_boxplot(outlier.size = 0.5) +
+  geom_boxplot(outlier.size = 0.5, lwd = 0.1) +
   annotate("text",
            x = 1.5,
            y = 250,

### 2020-12-17 analysis/paper/paper.Rmd
commit to merge

--- 2020-12-15 analysis/paper/paper.Rmd
@@
 author:
-  - Gayoung Park
-  - Ben Marwick
+  - Author Gayoung Park
+  - Author Ben Marwick
 date: "`r format(Sys.time(), '%d %B, %Y')`"
 output: 
-    bookdown::word_document2: 
+    bookdown::html_document2: 
       fig_caption: yes
@@
 knitr::include_graphics(here::here("analysis/figures/001-site-map.png"))
-
 ```

### 2020-12-17 analysis/paper/paper.Rmd

--- 2020-12-16 analysis/paper/paper.Rmd
@@
 
-```{r korea-map, fig.cap="Korean Paleolithic sites mentioned in the text.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/001-draw-the-map.R"))}
-quietly(source(here::here("analysis/paper/001-draw-the-map.R")))
-knitr::include_graphics(here::here("analysis/figures/001-site-map.png"))
-```
-
 After the first excavation of a Paleolithic site in Korea at Seokjang-ri site in the 1960s, more than 200 Paleolithic sites have been discovered in South Korea [@lee2019were].
@@
 We obtained the data from published excavation reports held by provincial museums.
+
+```{r korea-map, fig.cap="Korean Paleolithic sites mentioned in the text.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/001-draw-the-map.R"))}
+quietly(source(here::here("analysis/paper/001-draw-the-map.R")))
+knitr::include_graphics(here::here("analysis/figures/001-site-map.png"))
+```
 
@@
 D: The relationship between MAT and site elevation.
-The blue line is a linear regression on elevation and MAT with the grey area showing the 95% confidence interval.t.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/007-climate-model-plots.R"))}
+The blue line is a linear regression on elevation and MAT with the grey area showing the 95% confidence interval.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/007-climate-model-plots.R"))}
 quietly(source(here::here("analysis/paper/007-climate-model-plots.R")))

### 2020-12-21 analysis/paper/paper.Rmd
added stone artifact figures

@@
 
-The appearance of stemmed points in the Late Pleistocene (\~40-35 ka) in Korea is often assumed to have transformed forager lifeways because it reflected more specialized hunting practices [@seong2008tanged; @chang2013human].
+The appearance of stemmed points (Figure \@ref(fig:stemmed-points)) in the Late Pleistocene (\~40-35 ka) in Korea is often assumed to have transformed forager lifeways because it reflected more specialized hunting practices [@seong2008tanged; @chang2013human].
 Stemmed points are often called as tanged points in the western hemisphere, but we use 'stemmed point' to distinguish from Bronze Age stone projectile points that have long been called 'tanged points' in Korea.
@@
 We examined stone artefacts from 22 sites in South Korea to investigate the changes in human behaviour associated with the appearance stemmed points, and the blade industry that followed.
+
+```{r stemmed-points, fig.cap = "Stemmed points from Yongsandong site"}
+knitr::include_graphics(here::here("analysis/figures/YS_SP.png"))
+```
 
@@
 
-```{r volume-density, fig.cap="The main plot shows the artefact volumetric density over time.
+```{r volume-density, fig.cap="Artefact volumetric density over time.
 The size of data points represent the total number of artefacts from each assemblage and the colour indicates the presence of stemmed points.
@@
 
-```{r volume-density-retouch, fig.cap="Main plot shows relationship between artefact density and retouch frequency.
+```{r volume-density-retouch, fig.cap="Relationship between artefact density and retouch frequency.
 The colour of the data points indicate the age of the assemblage, point size indicates assemblage size, and shape indicates presence or absence of stemmed points.
@@
 
-```{r artefact-composition, fig.cap="This plot shows the composition of each assemblage.
+```{r artefact-composition, fig.cap="The composition of each assemblage.
 We excluded artefacts related to manufacturing processes such as pebbles, hammers, flakes and debris, artefacts appeared only in a few assemblages including point, beak shaped, and awl, and unknown and unfinished pieces.
@@
 knitr::include_graphics(here::here("analysis/figures/004-artefact-types.png"))
+```
+
+```{r SAR-all, fig.cap = "Stone artefacts from Songamri site. 1. Blade core, 2~4. Blades, 5~6. Stemmed points, 7. Side scraper, 8. Chopper, 9~10. Side scrapers."}
+knitr::include_graphics(here::here("analysis/figures/SAR_all.png"))
 ```
@@
 
-```{r raw-materials, fig.cap="This plot shows the composition of raw material type in each assemblage.
+```{r raw-materials, fig.cap="The composition of raw material type in each assemblage.
 We excluded raw materials that are included in less than 5 assemblages including crystal, basalt, iron_ore, slate, limestone, granite, gneiss, tuff, and unidentified ones.
@@
 
-```{r site-elevation, fig.cap="The main plot shows the site elevation by the age of the deposit.
+```{r site-elevation, fig.cap="The site elevation by the age of the deposit.
 Colour indicates presence/absence of stemmed points.

### 2021-01-04 analysis/paper/paper.Rmd
added citation and SPD figure/ title inside

@@
 We explored SPD to test for correlations between population, environmental risk and technological change.
-We used the R package 'rcarbon' [@cremaanalysing] to perform statistical analyses and conduct model testing using Monte Carlo methods.
+We used the R package 'rcarbon' [@cremaanalysing] to perform statistical analyses and conduct model testing using Monte Carlo methods [@timpson2014reconstructing; @crema2016summed; @palmisano2020holocene].
 We generated exponential, uniform, linear, and logistic models and evaluated the model goodness-of-fit by applying Akaike's Information Criterion (AIC) as a selection criterion [@sakamoto1986akaike; @bevan2017holocene; @riris2019widespread].

### 2021-01-04 analysis/paper/008-summed-probability-distributions.R

@@ -170,16 +170,14 @@ aic_log <- -363176.8
 df_lin <- 3
 aic_lin <- -368425.3
 
+#make a string with the model stats for each plot title : basic figure
 par(mar=c(2,2,2,2))
 par(mfrow=c(4,1))
 
-#make a string with the model stats for each plot title
 plot(exp_test, main = paste0("Exponential Model (df = ", df_exp, ", AIC = ", aic_exp,")"))
 plot(uni_test, main = "Uniform Model")
 plot(lin_test, main = paste0("Linear Model (df = ", df_lin, ", AIC = ", aic_lin,")"))
 plot(log_test, main = paste0("Logistic Model (df = ", df_log, ", AIC = ", aic_log,")"))
-
-#dev.off()
 
 png(here::here("analysis/figures/008-summed-probability-distribution-models.png"),
     h = 5,
@@ -187,10 +185,70 @@ png(here::here("analysis/figures/008-summed-probability-distribution-models.png"
     units = "px")
 #ggsave(here::here("analysis/figures/008-summed-probability-distribution-models.png"))
 
+dev.off()
+
+#
 
 
-#save a table of AIC result as csv
-# write.csv(df_aic, file="analysis/figures/008-SPD-AIC.csv", row.names = FALSE)
+# Codes from https://zenodo.org/record/4322979#.X-4bH-lKgWo
+# fig 2
+# layout(matrix(c(1,2,3,4,5,6,7), 7, 1, byrow=TRUE), widths=6, heights=c(1,1,1,1,1,1,1.8))
+# par(mar=c(0, 1, 0, 1))
+# par(yaxs="i")
+# par(xaxs="i")
+#
+ymax <- max(ko.spd.smoothed$grid$PrDens)
 
-# knitr::kable(df_aic)
+plot(exp_test, ylim=c(0,ymax), xlim=c(50000,10000), drawaxes=FALSE )
+legend(x=49000, ymax*0.94, legend=c("SPD", "95% MC envelope","positive deviation","negative deviation"),col=c( "black","lightgrey",rgb(0.7,0,0,0.2),rgb(0,0,0.7,0.2)), lwd=c(0.5,5,5,5), cex=0.6, bg="white", title="")
[…]

### 2021-01-04 analysis/paper/008-summed-probability-distributions.R
remove the x-axis ticks and labels for all except the bottom plot

@@ -187,68 +187,82 @@ png(here::here("analysis/figures/008-summed-probability-distribution-models.png"
 
 dev.off()
 
-#
-
-
+#move the title inside the plot area
 # Codes from https://zenodo.org/record/4322979#.X-4bH-lKgWo
-# fig 2
-# layout(matrix(c(1,2,3,4,5,6,7), 7, 1, byrow=TRUE), widths=6, heights=c(1,1,1,1,1,1,1.8))
-# par(mar=c(0, 1, 0, 1))
-# par(yaxs="i")
-# par(xaxs="i")
-#
 ymax <- max(ko.spd.smoothed$grid$PrDens)
 
+par(mar=c(2,2,2,2))
+par(mfrow=c(4,1))
+
 plot(exp_test, ylim=c(0,ymax), xlim=c(50000,10000), drawaxes=FALSE )
-legend(x=49000, ymax*0.94, legend=c("SPD", "95% MC envelope","positive deviation","negative deviation"),col=c( "black","lightgrey",rgb(0.7,0,0,0.2),rgb(0,0,0.7,0.2)), lwd=c(0.5,5,5,5), cex=0.6, bg="white", title="")
+legend(x=49000, ymax*0.94, legend=c("radiocarbon ages", "95% MC envelope","positive deviation","negative deviation"),col=c( "black","lightgrey",rgb(0.7,0,0,0.2),rgb(0,0,0.7,0.2)), lwd=c(0.5,5,5,5), cex=0.6, bg="white", title="")
 text(x=48000, ymax*0.90, labels="Exponential Model", font=2, cex=0.6, adj=c(0,0.7))
 text(x=48000, ymax*0.54, cex=0.5, font=2, adj=c(0,0),labels=paste0("(df = ", df_exp, ", AIC = ", aic_exp,")"))
 box()
 xticks <- seq(50000,10000,-10000)
 axis(side=1, at=xticks, labels=xticks, cex.axis=0.8)
 mtext("Years cal BP",1, 2.3, at=34000, adj=0, font=2, cex=0.8)
+plot(uni_test, ylim=c(0,ymax), xlim=c(50000,10000), drawaxes=FALSE )
[…]

### 2021-01-04 analysis/paper/008-summed-probability-distributions.R
knit the new changes

@@ -171,64 +171,58 @@ df_lin <- 3
 aic_lin <- -368425.3
 
 #make a string with the model stats for each plot title : basic figure
-par(mar=c(2,2,2,2))
-par(mfrow=c(4,1))
+# par(mar=c(2,2,2,2))
+# par(mfrow=c(4,1))
+#
+# plot(exp_test, main = paste0("Exponential Model (df = ", df_exp, ", AIC = ", aic_exp,")"))
+# plot(uni_test, main = "Uniform Model")
+# plot(lin_test, main = paste0("Linear Model (df = ", df_lin, ", AIC = ", aic_lin,")"))
+# plot(log_test, main = paste0("Logistic Model (df = ", df_log, ", AIC = ", aic_log,")"))
 
-plot(exp_test, main = paste0("Exponential Model (df = ", df_exp, ", AIC = ", aic_exp,")"))
-plot(uni_test, main = "Uniform Model")
-plot(lin_test, main = paste0("Linear Model (df = ", df_lin, ", AIC = ", aic_lin,")"))
-plot(log_test, main = paste0("Logistic Model (df = ", df_log, ", AIC = ", aic_log,")"))
 
-png(here::here("analysis/figures/008-summed-probability-distribution-models.png"),
-    h = 5,
-    w = 4,
-    units = "px")
-#ggsave(here::here("analysis/figures/008-summed-probability-distribution-models.png"))
-
-dev.off()
 
 #move the title inside the plot area
 # Codes from https://zenodo.org/record/4322979#.X-4bH-lKgWo
 ymax <- max(ko.spd.smoothed$grid$PrDens)
 
-par(mar=c(2,2,2,2))
-par(mfrow=c(4,1))
-
-plot(exp_test, ylim=c(0,ymax), xlim=c(50000,10000), drawaxes=FALSE )
-legend(x=49000, ymax*0.94, legend=c("radiocarbon ages", "95% MC envelope","positive deviation","negative deviation"),col=c( "black","lightgrey",rgb(0.7,0,0,0.2),rgb(0,0,0.7,0.2)), lwd=c(0.5,5,5,5), cex=0.6, bg="white", title="")
-text(x=48000, ymax*0.90, labels="Exponential Model", font=2, cex=0.6, adj=c(0,0.7))
[…]

### 2021-01-04 analysis/paper/008-summed-probability-distributions.R
saved image doesn't change

@@ -259,7 +259,7 @@ png(here::here("analysis/figures/008-summed-probability-distribution-models.png"
     w = 4,
     units = "px")
 
-dev.off()
+#dev.off()
 
 
 

### 2021-01-12 analysis/paper/paper.Rmd
remove the x-axis ticks and labels for all except the bottom plot
please don't hard-code critical results of the analysis

@@
 
-```{r spd-model, fig.cap="Summed probability distribution of 108 radiocarbon dates.
+fig.cap="Summed probability distribution of 108 radiocarbon dates.
 The black solid line represents actual radiocarbon ages.
-The grey shaded region shows the critical envelope that encompasses the 95% confidence interval for the null models.
+The grey shaded region shows the Monte Carlo envelope that encompasses the 95% confidence interval for the null models.
 The red and blue vertical bands highlight the portions of the SPD where positive and negative deviations are detected.
@@
 The best fit model has the lowest AIC score.
-In this case, the linear model is the best fit, with the next-best fit (exponential) having a higher AIC score by a minor margin at -48.6.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/008-summed-probability-distributions.R"))}
+In this case, the linear model is the best fit, with the next-best fit (exponential) having a higher AIC score by a minor margin at -48.6."
+
+```{r spd-model,  cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/008-summed-probability-distributions.R"))}
 quietly(source(here::here("analysis/paper/008-summed-probability-distributions.R")))

### 2021-01-12 analysis/paper/008-summed-probability-distributions.R

@@ -153,22 +153,24 @@ df_aic <- data.frame (Model = c("logistic",
                                # "uniform"
                                 ),
                      df = c(aic$df),
-                     AIC = c(aic$AIC)) %>% arrange(-AIC)
+                     AIC = c(aic$AIC)) %>%
+  arrange(-AIC)
 # Model df       AIC
-# logistic  4 -363176.8
-#  exponential  3 -368376.7
-#     linear  3 -368425.3
+#  logistic  4     -363176.8
+#  exponential  3  -368376.7
+#  linear  3       -368425.3
 
 
 ## plots for all three models.
-df_exp <- 3
-aic_exp <- -368376.7
 
-df_log <- 4
-aic_log <- -363176.8
+df_exp <-  df_aic %>% filter(Model == 'exponential') %>% pull(df)
+aic_exp <- df_aic %>% filter(Model == 'exponential') %>% pull(AIC) # -368376.7
 
-df_lin <- 3
-aic_lin <- -368425.3
+df_log <-  df_aic %>% filter(Model == 'logistic') %>% pull(df) # 4
+aic_log <- df_aic %>% filter(Model == 'logistic') %>% pull(AIC) # -363176.8
+
+df_lin <-  df_aic %>% filter(Model == 'linear') %>% pull(df) # 3
+aic_lin <- df_aic %>% filter(Model == 'linear') %>% pull(AIC) # -368425.3
 
 #make a string with the model stats for each plot title : basic figure
 # par(mar=c(2,2,2,2))
@@ -185,74 +187,184 @@ aic_lin <- -368425.3
 # Codes from https://zenodo.org/record/4322979#.X-4bH-lKgWo
 ymax <- max(ko.spd.smoothed$grid$PrDens)
 
-# par(mar=c(2,2,2,2))
-# par(mfrow=c(4,1))
-#
-# plot(exp_test, ylim=c(0,ymax), xlim=c(50000,10000), drawaxes=FALSE )
[…]

### 2021-01-12 analysis/paper/008-summed-probability-distributions.R
We can save the SPD multi-panel plot to png now.

@@ -189,6 +189,15 @@ ymax <- max(ko.spd.smoothed$grid$PrDens)
 
 # remove the x-axis
 
+# start the file that will hold the plot, then draw the plot
+# after this...
+png(here::here("analysis/figures/008-summed-probability-distribution-models.png"),
+    h = 15,
+    w = 13,
+    res = 1000,
+    units = "cm")
+
+# begin drawing the plot
 layout(
   matrix(c(1, 2, 3, 4), 4, 1, byrow = TRUE),
   widths = 6,
@@ -366,12 +375,10 @@ mtext(
   cex = 0.8
 )
 
-png(here::here("analysis/figures/008-summed-probability-distribution-models.png"),
-    h = 5,
-    w = 4,
-    units = "px")
+# now close the connection to the PNG file so we can take a look at it
+dev.off()
 
-#dev.off()
+# the figure is 008-summed-probability-distribution-models.png
 
 
 

### 2021-01-15 analysis/paper/paper.Rmd
adjust legand for SPD and knit rmd file

--- 2021-01-04 analysis/paper/paper.Rmd
@@
 ```
+
+# Keywords {-}
+Human Behavioural Ecology, Stemmed points, Mobility, Site occupation, Landscape use, Summed Probability Distribution, Pleistocene, Korean Late Paleolithic
+
 # Introduction {-}
@@
 The black solid line represents actual radiocarbon ages.
-The grey shaded region shows the critical envelope that encompasses the 95% confidence interval for the null models.
+The grey shaded region shows the Monte Carlo envelope that encompasses the 95% confidence interval for the null models.
 The red and blue vertical bands highlight the portions of the SPD where positive and negative deviations are detected.
@@
 
+# CRediT authorship contribution statement {-}
+Gayoung Park: Conceptualization, Data curation, Methodology, Software, Validation, Formal analysis, Resources, Writing - original draft, Visualization, Project administration, Funding acquisition.
+
+Ben Marwick: Methodology, Formal analysis, Investigation, Software, Validation, Visualization, Supervision, Project administration, Funding acquisition.
+
+# Funding source {-}
+This work was supported by Quaternary Research Center funding, and travel grants from the Department of Anthropology and Center for Statistics and the Social Sciences at University of Washington.
+
 # Acknowledgements {-}

### 2021-01-15 analysis/paper/008-summed-probability-distributions.R

@@ -239,12 +239,13 @@ text(
 # draw Exp plot: model stats
 text(
   x = 48000,
-  ymax * 0.8,
+  ymax * 0.49,
   cex = 0.5,
   font = 2,
   adj = c(0, 0),
-  labels = paste0("df = ", df_exp, ", AIC = ", aic_exp)
+  labels = paste0("(df = ", df_exp, ", AIC = ", aic_exp, ")")
 )
+
 box()
 # draw Uni plot
 par(mar = c(0, 1, 0, 1))
@@ -280,6 +281,8 @@ text(
   adj = c(0, 0.7)
 )
 box()
+
+# draw lin plot
 par(mar = c(0, 1, 0, 1))
 plot(
   lin_test,
@@ -351,7 +354,7 @@ text(
 )
 text(
   x = 48000,
-  ymax * 0.54,
+  ymax * 0.47,
   cex = 0.5,
   font = 2,
   adj = c(0, 0),

### 2021-01-16 analysis/paper/008-summed-probability-distributions.R
fixed spd

@@ -243,7 +243,7 @@ text(
   cex = 0.5,
   font = 2,
   adj = c(0, 0),
-  labels = paste0("(df = ", df_exp, ", AIC = ", aic_exp, ")")
+  labels = paste0("AIC = ", aic_exp)
 )
 
 box()
@@ -319,8 +319,10 @@ text(
   cex = 0.5,
   font = 2,
   adj = c(0, 0),
-  labels = paste0("(df = ", df_lin, ", AIC = ", aic_exp, ")")
+  labels = paste0("AIC = ", aic_lin)
 )
+
+
 box()
 par(mar = c(6, 1, 0, 1))
 plot(
@@ -358,7 +360,7 @@ text(
   cex = 0.5,
   font = 2,
   adj = c(0, 0),
-  labels = paste0("(df = ", df_log, ", AIC = ", aic_exp, ")")
+  labels = paste0("AIC = ", aic_log)
 )
 box()
 xticks <- seq(50000, 10000, -10000)

### 2021-01-16 analysis/paper/008-summed-probability-distributions.R
rearranged spd figure

@@ -203,8 +203,54 @@ layout(
   widths = 6,
   heights = c(1.8, 1.8, 1.8, 2.6)
 )
-# draw Exp plot
+
+# first plot: linear
 par(mar = c(0, 1, 1, 1)) #c(bottom, left, top, right)
+
+plot(
+  lin_test,
+  ylim = c(0, ymax),
+  xlim = c(50000, 10000),
+  drawaxes = FALSE
+)
+legend(
+  x = 49000,
+  ymax * 0.94,
+  legend = c(
+    "radiocarbon ages",
+    "95% MC envelope",
+    "positive deviation",
+    "negative deviation"
+  ),
+  col = c("black", "lightgrey", rgb(0.7, 0, 0, 0.2), rgb(0, 0, 0.7, 0.2)),
+  lwd = c(0.5, 5, 5, 5),
+  cex = 0.6,
+  bg = "white",
+  title = ""
+)
+text(
+  x = 48000,
+  ymax * 0.90,
+  labels = "Linear Model",
+  font = 2,
+  cex = 0.6,
+  adj = c(0, 0.7)
+)
+text(
+  x = 41000,
+  ymax * 0.90,
+  cex = 0.6,
+  font = 2,
+  adj = c(0, 0),
+  labels = paste0("AIC = ", round(aic_lin, digits = 0))
+)
+
+
+box()
+
+# Second plot: exponential
+par(mar = c(0, 1, 0, 1))
+
 plot(
   exp_test,
   ylim = c(0, ymax),
@@ -238,17 +284,68 @@ text(
 )
 # draw Exp plot: model stats
 text(
-  x = 48000,
-  ymax * 0.49,
-  cex = 0.5,
+  x = 41000,
+  ymax * 0.90,
+  cex = 0.6,
   font = 2,
   adj = c(0, 0),
-  labels = paste0("AIC = ", aic_exp)
+  labels = paste0("AIC = ", round(aic_exp, digits = 0))
+)
+
+
+
+box()
+
+# Third plot: logistic
+par(mar = c(0, 1, 0, 1))
+
+plot(
+  log_test,
+  ylim = c(0, ymax),
+  xlim = c(50000, 10000),
+  drawaxes = FALSE
+)
+
+legend(
+  x = 49000,
+  ymax * 0.94,
+  legend = c(
+    "radiocarbon ages",
+    "95% MC envelope",
[…]

### 2021-04-01 analysis/paper/paper.Rmd
Editing the paper

@@
 abstract: |
-  Human behavioural ecological models use stone artefacts as a source of information for understanding forager mobility.
-We use these models to examine stone artefacts from 22 sites in Korea to investigate mobility and site occupation during the Late Pleistocene.
-This study focuses on two questions: how did the introduction of stemmed points change forager' mobility and landscape use?
-And how did this new technology change the way people use habitation sites?
-We analyse artefact volumetric density, retouch frequency, composition of toolkits, artefact raw materials, environmental and demographic contexts.
+  Human behavioral ecological models, under the umbrella of cultural evolution, use stone artefacts as a source of information for understanding forager subsistence strategies.
+We use these models to examine stone artefacts from 22 sites in Korea to investigate mobility and site occupation patterns during the Late Pleistocene.
+This study focuses on two questions: what changes in foragers' landscape use were associated with the introduction of the new tools?
+And what changes in the way people use habitation sites and mobility were associated with the new technology?
+We conduct quantitative analyses including artefact volumetric density, retouch frequency, composition of toolkits, and artefact raw materials, and explore environmental and demographic contexts by applying paleoclimate simulation and summed probability distribution models.
 Quartz and side scrapers, in addition to cores and choppers, remain dominant in assemblages before and after the introduction of stemmed points throughout the Late Pleistocene.
@@
 The appearance of stemmed points (Figure \@ref(fig:stemmed-points)) in the Late Pleistocene (\~40-35 ka) in Korea is often assumed to have transformed forager lifeways because it reflected more specialized hunting practices [@seong2008tanged; @chang2013human].
-Stemmed points are often called as tanged points in the western hemisphere, but we use 'stemmed point' to distinguish from Bronze Age stone projectile points that have long been called 'tanged points' in Korea.
-The stemmed point is a projectile point made out of elongated flake or blade with slight retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft.
+Stemmed points are often called tanged points in the western hemisphere, but we use 'stemmed point' to distinguish from Bronze Age stone projectile points that have long been called 'tanged points' in Korea.
+The stemmed point is a projectile point made from an elongated flake or blade with slight retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft.
 Stemmed points are likely to have been projectile weapon tips because the stem of these tools is considered to represent hafting technology that joins the lithic to a wooden shaft.
@@
 
-Previous work on this technological change in Korea has focussed on issues of the transition from the Early to the Late Paleolithic and modern human dispersals [@norton2009evolution; @bae2010origin].
+Previous work on this technological change in Korea has focused on issues of the transition from the Early to the Late Paleolithic and modern human dispersals [@norton2009evolution; @bae2010origin].
 To our knowledge, the implications of the appearance of stemmed points for forager mobility and site use have not been explored in detail.
-Here, we attempt to identify patterns of mobility through analysis of the stone artefact assemblages using Human Behavioural Ecology (HBE) models.
+Here, we attempt to identify patterns of mobility and site occupation through analysis of the stone artefact assemblages using Human Behavioural Ecology (HBE) concepts.
 Drawing on established relationships between lithic technology and forager mobility [@shott1986settlement; @Hiscock_1994; @kuhn1994formal; @kuhn2004upper; @kuhn2015artifacts; @kuhn2016moves; @capriles2018mobility], we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that are part of a broader strategy of moving frequently and further, possibly as an adaptation to environmental or population changes.
@@
 In the Eastern hemisphere, major technological innovations during the Late Pleistocene are restricted to Northeast Asia, such as northern China, Japan and Korea.
-These innovations include blade and micro technology, high frequencies of retouched blade tools, several novel tools such as projectile points, stemmed points, end scrapers, burins, denticulates, etc, and using high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary].
+These innovations include blade technology, high frequencies of retouched blade tools, several novel tools such as projectile points, stemmed points, end scrapers, burins, denticulates, etc, and using high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary].
 Previous work has argued that the appearance of new stone artefact technologies in this region may be linked to modern human dispersals [@seong2008tanged; @bae2010origin; @bae2017origin].
@@
 In addition, more than 10,000 animal fossils and bone tools including a tool for fishing nets, an awl, and two needles were excavated [@pei2012shuidonggou; @gao2014discovery; @zhang2016bone].
-Stemmed points are not part of the Shuidonggou assemblages, but microblade technology is argued to have appeared in China after 29ka BP, related to changes in mobility associated with colder climates towards LGM [@yi2014significance].
 Shuidonggou shows that the blade component came from Siberia and/or Mongolia at 41 ka (SDG Locality1 and 9), advanced core and flake tools were likely to have been locally developed at 33 ka (2 and 8), and at 10.8 ka microblade technology appears but the origin is uncertain (12) [@yi2014significance; @li2019history].
+The microblade technology is argued to have appeared related to changes in mobility associated with colder climates towards LGM [@yi2014significance].
 
-Late Pleistocene technological innovations in Japan appeared in different regions of the archipelago around 38,000 years ago accompanied by remains of Homo sapiens found in Okinawa Island (c.36,000 cal BP) [@yamaoka2012use; @izuho201421].
-These new technologies include trap pits, cobble concentrations, hearths, charcoal concentrations, and toolkits including trapezoids, pointed-shaped backed blades, backed points, burins, end scrapers, side scrapers, wedges, beak-shaped tools, axes, edge-ground axes, hammerstones, cobble tools, and anvils [@bae2017origin; @izuho201421].
+The Japanese Early Upper Paleolithic (EUP) technologies appeared in different regions of the archipelago around 38,000 years ago accompanied by remains of Homo sapiens found in Okinawa Island (c.36,000 cal BP) [@yamaoka2012use; @izuho201421].
+These technologies include trap pits, cobble concentrations, hearths, charcoal concentrations, and toolkits including trapezoids, pointed-shaped backed blades, backed points, burins, end scrapers, side scrapers, wedges, beak-shaped tools, axes, edge-ground axes, hammerstones, cobble tools, and anvils [@bae2017origin; @izuho201421].
 These innovations have been interpreted as evidence of new foraging methods such as watercraft (marine transport of obsidian from the Kozu Island) and bow-and-arrow technology, driven by increased population and ecological changes [@yamaoka2012use; @bae2017origin; @nakazawa2018quaternary; @morisaki2019early].
@@
 Stemmed points appeared in both regions after the case of Korean Peninsula.
-The oldest stemmed points so far is from the Yonghodong site dated to 38.5 ka and made on elongated flakes [@seong2009emergence; @seong2015diversity; @bae2017origin].
+The oldest stemmed points so far are from the Yonghodong site dated to 38.5 ka and made on elongated flakes [@seong2009emergence; @seong2015diversity; @bae2017origin].
 After blades appeared at c.
-27 ka and were adopted to the stemmed points, the shape of points became more standardized with one or two ridges on the dorsal side and triangular cross section.
-Replacing flake to blade as a blank of the points led to an increase in quantity over time.
+27 ka and were used to make the stemmed points, the shape of points became more standardized with one or two ridges on the dorsal side and triangular cross section.
+Replacing flakes with blades as blanks of the stemmed points led to an increase in quantity of the points over time.
 In addition to the stemmed points, flake tools became more complex, with a greater diversity in both type and size along with continuously used core tools during the Late Pleistocene [@bae2010origin; @lee2013current; @Lee_2016].
-There are several attempts to explain the technological transition and they can be summarized into two competing models: 'heterogenic' migration [@bae2010origin] or *in situ* evolution [@seong2009emergence].
-The migration model argues that the new blade industry including stemmed points, and the earlier coarse flake tradition including large cores, polyhedrons, choppers, and even handaxes, came from different origins as the result of continuing influx of model human migration from two routes.
-Bae (2010) explains that blade and stemmed points arrived in the Korean peninsula from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP with the earliest evidence of blade in those regions.
-He assumes that core and flake industry come from southern China based on the existence of similar assemblages in both Korea and southern China in addition to a genetic analysis of the Y-chromosome of the modern population [@bae2013early; @lee2013current].
+There are several attempts to explain this technological transition and they can be summarized into two competing models: 'heterogenic' migration [@bae2010origin] or *in situ* evolution [@seong2009emergence].
+The migration model argues that the new blade industry including stemmed points, and the earlier coarse flake tradition including large cores, polyhedrons, choppers, and even handaxes, came from different origins as the result of continuing influx of modern human migration from two routes.
+Bae (2010) explains that blade and stemmed points arrived in the Korean peninsula from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP with the earliest evidence of blades in those regions.
+He assumes that core and flake industry come from southern China based on the existence of similar assemblages in both Korea and southern China.
+In addition, the genetic studies of two “Asian-specific” Y-chromosome haplogroups (O3-M122, D-M174), mtDNA and autosomal SNPs, and other laboratories verify that modern humans initially arrive in southern China and the population migrated to northward [@bae2013early; @lee2013current].
 The '*in situ*' model claims that stemmed points and other Late Paleolithic assemblages including blade industries autonomously emerged in the south of the Korean peninsula with no external influence [@seong2008tanged].
-This claim is supported by the earliest appearance of stemmed points among Northeast Asia, such as Hwadae-ri, Hopyeong-dong, and Yonghodong sites dated to 40-35ka BP [@seong2009emergence].
-Seong(2009) uses the increased blade-to-flake ratios on lithic assemblages to support the *in situ* model based on the premise that the blade industry represents new technology while flakes indicate a continuing local one.\
-
-<!-- need to put two images,  one for stemmed points, one for overall lithic assemblages -->
+This claim is supported by Korea having the earliest appearance of stemmed points in Northeast Asia, at sites such as Hwadaeri, Hopyeongdong, and Yonghodong dated to 40-35ka BP [@seong2009emergence].
+Seong(2009) uses the increased blade-to-flake ratios on lithic assemblages to support the *in situ* model based on the premise that the blade industry represents new technology while flakes indicate a continuing local one.
 
@@
 
-Our brief review of Late Pleistocene technological innovations in Northeast Asia shows that previous work has largely focussed on timing, location, origins and description of these new tools, but modelling the land-use behaviours that generated these assemblages has rarely been undertaken.
+Our brief review of Late Pleistocene technological innovations in Northeast Asia shows that previous work has largely focussed on timing, location, origins and description of these new tools, but modelling the land-use behaviours associated with these assemblages has rarely been undertaken [@seong2008tanged; @seong2009emergence; @bae2010origin; @lee2013current; @Lee_2016].
 We focus on the appearance of stemmed points in Korea because these are a new technology that represents a new hunting method.
@@
 We draw on behavioural ecological theory to model the effects of this technological change.
-Behavioural ecology theory offers structured and testable models based on optimality assumptions that stand on economic rationality and environmental knowledge [@winterhalder1992evolutionary; @prentiss2019handbook].
-The patch choice model and marginal value theorem predict that forager mobility and travel times will increase when resource patch productivity decreases [@smith1991inujjuamiunt; @smith1983anthropological; @bettinger1997evolutionary; @wolverton2015patch; @llano2015optimal].
-We can also model individual artefact types as a kind of resource patch.
-Foragers may find a more costly and complicated technology (such as stemmed points relative to flakes) optimal if they have a long cumulative time in use, if they are maintainable and reusable in a landscape where lithic raw material supply is uncertain.
+Behavioural ecology theory offers structured models and testable hypotheses based on optimality assumptions that stand on economic rationality and environmental knowledge [@winterhalder1992evolutionary; @prentiss2019handbook].
+The patch choice model built around the marginal value theorem predicts that forager mobility between patches and inter-patch travel times will increase when resource patch productivity decreases [@smith1991inujjuamiunt; @smith1983anthropological; @bettinger1997evolutionary; @wolverton2015patch; @llano2015optimal].
+We can model individual artefact types as a kind of resource patch.
+Foragers may find a more costly and complicated technology (such as stemmed points relative to flakes) optimal if they have a long cumulative time in use, if they are maintainable and reusable in a landscape where lithic raw material supply is uncertain [@torrence1983time; @bleed1986optimal].
 That is, they may move to a patch where they can stay longer (i.e.
@@
 We ask two questions to investigate the period around the introduction of stemmed points in Korea.
-First, how did the introduction of the new tools change foragers' landscape use?
+First, what changes in foragers' landscape use were associated with the introduction of the new tools?
 Specifically, was the use of stemmed points associated with occupation of marginal or extreme environments?
@@
 
-Second, we ask how the new technology changed the way people use habitation sites and mobility.
-After the appearance of these new tools, did people tend to stay in one location for longer or shorter periods, perhaps for specific purposes?
+Second, we ask what changes in the way people use habitation sites and mobility were associated with the new technology?
+After the appearance of these new tools, did people tend to be more or less mobile?
 The stone artefact assemblage from a site can inform us how long individuals or groups stayed and give insights into their activities [@binford1979organization; @kelly1995foraging; @kuhn2016moves; @holdaway2019surface].
-For example, stone artifact density and the frequency of retouched pieces (scaled to the volume of excavated sediment) can be used as proxies to represent occupation patterns [@barton2011modeling; @clark2017lithics].
+For example, stone artefact density and the frequency of retouched pieces (scaled to the volume of excavated sediment) can be used as proxies to represent occupation patterns [@barton2011modeling; @clark2017lithics].
 Assemblages with low density but a high proportion of retouched and backed pieces may indicate the remains of 'short-term camp', which is a site for small and ephemeral overnight camp or limited activity station.
-On the other hand, the combination of high density and low retouched pieces represent logistically organized basecamp, which is a site with greater residential stability, long site occupation and occupied by larger groups [@clark2017lithics; @clark2019landscapes].
+On the other hand, the combination of high density and low retouched pieces might be expected at logistically organized basecamp, which is a site with greater residential stability, long site occupation and occupied by larger groups [@clark2017lithics; @clark2019landscapes].
 
@@
 After the first excavation of a Paleolithic site in Korea at Seokjang-ri site in the 1960s, more than 200 Paleolithic sites have been discovered in South Korea [@lee2019were].
-We selected sites dated to before and after the transition period (40-35 ka) to analyze the process of technological change.
-This resulted in a sample of analysed 33 assemblages from 22 sites spanning 49-24 ka (Figure \@ref(fig:korea-map)).
+We selected sites dated to before and after the transition period (40-35 ka) to analyse the process of technological change.
+This resulted in a sample of 33 assemblages from 22 sites spanning 49-24 ka (Figure \@ref(fig:korea-map)).
 We identified multiple assemblages in a site where culturally sterile deposits separated artefact-bearing deposits, or where stratigraphic units could be identified by major differences in the texture and composition of the sedimentary deposit containing the stone artefacts.
-For example, the Hwadae-ri site has three cultural layers.
+For example, the Hwadaeri site has three cultural layers.
 The lowest horizon at the bottom, dated to 39,000±1400 BP by OSL, contains coarse flake tools made of vein quartz and quartzite.
@@
 
-The 33 assemblages consist of cores, blade cores, blades, flakes, debris, hammers, choppers, plans, polyhedrals, side scrapers, end scrapers, notches, cleavers, stemmed points, awls, denticulates, burins, handaxes, blanks, knives, handadzes, peaks, flatters, flakers etc.
+The 33 assemblages consist of (non-blade) cores, blade cores, blades, flakes, debris, hammers, choppers, planes, polyhedrals, side scrapers, end scrapers, notches, cleavers, stemmed points, awls, denticulates, burins, handaxes, blanks, knives, handadzes, peaks, flatters, flakers etc.
 We excluded some artefacts from our sample because of uncertainty about their typology.
@@
 The retouch frequency is the number of retouched tools divided by the total number of artefacts in an assemblage.
-This value represents mobility and duration of site occupation based on the concept that tool reduction is an tactic to extend the use-life of a tool by generating sharp, usable edges while minimizing the cost of transporting tools or bulky raw materials by reducing the total load weight [@Kuhn_1990; @buvit2014late; @clark2019landscapes].
+This value represents mobility and duration of site occupation based on the assumption that tool reduction is a tactic to extend the use-life of a tool by generating sharp, usable edges while minimizing the cost of transporting tools or bulky raw materials by reducing the total load weight [@Kuhn_1990; @buvit2014late; @clark2019landscapes].
 In other words, producing and carrying highly retouched tools will increase portability and efficiency by decreasing the size of both individual tools and the assemblages so that hunter-gatherers could move often or further, or both [@kuhn1991unpacking; @kuhn1994formal; @kuhn2004upper; @andrefsky1994raw; @davies2018modeling].
-Combining these two proxies, the artefact volumetric density and the retouch frequency, we expect to determine whether or not the hunter-gatherers were residentially mobile or logistically organized.
+Combining these two proxies, the artefact volumetric density and the retouch frequency, we infer whether or not the hunter-gatherers were residentially mobile or logistically organized.
 If the assemblage in a site has a high artefact density and small proportion of retouch, we interpret it as a more expedient assemblage that represents "base camps" or "residences", while less density with higher proportions of retouch shows a more curated assemblage and indicates either residential mobility or certain task groups away from residential sites [@riel2004late; @clark2017lithics].
@@
 
-Studying the raw material consumption patterns is based on an assumption that stone raw materials were transported and consumed to optimize mobility and minimize risk of not having artefacts when they were needed [@andrefsky1994raw; @brantingham2000raw; @brantingham2003neutral; @yue2020lithic].
+Our study of raw material consumption patterns assumes that tool stone was transported and consumed to optimize mobility and minimize risk of not having artefacts when they were needed [@andrefsky1994raw; @brantingham2000raw; @brantingham2003neutral; @yue2020lithic].
 The representation of raw materials in an assemblage can be an indicator of site occupation patterns.
@@
 HBE models require environmental proxies to build hypotheses of behavioural change influenced by the surrounding environment.
-However, there are few paleoenvironmental proxies in Korean relevant to our target period.
-For example, geoarchaeological observations suggest that environment of the period was "overall cooler and drier during MIS 3" [@Han_2008; @seong2008tanged; @Choi_2011; @chang2013human; @im2015study; @bak2017late].
+However, there are few paleoenvironmental proxies in Korea relevant to our target period.
+For example, geoarchaeological observations simply suggest that environment of the period was overall cooler and drier during MIS 3 [@Han_2008; @seong2008tanged; @Choi_2011; @chang2013human; @im2015study; @bak2017late].
 To supplement the limited availability of high-resolution local proxies, we used climate information from the simulated data set of paleoclimate including global monthly temperature, covering the last 120,000 years [@beyer2020high].
-We applied information of site location including elevation, latitude and longitude to the simulated data set to get the annual temperature of our research period and area.
+We applied information of site location (all open-sites) including elevation, latitude and longitude to the simulated data set to get the annual temperature of our research period and area.
 Besides the climate analysis, we analysed site elevation as a proxy to represent different environments at local scales, to identify how these impact mobility and site occupation.
-We mainly focused on the elevation of the sites before and after the appearance of new technology and its relationship with the stemmed points.
+We mainly focused on the elevation pattern of the sites before and after the appearance of new technology and its relationship with the stemmed points.
 
 We used radiocarbon dates as a reference for population level.
-Summed probability distributions (SPD) of radiocarbon dates are often used to understand temporal trends in past human populations [@rick1987dates; @bamforth2012radiocarbon; @shennan2013regional; @timpson2014reconstructing; @contreras2014summed; @riris2019widespread].
-Despite limitations of the validity of SPD [@williams2012use; @bamforth2012radiocarbon], including the lack of excavated sites in certain periods, sample size, and calibration effects, this method may be helpful as a proxy for a first approximation of population fluctuations.
+Summed probability distributions (SPD) of radiocarbon dates are increasingly used to infer temporal trends in past human populations [@rick1987dates; @bamforth2012radiocarbon; @shennan2013regional; @timpson2014reconstructing; @contreras2014summed; @riris2019widespread].
+Despite limitations of the validity of SPD [@williams2012use; @bamforth2012radiocarbon], including the lack of excavated sites in certain periods, sample size, and calibration effects, this method may allow a first approximation of population fluctuations.
 Although this is often used with samples of thousands of ages, this method can also be useful for small sample sizes.
@@
 There is a debate about whether past population growth has played a role in technological change [@shennan2001demography; @kline2010population].
-@shennan2009pattern observed a correlation between population decrease and technological simplification in their study of lithic arrowheads from the southern Scandinavian Mesolithic.
-They found that the lithics became less complex, and point shape changed as the population level dropped.
+@shennan2009pattern observed a correlation between population decrease and technological simplification in his study of lithic arrowheads from the southern Scandinavian Mesolithic.
+He found that the lithics became less complex, and point shape changed as the population level dropped.
 On the other hand, @collard2013population argue that regional archaeological data do not support the population-driven models [@buchanan2016drivers].
 For example, @buchanan2016drivers concluded that changes in point types from \>13,000 years ago to 400 years ago in Texas were related to environmental risk, and not dependent on population size.
-We explored SPD to test for correlations between population, environmental risk and technological change.
+We explored SPD to test for correlations between population, environmental proxies and technological change.
 We used the R package 'rcarbon' [@cremaanalysing] to perform statistical analyses and conduct model testing using Monte Carlo methods [@timpson2014reconstructing; @crema2016summed; @palmisano2020holocene].
@@
 In addition to side scrapers, cores also exist in all assemblages often as a high proportion.
-The proportion of choppers decreased towards in more recent assemblages.
+The proportion of choppers decreased in more recent assemblages.
 Blades first appeared in Gawolri-A at 43 ka and then often found in assemblages thereafter.
@@
 Through MIS 3, the temperature gradually decreased until the Last Glacial Maximum (LGM, 26.5-20 ka) [@clark2009last].
-The first appearance of stemmed points occurs in the middle of the decreasing MAT trend in MIS 3, at 40-35ka.
+The first appearance of stemmed points occurred in the middle of the decreasing MAT trend in MIS 3, at 40-35ka.
 The MAT of MIS 2, including LGM, was relatively stable.
@@
 Figure 7D shows a negative relationship between temperature and elevation (p = 0.073).
-For example, Gihwa-ri Cave site, with one of the lowest MAT distributions, is located in the highest elevation around 300m above sea level.
-Jeungsan-ri is located at a lower elevation and has one of the highest MAT distributions.
+For example, Gihwari Cave site, with one of the lowest MAT distributions, is located in the highest elevation around 300m above sea level.
+Jeungsanri is located at a lower elevation and has one of the highest MAT distributions.
 
@@
 
-Our study focused on two questions: how did the introduction of the new tools change foragers' mobility and landscape use; and how did the new technology change the way people use habitation sites?
+Our study focused on two questions: what changes in foragers' landscape use were associated with the introduction of the new tools?; what changes in the way people use habitation sites and mobility were associated with the new technology?
 Overall, our results show that assemblages with stemmed points tend to have higher artefact densities and lower proportions of retouched pieces.
-Quartz and side scrapers, in addition to cores and choppers, remain dominant in assemblages before and after the introduction of stemmed points.
+Quartz and side scrapers, in addition to cores and choppers, remain dominant in assemblages throughout the late Pleistocene.
 The environmental context of this technological innovation was a gradual decrease in temperature into the LGM.
+And the population increase followed after the appearance of stemmed points.
 
@@
 
-We examined the artefact volumetric density and retouch frequency to understand the difference in mobility strategies and site occupation patterns before and after the introduction of stemmed points.
+We examined the artefact volumetric density and retouch frequency to understand the difference in mobility strategies and site occupation patterns associated with the introduction of stemmed points.
 The analyses are based on the premise that highly mobile groups would carry lightweight toolkits with more retouched and easily replaceable composite tools, and leave a smaller amount of artefacts on site as a result of short-term occupations.
-These curated technologies invest time and effort in stone artefact manufacturing and maintenance to optimize travel costs for groups using high mobility or special purpose forays.
-On the other hand, less mobile groups would produce fewer retouched tools and more simple tools, and leave higher densities of artefacts generated from long-term occupations.
+These curated technologies invest time and effort in stone artefact manufacturing and maintenance to minimize travel costs of carrying the tools for groups using high mobility or special purpose forays.
+On the other hand, less mobile groups would produce fewer retouched tools and leave higher densities of artefacts generated from long-term occupations.
 These expedient technologies are more typical for groups with greater residential stability who can make, use, abandon, and remake their tools frequently in the same location [@binford1979organization; @torrence1989time; @kuhn1994formal; @meignen2006middle; @kuhn2014mousterian; @vaquero2018searching; @clark2019landscapes].
@@
 An important quality of composite tools is that the user can easily replace damaged parts, contributing to an increased maintainability of the tool [@kuhn2015artifacts; @cardillo2010some].
-Combining the stone projective with a wooden shaft further materials enhances functionality of the tool such as improving penetration by increasing weight [@browne1940projectile].
+Combining the stone projective with a wooden shaft further materials enhances functionality of the tool such as improving penetration by increasing weight and sharpness [@browne1940projectile].
 We expected that the functional advantage of the new composite tool might have allowed foragers to be more active in less productive landscapes.
-Previous studies simply describe the environment of the research period as "overall cooler and drier during MIS 3" but lack detail due to a lack of direct paleoenvironmental proxies in Korea [@Han_2008; @seong2008tanged; @Choi_2011; @chang2013human; @im2015study; @bak2017late].
+Previous studies simply describe the environment of the research period as overall cooler and drier during MIS 3 but lack detail due to a lack of direct paleoenvironmental proxies in Korea [@Han_2008; @seong2008tanged; @Choi_2011; @chang2013human; @im2015study; @bak2017late].
 To explore environmental and demographic context related to the technological transition, we examined the distribution of site elevations, and simulated MAT during the Late Pleistocene.
-The results show that groups with stemmed points occupied higher elevations, where the MAT was lower.
+The results show that groups with stemmed points reached higher elevations, where the MAT was lower.
 This supports our hypothesis that stemmed points supported occupation of more marginal habitats.
@@
 Bae claims that stemmed points are the result of the continuous influx of modern humans from a Siberian migration [@bae2010origin].
-Our results are consistent with this scenario, showing a gradual increase in population after the appearance of stemmed points which may reflect a continuous influx of migrants over a long period.
+Our results are partly consistent with this scenario, showing a gradual increase in population after the appearance of stemmed points.
 
@@
 To understand the change in mobility and site occupation related to the appearance of new technologies during the late Pleistocene in Korea, we examined stone artefacts from 22 sites dated to 49-24 ka.
-The results show that the forager groups with stemmed points tended to be located at higher elevations, which means the use of stemmed points was associated with occupation of marginal or extreme environments.
+The results show that the forager groups with stemmed points tended to be located at higher elevations.
+The temperature gradually decreased during the technological transition and after the appearance of stemmed points.
+Overall, the result could indicate the use of stemmed points was associated with occupation of marginal or extreme environments.
 In addition, the groups with stemmed points had expedient technologies which reflect residential and less mobile land use patterns.
-The temperature gradually decreased during the technological transition and after the appearance of stemmed points, population gradually increased.
 
-Our findings are limited by the relatively small number of sites and radiocarbon ages currently available.
-The absence of detailed stratigraphic data limit the chronological resolution for the assemblages in our sample.
+Our findings are limited by the relatively small number of sites and radiocarbon ages currently available The absence of detailed stratigraphic data limit the chronological resolution for the assemblages in our sample.
 Furthermore, we currently lack functional data on individual artefact assemblages to support our conclusions about site use.

### 2021-04-12 analysis/paper/paper.Rmd
Editing text in rmd file

@@
 abstract: |
-  Human behavioral ecological models, under the umbrella of cultural evolution, use stone artefacts as a source of information for understanding forager subsistence strategies.
-We use these models to examine stone artefacts from 22 sites in Korea to investigate mobility and site occupation patterns during the Late Pleistocene.
-This study focuses on two questions: what changes in foragers' landscape use were associated with the introduction of the new tools?
+  We use models from human behavioral ecology to examine stone artefacts from 22 sites in Korea to investigate mobility and site occupation patterns during the Late Pleistocene when new tools, such as stemmed points and blades, appeared in the archaeological record.
+We focus on two questions: what changes in foragers’ landscape use were associated with the introduction of new tools?
 And what changes in the way people use habitation sites and mobility were associated with the new technology?
-We conduct quantitative analyses including artefact volumetric density, retouch frequency, composition of toolkits, and artefact raw materials, and explore environmental and demographic contexts by applying paleoclimate simulation and summed probability distribution models.
-Quartz and side scrapers, in addition to cores and choppers, remain dominant in assemblages before and after the introduction of stemmed points throughout the Late Pleistocene.
+To answer these questions we present quantitative analyses of artefact volumetric density, retouch frequency, composition of toolkits, and artefact raw materials.
+We explore the environmental and demographic contexts by applying paleoclimate simulations and summed probability distribution models.
+We find that quartz and side scrapers, in addition to cores and choppers, remain dominant in assemblages before and after the introduction of stemmed points throughout the Late Pleistocene.
 Our results show that forager groups using stemmed points were associated with occupation of marginal or extreme environments.
-In addition, groups with stemmed points were associated with expedient technologies indicating residential and less mobile behaviours.
+In addition, groups with stemmed points were associated with expedient technologies, indicating residential and less mobile behaviours.
 The environmental context of this technological innovation was a gradual decrease in temperature into the LGM.
@@
 
-The appearance of stemmed points (Figure \@ref(fig:stemmed-points)) in the Late Pleistocene (\~40-35 ka) in Korea is often assumed to have transformed forager lifeways because it reflected more specialized hunting practices [@seong2008tanged; @chang2013human].
-Stemmed points are often called tanged points in the western hemisphere, but we use 'stemmed point' to distinguish from Bronze Age stone projectile points that have long been called 'tanged points' in Korea.
+The appearance of stemmed points (Figure \@ref(fig:stemmed-points)) in the Late Pleistocene (ca.
+42-35 ka) in Korea is often assumed to have transformed forager lifeways because it was thought to have indicated more specialized hunting practices [@seong2008tanged; @chang2013human].
 The stemmed point is a projectile point made from an elongated flake or blade with slight retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft.
-Stemmed points are likely to have been projectile weapon tips because the stem of these tools is considered to represent hafting technology that joins the lithic to a wooden shaft.
-If they were projectile weapon tips, stemmed points represent the first appearance of hafting technology in Korea.
-Most Korean Paleolithic sites do not preserve organic remains, so we lack direct fauna evidence on hunting and subsistence strategies.
-Here we explore questions about the implications of the appearance of this new technology using a behavioural ecological framework and stone artefacts.
-We explore changes in the way people occupied the landscape, the way people used sites, and impact of social and ecological environment.
-We examined stone artefacts from 22 sites in South Korea to investigate the changes in human behaviour associated with the appearance stemmed points, and the blade industry that followed.
+These types of artefacts are often called tanged points in the western hemisphere [@serwatka20162d; @ivanovaite2020all], but we use ‘stemmed point’ here to distinguish from Bronze Age stone projectile points that have long been called ‘tanged points’ in Korea.
+Stemmed points are likely to have been projectile weapon tips because the stem of these tools is considered to represent a hafting projection used for joining the lithic to a wooden shaft.
+If these artefacts were projectile weapon tips, then stemmed points represent the first appearance of hafting technology in Korea.
+Most Korean Paleolithic sites do not preserve organic remains, so we lack direct fauna evidence on hunting and subsistence strategies, and stone artefacts are our only window into these behaviours.
+Here we use a behavioural ecological framework to explore questions about the implications of the appearance of this new technology.
+We used data from 22 sites in South Korea to explore changes in the way people occupied the landscape, the way people used sites, and the of the social and ecological contexts associated with the appearance stemmed points, and the blade industry that followed.
 
@@
 Here, we attempt to identify patterns of mobility and site occupation through analysis of the stone artefact assemblages using Human Behavioural Ecology (HBE) concepts.
-Drawing on established relationships between lithic technology and forager mobility [@shott1986settlement; @Hiscock_1994; @kuhn1994formal; @kuhn2004upper; @kuhn2015artifacts; @kuhn2016moves; @capriles2018mobility], we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that are part of a broader strategy of moving frequently and further, possibly as an adaptation to environmental or population changes.
+Drawing on established relationships between lithic technology and forager mobility [@shott1986settlement; @Hiscock_1994; @kuhn1994formal; @kuhn2004upper; @kuhn2015artifacts; @kuhn2016moves; @capriles2018mobility], we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that were part of a broader strategy of moving frequently and further, possibly as an adaptation to environmental or population changes.
 
-# Background {-}
+# Late Pleistocene Lithic Technologies in East Asia {-}
 
-In the Eastern hemisphere, major technological innovations during the Late Pleistocene are restricted to Northeast Asia, such as northern China, Japan and Korea.
-These innovations include blade technology, high frequencies of retouched blade tools, several novel tools such as projectile points, stemmed points, end scrapers, burins, denticulates, etc, and using high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary].
+In the Eastern hemisphere, the most striking technological innovations during the Late Pleistocene are found in Northeast Asia, such as northern China, Japan and Korea.
+These innovations include blade technology, high frequencies of retouched blade tools, several novel tools such as projectile points, stemmed points, end scrapers, burins, denticulates, etc, and frequent use of high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary].
 Previous work has argued that the appearance of new stone artefact technologies in this region may be linked to modern human dispersals [@seong2008tanged; @bae2010origin; @bae2017origin].
@@
 The Shuidonggou site in northern China is an important example of the transition from cobble to blade tool industries in the Late Pleistocene of Northeast Asia [@Brantingham_Perreault_2010; @Gao_Zhang_Yang_Shen_Wu_2010; @pei2012shuidonggou].
-Shuidonggou Locality SDG 2, which dates to around c.
-32,000 years, contains fauna specimens, ostrich eggshell beads, and stone artefacts including blades, retouched flakes, cores, and debris along with well-preserved hearths.
-In Locality 9 dated to c.
-29,000 years, there is a small scatter of stone artefacts consisting of blades, Levallois flakes, cores and other retouched flakes.
-The recent excavation of Locality 12 shows that its archaeological context, dated to 13,078--13,296 cal BP, includes ground stone and more than 30,000 microlithics made out of a variety of raw materials such as fine-grained and siliceous rocks.
+Shuidonggou Locality SDG 2, which dates to around ca.
+32ka, contains fauna specimens, ostrich eggshell beads, and stone artefacts including blades, retouched flakes, cores, and debris along with well-preserved hearths.
+In Locality 9 dated to ca.
+29ka, there is a small scatter of stone artefacts consisting of blades, Levallois flakes, cores and other retouched flakes.
+Excavations Locality 12 shows that its archaeological context, dated to 13,078--13,296 cal BP, includes ground stone and more than 30,000 microlithics made out of a variety of raw materials such as fine-grained and siliceous rocks.
 In addition, more than 10,000 animal fossils and bone tools including a tool for fishing nets, an awl, and two needles were excavated [@pei2012shuidonggou; @gao2014discovery; @zhang2016bone].
-Shuidonggou shows that the blade component came from Siberia and/or Mongolia at 41 ka (SDG Locality1 and 9), advanced core and flake tools were likely to have been locally developed at 33 ka (2 and 8), and at 10.8 ka microblade technology appears but the origin is uncertain (12) [@yi2014significance; @li2019history].
-The microblade technology is argued to have appeared related to changes in mobility associated with colder climates towards LGM [@yi2014significance].
+Shuidonggou shows that although stemmed points are not present at all, a blade technology came from Siberia and/or Mongolia at 41 ka (SDG Locality1 and 9).
+Advanced core and flake tools were likely to have been locally developed at 33 ka (SDG Locality 2 and 8), and at 10.8 ka microblade technology appears, but the origin is uncertain (SDG Locality 12) [@yi2014significance; @li2019history].
+The microblade technology is argued to have appeared related to changes in mobility associated with colder climates towards the Last Glacial Maximum (LGM) [@yi2014significance].
 
-The Japanese Early Upper Paleolithic (EUP) technologies appeared in different regions of the archipelago around 38,000 years ago accompanied by remains of Homo sapiens found in Okinawa Island (c.36,000 cal BP) [@yamaoka2012use; @izuho201421].
+In Japan, Early Upper Paleolithic (EUP) technologies appeared in different regions of the archipelago around 38ka, accompanied by remains of Homo sapiens found in Okinawa Island (ca.
+36ka) [@yamaoka2012use; @izuho201421].
 These technologies include trap pits, cobble concentrations, hearths, charcoal concentrations, and toolkits including trapezoids, pointed-shaped backed blades, backed points, burins, end scrapers, side scrapers, wedges, beak-shaped tools, axes, edge-ground axes, hammerstones, cobble tools, and anvils [@bae2017origin; @izuho201421].
@@
 Hafted trapezoids were likely multifunctional tools adapted to the specific environmental settings of the different Japanese Islands [@ono2002radiocaron; @yamaoka2012use].
-Ishinomoto 8-ku is one of the earliest sites dated to 39,690--34,790 cal BP, located in Takuma upland of Kumamoto Prefecture.
+An important example is Ishinomoto 8-ku, which is one of the earliest sites, dated to 39,690--34,790 cal BP, located in Takuma upland of Kumamoto Prefecture.
 A total of 500 stone artefacts were discovered including trapezoids, side scrapers, flakes, flake cores, edge-ground axes, and cobble tools.
-The main lithic raw materials are chert and andesite [@izuho201421].
-Stemmed points appear much later, dated to between about 15,500 and 13,800 cal BP [@ono2002radiocaron; @tsutsumi2007dynamics].
+The main lithic raw materials are Chert and Andesite [@izuho201421].
+Stemmed points appear in Japan much later than in Korea, dated to between about 15,500 and 13,800 cal BP [@ono2002radiocaron; @tsutsumi2007dynamics].
 
-What makes the Korean Late Pleistocene technological transition distinctive from what we see at Shuidonggou and in Japan is that the earliest signs of new technologies in Korean assemblages are stemmed points, followed by blade technologies.
-Stemmed points appeared in both regions after the case of Korean Peninsula.
-The oldest stemmed points so far are from the Yonghodong site dated to 38.5 ka and made on elongated flakes [@seong2009emergence; @seong2015diversity; @bae2017origin].
-After blades appeared at c.
-27 ka and were used to make the stemmed points, the shape of points became more standardized with one or two ridges on the dorsal side and triangular cross section.
+What makes the Korean Late Pleistocene technological transition distinctive from what we see at Shuidonggou and in Japan?
+The key difference is that the earliest signs of new technologies in Korean assemblages are stemmed points, followed by blade technologies.
+Stemmed points appeared in northern China and Japan after they appear in the Korean Peninsula.
+The oldest stemmed points so far in Korea are from the Bonggok site dated to ca.
+41.5 ka, and made on elongated flakes [@seong2009emergence; @seong2015diversity; @bae2017origin].
+After blades became widespread in Korea at ca.
+27 ka, and were used to make the stemmed points, the shape of points became more standardized with one or two ridges on the dorsal side and triangular cross section.
 Replacing flakes with blades as blanks of the stemmed points led to an increase in quantity of the points over time.
 In addition to the stemmed points, flake tools became more complex, with a greater diversity in both type and size along with continuously used core tools during the Late Pleistocene [@bae2010origin; @lee2013current; @Lee_2016].
-There are several attempts to explain this technological transition and they can be summarized into two competing models: 'heterogenic' migration [@bae2010origin] or *in situ* evolution [@seong2009emergence].
+
+There have been several attempts to explain this technological transition and they can be summarized into two competing models: 'heterogenic' migration [@bae2010origin] or *in situ* evolution [@seong2009emergence].
 The migration model argues that the new blade industry including stemmed points, and the earlier coarse flake tradition including large cores, polyhedrons, choppers, and even handaxes, came from different origins as the result of continuing influx of modern human migration from two routes.
-Bae (2010) explains that blade and stemmed points arrived in the Korean peninsula from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP with the earliest evidence of blades in those regions.
+Bae (2010) explains that blade and stemmed points arrived in the Korean peninsula from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka with the earliest evidence of blades in those regions.
 He assumes that core and flake industry come from southern China based on the existence of similar assemblages in both Korea and southern China.
-In addition, the genetic studies of two “Asian-specific” Y-chromosome haplogroups (O3-M122, D-M174), mtDNA and autosomal SNPs, and other laboratories verify that modern humans initially arrive in southern China and the population migrated to northward [@bae2013early; @lee2013current].
-The '*in situ*' model claims that stemmed points and other Late Paleolithic assemblages including blade industries autonomously emerged in the south of the Korean peninsula with no external influence [@seong2008tanged].
-This claim is supported by Korea having the earliest appearance of stemmed points in Northeast Asia, at sites such as Hwadaeri, Hopyeongdong, and Yonghodong dated to 40-35ka BP [@seong2009emergence].
-Seong(2009) uses the increased blade-to-flake ratios on lithic assemblages to support the *in situ* model based on the premise that the blade industry represents new technology while flakes indicate a continuing local one.
+In addition, the genetic studies of two “Asian-specific” Y-chromosome haplogroups (O3-M122, D-M174), mtDNA and autosomal SNPs, and other analyses verify that modern humans initially arrive in southern China and the population migrated to northward [@bae2013early; @lee2013current].
+
+The alternative model, which we call the '*in situ*' model claims that stemmed points and other Late Paleolithic assemblages including blade industries autonomously emerged in the south of the Korean peninsula, with no apparent external influence [@seong2008tanged].
+This claim is supported by Korea having the earliest appearance of stemmed points in Northeast Asia, at sites such as Bonggok, Hwadaeri, Hopyeongdong, and Yonghodong dated to approximately 42-35ka BP [@seong2009emergence].
+Seong(2009) presents the increased blade-to-flake ratios of lithic assemblages after the appearance of stemmed points to support the *in situ* model, based on the premise that the blade industry represents a new exogenous technology while flakes indicate a continuing local one.
 
@@
 
-Our brief review of Late Pleistocene technological innovations in Northeast Asia shows that previous work has largely focussed on timing, location, origins and description of these new tools, but modelling the land-use behaviours associated with these assemblages has rarely been undertaken [@seong2008tanged; @seong2009emergence; @bae2010origin; @lee2013current; @Lee_2016].
-We focus on the appearance of stemmed points in Korea because these are a new technology that represents a new hunting method.
+Our brief review of Late Pleistocene technological innovations in Northeast Asia shows that previous work has largely focussed on timing, location, origins and description of these new tools [@seong2008tanged; @seong2009emergence; @bae2010origin; @lee2013current; @Lee_2016], but modelling the land-use behaviours associated with these assemblages has rarely been undertaken.
+We focus on the appearance of stemmed points in Korea because these are a new technology that likely represented new hunting behaviours.
 This is because the hafting implies throwing techniques that can reach long-distance targets [@Keeley_1982; @kuhn2015artifacts].
 We draw on behavioural ecological theory to model the effects of this technological change.
-Behavioural ecology theory offers structured models and testable hypotheses based on optimality assumptions that stand on economic rationality and environmental knowledge [@winterhalder1992evolutionary; @prentiss2019handbook].
-The patch choice model built around the marginal value theorem predicts that forager mobility between patches and inter-patch travel times will increase when resource patch productivity decreases [@smith1991inujjuamiunt; @smith1983anthropological; @bettinger1997evolutionary; @wolverton2015patch; @llano2015optimal].
-We can model individual artefact types as a kind of resource patch.
-Foragers may find a more costly and complicated technology (such as stemmed points relative to flakes) optimal if they have a long cumulative time in use, if they are maintainable and reusable in a landscape where lithic raw material supply is uncertain [@torrence1983time; @bleed1986optimal].
+Behavioural ecology theory offers structured models and testable hypotheses based on optimality assumptions derived from principles of economic rationality and environmental adaptation [@winterhalder1992evolutionary; @prentiss2019handbook].
+Specifically, we use the patch choice model, which is built around the marginal value theorem, to predict that forager mobility between patches and inter-patch travel times will increase when resource patch productivity decreases [@smith1991inujjuamiunt; @smith1983anthropological; @bettinger1997evolutionary; @wolverton2015patch; @llano2015optimal].
+We model individual artefact types as a kind of resource patch.
+Foragers may find a more costly and complicated technology (such as stemmed points, relative to flakes) optimal if they have a long cumulative time in use, if they are maintainable and reusable in a landscape where lithic raw material supply is uncertain [@torrence1983time; @bleed1986optimal].
 That is, they may move to a patch where they can stay longer (i.e.
@@
 We ask two questions to investigate the period around the introduction of stemmed points in Korea.
-First, what changes in foragers' landscape use were associated with the introduction of the new tools?
+First, what changes in foragers’ landscape use were associated with the introduction of the new tools?
 Specifically, was the use of stemmed points associated with occupation of marginal or extreme environments?
-Our assumption here is that this new technology might have allowed foragers to explore and sojourn in less productive landscapes by increasing their mobility.
-In Australia during the mid-Holocene, Hiscock et al.
-argue that the first appearance of microlithics in many locations represent a portable and multi-functional toolkit that minimize travel expenses and increase tool readiness to adapt to patchy and unpredictable resources associated with environmental change [@Hiscock_1994; @Hiscock_Clarkson_Mackay_2011].Similarly, in Japan, Morisaki et al.
-[@Morisaki_Izuho_Terry_Sato_2015] argue that the transition from trapezoid to the blades and projectile points around 25 ka cal BP in north Paleo-Honshu Island enabled the foragers to extend their occupation into cold grassland landscapes.
+Our assumption here is that this new technology, and the ability to reach long-distance targets, might have allowed foragers to explore and sojourn in less productive landscapes by increasing their mobility.
+For example, in Australia during the mid-Holocene, Hiscock et al.
+argue that the first appearance of microlithics in many locations represent a portable and multi-functional toolkit that minimize travel expenses and increase tool readiness to adapt to patchy and unpredictable resources associated with environmental change [@Hiscock_1994; @Hiscock_Clarkson_Mackay_2011].
+Similarly, in Japan, [@Morisaki_Izuho_Terry_Sato_2015] argue that the transition from trapezoid to blades and projectile points around 25 ka in north Paleo-Honshu Island enabled the foragers to extend their occupation into cold grassland landscapes.
 
-Second, we ask what changes in the way people use habitation sites and mobility were associated with the new technology?
+Second, we ask what changes in the way people used habitation sites and mobility were associated with the new technology?
 After the appearance of these new tools, did people tend to be more or less mobile?
@@
 After the first excavation of a Paleolithic site in Korea at Seokjang-ri site in the 1960s, more than 200 Paleolithic sites have been discovered in South Korea [@lee2019were].
-We selected sites dated to before and after the transition period (40-35 ka) to analyse the process of technological change.
+We selected sites dated to before and after the transition period (42-35 ka) to analyse the process of technological change.
 This resulted in a sample of 33 assemblages from 22 sites spanning 49-24 ka (Figure \@ref(fig:korea-map)).
@@
 If population dynamics were a key driver, then we expect population to be high at the time the new technology first appears.
-However, with the first stemmed points appearing at 38 ka, well before the observed peak in population, we conclude that population dynamics were not a major mechanism in the appearance of this new technology.
+However, with the first stemmed points appearing at 41.5 ka, well before the observed peak in population, we conclude that population dynamics were not a major mechanism in the appearance of this new technology.
 Bae claims that stemmed points are the result of the continuous influx of modern humans from a Siberian migration [@bae2010origin].

### 2021-04-12 analysis/paper/Trial-combine-spd-mat.R

@@ -1,1 +1,0 @@
-#Try to combine SPD + MAT

### 2021-04-12 analysis/paper/paper.Rmd
Finish major text editing

@@
 abstract: |
-  We use models from human behavioral ecology to examine stone artefacts from 22 sites in Korea to investigate mobility and site occupation patterns during the Late Pleistocene when new tools, such as stemmed points and blades, appeared in the archaeological record.
+  We use models from human behavioral ecology to examine stone artifacts from 22 sites in Korea to investigate mobility and site occupation patterns during the Late Pleistocene when new tools, such as stemmed points and blades, appeared in the archaeological record.
 We focus on two questions: what changes in foragers’ landscape use were associated with the introduction of new tools?
 And what changes in the way people use habitation sites and mobility were associated with the new technology?
-To answer these questions we present quantitative analyses of artefact volumetric density, retouch frequency, composition of toolkits, and artefact raw materials.
+To answer these questions we present quantitative analyses of artifact volumetric density, retouch frequency, composition of toolkits, and artifact raw materials.
 We explore the environmental and demographic contexts by applying paleoclimate simulations and summed probability distribution models.
@@
 # Keywords {-}
-Human Behavioural Ecology, Stemmed points, Mobility, Site occupation, Landscape use, Summed Probability Distribution, Pleistocene, Korean Late Paleolithic
+Human Behavioral Ecology, Stemmed points, Mobility, Site occupation, Landscape use, Summed Probability Distribution, Pleistocene, Korean Late Paleolithic
 
@@
 The stemmed point is a projectile point made from an elongated flake or blade with slight retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft.
-These types of artefacts are often called tanged points in the western hemisphere [@serwatka20162d; @ivanovaite2020all], but we use ‘stemmed point’ here to distinguish from Bronze Age stone projectile points that have long been called ‘tanged points’ in Korea.
+These types of artifacts are often called tanged points in the western hemisphere [@serwatka20162d; @ivanovaite2020all], but we use ‘stemmed point’ here to distinguish from Bronze Age stone projectile points that have long been called ‘tanged points’ in Korea.
 Stemmed points are likely to have been projectile weapon tips because the stem of these tools is considered to represent a hafting projection used for joining the lithic to a wooden shaft.
-If these artefacts were projectile weapon tips, then stemmed points represent the first appearance of hafting technology in Korea.
-Most Korean Paleolithic sites do not preserve organic remains, so we lack direct fauna evidence on hunting and subsistence strategies, and stone artefacts are our only window into these behaviours.
-Here we use a behavioural ecological framework to explore questions about the implications of the appearance of this new technology.
+If these artifacts were projectile weapon tips, then stemmed points represent the first appearance of hafting technology in Korea.
+Most Korean Paleolithic sites do not preserve organic remains, so we lack direct fauna evidence on hunting and subsistence strategies, and stone artifacts are our only window into these behaviors.
+Here we use a behavioral ecological framework to explore questions about the implications of the appearance of this new technology.
 We used data from 22 sites in South Korea to explore changes in the way people occupied the landscape, the way people used sites, and the of the social and ecological contexts associated with the appearance stemmed points, and the blade industry that followed.
@@
 To our knowledge, the implications of the appearance of stemmed points for forager mobility and site use have not been explored in detail.
-Here, we attempt to identify patterns of mobility and site occupation through analysis of the stone artefact assemblages using Human Behavioural Ecology (HBE) concepts.
+Here, we attempt to identify patterns of mobility and site occupation through analysis of the stone artifact assemblages using Human Behavioral Ecology (HBE) concepts.
 Drawing on established relationships between lithic technology and forager mobility [@shott1986settlement; @Hiscock_1994; @kuhn1994formal; @kuhn2004upper; @kuhn2015artifacts; @kuhn2016moves; @capriles2018mobility], we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that were part of a broader strategy of moving frequently and further, possibly as an adaptation to environmental or population changes.
@@
 These innovations include blade technology, high frequencies of retouched blade tools, several novel tools such as projectile points, stemmed points, end scrapers, burins, denticulates, etc, and frequent use of high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary].
-Previous work has argued that the appearance of new stone artefact technologies in this region may be linked to modern human dispersals [@seong2008tanged; @bae2010origin; @bae2017origin].
+Previous work has argued that the appearance of new stone artifact technologies in this region may be linked to modern human dispersals [@seong2008tanged; @bae2010origin; @bae2017origin].
 
@@
 Shuidonggou Locality SDG 2, which dates to around ca.
-32ka, contains fauna specimens, ostrich eggshell beads, and stone artefacts including blades, retouched flakes, cores, and debris along with well-preserved hearths.
+32 ka, contains fauna specimens, ostrich eggshell beads, and stone artifacts including blades, retouched flakes, cores, and debris along with well-preserved hearths.
 In Locality 9 dated to ca.
-29ka, there is a small scatter of stone artefacts consisting of blades, Levallois flakes, cores and other retouched flakes.
+29 ka, there is a small scatter of stone artifacts consisting of blades, Levallois flakes, cores and other retouched flakes.
 Excavations Locality 12 shows that its archaeological context, dated to 13,078--13,296 cal BP, includes ground stone and more than 30,000 microlithics made out of a variety of raw materials such as fine-grained and siliceous rocks.
@@
 
-In Japan, Early Upper Paleolithic (EUP) technologies appeared in different regions of the archipelago around 38ka, accompanied by remains of Homo sapiens found in Okinawa Island (ca.
-36ka) [@yamaoka2012use; @izuho201421].
+In Japan, Early Upper Paleolithic (EUP) technologies appeared in different regions of the archipelago around 38 ka, accompanied by remains of Homo sapiens found in Okinawa Island (ca.
+36 ka) [@yamaoka2012use; @izuho201421].
 These technologies include trap pits, cobble concentrations, hearths, charcoal concentrations, and toolkits including trapezoids, pointed-shaped backed blades, backed points, burins, end scrapers, side scrapers, wedges, beak-shaped tools, axes, edge-ground axes, hammerstones, cobble tools, and anvils [@bae2017origin; @izuho201421].
@@
 An important example is Ishinomoto 8-ku, which is one of the earliest sites, dated to 39,690--34,790 cal BP, located in Takuma upland of Kumamoto Prefecture.
-A total of 500 stone artefacts were discovered including trapezoids, side scrapers, flakes, flake cores, edge-ground axes, and cobble tools.
+A total of 500 stone artifacts were discovered including trapezoids, side scrapers, flakes, flake cores, edge-ground axes, and cobble tools.
 The main lithic raw materials are Chert and Andesite [@izuho201421].
@@
 The alternative model, which we call the '*in situ*' model claims that stemmed points and other Late Paleolithic assemblages including blade industries autonomously emerged in the south of the Korean peninsula, with no apparent external influence [@seong2008tanged].
-This claim is supported by Korea having the earliest appearance of stemmed points in Northeast Asia, at sites such as Bonggok, Hwadaeri, Hopyeongdong, and Yonghodong dated to approximately 42-35ka BP [@seong2009emergence].
+This claim is supported by Korea having the earliest appearance of stemmed points in Northeast Asia, at sites such as Bonggok, Hwadaeri, Hopyeongdong, and Yonghodong dated to approximately 42-35 ka BP [@seong2009emergence].
 Seong(2009) presents the increased blade-to-flake ratios of lithic assemblages after the appearance of stemmed points to support the *in situ* model, based on the premise that the blade industry represents a new exogenous technology while flakes indicate a continuing local one.
@@
 
-Our brief review of Late Pleistocene technological innovations in Northeast Asia shows that previous work has largely focussed on timing, location, origins and description of these new tools [@seong2008tanged; @seong2009emergence; @bae2010origin; @lee2013current; @Lee_2016], but modelling the land-use behaviours associated with these assemblages has rarely been undertaken.
-We focus on the appearance of stemmed points in Korea because these are a new technology that likely represented new hunting behaviours.
+Our brief review of Late Pleistocene technological innovations in Northeast Asia shows that previous work has largely focused on timing, location, origins and description of these new tools [@seong2008tanged; @seong2009emergence; @bae2010origin; @lee2013current; @Lee_2016], but modeling the land-use behaviors associated with these assemblages has rarely been undertaken.
+We focus on the appearance of stemmed points in Korea because these are a new technology that likely represented new hunting behaviors.
 This is because the hafting implies throwing techniques that can reach long-distance targets [@Keeley_1982; @kuhn2015artifacts].
-We draw on behavioural ecological theory to model the effects of this technological change.
-Behavioural ecology theory offers structured models and testable hypotheses based on optimality assumptions derived from principles of economic rationality and environmental adaptation [@winterhalder1992evolutionary; @prentiss2019handbook].
+We draw on behavioral ecological theory to model the effects of this technological change.
+Behavioral ecology theory offers structured models and testable hypotheses based on optimality assumptions derived from principles of economic rationality and environmental adaptation [@winterhalder1992evolutionary; @prentiss2019handbook].
 Specifically, we use the patch choice model, which is built around the marginal value theorem, to predict that forager mobility between patches and inter-patch travel times will increase when resource patch productivity decreases [@smith1991inujjuamiunt; @smith1983anthropological; @bettinger1997evolutionary; @wolverton2015patch; @llano2015optimal].
-We model individual artefact types as a kind of resource patch.
+We model individual artifact types as a kind of resource patch.
 Foragers may find a more costly and complicated technology (such as stemmed points, relative to flakes) optimal if they have a long cumulative time in use, if they are maintainable and reusable in a landscape where lithic raw material supply is uncertain [@torrence1983time; @bleed1986optimal].
 That is, they may move to a patch where they can stay longer (i.e.
-an artefact type that has a long and extendible use-life) when they are not certain of the productivity and travel times of other patches [@kuhn2015artifacts].
+an artifact type that has a long and extendible use-life) when they are not certain of the productivity and travel times of other patches [@kuhn2015artifacts].
 
@@
 After the appearance of these new tools, did people tend to be more or less mobile?
-The stone artefact assemblage from a site can inform us how long individuals or groups stayed and give insights into their activities [@binford1979organization; @kelly1995foraging; @kuhn2016moves; @holdaway2019surface].
-For example, stone artefact density and the frequency of retouched pieces (scaled to the volume of excavated sediment) can be used as proxies to represent occupation patterns [@barton2011modeling; @clark2017lithics].
+The stone artifact assemblage from a site can inform us how long individuals or groups stayed and give insights into their activities [@binford1979organization; @kelly1995foraging; @kuhn2016moves; @holdaway2019surface].
+For example, stone artifact density and the frequency of retouched pieces (scaled to the volume of excavated sediment) can be used as proxies to represent occupation patterns [@barton2011modeling; @clark2017lithics].
 Assemblages with low density but a high proportion of retouched and backed pieces may indicate the remains of 'short-term camp', which is a site for small and ephemeral overnight camp or limited activity station.
@@
 
-## Sites and dates {-}
+## Archaeological Sites {-}
 
 After the first excavation of a Paleolithic site in Korea at Seokjang-ri site in the 1960s, more than 200 Paleolithic sites have been discovered in South Korea [@lee2019were].
-We selected sites dated to before and after the transition period (42-35 ka) to analyse the process of technological change.
+We selected sites dated to before and after the transition period (42-35 ka) to analyze the process of technological change.
 This resulted in a sample of 33 assemblages from 22 sites spanning 49-24 ka (Figure \@ref(fig:korea-map)).
-We identified multiple assemblages in a site where culturally sterile deposits separated artefact-bearing deposits, or where stratigraphic units could be identified by major differences in the texture and composition of the sedimentary deposit containing the stone artefacts.
+We identified multiple assemblages in a site where culturally sterile deposits separated artifact-bearing deposits, or where stratigraphic units could be identified by major differences in the texture and composition of the sedimentary deposit containing the stone artifacts.
 For example, the Hwadaeri site has three cultural layers.
@@
 
-## Assemblages {-}
+## Stone Artifact Assemblages {-}
 
 The 33 assemblages consist of (non-blade) cores, blade cores, blades, flakes, debris, hammers, choppers, planes, polyhedrals, side scrapers, end scrapers, notches, cleavers, stemmed points, awls, denticulates, burins, handaxes, blanks, knives, handadzes, peaks, flatters, flakers etc.
-We excluded some artefacts from our sample because of uncertainty about their typology.
-The stone artefacts from all of these sites are stored in museums located throughout South Korea, and have been briefly described in the excavation reports.
-We have visited most of these collections to obtain permission to study the materials.
+We excluded some artifacts from our sample because of uncertainty about their typology.
+The stone artifacts from all of these sites are stored in museums located throughout South Korea, and have been briefly described in the excavation reports.
+We have visited most of these collections to obtain permission to study the materials directly.
 
-## Statistical methods {-}
+## Methods for Analyzing Stone Artifact Assemblages {-}
 
-To answer the two research questions, we compiled information about artefact density, retouch frequency, toolkit composition, raw materials, site elevation, and radiocarbon ages.
+To answer our two research questions, we compiled information about artifact density, retouch frequency, toolkit composition, raw materials, site elevation, and radiocarbon ages.
 
-The artefact volumetric density is defined as the total number of artefacts per cubic meter of excavated sediment, and serves as a proxy for the accumulation ratio of artefacts [@clark2017lithics; @clark2019landscapes].
+Artifact volumetric density is defined as the total number of artifacts per cubic meter of excavated sediment, and serves as a proxy for the accumulation ratio of artifacts [@clark2017lithics; @clark2019landscapes].
 We used these density values to evaluate mobility and land-use practices of hunter-gatherers during the Late Pleistocene.
-The retouch frequency is the number of retouched tools divided by the total number of artefacts in an assemblage.
+Retouch frequency is the number of retouched tools divided by the total number of artifacts in an assemblage.
 This value represents mobility and duration of site occupation based on the assumption that tool reduction is a tactic to extend the use-life of a tool by generating sharp, usable edges while minimizing the cost of transporting tools or bulky raw materials by reducing the total load weight [@Kuhn_1990; @buvit2014late; @clark2019landscapes].
-In other words, producing and carrying highly retouched tools will increase portability and efficiency by decreasing the size of both individual tools and the assemblages so that hunter-gatherers could move often or further, or both [@kuhn1991unpacking; @kuhn1994formal; @kuhn2004upper; @andrefsky1994raw; @davies2018modeling].
-Combining these two proxies, the artefact volumetric density and the retouch frequency, we infer whether or not the hunter-gatherers were residentially mobile or logistically organized.
-If the assemblage in a site has a high artefact density and small proportion of retouch, we interpret it as a more expedient assemblage that represents "base camps" or "residences", while less density with higher proportions of retouch shows a more curated assemblage and indicates either residential mobility or certain task groups away from residential sites [@riel2004late; @clark2017lithics].
+In other words, producing and carrying highly retouched tools increased tool portability and efficiency by decreasing the size of both individual tools and the assemblages so that hunter-gatherers could move often or further, or both [@kuhn1991unpacking; @kuhn1994formal; @kuhn2004upper; @andrefsky1994raw; @davies2018modeling].
+Combining these two proxies, the artifact volumetric density and the retouch frequency, we infer whether or not the hunter-gatherers were residentially mobile or logistically organized.
+If the assemblage in a site has a high artifact density and small proportion of retouch, we interpret it as a more expedient assemblage that represents "base camps" or "residences", while lower artifact densities with higher proportions of retouched pieces shows a more curated assemblage and indicates either residential mobility or certain task groups away from residential sites [@riel2004late; @clark2017lithics].
 
-We calculated toolkit composition by analysing tool and debitage frequencies to identify changes in the assemblages during the transition period.
+We calculated toolkit composition by analyzing tool and debitage frequencies to identify changes in the assemblages during the study period.
 Toolkit composition helps us answer the question of how site occupation patterns changed before and after the appearance of new technology.
-The type and proportion of retouched tools and knapping products, including debris and hammerstones, could give information about settlement patterns based on a premise that more activities are conducted with a wider range of tool types for longer occupation [@buvit2014late; @garcia2017occupational; @centi2019bulb].
-For instance, small sized assemblages with a few types of tools represent short duration of site occupation [@shott2010size; @buvit2014late] .
+The type and proportion of retouched tools and knapping products, including debris and hammerstones, provide information about settlement patterns based on a premise that more activities are conducted with a wider range of tool types for longer occupation [@buvit2014late; @garcia2017occupational; @centi2019bulb].
+For example, we interpret small sized assemblages with a low diversity of tools as representing a short duration of site occupation [@shott2010size; @buvit2014late].
 
-Our study of raw material consumption patterns assumes that tool stone was transported and consumed to optimize mobility and minimize risk of not having artefacts when they were needed [@andrefsky1994raw; @brantingham2000raw; @brantingham2003neutral; @yue2020lithic].
+Our study of raw material consumption patterns assumes that tool stone was transported and consumed to optimize mobility and minimize risks of not having artifacts when they were needed [@andrefsky1994raw; @brantingham2000raw; @brantingham2003neutral; @yue2020lithic].
 The representation of raw materials in an assemblage can be an indicator of site occupation patterns.
-For example, a large variability of raw materials in a site can suggest long-term settlement, while varied and exotic raw materials can indicate higher mobility [@valde2020models].
-Here we want to know the specific raw materials that relate to the appearance of a new technology, and understanding site occupation patterns by examining changes in the proportions of raw material types in the assemblages.
+For example, a large variability of raw materials in a site can suggest long-term settlement, while a small assemblage of exotic raw materials can indicate higher mobility [@valde2020models].
+Here we want to know the specific raw materials that relate to the appearance of a new technology, and investigate site occupation patterns by examining changes in the proportions of raw material types in the assemblages.
 
-HBE models require environmental proxies to build hypotheses of behavioural change influenced by the surrounding environment.
-However, there are few paleoenvironmental proxies in Korea relevant to our target period.
+## Methods for Analyzing Environmental Context {-}
+
+In addition to these data on human behaviour from the stone artifact assemblages, our HBE models require environmental proxies to build hypotheses of behavioural change influenced by the surrounding environment.
+However, there are few paleoenvironmental proxies in Korea relevant to our study period.
 For example, geoarchaeological observations simply suggest that environment of the period was overall cooler and drier during MIS 3 [@Han_2008; @seong2008tanged; @Choi_2011; @chang2013human; @im2015study; @bak2017late].
-To supplement the limited availability of high-resolution local proxies, we used climate information from the simulated data set of paleoclimate including global monthly temperature, covering the last 120,000 years [@beyer2020high].
-We applied information of site location (all open-sites) including elevation, latitude and longitude to the simulated data set to get the annual temperature of our research period and area.
-Besides the climate analysis, we analysed site elevation as a proxy to represent different environments at local scales, to identify how these impact mobility and site occupation.
-We mainly focused on the elevation pattern of the sites before and after the appearance of new technology and its relationship with the stemmed points.
+To supplement the limited availability of high-resolution local proxies, we used climate information from a simulated data set of paleoclimate including global monthly temperature, covering the last 120,000 years [@beyer2020high].
+We extracted annual temperature values from the data generated by @beyer2020high for all site locations in our sample.
+In addition, we extracted site elevation values from a raster, as a proxy to represent different environments at local scales, to identify how these impact mobility and site occupation, and their relationship with the stemmed points.
 
-We used radiocarbon dates as a reference for population level.
+## Methods for Analyzing Demographic Context {-}
+
+We used radiocarbon dates as a proxy for human population to infer the demographic context of technological change.
 Summed probability distributions (SPD) of radiocarbon dates are increasingly used to infer temporal trends in past human populations [@rick1987dates; @bamforth2012radiocarbon; @shennan2013regional; @timpson2014reconstructing; @contreras2014summed; @riris2019widespread].
-Despite limitations of the validity of SPD [@williams2012use; @bamforth2012radiocarbon], including the lack of excavated sites in certain periods, sample size, and calibration effects, this method may allow a first approximation of population fluctuations.
+Despite concerns about limitations of the validity of SPD [@williams2012use; @bamforth2012radiocarbon; @carleton2021evaluating], several studies have shown good agreement between SPDs and other archaeological indicators (e.g.
+site counts, settlement size) so this method may allow a first approximation of population fluctuations [@french2015upper; @palmisano2017comparing].
 Although this is often used with samples of thousands of ages, this method can also be useful for small sample sizes.
-For instance, Timpson et al.
-apply SPD to 93 ages from Eastern Middle Sweden and demonstrate that the result is equivalent to a larger sample size (n= 243) [@timpson2014reconstructing].
+For instance, @timpson2014reconstructing apply SPD to 93 ages from Eastern Middle Sweden and demonstrate that the result is equivalent to a larger sample size (n= 243).
 We used 108 radiocarbon ages from 22 sites to examine connections between population trends during the Late Pleistocene in Korea and technological innovation.
@@
 He found that the lithics became less complex, and point shape changed as the population level dropped.
-On the other hand, @collard2013population argue that regional archaeological data do not support the population-driven models [@buchanan2016drivers].
+On the other hand, @collard2013population argue that regional archaeological data do not support the population-driven models.
 For example, @buchanan2016drivers concluded that changes in point types from \>13,000 years ago to 400 years ago in Texas were related to environmental risk, and not dependent on population size.
 We explored SPD to test for correlations between population, environmental proxies and technological change.
-We used the R package 'rcarbon' [@cremaanalysing] to perform statistical analyses and conduct model testing using Monte Carlo methods [@timpson2014reconstructing; @crema2016summed; @palmisano2020holocene].
+We used the R package 'rcarbon' [@cremaanalysing] to perform statistical analyzes and conduct model testing using Monte Carlo methods [@timpson2014reconstructing; @crema2016summed; @palmisano2020holocene].
 We generated exponential, uniform, linear, and logistic models and evaluated the model goodness-of-fit by applying Akaike's Information Criterion (AIC) as a selection criterion [@sakamoto1986akaike; @bevan2017holocene; @riris2019widespread].
 
-All analyses and visualization are performed in the R Environment [@team2017r].
-Our R code and data are available on the following GitHub repository: <https://github.com/parkgayoung/koreapaleolithicmobilityoccupation> as well as on the Open Science Framework: <https://osf.io/yagd2/> to enable full reproducibility [@marwick2017computational].
+## Reproducibility and Open Source Materials {-}
+
+The entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is included in the Supplementary Online Materials at http://doi.org/osf.io/yagd2/ to enable re-use of materials and improve reproducibility and transparency [@Marwick2017].
+Also in this version-controlled compendium [@marwick2018packaging] are the raw data for all the visualizations and tests reported here.
+All of the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this repository.
+The code is released under the MIT license, the data as CC-0, and figures as CC-BY, to enable maximum re-use.
 
@@
 
-## Artefact volumetric density and retouch frequency {-}
+## Artifact Volumetric Density and Retouch Frequency {-}
 
-Figure \@ref(fig:volume-density) shows that there is no clear temporal pattern of changes in artefact density, although the density of some assemblages are two or three times higher than others, especially after 30ka.
-The inset figure shows that assemblages with stemmed points have higher artefact densities than the assemblages without.
-With the newer tools, hunter-gatherers may have stayed in the same location for longer durations, and we may be seeing site occupation patterns in this sample that represent the use of both base camps or residences, with the artefact higher densities, and more briefly occupied camps with lower densities.
+Figure \@ref(fig:volume-density) shows that there is no strong directional temporal pattern of changes in artifact density.
+The upper bound of artifact density increases in more recent periods.
+For example, the density of some assemblages are two or three times higher than others, especially after 30 ka.
+The inset figure shows that assemblages with stemmed points have higher artifact densities than the assemblages without.
+With the newer tools, hunter-gatherers may have stayed in the same location for longer durations.
+We may be seeing site occupation patterns in this sample that represent the long term trend of the use of both base camps or residences, represented by the artifact higher densities, and more briefly occupied camps with lower densities.
+The most striking result here is that maximum size of base camps increases in more recent periods.
 
-```{r volume-density, fig.cap="Artefact volumetric density over time.
-The size of data points represent the total number of artefacts from each assemblage and the colour indicates the presence of stemmed points.
-The inset plot shows direct comparison of artefact densities for assemblages with and without stemmed points.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/002-artifact-volumetric-density.R"))}
+```{r volume-density, fig.cap="Artifact volumetric density over time.
+The size of data points represent the total number of artifacts from each assemblage and the colour indicates the presence of stemmed points.
+The inset plot shows direct comparison of artifact densities for assemblages with and without stemmed points.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/002-artifact-volumetric-density.R"))}
 quietly(source(here::here("analysis/paper/002-artifact-volumetric-density.R")))
@@
 
-Figure \@ref(fig:volume-density-retouch) shows a strong pattern of artefacts from less dense assemblages having higher retouch frequencies while more dense assemblages have lower retouch frequencies (p=0.043), further showing a spectrum of site functions in this sample.
-As for artefact density, there are no clear chronological trends in retouch frequencies.
-The assemblages containing stemmed points tend to have fewer retouched pieces compared to assemblages without stemmed points (p=0.011).
+Figure \@ref(fig:volume-density-retouch) shows a strong pattern of artifacts from less dense assemblages having higher retouch frequencies while more dense assemblages have lower retouch frequencies (p=0.043), further showing a spectrum of site functions in this sample.
+As we saw for artifact density, there are no clear directional chronological trends in retouch frequencies.
+Assemblages containing stemmed points tend to have substantially fewer retouched pieces compared to assemblages without stemmed points (p=0.011).
 
-```{r volume-density-retouch, fig.cap="Relationship between artefact density and retouch frequency.
+```{r volume-density-retouch, fig.cap="Relationship between artifact density and retouch frequency.
 The colour of the data points indicate the age of the assemblage, point size indicates assemblage size, and shape indicates presence or absence of stemmed points.
@@
 
-## Toolkit composition {-}
+## Toolkit Composition {-}
 
-Figure \@ref(fig:artefact-composition) shows that side scrapers are the major part of most assemblages and are present in all assemblages.
+In \@ref(fig:artifact-composition) we see that side scrapers are the major part of most assemblages and are present in all assemblages.
 For example, more than 70 percent of the two assemblages from Hwaderi are side scrapers.
-After around 31 ka, the proportion of side scrapers declines.
-In addition to side scrapers, cores also exist in all assemblages often as a high proportion.
-The proportion of choppers decreased in more recent assemblages.
-Blades first appeared in Gawolri-A at 43 ka and then often found in assemblages thereafter.
-Stemmed points first appear in Bonggok around 41.5ka and were typically found in assemblages that contain blades, except for Songamri-A.
-Other artefacts, including burins, denticulates, end scrapers, handaxes, knives, notches, planes, and polyhedrals made up a small proportion of the assemblages.
+The proportion of side scrapers shows a striking trend over time, rising to a peak at 38-39 ka, and declining thereafter.
+In addition to side scrapers, cores also exist in all assemblages, often as a high proportion.
+The proportion of choppers is lower in more recent assemblages.
+Stemmed points first appeared in Bonggok around 41.5 ka and were typically found in assemblages that contain blades, except for Songamri-A.
+The earliest blades are also recorded from Bonggok but it is not guaranteed that those artifacts as well as other earlier blades are actual blades or elongate flakes without accompanying blade core.
+Blade cores first appeared in Suyanggae-C at 34.3 ka.
+Other artifacts, including burins, denticulates, end scrapers, handaxes, knives, notches, planes, and polyhedrals made up a small proportion of the assemblages and show no directional chronological change.
 
-```{r artefact-composition, fig.cap="The composition of each assemblage.
-We excluded artefacts related to manufacturing processes such as pebbles, hammers, flakes and debris, artefacts appeared only in a few assemblages including point, beak shaped, and awl, and unknown and unfinished pieces.
+```{r artifact-composition, fig.cap="The composition of each assemblage.
+We excluded artifacts related to manufacturing processes such as pebbles, hammers, flakes and debris, and artifacts that appeared only in a few assemblages including points, beak shaped pieces, and awls, and unknown pieces.
 The colour represents different types of tools and the assemblages are placed in chronological order.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/004-artifact-composition.R"))}
 quietly(source(here::here("analysis/paper/004-artifact-composition.R")))
-knitr::include_graphics(here::here("analysis/figures/004-artefact-types.png"))
+knitr::include_graphics(here::here("analysis/figures/004-artifact-types.png"))
 ```
 
-```{r SAR-all, fig.cap = "Stone artefacts from Songamri site. 1. Blade core, 2~4. Blades, 5~6. Stemmed points, 7. Side scraper, 8. Chopper, 9~10. Side scrapers."}
+```{r SAR-all, fig.cap = "Stone artifacts from Songamri site. 1. Blade core, 2~4. Blades, 5~6. Stemmed points, 7. Side scraper, 8. Chopper, 9~10. Side scrapers."}
 knitr::include_graphics(here::here("analysis/figures/SAR_all.png"))
@@
 
-## Raw materials {-}
+## Raw Materials {-}
 
 Quartz, quartz vein and quartzites were the most frequently found raw materials, and they were constantly used throughout the Late Pleistocene (Figure \@ref(fig:raw-materials)).
-For example, both Galdamri-A, dated to 44ka with a small assemblage, and Hopyeongdong-B, dated to 27.5ka with a larger assemblage, consist of only the quartz related materials.
-The use of chert, hornfels, rhyolite, and shale increased after 41.5 ka.
-All artefacts in Sageunri are made of rhyolite.
-Hornfels are dominant at Yongsandong and Yongbang sites at 25 ka.
+For example, both Galdamri-A, dated to 44 ka with a small assemblage, and Hopyeongdong-B, dated to 27.5 ka with a larger assemblage, consist of only the quartz related materials.
+The use of chert, hornfels, rhyolite, and shale increased after 41 ka, but these raw materials remained a small proportion of assemblages until 25 ka.
+For example, hornfels are dominant at Yongsandong and Yongbang, dated to 25 ka.
 Chert, hornfels, rhyolite, and shale are very suitable to make elongate blades and stemmed points because of the predictability of flaking afforded by their fine-grained texture.
@@
 
-## Environmental context {-}
+## Environmental Context {-}
 
 Figure \@ref(fig:site-elevation) shows that people occupied sites located at similar elevations, around 0 to 200 meters above sea level.
-However, there is a substantial difference in elevation between sites with stemmed points and sites without.
+As we saw for artifact density, there is an increase in the upper bound of site elevation in more recent times.
+The most striking pattern in the elevation data is the substantial difference in elevation between sites with stemmed points and sites without.
 The distribution of elevations of sites without stemmed points is generally lower than the elevations of sites with stemmed points.
-Although not a statistically significant difference in elevations (p = 0.102), this may indicate that forager groups who used stemmed points were able to occupy higher altitudes, while the groups without tended to prefer lower altitudes.
-Sites without stemmed points reach a maximum elevation of about 150 m at around 39ka.
-Sites containing stemmed points reach a maximum elevation of 175 m at 28ka.
+Although not a statistically significant difference in elevations (p = 0.102), this may indicate that forager groups who used stemmed points were generally more able to occupy higher altitudes, while the groups without stemmed points tended to prefer lower altitudes.
+According to our locally weighted regression, sites without stemmed points reach a maximum elevation of about 150 m at around 39 ka.
+Sites containing stemmed points reach a maximum elevation of 175 m at 28 ka.
 
@@
 Among the sites in our sample there is a variation of about 5 degrees in the mean annual temperature (MAT), mostly controlled by elevation (Figure \@ref(fig:climate-model)).
+Through MIS 3, the temperature gradually decreased until the Last Glacial Maximum (LGM, 26.5-20 ka) (Clark et al., 2009).
+The MAT of MIS 2, including LGM, was relatively stable.
+The temperature increased again from late MIS 2 towards MIS 1.
 Compared to the east side of the Korean Peninsula, the west side tends to be relatively warmer.
@@
 Gihwari and Gigok sites have the lowest (3℃) and Jeungsan-ri site has the highest MAT (8℃).
-The mean MAT for all sites fluctuated within 4℃ between 50ka and 10ka.
-Through MIS 3, the temperature gradually decreased until the Last Glacial Maximum (LGM, 26.5-20 ka) [@clark2009last].
-The first appearance of stemmed points occurred in the middle of the decreasing MAT trend in MIS 3, at 40-35ka.
-The MAT of MIS 2, including LGM, was relatively stable.
-The temperature increased again from late MIS 2 towards MIS 1.
+The mean MAT for all sites fluctuated within 4℃ between 50 ka and 10 ka.
+The first appearance of stemmed points occurred in the middle of the decreasing MAT trend in MIS 3, at 40-35 ka.
 Figure 7D shows a negative relationship between temperature and elevation (p = 0.073).
@@
 
-```{r climate-model, fig.cap="Mean Annual Temperature (MAT) of the Korean Paleolithic assemblages mentioned in this research.
-A: Site locations on the map and MAT.
+```{r climate-model, fig.cap="Mean Annual Temperature (MAT) of the Korean Paleolithic assemblages mentioned in the text.
+A: Site locations and their MAT.
 B: MAT distributions for each site during the period they were occupied.
-C: MAT from 50ka to 10ka.
+C: MAT from 50 ka to 10 ka.
 The grey lines indicate the MAT for each site and the black line is the mean temperature of all sites.
-The light grey area in the middle of the plot indicates the duration of MIS 2, separating MIS 3 and 1, and dark grey area represents the duration of LGM.
+The light grey area in the middle of the plot indicates the duration of MIS 2, and the dark grey area represents the duration of the  LGM.
 D: The relationship between MAT and site elevation.
@@
 
-## Demographic context {-}
+## Demographic Context {-}
 
-Figure \@ref(fig:spd-model) shows the four theoretical null models of demographic change for the period 50ka to 10ka, exponential, uniform, linear, and logistic, generated from the radiocarbon dates.
-Statistical significance of each model is computed using 200 simulations and the global p-values are 0.004, 0.004, 0.009, and 0.004, respectively.
-In this case, the linear model is the best fit with lower value verified by AIC.
-However, the other models have similar patterns with a minor margin at -48.6.
-The radiocarbon dates fit well with the overall trajectory of the null models (grey shading) with both positive and negative deviations.
-SPD shows a steady growth until 35ka, stays at the peak with minor deviation 32ka, and then starts declining with frequent temporary fluctuations (relatively sharp).
-The beginning and ending of the SPD indicates significant negative deviation with low density of radiocarbon dates.
+Figure \@ref(fig:spd-model) shows the summed probability distribution (SPD) of the radiocarbon ages from the sites in our sample overlaid with distributions (indicated by the grey envelopes) generated from four models of demographic change for the period 50 ka to 10 ka (exponential, uniform, linear, and logistic).
+The statistical significance of each of the models we fit to the observed SPD was computed using 200 simulations, and the global p-values are 0.004, 0.004, 0.009, and 0.004, respectively.
+These low p-values indicates significant deviations in the observed SPD from the modeled distributions.
+Overall, the linear model is the best fit with the observed SPD, having the lowest AIC value of the three models where we could compute this.
+We can say that the population did not grow or decline strongly logistically, exponentially or linearly during the Late Pleistocene, but instead followed a noisy distribution.
+The most striking results here are the SPD showing a steady growth rising to a peak at about 35 ka, indicated by the significant deviations from all four null models (shaded in red).
+The SPD remains high with minor deviations until declining to a lower level at 32 ka.
+After 32 ka the SPD remains generally low, with frequent temporary fluctuations.
 
@@
 The best fit model has the lowest AIC score.
-In this case, the linear model is the best fit, with the next-best fit (exponential) having a higher AIC score by a minor margin at -48.6.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/008-summed-probability-distributions.R"))}
+In this case, the linear model is the best fit for the observed data", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/008-summed-probability-distributions.R"))}
 quietly(source(here::here("analysis/paper/008-summed-probability-distributions.R")))
@@
 
-Our study focused on two questions: what changes in foragers' landscape use were associated with the introduction of the new tools?; what changes in the way people use habitation sites and mobility were associated with the new technology?
-Overall, our results show that assemblages with stemmed points tend to have higher artefact densities and lower proportions of retouched pieces.
+Our study focused on two questions: (1) what changes in foragers’ landscape use were associated with the introduction of the new tools?; and (2) what changes in people’s mobility and use of habitation sites were associated with the new technology?
+Overall, our results show that assemblages with stemmed points tend to have higher artifact densities and lower proportions of retouched pieces.
 Quartz and side scrapers, in addition to cores and choppers, remain dominant in assemblages throughout the late Pleistocene.
@@
 
-## Artefact volumetric density and retouch frequency {-}
+## Artifact Volumetric Density and Retouch Frequency {-}
 
-We examined the artefact volumetric density and retouch frequency to understand the difference in mobility strategies and site occupation patterns associated with the introduction of stemmed points.
-The analyses are based on the premise that highly mobile groups would carry lightweight toolkits with more retouched and easily replaceable composite tools, and leave a smaller amount of artefacts on site as a result of short-term occupations.
-These curated technologies invest time and effort in stone artefact manufacturing and maintenance to minimize travel costs of carrying the tools for groups using high mobility or special purpose forays.
-On the other hand, less mobile groups would produce fewer retouched tools and leave higher densities of artefacts generated from long-term occupations.
+We examined the artifact volumetric density and retouch frequency to understand the difference in mobility strategies and site occupation patterns associated with the introduction of stemmed points.
+The analyses are based on the premise that highly mobile groups carried lightweight toolkits with more retouched and easily replaceable composite tools, and left smaller amounts of artifacts at sites as a result of short-term occupations.
+These curated technologies reflect an investment of time and effort in stone artifact manufacturing and maintenance to minimize travel costs of carrying the tools for high mobility or special purpose foray groups.
+On the other hand, less mobile groups produced fewer retouched tools and left higher densities of artifacts generated from long-term occupations.
 These expedient technologies are more typical for groups with greater residential stability who can make, use, abandon, and remake their tools frequently in the same location [@binford1979organization; @torrence1989time; @kuhn1994formal; @meignen2006middle; @kuhn2014mousterian; @vaquero2018searching; @clark2019landscapes].
-Our data support the concept of the curated and expedient technology by showing a strong pattern of artefacts from less dense assemblages having higher retouch frequencies, while more dense assemblages have lower retouch frequencies.
-Our results show that foragers with stemmed points tended to stay in the same site for longer periods with higher artefact densities and fewer retouched pieces, compared to assemblages without stemmed points.
+Our data support the general concept of the curated and expedient technology by showing a strong pattern of artifacts from less dense assemblages having higher retouch frequencies, while more dense assemblages have lower retouch frequencies.
+Our results show that foragers with stemmed points tended to stay in the same site for longer periods with higher artifact densities and fewer retouched pieces, compared to assemblages without stemmed points.
 In general, the new technology of stemmed points was more associated with expedient technological strategies.
 
-It could be assumed that foragers with stemmed points were highly mobile groups and chose curated technologies since the stemmed points were likely used for hunting activities [@chang2013human; @lee2019were].
-However, our results show the opposite that the new technology was more often associated with expedient technological strategies.
-Furthermore, our results show that different sites represent different occupation patterns during the time of technological transition, making it difficult to characterise this period with a single land use strategy.
+If stemmed points were primarily a hunting tool, as prior work has assumed, we might expect foragers to have been using stemmed points in highly mobile groups and stemmed points mostly found in small assemblages at task-specific sites [@chang2013human; @lee2019were].
+However, our results show the opposite, namely that this new technology was more often associated with expedient technological strategies.
+Furthermore, our results show that different sites represent different occupation patterns during the time of technological transition, making it difficult to characterize this period with a single land use strategy.
 Seong claims that transitions from the expedient to curated technology in the Late Pleistocene were neither a straightforward nor unilineal processes and diverse site types can be more related to different occupation purposes including hunting camps, limited activity stations, caches, and so forth [@seong2015diversity].
-Our findings support Seong, with some sites such as Bonggok and Yonghodong containing only 2 stemmed points, while Yongsandong has 38 points including broken tips and a base.
-More than 70% of Yongsandong stemmed points were unearthed, dominating the tool kit with 233 blades and other byproducts related to lithic manufacturing including cores and debris made of the same raw materials [@kim2004yongsan; @bae2012nature].
+Our findings support Seong, with some sites such as Bonggok and Yonghodong containing only two stemmed points, while Yongsandong has 38 points including broken tips and a base.
+Yongsandong stemmed points are made from blades, dominating the tool kit with 233 blades and other byproducts related to lithic manufacturing including cores and debris made of the same raw materials (i.e.
+hornfels) [@kim2004yongsan; @bae2012nature].
 In the specific case of Yongsandong it seems likely that stemmed points were manufactured for hunting purposes, which dominated the function of the site.
@@
 
-## Toolkit composition and raw materials {-}
+## Toolkit Composition and Raw Materials {-}
 
-The composition of toolkits during the Late Pleistocene of Korea has been described within multiple chronological sub-periods.
-Seong divides five successive assemblage types for the Late/Upper Paleolithic in Korea: (1) quartzite and quartz vein artefacts; (2) mostly small quartzite and quartz vein artefacts with some large artefacts such as cores and choppers; (3) stemmed points dominant; (4) typical blade assemblages including stemmed points; and (5) microblade after 30ka [@seong2015diversity].
+Previous work on toolkit composition during the Late Pleistocene of Korea attempted to divide it up into multiple chronological sub-periods.
+Seong divides five successive assemblage types for the Late/Upper Paleolithic in Korea: (1) quartzite and quartz vein artifacts; (2) mostly small quartzite and quartz vein artifacts with some large artifacts such as cores and choppers; (3) stemmed points dominant; (4) typical blade assemblages including stemmed points; and (5) microblade after 30 ka [@seong2015diversity].
 Lee focuses on the Honam region in southwestern Korea and divides the same time frame into two phases; 'core tool industry with large flake knapping' during early MIS 3 (59−40 ka), and 'blade industry with tanged point chiefly' during late MIS (40−24 ka) [@lee2012characteristics].
@@
 Our results, covering 49-24 ka and corresponding to the first three types of Seong's divisions, support the continuation of core tools abundantly throughout the Late Pleistocene.
-In addition, our data show that side scrapers stayed as major tools while other tools occupied only a small portion of the assemblages.
+In addition, our data show that side scrapers remained as major tools while other tools occupied only a small portion of the assemblages.
 As stemmed points and blades increased in the toolkits, both choppers and side scrapers went through minor decreases but still remained abundant.
@@
 Our data also show that quartz, quartz vein and quartzite were consistently dominant and the use of fine-grained materials including chert, hornfels, rhyolite, and shale, increased after 41.5 ka, around the time that stemmed points appear.
-Projectile points, such as stemmed points, need high quality materials which enable high flaking quality, durability and effectiveness of an edge, as well as the creation of well-defined outline forms [@bamforth2009projectile].
+Projectile points, such as stemmed points, benefit from high quality materials which enable high flaking quality, durability and effectiveness of an edge, as well as the creation of well-defined outline forms [@bamforth2009projectile].
 The increased use of finer-grained raw materials likely changed forager mobility patterns to ensure their movements over the landscape supplied them with the materials needed to make stemmed points.
 
-## Environmental and demographic contexts {-}
+## Environmental and Demographic Contexts {-}
 
-We hypothesized that the introduction of stemmed points might reflect a preference for more portable and efficient hunting tools, as an adaptation to environmental or population changes.
-As the first composite tool appearing on the Korean Peninsula, stemmed points represent a major change in stone artefact technology.
+We hypothesized that the introduction of stemmed points might reflect a preference for more portable and efficient hunting tools, as an adaptation to environmental or demographic changes.
+As the first composite tool appearing on the Korean Peninsula, stemmed points represent a major change in stone artifact technology.
 An important quality of composite tools is that the user can easily replace damaged parts, contributing to an increased maintainability of the tool [@kuhn2015artifacts; @cardillo2010some].
-Combining the stone projective with a wooden shaft further materials enhances functionality of the tool such as improving penetration by increasing weight and sharpness [@browne1940projectile].
-We expected that the functional advantage of the new composite tool might have allowed foragers to be more active in less productive landscapes.
-Previous studies simply describe the environment of the research period as overall cooler and drier during MIS 3 but lack detail due to a lack of direct paleoenvironmental proxies in Korea [@Han_2008; @seong2008tanged; @Choi_2011; @chang2013human; @im2015study; @bak2017late].
-To explore environmental and demographic context related to the technological transition, we examined the distribution of site elevations, and simulated MAT during the Late Pleistocene.
-The results show that groups with stemmed points reached higher elevations, where the MAT was lower.
-This supports our hypothesis that stemmed points supported occupation of more marginal habitats.
+Combining the stone projective with a wooden shaft further materials enhances functionality of the tool by improving penetration by increasing weight and sharpness [@browne1940projectile].
+We expected that the functional advantage of the new composite tool would have allowed foragers to be more effective in less productive landscapes.
+To explore the environmental and demographic contexts related to the technological transition, we examined the distribution of site elevations, and simulated MAT during the Late Pleistocene.
+The results show that groups with stemmed points generally reached higher elevations, where the MAT was generally lower.
+This supports our hypothesis that stemmed points were associated with the occupation of more marginal habitats.
 Through MIS 3 the temperature gradually decreased until the LGM, suggesting that stemmed points may have been part of a suite of adaptations to cooler temperatures.
@@
 
-We investigated forager population dynamics to assess if population change might be a relevant mechanism to explain the appearance of stemmed points.
-The output of our SPD models indicate population increasing until 35 ka and remaining at the peak with minor deviations for about 3,000 years.
+We investigated forager population dynamics to assess if demographic change might be a relevant mechanism to explain the appearance of stemmed points.
+The output of our SPD models indicate population increasing until 35 ka and remaining at the peak with minor deviations and then declining after 31 ka.
 If population dynamics were a key driver, then we expect population to be high at the time the new technology first appears.
@@
 Our results are partly consistent with this scenario, showing a gradual increase in population after the appearance of stemmed points.
+Given the pattern we observe in the timing of the appearance of stemmed points preceding an increase in population, we might speculate that their appearance was a mechanism enabling an increase in population by making marginal habitats more viable.
 
@@
 
-To understand the change in mobility and site occupation related to the appearance of new technologies during the late Pleistocene in Korea, we examined stone artefacts from 22 sites dated to 49-24 ka.
+To understand the change in mobility and site occupation related to the appearance of new technologies during the late Pleistocene in Korea, we examined stone artifacts from 22 sites dated to 49-24 ka.
 The results show that the forager groups with stemmed points tended to be located at higher elevations.
 The temperature gradually decreased during the technological transition and after the appearance of stemmed points.
-Overall, the result could indicate the use of stemmed points was associated with occupation of marginal or extreme environments.
+Overall, the result indicates that the use of stemmed points was associated with occupation of marginal or extreme environments.
 In addition, the groups with stemmed points had expedient technologies which reflect residential and less mobile land use patterns.
@@
 Our findings are limited by the relatively small number of sites and radiocarbon ages currently available The absence of detailed stratigraphic data limit the chronological resolution for the assemblages in our sample.
-Furthermore, we currently lack functional data on individual artefact assemblages to support our conclusions about site use.
-With future work we may be better able to distinguish a high density of stone artefacts as results from a long-term/small group occupation versus a short-term/large group occupation.
+Furthermore, we currently lack functional data on individual artifact assemblages to support our conclusions about site use.
+With future work we may be better able to distinguish a high density of stone artifacts as results from a long-term/small group occupation versus a short-term/large group occupation.
 Further research should focus on detailed assemblage analysis at each site to better understand foragers mobility and site use strategies.

### 2021-04-12 analysis/paper/paper.Rmd
comment out some code in Rmd for fast knitting

--- 2021-04-01 analysis/paper/paper.Rmd
@@
 quietly(source(here::here("analysis/paper/004-artifact-composition.R")))
-knitr::include_graphics(here::here("analysis/figures/004-artefact-types.png"))
+knitr::include_graphics(here::here("analysis/figures/004-artifact-types.png"))
 ```
@@
 The blue line is a linear regression on elevation and MAT with the grey area showing the 95% confidence interval.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/007-climate-model-plots.R"))}
-quietly(source(here::here("analysis/paper/007-climate-model-plots.R")))
+# quietly(source(here::here("analysis/paper/007-climate-model-plots.R")))
 knitr::include_graphics(here::here("analysis/figures/007-climate-model-sites-panel-plot.png"))
@@
 In this case, the linear model is the best fit, with the next-best fit (exponential) having a higher AIC score by a minor margin at -48.6.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/008-summed-probability-distributions.R"))}
-quietly(source(here::here("analysis/paper/008-summed-probability-distributions.R")))
+# quietly(source(here::here("analysis/paper/008-summed-probability-distributions.R")))
 knitr::include_graphics(here::here("analysis/figures/008-summed-probability-distribution-models.png"))

### 2021-04-12 analysis/paper/paper.Rmd
resolve merge conflict

--- 2021-04-12 analysis/paper/paper.Rmd
@@
 In this case, the linear model is the best fit for the observed data", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/008-summed-probability-distributions.R"))}
-quietly(source(here::here("analysis/paper/008-summed-probability-distributions.R")))
+# quietly(source(here::here("analysis/paper/008-summed-probability-distributions.R")))
 knitr::include_graphics(here::here("analysis/figures/008-summed-probability-distribution-models.png"))

### 2021-04-12 analysis/paper/paper.Rmd
add some inline R code

--- 2021-04-12 analysis/paper/paper.Rmd
@@
 
+```{r korea-map, fig.cap="Korean Paleolithic sites mentioned in the text.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/001-draw-the-map.R"))}
+quietly(source(here::here("analysis/paper/001-draw-the-map.R")))
+knitr::include_graphics(here::here("analysis/figures/001-site-map.png"))
+```
+
 After the first excavation of a Paleolithic site in Korea at Seokjang-ri site in the 1960s, more than 200 Paleolithic sites have been discovered in South Korea [@lee2019were].
 We selected sites dated to before and after the transition period (42-35 ka) to analyze the process of technological change.
-This resulted in a sample of 33 assemblages from 22 sites spanning 49-24 ka (Figure \@ref(fig:korea-map)).
+This resulted in a sample of 33 assemblages from `r nrow(korean_archaeological_site_locations)` sites spanning 49-24 ka (Figure \@ref(fig:korea-map)).
 We identified multiple assemblages in a site where culturally sterile deposits separated artifact-bearing deposits, or where stratigraphic units could be identified by major differences in the texture and composition of the sedimentary deposit containing the stone artifacts.
@@
 We obtained the data from published excavation reports held by provincial museums.
-
-```{r korea-map, fig.cap="Korean Paleolithic sites mentioned in the text.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/001-draw-the-map.R"))}
-quietly(source(here::here("analysis/paper/001-draw-the-map.R")))
-knitr::include_graphics(here::here("analysis/figures/001-site-map.png"))
-```
 
@@
 
+```{r spd-model-prep,, cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/008-summed-probability-distributions.R"))}
+quietly(source(here::here("analysis/paper/008-summed-probability-distributions.R")))
+```
+
 We used radiocarbon dates as a proxy for human population to infer the demographic context of technological change.
@@
 Although this is often used with samples of thousands of ages, this method can also be useful for small sample sizes.
-For instance, @timpson2014reconstructing apply SPD to 93 ages from Eastern Middle Sweden and demonstrate that the result is equivalent to a larger sample size (n= 243).
-We used 108 radiocarbon ages from 22 sites to examine connections between population trends during the Late Pleistocene in Korea and technological innovation.
+For instance, @timpson2014reconstructing apply SPD to 93 ages from Eastern Middle Sweden and demonstrate that the result is equivalent to a larger sample size (n = 243).
+We used `r nrow(dates_clean)` radiocarbon ages from `r nrow(korean_archaeological_site_locations)` sites to examine connections between population trends during the Late Pleistocene in Korea and technological innovation.
 
@@
 
-Figure \@ref(fig:site-elevation) shows that people occupied sites located at similar elevations, around 0 to 200 meters above sea level.
-As we saw for artifact density, there is an increase in the upper bound of site elevation in more recent times.
-The most striking pattern in the elevation data is the substantial difference in elevation between sites with stemmed points and sites without.
-The distribution of elevations of sites without stemmed points is generally lower than the elevations of sites with stemmed points.
-Although not a statistically significant difference in elevations (p = 0.102), this may indicate that forager groups who used stemmed points were generally more able to occupy higher altitudes, while the groups without stemmed points tended to prefer lower altitudes.
-According to our locally weighted regression, sites without stemmed points reach a maximum elevation of about 150 m at around 39 ka.
-Sites containing stemmed points reach a maximum elevation of 175 m at 28 ka.
-
 ```{r site-elevation, fig.cap="The site elevation by the age of the deposit.
@@
 
-Among the sites in our sample there is a variation of about 5 degrees in the mean annual temperature (MAT), mostly controlled by elevation (Figure \@ref(fig:climate-model)).
-Through MIS 3, the temperature gradually decreased until the Last Glacial Maximum (LGM, 26.5-20 ka) (Clark et al., 2009).
-The MAT of MIS 2, including LGM, was relatively stable.
-The temperature increased again from late MIS 2 towards MIS 1.
-Compared to the east side of the Korean Peninsula, the west side tends to be relatively warmer.
-The range of MAT at the sites over our study period is 2-10℃.
-Gihwari and Gigok sites have the lowest (3℃) and Jeungsan-ri site has the highest MAT (8℃).
-The mean MAT for all sites fluctuated within 4℃ between 50 ka and 10 ka.
-The first appearance of stemmed points occurred in the middle of the decreasing MAT trend in MIS 3, at 40-35 ka.
-Figure 7D shows a negative relationship between temperature and elevation (p = 0.073).
-For example, Gihwari Cave site, with one of the lowest MAT distributions, is located in the highest elevation around 300m above sea level.
-Jeungsanri is located at a lower elevation and has one of the highest MAT distributions.
+Figure \@ref(fig:site-elevation) shows that people occupied sites located at similar elevations, around 0 to 200 meters above sea level.
+As we saw for artifact density, there is an increase in the upper bound of site elevation in more recent times.
+The most striking pattern in the elevation data is the substantial difference in elevation between sites with stemmed points and sites without.
+The distribution of elevations of sites without stemmed points is generally lower than the elevations of sites with stemmed points.
+Although not a statistically significant difference in elevations (`r apa::apa(elev_sp_ttest)`), this may indicate that forager groups who used stemmed points were generally more able to occupy higher altitudes, while the groups without stemmed points tended to prefer lower altitudes.
+According to our locally weighted regression, sites without stemmed points reach a maximum elevation of about 150 m at around 39 ka.
+Sites containing stemmed points reach a maximum elevation of 175 m at 28 ka.
 
@@
 
+Among the sites in our sample there is a variation of about 5 degrees in the mean annual temperature (MAT), mostly controlled by elevation (Figure \@ref(fig:climate-model)).
+Through MIS 3, the temperature gradually decreased until the Last Glacial Maximum (LGM, 26.5-20 ka) (Clark et al., 2009).
+The MAT of MIS 2, including LGM, was relatively stable.
+The temperature increased again from late MIS 2 towards MIS 1.
+Compared to the east side of the Korean Peninsula, the west side tends to be relatively warmer.
+The range of MAT at the sites over our study period is 2-10℃.
+Gihwari and Gigok sites have the lowest (3℃) and Jeungsan-ri site has the highest MAT (8℃).
+The mean MAT for all sites fluctuated within 4℃ between 50 ka and 10 ka.
+The first appearance of stemmed points occurred in the middle of the decreasing MAT trend in MIS 3, at 40-35 ka.
+Figure 7D shows a negative relationship between temperature and elevation.
+For example, Gihwari Cave site, with one of the lowest MAT distributions, is located in the highest elevation around 300m above sea level.
+Jeungsanri is located at a lower elevation and has one of the highest MAT distributions.
+
 ## Demographic Context {-}
@@
 Figure \@ref(fig:spd-model) shows the summed probability distribution (SPD) of the radiocarbon ages from the sites in our sample overlaid with distributions (indicated by the grey envelopes) generated from four models of demographic change for the period 50 ka to 10 ka (exponential, uniform, linear, and logistic).
-The statistical significance of each of the models we fit to the observed SPD was computed using 200 simulations, and the global p-values are 0.004, 0.004, 0.009, and 0.004, respectively.
-These low p-values indicates significant deviations in the observed SPD from the modeled distributions.
+The fit of models to our observed SPD was computed using 1000 simulations, and the resulting global p-values indicate significant deviations in the observed SPD from the modeled distributions.
 Overall, the linear model is the best fit with the observed SPD, having the lowest AIC value of the three models where we could compute this.
@@
 
-```{r spd-model, fig.cap="Summed probability distribution of 108 radiocarbon dates.
+```{r spd-model, fig.cap=paste0("Summed probability distribution of ", nrow(dates_clean), " radiocarbon dates.
 The black solid line represents actual radiocarbon ages.
@@
 The best fit model has the lowest AIC score.
-In this case, the linear model is the best fit for the observed data", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/008-summed-probability-distributions.R"))}
-quietly(source(here::here("analysis/paper/008-summed-probability-distributions.R")))
+In this case, the linear model is the best fit for the observed data"), cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/009-summed-probability-distributions.R"))}
+quietly(source(here::here("analysis/paper/009-summed-probability-distributions.R")))
 knitr::include_graphics(here::here("analysis/figures/008-summed-probability-distribution-models.png"))

### 2021-04-12 analysis/paper/008-summed-probability-distributions.R

@@ -15,415 +15,3 @@ dates_clean <-
          age >= 10000,
          method == 'AMS')
 
-# devtools::install_github('ahb108/rcarbon')
-library(rcarbon)
-dates_calibrated <-
-  calibrate(x = dates_clean$age,
-            errors = dates_clean$error,
-            calCurves = 'intcal20')
-
-dates_calibrated_spd <-
-  spd(dates_calibrated,
-      timeRange = c(50000, 10000))
-# plot(dates_calibrated_spd)
-
-# Testing Observed SPDs against theoretical models
-## recalibrate dates without normalisation to avoid artificial peaks
-
-spd_dates <-
-  calibrate(
-    x = dates_clean$age,
-    errors = dates_clean$error,
-    normalised = FALSE,
-    verbose = F
-  )
-
-#-------------------------------------------------
-### this section of code takes long time, fyi
-nsim = 10000 # set to 10 while developing, then to 10000 for the final run
-
-# exponential model
-exp_test <-
-  modelTest(
-    spd_dates,
-    errors = dates_clean$error,
-    timeRange = c(50000, 10000),
-    model = 'exponential',
-    nsim = nsim,
-    ncores = 3
-  )
-
-# uniform model, does not produce anything for $fitobject
-uni_test <-
-  modelTest(
-    spd_dates,
-    errors = dates_clean$error,
-    timeRange = c(50000, 10000),
-    model = 'uniform',
-    nsim = nsim,
-    ncores = 3
-  )
-
-
-# linear model
-lin_test <-
-  modelTest(
-    spd_dates,
-    errors = dates_clean$error,
-    timeRange = c(50000, 10000),
-    model = 'linear',
-    nsim = nsim,
-    ncores = 3
-  )
-
-
-# logistical model
-
-# rcarbon pkg version of logistical growth model, accprding to instructions here
-# https://cran.r-project.org/web/packages/rcarbon/vignettes/rcarbon.html#testing-against-custom-growth-models
-
-ko.bins = binPrep(sites=dates_clean$ID,
-                  ages=dates_clean$age,
-                  h=100)
-
-# Generate a smoothed SPD
-ko.spd.smoothed = spd(spd_dates,
[…]

### 2021-04-12 analysis/paper/009-summed-probability-distributions.R

@@ -1,0 +1,418 @@
+suppressPackageStartupMessages(library(tidyverse))
+library(glue)
+library(ggrepel)
+library(here)
+
+
+# devtools::install_github('ahb108/rcarbon')
+library(rcarbon)
+dates_calibrated <-
+  calibrate(x = dates_clean$age,
+            errors = dates_clean$error,
+            calCurves = 'intcal20')
+
+dates_calibrated_spd <-
+  spd(dates_calibrated,
+      timeRange = c(50000, 10000))
+# plot(dates_calibrated_spd)
+
+# Testing Observed SPDs against theoretical models
+## recalibrate dates without normalisation to avoid artificial peaks
+
+spd_dates <-
+  calibrate(
+    x = dates_clean$age,
+    errors = dates_clean$error,
+    normalised = FALSE,
+    verbose = F
+  )
+
+#-------------------------------------------------
+### this section of code takes long time, fyi
+nsim = 10 # set to 10 while developing, then to 1000 for the final run
+
+# exponential model
+exp_test <-
+  modelTest(
+    spd_dates,
+    errors = dates_clean$error,
+    timeRange = c(50000, 10000),
+    model = 'exponential',
+    nsim = nsim,
+    ncores = 3
+  )
+
+# uniform model, does not produce anything for $fitobject
+uni_test <-
+  modelTest(
+    spd_dates,
+    errors = dates_clean$error,
+    timeRange = c(50000, 10000),
+    model = 'uniform',
+    nsim = nsim,
+    ncores = 3
+  )
+
+
+# linear model
+lin_test <-
+  modelTest(
+    spd_dates,
+    errors = dates_clean$error,
+    timeRange = c(50000, 10000),
+    model = 'linear',
+    nsim = nsim,
+    ncores = 3
+  )
+
+
+# logistical model
+
+# rcarbon pkg version of logistical growth model, accprding to instructions here
+# https://cran.r-project.org/web/packages/rcarbon/vignettes/rcarbon.html#testing-against-custom-growth-models
+
+ko.bins = binPrep(sites=dates_clean$ID,
+                  ages=dates_clean$age,
+                  h=100)
+
+# Generate a smoothed SPD
+ko.spd.smoothed = spd(spd_dates,
[…]

### 2021-04-12 analysis/paper/paper.Rmd
replace some hard coded values with apa::apa()

@@
 
-Figure \@ref(fig:volume-density-retouch) shows a strong pattern of artifacts from less dense assemblages having higher retouch frequencies while more dense assemblages have lower retouch frequencies (p=0.043), further showing a spectrum of site functions in this sample.
-As we saw for artifact density, there are no clear directional chronological trends in retouch frequencies.
-Assemblages containing stemmed points tend to have substantially fewer retouched pieces compared to assemblages without stemmed points (p=0.011).
-
 ```{r volume-density-retouch, fig.cap="Relationship between artifact density and retouch frequency.
@@
 ```
+
+Figure \@ref(fig:volume-density-retouch) shows a strong pattern of artifacts from less dense assemblages having higher retouch frequencies while more dense assemblages have lower retouch frequencies (`r apa::apa(den_sp_ttest)`), further showing a spectrum of site functions in this sample.
+As we saw for artifact density, there are no clear directional chronological trends in retouch frequencies.
+Assemblages containing stemmed points tend to have substantially fewer retouched pieces compared to assemblages without stemmed points (`r apa::apa(retouch_sp_ttest)`).
 

### 2021-04-12 analysis/paper/paper.Rmd
edit osf link and citation

@@
 
-The entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is included in the Supplementary Online Materials at http://doi.org/osf.io/yagd2/ to enable re-use of materials and improve reproducibility and transparency [@Marwick2017].
+The entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is included in the Supplementary Online Materials at https://osf.io/jxwy5/ to enable re-use of materials and improve reproducibility and transparency [@Marwick2017].
 Also in this version-controlled compendium [@marwick2018packaging] are the raw data for all the visualizations and tests reported here.

### 2021-04-12 analysis/paper/paper.Rmd
Wrote acknowledgements. Added missed codes (moved to the new spd script) to the main spd script to run all the codes

@@
 
-<!--The following line inserts a page break when the output is MS Word. For page breaks in PDF, use \newpage on its own line.  -->
+We thank Yongjoon Chang, Seonsook Park, Jaeyoung Kim, Soonmi Lee, Taeyoung Kim, and the staff in Korean national museums, and Sangtaek Lim, Chuntaek Seong, Sangmin Cha, Sooa Kim, Gigil Lee, Myeongjin Choi, and Chanho Park in local archaeological institutions and University museums in Korea for their assistance and support in accessing to the data used in this paper and accommodating our trip to their collections.
+This work was supported by grant award of Quaternary Research Center and travel grands from the Department of Anthropology and Center for Statistics and the Social Sciences at University of Washington.
+We appreciate insightful feedback on an early draft from Ben Fitzhugh and Peter Lape.
 

### 2021-04-12 analysis/paper/009-summed-probability-distributions.R

@@ -3,6 +3,17 @@ library(glue)
 library(ggrepel)
 library(here)
 
+dates <-
+  read.csv(here("analysis/data/raw_data/Radiocarbon dates.csv"))
+
+dates_clean <-
+  dates %>%
+  mutate(age = parse_number(as.character(age)),
+         error = parse_number(as.character(error))) %>%
+  # filter age range here
+  filter(age < 50500,
+         age >= 10000,
+         method == 'AMS')
 
 # devtools::install_github('ahb108/rcarbon')
 library(rcarbon)

### 2021-04-12 analysis/paper/009-summed-probability-distributions.R
Tried to put pvalue for linear model. The spd plot wasn't saved

@@ -2,6 +2,7 @@ suppressPackageStartupMessages(library(tidyverse))
 library(glue)
 library(ggrepel)
 library(here)
+library(cowplot)
 
 dates <-
   read.csv(here("analysis/data/raw_data/Radiocarbon dates.csv"))
@@ -118,10 +119,10 @@ log_test  <- modelTest(spd_dates,
 
 
 # look at p-values for the models
-exp_test$pval
-uni_test$pval
-lin_test$pval
-log_test$pval
+exp_test$pval #0.09090909
+uni_test$pval #0.09090909
+lin_test$pval #0.09090909
+log_test$pval #0.09090909
 # they are all the same!
 
 ## rank models by AIC score with subset of ages
@@ -192,7 +193,7 @@ ymax <- max(ko.spd.smoothed$grid$PrDens)
 
 
 # define a function that emits the desired plot
-p1 <- function() {
+#p1 <- function() {
 
 # begin drawing the plot
 layout(
@@ -241,6 +242,15 @@ text(
   adj = c(0, 0),
   labels = paste0("AIC = ", round(aic_lin, digits = 0))
 )
+text(
+  x = 41000,
+  ymax * 0.80,
+  cex = 0.6,
+  font = 2,
+  adj = c(0, 0),
+  labels = paste0("P-value = ", round(lin_test$pval, digits = 6))
+)
+
 
 
 box()
@@ -394,9 +404,9 @@ mtext(
   font = 2,
   cex = 0.8
 )
-}
+#}
 
-ggdraw(p1)
+#ggdraw(p1)
 
 ggsave(here::here("analysis/figures/008-summed-probability-distribution-models.png"),
        h = 15,

### 2021-04-12 analysis/paper/009-summed-probability-distributions.R
Uncommented p1 and ggdraw code

@@ -193,7 +193,7 @@ ymax <- max(ko.spd.smoothed$grid$PrDens)
 
 
 # define a function that emits the desired plot
-#p1 <- function() {
+p1 <- function() {
 
 # begin drawing the plot
 layout(
@@ -248,7 +248,7 @@ text(
   cex = 0.6,
   font = 2,
   adj = c(0, 0),
-  labels = paste0("P-value = ", round(lin_test$pval, digits = 6))
+  labels = paste0("p = ", round(lin_test$pval, digits = 3))
 )
 
 
@@ -404,9 +404,9 @@ mtext(
   font = 2,
   cex = 0.8
 )
-#}
+}
 
-#ggdraw(p1)
+ggdraw(p1)
 
 ggsave(here::here("analysis/figures/008-summed-probability-distribution-models.png"),
        h = 15,

### 2021-04-13 analysis/paper/009-summed-probability-distributions.R
added p values for spd figure

@@ -299,7 +299,14 @@ text(
   labels = paste0("AIC = ", round(aic_exp, digits = 0))
 )
 
-
+text(
+  x = 41000,
+  ymax * 0.80,
+  cex = 0.6,
+  font = 2,
+  adj = c(0, 0),
+  labels = paste0("p = ", round(exp_test$pval, digits = 3))
+)
 
 box()
 
@@ -346,6 +353,14 @@ text(
   adj = c(0, 0),
   labels = paste0("AIC = ", round(aic_log, digits = 0))
 )
+text(
+  x = 41000,
+  ymax * 0.80,
+  cex = 0.6,
+  font = 2,
+  adj = c(0, 0),
+  labels = paste0("p = ", round(log_test$pval, digits = 3))
+)
 
 box()
 
@@ -384,6 +399,16 @@ text(
   cex = 0.6,
   adj = c(0, 0.7)
 )
+
+text(
+  x = 41000,
+  ymax * 0.80,
+  cex = 0.6,
+  font = 2,
+  adj = c(0, 0),
+  labels = paste0("p = ", round(uni_test$pval, digits = 3))
+)
+
 box()
 
 

### 2021-04-14 analysis/paper/paper.Rmd
knit ok with SPD figure

@@
 
-Figure \@ref(fig:volume-density-retouch) shows a strong pattern of artifacts from less dense assemblages having higher retouch frequencies while more dense assemblages have lower retouch frequencies (`r apa::apa(den_sp_ttest)`), further showing a spectrum of site functions in this sample.
+Figure \@ref(fig:volume-density-retouch) shows a strong pattern of artifacts from less dense assemblages having higher retouch frequencies while more dense assemblages have lower retouch frequencies (`r den_sp_ttest_str`), further showing a spectrum of site functions in this sample.
 As we saw for artifact density, there are no clear directional chronological trends in retouch frequencies.
-Assemblages containing stemmed points tend to have substantially fewer retouched pieces compared to assemblages without stemmed points (`r apa::apa(retouch_sp_ttest)`).
+Assemblages containing stemmed points tend to have substantially fewer retouched pieces compared to assemblages without stemmed points (`r retouch_sp_ttest_str`).
 
@@
 The distribution of elevations of sites without stemmed points is generally lower than the elevations of sites with stemmed points.
-Although not a statistically significant difference in elevations (`r apa::apa(elev_sp_ttest)`), this may indicate that forager groups who used stemmed points were generally more able to occupy higher altitudes, while the groups without stemmed points tended to prefer lower altitudes.
+Although not a statistically significant difference in elevations (`r elev_sp_ttest_str`), this may indicate that forager groups who used stemmed points were generally more able to occupy higher altitudes, while the groups without stemmed points tended to prefer lower altitudes.
 According to our locally weighted regression, sites without stemmed points reach a maximum elevation of about 150 m at around 39 ka.
@@
 In this case, the linear model is the best fit for the observed data"), cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/009-summed-probability-distributions.R"))}
-quietly(source(here::here("analysis/paper/009-summed-probability-distributions.R")))
+# quietly(source(here::here("analysis/paper/009-summed-probability-distributions.R")))
 knitr::include_graphics(here::here("analysis/figures/008-summed-probability-distribution-models.png"))

### 2021-04-14 analysis/paper/009-summed-probability-distributions.R

@@ -41,7 +41,7 @@ spd_dates <-
 
 #-------------------------------------------------
 ### this section of code takes long time, fyi
-nsim = 10 # set to 10 while developing, then to 1000 for the final run
+nsim = 1000 # set to 10 while developing, then to 1000 for the final run
 
 # exponential model
 exp_test <-

### 2021-04-14 analysis/paper/paper.Rmd
Fixed wrong p value on rmd file

@@
 
-Figure \@ref(fig:volume-density-retouch) shows a strong pattern of artifacts from less dense assemblages having higher retouch frequencies while more dense assemblages have lower retouch frequencies (`r den_sp_ttest_str`), further showing a spectrum of site functions in this sample.
+Figure \@ref(fig:volume-density-retouch) shows a strong pattern of artifacts from less dense assemblages having higher retouch frequencies while more dense assemblages have lower retouch frequencies (`r combined_retouch_density_ttest_str`), further showing a spectrum of site functions in this sample.
 As we saw for artifact density, there are no clear directional chronological trends in retouch frequencies.

### 2021-04-14 analysis/paper/003-artifact-volumetric-density-retouch-frequency.R

@@ -94,10 +94,16 @@ kas_sites_retouch_density_corr <-
   cor.test(Assemblage_info_retouch_density_ages_prop$prop_retouched,
            Assemblage_info_retouch_density_ages_prop$artefact_density)
 
-r_value <- unname(kas_sites_retouch_density_corr$estimate)
-p_value <- unname(kas_sites_retouch_density_corr$p.value )
-t_value <- unname(kas_sites_retouch_density_corr$statistic )
-df_value <- unname(kas_sites_retouch_density_corr$parameter )
+r_value <- round(unname(kas_sites_retouch_density_corr$estimate), 3)
+p_value <- round(unname(kas_sites_retouch_density_corr$p.value ), 3)
+t_value <- round(unname(kas_sites_retouch_density_corr$statistic ), 3)
+df_value <- round(unname(kas_sites_retouch_density_corr$parameter ), 3)
+
+
+# t(degress of freedom) = the t statistic, p = p value.
+combined_retouch_density_ttest_str <-
+  paste0("t(", df_value, ") = ", t_value, ", p = ", p_value)
+
 
 size <-  6
 Assemblage_info_retouch_density_ages_prop_main_plot <-

### 2021-04-15 analysis/paper/003-artifact-volumetric-density-retouch-frequency.R
First commit to check my new computer

@@ -104,7 +104,6 @@ df_value <- round(unname(kas_sites_retouch_density_corr$parameter ), 3)
 combined_retouch_density_ttest_str <-
   paste0("t(", df_value, ") = ", t_value, ", p = ", p_value)
 
-
 size <-  6
 Assemblage_info_retouch_density_ages_prop_main_plot <-
   ggplot(Assemblage_info_retouch_density_ages_prop,

### 2021-09-03 analysis/paper/paper.Rmd
Editing the manuscipt based on Reviewer2's comments

@@
 If these artifacts were projectile weapon tips, then stemmed points represent the first appearance of hafting technology in Korea.
-Most Korean Paleolithic sites do not preserve organic remains, so we lack direct fauna evidence on hunting and subsistence strategies, and stone artifacts are our only window into these behaviors.
+Most open sites in Korea do not preserve organic remains due to acidic soil.
+There are a few limestone cave sites containing human fossils or faunal remains (e.g.
+Durubong and Jeommal Cave), which located in constrained area to access in North Korea or have been rarely analyzed to directly apply to other archaeological research[@bae2013paleolithic].
+Therefore stone artifacts are our only window to examine hunting and subsistence strategies.
 Here we use a behavioral ecological framework to explore questions about the implications of the appearance of this new technology.
@@
 
-Previous work on this technological change in Korea has focused on issues of the transition from the Early to the Late Paleolithic and modern human dispersals [@norton2009evolution; @bae2010origin].
+Previous work on this technological change in Korea has focused on issues of the transition from the Early to the Late Paleolithic and modern human dispersals [@norton2009evolution; @bae2010origin; @bae2017late].
 To our knowledge, the implications of the appearance of stemmed points for forager mobility and site use have not been explored in detail.
 Here, we attempt to identify patterns of mobility and site occupation through analysis of the stone artifact assemblages using Human Behavioral Ecology (HBE) concepts.
-Drawing on established relationships between lithic technology and forager mobility [@shott1986settlement; @Hiscock_1994; @kuhn1994formal; @kuhn2004upper; @kuhn2015artifacts; @kuhn2016moves; @capriles2018mobility], we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that were part of a broader strategy of moving frequently and further, possibly as an adaptation to environmental or population changes.
+Drawing on established relationships between lithic technology and forager mobility [@shott1986settlement; @Hiscock_1994; @kuhn1994formal; @kuhn2004upper; @kuhn2015artifacts; @kuhn2016moves; @capriles2018mobility; @seong2015diversity; @seong2008tanged], we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that were part of a broader strategy of moving frequently and further, possibly as an adaptation to environmental or population changes.
 
@@
 In the Eastern hemisphere, the most striking technological innovations during the Late Pleistocene are found in Northeast Asia, such as northern China, Japan and Korea.
-These innovations include blade technology, high frequencies of retouched blade tools, several novel tools such as projectile points, stemmed points, end scrapers, burins, denticulates, etc, and frequent use of high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary].
-Previous work has argued that the appearance of new stone artifact technologies in this region may be linked to modern human dispersals [@seong2008tanged; @bae2010origin; @bae2017origin].
+These innovations include blade technology, high frequencies of retouched blade tools, several novel tools such as projectile points, stemmed points, end scrapers, burins, denticulates, ground stone tools etc, and frequent use of high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary; @bae2017late; @lee2017korean].
+Previous work has argued that the appearance of new stone artifact technologies in this region may be linked to modern human dispersals [@seong2008tanged; @bae2010origin; @bae2012nature; @bae2017origin].
+For example, the new dating results from Zhoukoudian Upper Cave from northern China (35.1–33.5 ka) and material evidence such as ornaments and hominin fossil show close relation to one of multiple dispersal events out of Africa into Eurasia    [@li2018re].
 
@@
 Shuidonggou Locality SDG 2, which dates to around ca.
-32 ka, contains fauna specimens, ostrich eggshell beads, and stone artifacts including blades, retouched flakes, cores, and debris along with well-preserved hearths.
+32 ka, contains faunal specimens, ostrich eggshell beads, and stone artifacts including blades, retouched flakes, cores, and debris along with well-preserved hearths.
 In Locality 9 dated to ca.
 29 ka, there is a small scatter of stone artifacts consisting of blades, Levallois flakes, cores and other retouched flakes.
-Excavations Locality 12 shows that its archaeological context, dated to 13,078--13,296 cal BP, includes ground stone and more than 30,000 microlithics made out of a variety of raw materials such as fine-grained and siliceous rocks.
+Excavations at Locality 12 shows that its archaeological context, dated to 13,078--13,296 cal BP, includes ground stone and more than 30,000 microlithics made out of a variety of raw materials such as fine-grained and siliceous rocks.
 In addition, more than 10,000 animal fossils and bone tools including a tool for fishing nets, an awl, and two needles were excavated [@pei2012shuidonggou; @gao2014discovery; @zhang2016bone].
@@
 A total of 500 stone artifacts were discovered including trapezoids, side scrapers, flakes, flake cores, edge-ground axes, and cobble tools.
-The main lithic raw materials are Chert and Andesite [@izuho201421].
+The main lithic raw materials are chert and andesite [@izuho201421].
 Stemmed points appear in Japan much later than in Korea, dated to between about 15,500 and 13,800 cal BP [@ono2002radiocaron; @tsutsumi2007dynamics].
@@
 
-Our study of raw material consumption patterns assumes that tool stone was transported and consumed to optimize mobility and minimize risks of not having artifacts when they were needed [@andrefsky1994raw; @brantingham2000raw; @brantingham2003neutral; @yue2020lithic].
+Our study of raw material consumption patterns assumes that stone tool was transported and utilized to optimize mobility and minimize risks of not having artifacts when they were needed [@andrefsky1994raw; @brantingham2000raw; @brantingham2003neutral; @yue2020lithic].
 The representation of raw materials in an assemblage can be an indicator of site occupation patterns.
@@
 The inset figure shows that assemblages with stemmed points have higher artifact densities than the assemblages without.
-With the newer tools, hunter-gatherers may have stayed in the same location for longer durations.
-We may be seeing site occupation patterns in this sample that represent the long term trend of the use of both base camps or residences, represented by the artifact higher densities, and more briefly occupied camps with lower densities.
+With the newer tools, hunter-gatherers may have stayed in the same location for longer duration.
+We may be seeing site occupation patterns in this sample that represent the long term trend of the use of both base camps or residences, represented by the higher artifact densities, and more briefly occupied camps with lower densities.
 The most striking result here is that maximum size of base camps increases in more recent periods.
@@
 Previous work on toolkit composition during the Late Pleistocene of Korea attempted to divide it up into multiple chronological sub-periods.
-Seong divides five successive assemblage types for the Late/Upper Paleolithic in Korea: (1) quartzite and quartz vein artifacts; (2) mostly small quartzite and quartz vein artifacts with some large artifacts such as cores and choppers; (3) stemmed points dominant; (4) typical blade assemblages including stemmed points; and (5) microblade after 30 ka [@seong2015diversity].
+Seong divides five successive assemblage types for the Late Paleolithic in Korea: (1) quartzite and quartz vein artifacts; (2) mostly small quartzite and quartz vein artifacts with some large artifacts such as cores and choppers; (3) stemmed points dominant; (4) typical blade assemblages including stemmed points; and (5) microblade after 30 ka [@seong2015diversity].
 Lee focuses on the Honam region in southwestern Korea and divides the same time frame into two phases; 'core tool industry with large flake knapping' during early MIS 3 (59−40 ka), and 'blade industry with tanged point chiefly' during late MIS (40−24 ka) [@lee2012characteristics].

### 2021-09-23 analysis/paper/007-climate-model-plots.R
commit to pull

@@ -60,8 +60,10 @@ site_locations_tbl %>%
   filter(between(year, -50000, -10000))
 
 # save this for later since it takes ages to get this out of the ncdf file!
-write_csv(site_locations_tbl_temps,
+write.csv(site_locations_tbl_temps,
           here::here("analysis/data/site_locations_tbl_temps.csv"))
+
+
 
 #----------------------------------------------------------------------
 base_size = 6

### 2021-09-24 analysis/paper/002-artifact-volumetric-density.R
Fixed ggsave background color issue

@@ -96,7 +96,10 @@ library(cowplot)
 ggdraw(density_sp_main_plot) +
   draw_plot(density_sp_sub_plot,
             .37, .62,
-            .32, .33)
+            .32, .33) +
+  theme(panel.background = element_rect(fill='white', colour="white"),
+                               plot.background = element_rect(fill='white', colour="white"))
+
 
 ggsave(here::here("analysis/figures/002-age-by-density.png"),
        width = 4.45,

### 2021-09-24 analysis/paper/003-artifact-volumetric-density-retouch-frequency.R

@@ -155,7 +155,10 @@ ggdraw(Assemblage_info_retouch_density_ages_prop_main_plot) +
             .35, .28) +
   draw_plot(retouch_sp_sub_plot,
             .46, .1,
-            .23, .28)
+            .23, .28) +
+  theme(panel.background = element_rect(fill='white', colour="white"),
+                               plot.background = element_rect(fill='white', colour="white"))
+
 
 
 ggsave(here::here("analysis/figures/003-retouch-by-density.png"),

### 2021-09-24 analysis/paper/004-artifact-composition.R

@@ -102,7 +102,9 @@ ggplot(kasa_long3,
   scale_fill_viridis_d(name = "Artifact type",
                        option = "D") +
   coord_flip() +
-  theme_minimal(base_size = 8)
+  theme_minimal(base_size = 8) +
+  theme(panel.background = element_rect(fill='white', colour="white"),
+        plot.background = element_rect(fill='white', colour="white"))
 
 ggsave(plot = kasa_long3_fill_plot,
          here::here("analysis/figures/004-artifact-types.png"),

### 2021-09-24 analysis/paper/005-raw-materials.R

@@ -83,7 +83,9 @@ ggplot(kasr_long2,
   scale_fill_viridis_d(name = "Raw material",
                        option = "C") +
   coord_flip() +
-  theme_minimal(base_size = 8)
+  theme_minimal(base_size = 8) +
+  theme(panel.background = element_rect(fill='white', colour="white"),
+                                    plot.background = element_rect(fill='white', colour="white"))
 
 ggsave(here::here("analysis/figures/005-raw-materials.png"),
        width = 4.45,

### 2021-09-24 analysis/paper/007-climate-model-plots.R

@@ -169,7 +169,6 @@ ggplot(site_locations_tbl_temps_periods_filtered) +
   xlab("") +
   ylab("Mean annual temperature (MAT, °C)") +
   theme_minimal(base_size = base_size)
-
 #----------------------------------------------------------------------
 
 # can we have a map also?
@@ -230,7 +229,6 @@ ggmap(map)  +
         legend.key.size = unit(0.3, "cm")) +
           xlab('') +
           ylab('')
-
 #----------------------------------------------------------------------
 
 library(ggpubr)
@@ -263,6 +261,7 @@ mat_elev_cor_plot <-
   theme(legend.position = c(0.75, 0.8),
         legend.key.size = unit(0.5, 'lines'))
 
+
 #----------------------------------------------------------------------
 
 # combine plots
@@ -285,7 +284,8 @@ bottom_row <-
 # combine rows
 plot_grid(top_row,
           bottom_row,
-          ncol = 1)
+          ncol = 1) +  theme(panel.background = element_rect(fill='white', colour="white"),
+                             plot.background = element_rect(fill='white', colour="white"))
 
 ggsave(here::here("analysis/figures/007-climate-model-sites-panel-plot.png"),
        scale = 1,

### 2021-09-24 analysis/paper/paper.Rmd
Able to save spd image in png

@@
 These technologies include trap pits, cobble concentrations, hearths, charcoal concentrations, and toolkits including trapezoids, pointed-shaped backed blades, backed points, burins, end scrapers, side scrapers, wedges, beak-shaped tools, axes, edge-ground axes, hammerstones, cobble tools, and anvils [@bae2017origin; @izuho201421].
-These innovations have been interpreted as evidence of new foraging methods such as watercraft (marine transport of obsidian from the Kozu Island) and bow-and-arrow technology, driven by increased population and ecological changes [@yamaoka2012use; @bae2017origin; @nakazawa2018quaternary; @morisaki2019early].
+These innovations have been interpreted as evidence of new foraging methods such as watercraft (marine transport of obsidian from the Kozu Island) and bow-and-arrow technology, driven by increased population and ecological changes and led to the megafaunal extinctions [@yamaoka2012use; @bae2017origin; @nakazawa2018quaternary; @morisaki2019early; @norton2010nature; @bae2017late; @nakazawa2017pleistocene; @kaifu2019palaeolithic].
 Hafted trapezoids were likely multifunctional tools adapted to the specific environmental settings of the different Japanese Islands [@ono2002radiocaron; @yamaoka2012use].
@@
 
-After the first excavation of a Paleolithic site in Korea at Seokjang-ri site in the 1960s, more than 200 Paleolithic sites have been discovered in South Korea [@lee2019were].
+After the first Paleolithic sites such as Dongkwanjin and Yondaebong were discovered during the mid-1930s, more than 200 Paleolithic sites have been discovered in South Korea [@lee2019were].
 We selected sites dated to before and after the transition period (42-35 ka) to analyze the process of technological change.

### 2021-09-24 analysis/paper/009-summed-probability-distributions.R

@@ -432,7 +432,11 @@ mtext(
 
 }
 
-ggdraw(p1)
+library(gridGraphics)
+ggdraw(p1) +
+  theme(panel.background = element_rect(fill='white', colour="white"),
+        plot.background = element_rect(fill='white', colour="white"))
+
 
 ggsave(here::here("analysis/figures/008-summed-probability-distribution-models.png"),
        h = 15,
@@ -447,7 +451,7 @@ ggsave(here::here("analysis/figures/008-summed-probability-distribution-models.p
 #     w = 13,
 #     res = 1000,
 #     units = "cm")
-#
+# #
 # p1
 #
 # # now close the connection to the PNG file so we can take a look at it

### 2021-09-27 analysis/paper/paper.Rmd
Excluded data for SPD, adjusted raw material and artifact composition plots

@@
 
-After the first Paleolithic sites such as Dongkwanjin and Yondaebong were discovered during the mid-1930s, more than 200 Paleolithic sites have been discovered in South Korea [@lee2019were].
+After the first Paleolithic sites such as Dongkwanjin and Yondaebong were discovered during the mid-1930s, more than 200 Paleolithic sites have been discovered in South Korea [@norton2000current; @lee2017korean; @lee2019were].
 We selected sites dated to before and after the transition period (42-35 ka) to analyze the process of technological change.
@@
 
-There is a debate about whether past population growth has played a role in technological change [@shennan2001demography; @kline2010population].
+There is a debate about whether past population growth has played a role in technological change [@shennan2001demography; @hosfield2005individuals;  @kline2010population].
 @shennan2009pattern observed a correlation between population decrease and technological simplification in his study of lithic arrowheads from the southern Scandinavian Mesolithic.
 He found that the lithics became less complex, and point shape changed as the population level dropped.
+By proposing the generalized null demographic model built on cultural transmission theory, @lycett2010demographic also claimed that relatively lower effective population sizes in East Asia could be strongly associated with a lower frequency of handaxe sites, a lower percentage of bifaces, and morphological difference between East Asian handaxes and classic Acheulean artifacts.
+They also suggested that a lack of Levallois technologies in East Asia might be similar to the previous example of handaxe.
 On the other hand, @collard2013population argue that regional archaeological data do not support the population-driven models.

### 2021-09-27 analysis/paper/004-artifact-composition.R

@@ -49,7 +49,8 @@ kasa_long1 <-
                   'flake',
                    'debris',
                    'beak_shaped',
-                  'awl')
+                  'awl',
+                  'anvil')
                    ) %>%
   # if percentage is <10%, call it 'other'
    mutate(artefact_type =   X1 #ifelse(as.character(X) == 'stemmed_point',

### 2021-09-27 analysis/paper/005-raw-materials.R

@@ -41,7 +41,13 @@ mydata_ages <-
       'limestone',
       'granite',
       'gneiss',
-      'tuff')) %>%
+      'tuff',
+      'amphibolite',
+      'andesite',
+      'grit',
+      'metamorphic-rock',
+      'quartzitic-gneiss',
+      'siliceous-schist')) %>%
   # if percentage is <10%, call it 'other'
   mutate(raw_material = X) %>%  #ifelse(percentage >= 10, as.character(X), "other")) %>%
   #mutate(raw_material = ifelse(raw_material == "etc.", "other", raw_material)) %>%

### 2021-09-27 analysis/paper/009-summed-probability-distributions.R

@@ -14,7 +14,17 @@ dates_clean <-
   # filter age range here
   filter(age < 50500,
          age >= 10000,
-         method == 'AMS')
+         method == 'AMS') %>%   #exclude OSL dates
+  filter(!site_name_original %in% c('Nobong',
+                                    'Geoduri',
+                                    'Geumneungdong',
+                                    'Janggi',
+                                    'Sanggari',
+                                    'Sinheug',
+                                    'Oeryang',
+                                    'Eosanri',
+                                    'Jeommal',
+                                    'Pyeongneungdong'))#exclude sites that have only one date
 
 # devtools::install_github('ahb108/rcarbon')
 library(rcarbon)

### 2021-09-27 analysis/paper/009-summed-probability-distributions.R
stash

@@ -4,17 +4,8 @@ library(ggrepel)
 library(here)
 library(cowplot)
 
-dates <-
-  read.csv(here("analysis/data/raw_data/Radiocarbon dates.csv"))
-
-dates_clean <-
-  dates %>%
-  mutate(age = parse_number(as.character(age)),
-         error = parse_number(as.character(error))) %>%
-  # filter age range here
-  filter(age < 50500,
-         age >= 10000,
-         method == 'AMS')
+# import the data
+source(here::here("analysis/paper/008-summed-probability-distributions.R"))
 
 # devtools::install_github('ahb108/rcarbon')
 library(rcarbon)

### 2021-09-27 analysis/paper/paper.Rmd

@@
 abstract: |
-  We use models from human behavioral ecology to examine stone artifacts from 22 sites in Korea to investigate mobility and site occupation patterns during the Late Pleistocene when new tools, such as stemmed points and blades, appeared in the archaeological record.
+  We use models from human behavioral ecology to examine stone artifacts from 23 sites in Korea to investigate mobility and site occupation patterns during the Late Pleistocene when new tools, such as stemmed points and blades, appeared in the archaeological record.
 We focus on two questions: what changes in foragers’ landscape use were associated with the introduction of new tools?
@@
 The environmental context of this technological innovation was a gradual decrease in temperature into the LGM.
-A population increase followed after the appearance of stemmed points.
+A population steadily increased before the appearance of stemmed points.
 
@@
 Here we use a behavioral ecological framework to explore questions about the implications of the appearance of this new technology.
-We used data from 22 sites in South Korea to explore changes in the way people occupied the landscape, the way people used sites, and the of the social and ecological contexts associated with the appearance stemmed points, and the blade industry that followed.
+We used data from 23 sites in South Korea to explore changes in the way people occupied the landscape, the way people used sites, and the of the social and ecological contexts associated with the appearance stemmed points, and the blade industry that followed.
 
@@
 
-Previous work on this technological change in Korea has focused on issues of the transition from the Early to the Late Paleolithic and modern human dispersals [@norton2009evolution; @bae2010origin; @bae2017late].
+Previous work on this technological change in Korea has focused on issues of the transition from the Early to the Late Paleolithic and modern human dispersals [@norton2009evolution; @bae2010origin; @seong2016eastern; @bae2017late].
 To our knowledge, the implications of the appearance of stemmed points for forager mobility and site use have not been explored in detail.
@@
 Stemmed points appeared in northern China and Japan after they appear in the Korean Peninsula.
-The oldest stemmed points so far in Korea are from the Bonggok site dated to ca.
+The stemmed points from the Bonggok site had been known as the oldest stemmed points dated to ca.
 41.5 ka, and made on elongated flakes [@seong2009emergence; @seong2015diversity; @bae2017origin].
@@
 In addition to the stemmed points, flake tools became more complex, with a greater diversity in both type and size along with continuously used core tools during the Late Pleistocene [@bae2010origin; @lee2013current; @Lee_2016].
+However, recently excavated Hajinri site contain stemmed points and blades dated to 42 ka, which is even older than Bonggok site.
 
@@
 Assemblages with low density but a high proportion of retouched and backed pieces may indicate the remains of 'short-term camp', which is a site for small and ephemeral overnight camp or limited activity station.
-On the other hand, the combination of high density and low retouched pieces might be expected at logistically organized basecamp, which is a site with greater residential stability, long site occupation and occupied by larger groups [@clark2017lithics; @clark2019landscapes].
+On the other hand, the combination of high density and low retouched pieces might be expected at logistically organized 'base camp', which is a site with greater residential stability, long site occupation and occupied by larger groups [@clark2017lithics; @clark2019landscapes].
 
@@
 We selected sites dated to before and after the transition period (42-35 ka) to analyze the process of technological change.
-This resulted in a sample of 33 assemblages from `r nrow(korean_archaeological_site_locations)` sites spanning 49-24 ka (Figure \@ref(fig:korea-map)).
+We acknowledge the limitations of directly accepting the dating results from excavation reports since the majority of sites in Korea are open-air localities which are more prone to be disturbed compared to cave sites.
+Therefore, we carefully excluded sites whose assemblages are not matched with their dating results.
+This resulted in a sample of 35 assemblages from `r nrow(korean_archaeological_site_locations)` sites spanning 49-24 ka (Figure \@ref(fig:korea-map)).
 We identified multiple assemblages in a site where culturally sterile deposits separated artifact-bearing deposits, or where stratigraphic units could be identified by major differences in the texture and composition of the sedimentary deposit containing the stone artifacts.
@@
 
-The 33 assemblages consist of (non-blade) cores, blade cores, blades, flakes, debris, hammers, choppers, planes, polyhedrals, side scrapers, end scrapers, notches, cleavers, stemmed points, awls, denticulates, burins, handaxes, blanks, knives, handadzes, peaks, flatters, flakers etc.
+The 35 assemblages consist of (non-blade) cores, blade cores, blades, flakes, debris, hammers, choppers, planes, polyhedrals, side scrapers, end scrapers, notches, cleavers, stemmed points, awls, denticulates, burins, handaxes, blanks, knives, handadzes, peaks, flatters, flakers etc.
 We excluded some artifacts from our sample because of uncertainty about their typology.
@@
 
-In addition to these data on human behaviour from the stone artifact assemblages, our HBE models require environmental proxies to build hypotheses of behavioural change influenced by the surrounding environment.
+In addition to these data on human behavior from the stone artifact assemblages, our HBE models require environmental proxies to build hypotheses of behavioral change influenced by the surrounding environment.
 However, there are few paleoenvironmental proxies in Korea relevant to our study period.
@@
 For instance, @timpson2014reconstructing apply SPD to 93 ages from Eastern Middle Sweden and demonstrate that the result is equivalent to a larger sample size (n = 243).
-We used `r nrow(dates_clean)` radiocarbon ages from `r nrow(korean_archaeological_site_locations)` sites to examine connections between population trends during the Late Pleistocene in Korea and technological innovation.
+In addition to the 93 radiocarbon ages from our targeted sites, we included all available dating results (540 ages) from other Korean Paleolithic sites that have younger and older assemblages.
+We compiled a total of 633 dates from 113 sites and applied a chronometric hygiene protocol to assess the reliability of radiocarbon determinations in relation to target events.
+Following the criteria from @napolitano2019reevaluating and @fitzpatrick2006critical, we kept ages from dating short-lived terrestrial material (i.e., plant remains) and charcoal, and having laboratory name and number.
+We excluded OSL dates and sites with only one date.
+We ended up using `r nrow(dates_clean)` radiocarbon ages from `r nrow(korean_archaeological_site_locations)` sites to examine connections between population trends during the Late Pleistocene in Korea and technological innovation.
 
@@
 The inset plot shows direct comparison of artifact densities for assemblages with and without stemmed points.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/002-artifact-volumetric-density.R"))}
-quietly(source(here::here("analysis/paper/002-artifact-volumetric-density.R")))
+# quietly(source(here::here("analysis/paper/002-artifact-volumetric-density.R")))
 knitr::include_graphics(here::here("analysis/figures/002-age-by-density.png"))
@@
 The proportion of choppers is lower in more recent assemblages.
-Stemmed points first appeared in Bonggok around 41.5 ka and were typically found in assemblages that contain blades, except for Songamri-A.
-The earliest blades are also recorded from Bonggok but it is not guaranteed that those artifacts as well as other earlier blades are actual blades or elongate flakes without accompanying blade core.
-Blade cores first appeared in Suyanggae-C at 34.3 ka.
+Stemmed points first appeared in Hajinri around 42 ka and were typically found in assemblages that contain blades, except for Songamri-A.
+The earliest blades and blade cores are also recorded from Hajinri but the date of this site can be unreliable because the stemmed points and blades are highly standardized and refined.
+Bonggok site, which had been known as the earliest site for stemmed points, also includes blades but it has not been guaranteed that those artifacts as well as other earlier blades are actual blades or elongate flakes without accompanying blade core.
+Except for Hajinri, blade cores first appeared in Suyanggae-C at 34.3 ka.
 Other artifacts, including burins, denticulates, end scrapers, handaxes, knives, notches, planes, and polyhedrals made up a small proportion of the assemblages and show no directional chronological change.
@@
 ```{r artifact-composition, fig.cap="The composition of each assemblage.
-We excluded artifacts related to manufacturing processes such as pebbles, hammers, flakes and debris, and artifacts that appeared only in a few assemblages including points, beak shaped pieces, and awls, and unknown pieces.
+We excluded artifacts related to manufacturing processes such as pebbles, hammers, flakes and debris, and artifacts that appeared only in a few assemblages including points, beak shaped pieces, awls, and anvils, and unknown pieces.
 The colour represents different types of tools and the assemblages are placed in chronological order.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/004-artifact-composition.R"))}
@@
 For example, both Galdamri-A, dated to 44 ka with a small assemblage, and Hopyeongdong-B, dated to 27.5 ka with a larger assemblage, consist of only the quartz related materials.
-The use of chert, hornfels, rhyolite, and shale increased after 41 ka, but these raw materials remained a small proportion of assemblages until 25 ka.
-For example, hornfels are dominant at Yongsandong and Yongbang, dated to 25 ka.
-Chert, hornfels, rhyolite, and shale are very suitable to make elongate blades and stemmed points because of the predictability of flaking afforded by their fine-grained texture.
-For example, Yongsandong has the largest number of stemmed points (n=40) among the assemblages studied here, and all stemmed points from Yongsandong are made from hornfels.
+The use of chert, hornfels, rhyolite, and shale increased after 42 ka, but these raw materials remained a small proportion of assemblages until 25 ka.
+They are very suitable to make elongate blades and stemmed points because of the predictability of flaking afforded by their fine-grained texture.
+For example, Hajinri has the largest number of stemmed points (n=86) among the assemblages studied here, and most stemmed points from Hajinri are made from shales.
 Other raw materials, such as porphyry and sandstone, were found in just a few assemblages in small proportions.
@@
 ```{r raw-materials, fig.cap="The composition of raw material type in each assemblage.
-We excluded raw materials that are included in less than 5 assemblages including crystal, basalt, iron_ore, slate, limestone, granite, gneiss, tuff, and unidentified ones.
+We excluded raw materials that are included in less than 5 assemblages including crystal, basalt, iron_ore, slate, limestone, granite, gneiss, tuff, amphibolite, andesite, grit, metamorphic-rock, quartzitic-gneiss, siliceous-schist, and unidentified ones.
 Hornfels are shown in 4 assemblages but since it occupies more than half of the Yongsandong and Yongband assemblages, we included it in the plot.
@@
 Sites containing stemmed points reach a maximum elevation of 175 m at 28 ka.
+We acknowledged that site elevation might not be the best proxy for the environmental context due to mountainous geography of Korea and possible disturbance during the site formation processes.
+However recent studies show that elevation makes huge impact on the diversity and ecology of plants and animals in Korea [@sohn2019effect; @park2020temporal] and thus enable us to explore the elevation as a environmental proxy.
 
@@
 
-Figure \@ref(fig:spd-model) shows the summed probability distribution (SPD) of the radiocarbon ages from the sites in our sample overlaid with distributions (indicated by the grey envelopes) generated from four models of demographic change for the period 50 ka to 10 ka (exponential, uniform, linear, and logistic).
+Figure \@ref(fig:spd-model) shows the summed probability distribution (SPD) of the radiocarbon ages from the sites in our sample overlaid with distributions (indicated by the grey envelopes) generated from four models of demographic change for the period 50 ka to 10 ka (linear, exponential, logistic, and uniform).
 The fit of models to our observed SPD was computed using 1000 simulations, and the resulting global p-values indicate significant deviations in the observed SPD from the modeled distributions.
@@
 We can say that the population did not grow or decline strongly logistically, exponentially or linearly during the Late Pleistocene, but instead followed a noisy distribution.
-The most striking results here are the SPD showing a steady growth rising to a peak at about 35 ka, indicated by the significant deviations from all four null models (shaded in red).
-The SPD remains high with minor deviations until declining to a lower level at 32 ka.
-After 32 ka the SPD remains generally low, with frequent temporary fluctuations.
+The most striking results here are the SPD showing a steady growth rising to a peak at about 43 ka, indicated by the positive deviations from the three null models (shaded in red), except for the logistic model.
+During the transition period, SPD hits another peak at about 35ka.
+Then the SPD goes down with minor deviations to a lower level at 31.5 ka.
+From 27 ka the SPD constantly fluctuates with temporary positive deviation until 20ka.
 
@@
 The red and blue vertical bands highlight the portions of the SPD where positive and negative deviations are detected.
-We applied logistic, exponential and linear to AIC along with the logistical growth model (fit) to evaluate each model [@riris2019widespread].
+We applied linear, exponential and logistic to AIC along with the logistical growth model (fit) to evaluate each model [@riris2019widespread].
 We excluded the uniform model because it has no fitted curve or surface to evaluate.
@@
 The environmental context of this technological innovation was a gradual decrease in temperature into the LGM.
-And the population increase followed after the appearance of stemmed points.
+And the population increase before the appearance of stemmed points.
 
@@
 These expedient technologies are more typical for groups with greater residential stability who can make, use, abandon, and remake their tools frequently in the same location [@binford1979organization; @torrence1989time; @kuhn1994formal; @meignen2006middle; @kuhn2014mousterian; @vaquero2018searching; @clark2019landscapes].
-Our data support the general concept of the curated and expedient technology by showing a strong pattern of artifacts from less dense assemblages having higher retouch frequencies, while more dense assemblages have lower retouch frequencies.
+As mentioned above, the sites we addressed here are all open-air localities with limited material remains to infer the clear view of subsistence strategies but our data still supports the general concept of the curated and expedient technology by showing a strong pattern of artifacts from less dense assemblages having higher retouch frequencies, while more dense assemblages have lower retouch frequencies.
 Our results show that foragers with stemmed points tended to stay in the same site for longer periods with higher artifact densities and fewer retouched pieces, compared to assemblages without stemmed points.
@@
 Previous studies claim that the existing tools, including choppers, polyhedrals, handaxes and cores were usually made of locally acquired quartzite, while the new tools including stemmed points and other blade assemblages were made of more fine-grained materials such as siliceous shale, hornfels, and obsidian which might be brought from distant sources [@chang2013human; @seong2015diversity].
-Our data also show that quartz, quartz vein and quartzite were consistently dominant and the use of fine-grained materials including chert, hornfels, rhyolite, and shale, increased after 41.5 ka, around the time that stemmed points appear.
+Our data also show that quartz, quartz vein and quartzite were consistently dominant and the use of fine-grained materials including chert, hornfels, rhyolite, and shale, increased after 42 ka, around the time that stemmed points appear.
 Projectile points, such as stemmed points, benefit from high quality materials which enable high flaking quality, durability and effectiveness of an edge, as well as the creation of well-defined outline forms [@bamforth2009projectile].
@@
 An important quality of composite tools is that the user can easily replace damaged parts, contributing to an increased maintainability of the tool [@kuhn2015artifacts; @cardillo2010some].
-Combining the stone projective with a wooden shaft further materials enhances functionality of the tool by improving penetration by increasing weight and sharpness [@browne1940projectile].
+Combining the stone projectile with a wooden shaft further materials enhances functionality of the tool by improving penetration by increasing weight and sharpness [@browne1940projectile].
 We expected that the functional advantage of the new composite tool would have allowed foragers to be more effective in less productive landscapes.
@@
 The results show that groups with stemmed points generally reached higher elevations, where the MAT was generally lower.
-This supports our hypothesis that stemmed points were associated with the occupation of more marginal habitats.
+This supports our hypothesis that stemmed points were associated with the occupation of more harsh habitats.
 Through MIS 3 the temperature gradually decreased until the LGM, suggesting that stemmed points may have been part of a suite of adaptations to cooler temperatures.
@@
 We investigated forager population dynamics to assess if demographic change might be a relevant mechanism to explain the appearance of stemmed points.
-The output of our SPD models indicate population increasing until 35 ka and remaining at the peak with minor deviations and then declining after 31 ka.
-If population dynamics were a key driver, then we expect population to be high at the time the new technology first appears.
-However, with the first stemmed points appearing at 41.5 ka, well before the observed peak in population, we conclude that population dynamics were not a major mechanism in the appearance of this new technology.
+The output of our SPD models indicate population increasing before the appearance of stemmed points and remaining at the peak with minor deviations until 35ka and then declining after 35 ka.
+If population dynamics were a key driver, then we expect population to see more frequent and constant positive deviations during the transition period.
+We conclude that population dynamics were not a major mechanism in the appearance of this new technology.
 Bae claims that stemmed points are the result of the continuous influx of modern humans from a Siberian migration [@bae2010origin].
-Our results are partly consistent with this scenario, showing a gradual increase in population after the appearance of stemmed points.
-Given the pattern we observe in the timing of the appearance of stemmed points preceding an increase in population, we might speculate that their appearance was a mechanism enabling an increase in population by making marginal habitats more viable.
+Our results are partly consistent with this scenario, showing a gradual increase in population before the appearance of stemmed points.
 
@@
 
-To understand the change in mobility and site occupation related to the appearance of new technologies during the late Pleistocene in Korea, we examined stone artifacts from 22 sites dated to 49-24 ka.
+To understand the change in mobility and site occupation related to the appearance of new technologies during the late Pleistocene in Korea, we examined stone artifacts from 23 sites dated to 49-24 ka.
 The results show that the forager groups with stemmed points tended to be located at higher elevations.
@@
 
-Our findings are limited by the relatively small number of sites and radiocarbon ages currently available The absence of detailed stratigraphic data limit the chronological resolution for the assemblages in our sample.
+Our findings are limited by the relatively small number of sites and radiocarbon ages currently available.
+The absence of detailed stratigraphic data limit the chronological resolution for the assemblages in our sample.
 Furthermore, we currently lack functional data on individual artifact assemblages to support our conclusions about site use.

### 2021-09-27 analysis/paper/009-summed-probability-distributions.R
resolve merge conflict

@@ -4,10 +4,8 @@ library(ggrepel)
 library(here)
 library(cowplot)
 
-<<<<<<< HEAD
+
 # import the data
-source(here::here("analysis/paper/008-summed-probability-distributions.R"))
-=======
 dates <-
   read.csv(here("analysis/data/raw_data/Radiocarbon dates_revision.csv"))
 
@@ -29,7 +27,7 @@ dates_clean <-
                                     'Eosanri',
                                     'Jeommal',
                                     'Pyeongneungdong'))#exclude sites that have only one date
->>>>>>> 9bd768474400b84f5496496025189e7c913c0f5f
+
 
 # devtools::install_github('ahb108/rcarbon')
 library(rcarbon)

### 2021-09-27 analysis/paper/008-summed-probability-distributions.R
replace data_clean in 008

@@ -6,6 +6,7 @@ library(here)
 dates <-
   read.csv(here("analysis/data/raw_data/Radiocarbon dates.csv"))
 
+
 dates_clean <-
   dates %>%
   mutate(age = parse_number(as.character(age)),
@@ -13,5 +14,14 @@ dates_clean <-
   # filter age range here
   filter(age < 50500,
          age >= 10000,
-         method == 'AMS')
-
+         method == 'AMS') %>%   #exclude OSL dates
+  filter(!site_name_original %in% c('Nobong',
+                                    'Geoduri',
+                                    'Geumneungdong',
+                                    'Janggi',
+                                    'Sanggari',
+                                    'Sinheug',
+                                    'Oeryang',
+                                    'Eosanri',
+                                    'Jeommal',
+                                    'Pyeongneungdong'))

### 2021-09-27 analysis/paper/009-summed-probability-distributions.R
simplify 008 and 009

@@ -4,30 +4,8 @@ library(ggrepel)
 library(here)
 library(cowplot)
 
-
-# import the data
-dates <-
-  read.csv(here("analysis/data/raw_data/Radiocarbon dates_revision.csv"))
-
-dates_clean <-
-  dates %>%
-  mutate(age = parse_number(as.character(age)),
-         error = parse_number(as.character(error))) %>%
-  # filter age range here
-  filter(age < 50500,
-         age >= 10000,
-         method == 'AMS') %>%   #exclude OSL dates
-  filter(!site_name_original %in% c('Nobong',
-                                    'Geoduri',
-                                    'Geumneungdong',
-                                    'Janggi',
-                                    'Sanggari',
-                                    'Sinheug',
-                                    'Oeryang',
-                                    'Eosanri',
-                                    'Jeommal',
-                                    'Pyeongneungdong'))#exclude sites that have only one date
-
+# read in spreadsheet and clean dates data
+source(here::here("analysis/paper/008-summed-probability-distributions.R"))
 
 # devtools::install_github('ahb108/rcarbon')
 library(rcarbon)

### 2021-09-27 analysis/paper/005-raw-materials.R
check that I can knit, yes, ok

@@ -32,7 +32,7 @@ mydata_ages <-
          total = sum(count, na.rm = TRUE)) %>%
   filter(!is.na(percentage)) %>%
     filter(!X %in% c(
-      #'gneiss', exclude raw materials cluded in less then 4 assemblages
+      #'gneiss', exclude raw materials included in less then 4 assemblages
       'crystal',
       'basalt',
       'etc.',

### 2021-09-27 analysis/paper/paper.Rmd
Change in 002 to knit RMD file

@@
 The inset plot shows direct comparison of artifact densities for assemblages with and without stemmed points.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/002-artifact-volumetric-density.R"))}
-# quietly(source(here::here("analysis/paper/002-artifact-volumetric-density.R")))
+quietly(source(here::here("analysis/paper/002-artifact-volumetric-density.R")))
 knitr::include_graphics(here::here("analysis/figures/002-age-by-density.png"))

### 2021-10-12 analysis/paper/paper.Rmd
Edited rmd file for revision, added files for submission

@@
 abstract: |
-  We use models from human behavioral ecology to examine stone artifacts from 23 sites in Korea to investigate mobility and site occupation patterns during the Late Pleistocene when new tools, such as stemmed points and blades, appeared in the archaeological record.
+  We use models from human behavioral ecology to examine stone artifacts from 23 sites in Korea to investigate mobility and site occupation patterns during the Late Pleistocene.
+This is an important period because new tools, such as stemmed points and blades, appeared in the archaeological record.
 We focus on two questions: what changes in foragers’ landscape use were associated with the introduction of new tools?
@@
 To answer these questions we present quantitative analyses of artifact volumetric density, retouch frequency, composition of toolkits, and artifact raw materials.
-We explore the environmental and demographic contexts by applying paleoclimate simulations and summed probability distribution models.
+We explore environmental and demographic contexts by applying paleoclimate simulations and summed probability distribution models.
 We find that quartz and side scrapers, in addition to cores and choppers, remain dominant in assemblages before and after the introduction of stemmed points throughout the Late Pleistocene.
-Our results show that forager groups using stemmed points were associated with occupation of marginal or extreme environments.
-In addition, groups with stemmed points were associated with expedient technologies, indicating residential and less mobile behaviours.
+Our results show that forager groups using stemmed points may have been associated with occupation of marginal or extreme environments.
+Groups with stemmed points were associated with expedient technologies, indicating residential and less mobile behaviors.
 The environmental context of this technological innovation was a gradual decrease in temperature into the LGM.
-A population steadily increased before the appearance of stemmed points.
+Population increased before the appearance of stemmed points.
 
@@
 42-35 ka) in Korea is often assumed to have transformed forager lifeways because it was thought to have indicated more specialized hunting practices [@seong2008tanged; @chang2013human].
-The stemmed point is a projectile point made from an elongated flake or blade with slight retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft.
+Stemmed points are projectile points made from an elongated flake or blade with slight retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft.
 These types of artifacts are often called tanged points in the western hemisphere [@serwatka20162d; @ivanovaite2020all], but we use ‘stemmed point’ here to distinguish from Bronze Age stone projectile points that have long been called ‘tanged points’ in Korea.
@@
 If these artifacts were projectile weapon tips, then stemmed points represent the first appearance of hafting technology in Korea.
-Most open sites in Korea do not preserve organic remains due to acidic soil.
-There are a few limestone cave sites containing human fossils or faunal remains (e.g.
+However, most open sites in Korea do not preserve organic remains due to acidic soil, so shafts have not been found.
+There are very few limestone cave sites containing human fossils or faunal remains (e.g.
 Durubong and Jeommal Cave), which located in constrained area to access in North Korea or have been rarely analyzed to directly apply to other archaeological research[@bae2013paleolithic].
-Therefore stone artifacts are our only window to examine hunting and subsistence strategies.
-Here we use a behavioral ecological framework to explore questions about the implications of the appearance of this new technology.
-We used data from 23 sites in South Korea to explore changes in the way people occupied the landscape, the way people used sites, and the of the social and ecological contexts associated with the appearance stemmed points, and the blade industry that followed.
+Therefore stone artifacts such as stemmed points are our only window to examine hunting and subsistence strategies.
+Here we apply a behavioral ecological framework to explore questions about the implications of the appearance of this new technology.
+We use data from 23 sites in South Korea to explore changes in the way people occupied the landscape, the way people used sites, and the social and ecological contexts associated with the appearance stemmed points, and the blade industry that followed.
 
@@
 To our knowledge, the implications of the appearance of stemmed points for forager mobility and site use have not been explored in detail.
-Here, we attempt to identify patterns of mobility and site occupation through analysis of the stone artifact assemblages using Human Behavioral Ecology (HBE) concepts.
+Here, we attempt to identify patterns of mobility and site occupation through analysis of the stone artifact assemblages using Human Behavioral Ecology (HBE) models.
 Drawing on established relationships between lithic technology and forager mobility [@shott1986settlement; @Hiscock_1994; @kuhn1994formal; @kuhn2004upper; @kuhn2015artifacts; @kuhn2016moves; @capriles2018mobility; @seong2015diversity; @seong2008tanged], we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that were part of a broader strategy of moving frequently and further, possibly as an adaptation to environmental or population changes.
@@
 
-In the Eastern hemisphere, the most striking technological innovations during the Late Pleistocene are found in Northeast Asia, such as northern China, Japan and Korea.
+In the Eastern hemisphere, some of the most striking technological innovations during the Late Pleistocene are found in Northeast Asia, such as northern China, Japan and Korea.
 These innovations include blade technology, high frequencies of retouched blade tools, several novel tools such as projectile points, stemmed points, end scrapers, burins, denticulates, ground stone tools etc, and frequent use of high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary; @bae2017late; @lee2017korean].
@@
 
-In Japan, Early Upper Paleolithic (EUP) technologies appeared in different regions of the archipelago around 38 ka, accompanied by remains of Homo sapiens found in Okinawa Island (ca.
+In Japan, Early Upper Paleolithic (EUP) technologies appeared in different regions of the archipelago around 38 ka, accompanied by remains of *Homo sapiens* found in Okinawa Island (ca.
 36 ka) [@yamaoka2012use; @izuho201421].
@@
 An important example is Ishinomoto 8-ku, which is one of the earliest sites, dated to 39,690--34,790 cal BP, located in Takuma upland of Kumamoto Prefecture.
-A total of 500 stone artifacts were discovered including trapezoids, side scrapers, flakes, flake cores, edge-ground axes, and cobble tools.
-The main lithic raw materials are chert and andesite [@izuho201421].
+A total of 500 stone artifacts, mostly chert and andesite, were discovered including trapezoids, side scrapers, flakes, flake cores, edge-ground axes, and cobble tools [@izuho201421].
 Stemmed points appear in Japan much later than in Korea, dated to between about 15,500 and 13,800 cal BP [@ono2002radiocaron; @tsutsumi2007dynamics].
@@
 The key difference is that the earliest signs of new technologies in Korean assemblages are stemmed points, followed by blade technologies.
-Stemmed points appeared in northern China and Japan after they appear in the Korean Peninsula.
-The stemmed points from the Bonggok site had been known as the oldest stemmed points dated to ca.
+This contrasts with China and Japan where stemmed points appear after their first appearance in the Korean Peninsula.
+Stemmed points from the Bonggok site are currently accepted as the oldest stemmed points, dated to ca.
 41.5 ka, and made on elongated flakes [@seong2009emergence; @seong2015diversity; @bae2017origin].
 After blades became widespread in Korea at ca.
-27 ka, and were used to make the stemmed points, the shape of points became more standardized with one or two ridges on the dorsal side and triangular cross section.
+27 ka, they were used to make the stemmed points, and the shape of points became more standardized with one or two ridges on the dorsal side and a triangular cross section.
 Replacing flakes with blades as blanks of the stemmed points led to an increase in quantity of the points over time.
 In addition to the stemmed points, flake tools became more complex, with a greater diversity in both type and size along with continuously used core tools during the Late Pleistocene [@bae2010origin; @lee2013current; @Lee_2016].
-However, recently excavated Hajinri site contain stemmed points and blades dated to 42 ka, which is even older than Bonggok site.
+A notable exception to this pattern is the recently excavated Hajinri site which is an outlier with both stemmed points and blades appearing together at 42 ka [@Lee_Woo_Lee_An_Yun_Park_Otani_Kim_Kim_Han_et].
 
@@
 
-The alternative model, which we call the '*in situ*' model claims that stemmed points and other Late Paleolithic assemblages including blade industries autonomously emerged in the south of the Korean peninsula, with no apparent external influence [@seong2008tanged].
+The alternative model, which we call the '*in situ*' model claims that stemmed points and other Late Paleolithic technologies, including blade industries, autonomously emerged in the south of the Korean peninsula, with no apparent external influence [@seong2008tanged].
 This claim is supported by Korea having the earliest appearance of stemmed points in Northeast Asia, at sites such as Bonggok, Hwadaeri, Hopyeongdong, and Yonghodong dated to approximately 42-35 ka BP [@seong2009emergence].
-Seong(2009) presents the increased blade-to-flake ratios of lithic assemblages after the appearance of stemmed points to support the *in situ* model, based on the premise that the blade industry represents a new exogenous technology while flakes indicate a continuing local one.
+Seong(2009) presents the increased blade-to-flake ratios of lithic assemblages after the appearance of stemmed points as support of the *in situ* model, based on the premise that the blade industry represents a new exogenous technology while flakes indicate a continuing local one.
 
-## Behavioural ecology and forager land use behaviours {-}
+## Behavioral ecology and forager land use behaviors {-}
 
-Our brief review of Late Pleistocene technological innovations in Northeast Asia shows that previous work has largely focused on timing, location, origins and description of these new tools [@seong2008tanged; @seong2009emergence; @bae2010origin; @lee2013current; @Lee_2016], but modeling the land-use behaviors associated with these assemblages has rarely been undertaken.
+Our brief review of Late Pleistocene technological innovations in Northeast Asia shows that previous work has largely focused on the timing, location, origins and description of these new tools [@seong2008tanged; @seong2009emergence; @bae2010origin; @lee2013current; @Lee_2016].
+Modeling the land-use behaviors associated with these assemblages has rarely been undertaken.
 We focus on the appearance of stemmed points in Korea because these are a new technology that likely represented new hunting behaviors.
@@
 We draw on behavioral ecological theory to model the effects of this technological change.
-Behavioral ecology theory offers structured models and testable hypotheses based on optimality assumptions derived from principles of economic rationality and environmental adaptation [@winterhalder1992evolutionary; @prentiss2019handbook].
-Specifically, we use the patch choice model, which is built around the marginal value theorem, to predict that forager mobility between patches and inter-patch travel times will increase when resource patch productivity decreases [@smith1991inujjuamiunt; @smith1983anthropological; @bettinger1997evolutionary; @wolverton2015patch; @llano2015optimal].
+Behavioral ecology theory is well established in international archaeology and offers structured models and testable hypotheses based on optimality assumptions derived from principles of economic rationality and environmental adaptation [@winterhalder1992evolutionary; @prentiss2019handbook].
+Specifically, we use the standard patch choice model, which is built around the widely-used marginal value theorem, to predict that forager mobility between patches and inter-patch travel times will increase when resource patch productivity decreases [@smith1991inujjuamiunt; @smith1983anthropological; @bettinger1997evolutionary; @wolverton2015patch; @llano2015optimal].
 We model individual artifact types as a kind of resource patch.
 Foragers may find a more costly and complicated technology (such as stemmed points, relative to flakes) optimal if they have a long cumulative time in use, if they are maintainable and reusable in a landscape where lithic raw material supply is uncertain [@torrence1983time; @bleed1986optimal].
-That is, they may move to a patch where they can stay longer (i.e.
-an artifact type that has a long and extendible use-life) when they are not certain of the productivity and travel times of other patches [@kuhn2015artifacts].
+That is, foragers may move to a patch where they can stay longer (i.e.
+prefer to make and use an artifact type that has a long and extendible use-life) when they are not certain of the productivity and travel times of other patches (i.e.
+the productivity and use-life of other artifact types) [@kuhn2015artifacts].
 
@@
 argue that the first appearance of microlithics in many locations represent a portable and multi-functional toolkit that minimize travel expenses and increase tool readiness to adapt to patchy and unpredictable resources associated with environmental change [@Hiscock_1994; @Hiscock_Clarkson_Mackay_2011].
-Similarly, in Japan, [@Morisaki_Izuho_Terry_Sato_2015] argue that the transition from trapezoid to blades and projectile points around 25 ka in north Paleo-Honshu Island enabled the foragers to extend their occupation into cold grassland landscapes.
+Similarly, in Japan, @Morisaki_Izuho_Terry_Sato_2015 argue that the transition from trapezoid to blades and projectile points around 25 ka in north Paleo-Honshu Island enabled the foragers to extend their occupation into cold grassland landscapes.
 
@@
 We selected sites dated to before and after the transition period (42-35 ka) to analyze the process of technological change.
-We acknowledge the limitations of directly accepting the dating results from excavation reports since the majority of sites in Korea are open-air localities which are more prone to be disturbed compared to cave sites.
-Therefore, we carefully excluded sites whose assemblages are not matched with their dating results.
-This resulted in a sample of 35 assemblages from `r nrow(korean_archaeological_site_locations)` sites spanning 49-24 ka (Figure \@ref(fig:korea-map)).
+It is important to note that the quality of excavation and reporting of Paleolithic sites in Korea varies greatly.
+Furthermore the stratigraphic integrity of the deposits is also variable, as the sites are open-air localities formed as loose accumulation of artifacts in fluviatile condition or within a colluvium developed at the edge of low-angled slope.
+To ensure our analysis employs reliable data, we applied a chronometric hygiene protocol (described below) to exclude sites with problematic chronologies and stratigraphy.
+After applying our protocol to the 200+ sites currently known, we produced a sample of 35 assemblages from `r nrow(korean_archaeological_site_locations)` sites spanning 49-24 ka (Figure \@ref(fig:korea-map)).
 We identified multiple assemblages in a site where culturally sterile deposits separated artifact-bearing deposits, or where stratigraphic units could be identified by major differences in the texture and composition of the sedimentary deposit containing the stone artifacts.
@@
 We excluded some artifacts from our sample because of uncertainty about their typology.
-The stone artifacts from all of these sites are stored in museums located throughout South Korea, and have been briefly described in the excavation reports.
+The stone artifacts from all of these sites are stored in museums located throughout South Korea, and have been briefly described in Korean language excavation reports.
 We have visited most of these collections to obtain permission to study the materials directly.
@@
 
-To answer our two research questions, we compiled information about artifact density, retouch frequency, toolkit composition, raw materials, site elevation, and radiocarbon ages.
+To answer our two research questions, we compiled information about artifact density, retouch frequency, toolkit composition, raw materials, site elevation, and radiocarbon ages for the sites in our sample.
 
@@
 Our study of raw material consumption patterns assumes that stone tool was transported and utilized to optimize mobility and minimize risks of not having artifacts when they were needed [@andrefsky1994raw; @brantingham2000raw; @brantingham2003neutral; @yue2020lithic].
-The representation of raw materials in an assemblage can be an indicator of site occupation patterns.
+It follows that the representation of raw materials in an assemblage can be an indicator of site occupation patterns.
 For example, a large variability of raw materials in a site can suggest long-term settlement, while a small assemblage of exotic raw materials can indicate higher mobility [@valde2020models].
-Here we want to know the specific raw materials that relate to the appearance of a new technology, and investigate site occupation patterns by examining changes in the proportions of raw material types in the assemblages.
+Here we quantify the specific raw materials that relate to the appearance of a new technology, and investigate site occupation patterns by examining changes in the proportions of raw material types in the assemblages.
 
@@
 We extracted annual temperature values from the data generated by @beyer2020high for all site locations in our sample.
-In addition, we extracted site elevation values from a raster, as a proxy to represent different environments at local scales, to identify how these impact mobility and site occupation, and their relationship with the stemmed points.
+In addition, we extracted site elevation values from a raster, as a proxy to represent environmental variation at local scales, to assist in identifying how these impact mobility and site occupation, and their relationship with the stemmed points.
 
@@
 We used radiocarbon dates as a proxy for human population to infer the demographic context of technological change.
-Summed probability distributions (SPD) of radiocarbon dates are increasingly used to infer temporal trends in past human populations [@rick1987dates; @bamforth2012radiocarbon; @shennan2013regional; @timpson2014reconstructing; @contreras2014summed; @riris2019widespread].
+Summed probability distributions (SPD) of radiocarbon dates are widely used by archaeologists to infer temporal trends in past human populations [@rick1987dates; @bamforth2012radiocarbon; @shennan2013regional; @timpson2014reconstructing; @contreras2014summed; @riris2019widespread].
 Despite concerns about limitations of the validity of SPD [@williams2012use; @bamforth2012radiocarbon; @carleton2021evaluating], several studies have shown good agreement between SPDs and other archaeological indicators (e.g.
@@
 We excluded OSL dates and sites with only one date.
-We ended up using `r nrow(dates_clean)` radiocarbon ages from `r nrow(korean_archaeological_site_locations)` sites to examine connections between population trends during the Late Pleistocene in Korea and technological innovation.
+This resulted in a sample of `r nrow(dates_clean)` radiocarbon ages from `r nrow(korean_archaeological_site_locations)` sites which we use to examine connections between population trends during the Late Pleistocene in Korea and technological innovation.
 
@@
 For example, @buchanan2016drivers concluded that changes in point types from \>13,000 years ago to 400 years ago in Texas were related to environmental risk, and not dependent on population size.
-We explored SPD to test for correlations between population, environmental proxies and technological change.
+
+We used SPD to test for correlations between population, environmental proxies and technological change.
 We used the R package 'rcarbon' [@cremaanalysing] to perform statistical analyzes and conduct model testing using Monte Carlo methods [@timpson2014reconstructing; @crema2016summed; @palmisano2020holocene].
-We generated exponential, uniform, linear, and logistic models and evaluated the model goodness-of-fit by applying Akaike's Information Criterion (AIC) as a selection criterion [@sakamoto1986akaike; @bevan2017holocene; @riris2019widespread].
+We generated exponential, uniform, linear, and logistic models and evaluated the model goodness-of-fit by applying Akaike's Information Criterion (AIC) as a ranking criterion [@sakamoto1986akaike; @bevan2017holocene; @riris2019widespread].
 
@@
 Stemmed points first appeared in Hajinri around 42 ka and were typically found in assemblages that contain blades, except for Songamri-A.
-The earliest blades and blade cores are also recorded from Hajinri but the date of this site can be unreliable because the stemmed points and blades are highly standardized and refined.
-Bonggok site, which had been known as the earliest site for stemmed points, also includes blades but it has not been guaranteed that those artifacts as well as other earlier blades are actual blades or elongate flakes without accompanying blade core.
-Except for Hajinri, blade cores first appeared in Suyanggae-C at 34.3 ka.
+The earliest blades and blade cores are also recorded from Hajinri but the date of this site is an unusual outlier because the stemmed points and blades are highly standardized and refined, which are only found at other sites much later in time.
+Bonggok site, which is the second-earliest site for stemmed points in our sample, also includes blades but it has not been demonstrated that those artifacts, as well as other earlier blades, are actual blades or elongate flakes without accompanying blade core.
+Aside from the outlier site Hajinri, blade cores first appeared in Suyanggae-C at 34.3 ka.
 Other artifacts, including burins, denticulates, end scrapers, handaxes, knives, notches, planes, and polyhedrals made up a small proportion of the assemblages and show no directional chronological change.
@@
 They are very suitable to make elongate blades and stemmed points because of the predictability of flaking afforded by their fine-grained texture.
-For example, Hajinri has the largest number of stemmed points (n=86) among the assemblages studied here, and most stemmed points from Hajinri are made from shales.
+For example, Hajinri has the largest number of stemmed points (n=86) among the assemblages studied here, and most stemmed points from Hajinri are made from shale.
 Other raw materials, such as porphyry and sandstone, were found in just a few assemblages in small proportions.
@@
 ```{r raw-materials, fig.cap="The composition of raw material type in each assemblage.
-We excluded raw materials that are included in less than 5 assemblages including crystal, basalt, iron_ore, slate, limestone, granite, gneiss, tuff, amphibolite, andesite, grit, metamorphic-rock, quartzitic-gneiss, siliceous-schist, and unidentified ones.
-Hornfels are shown in 4 assemblages but since it occupies more than half of the Yongsandong and Yongband assemblages, we included it in the plot.
+We excluded raw materials that are included in less than 5 assemblages including crystal, basalt, ironore, slate, limestone, granite, gneiss, tuff, amphibolite, andesite, grit, metamorphic-rock, quartzitic-gneiss, siliceous-schist, and unidentified ones.
+Hornfels are shown in four assemblages but since it occupies more than half of the Yongsandong and Yongband assemblages, we included it in the plot.
 The color represents different types of raw materials, and the assemblages are placed in chronological order.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/paper/005-raw-materials.R"))}
@@
 
-```{r site-elevation, fig.cap="The site elevation by the age of the deposit.
+```{r site-elevation, fig.cap="Site elevation by the age of the deposit.
 Colour indicates presence/absence of stemmed points.
@@
 
-Figure \@ref(fig:site-elevation) shows that people occupied sites located at similar elevations, around 0 to 200 meters above sea level.
+Figure \@ref(fig:site-elevation) shows that people occupied sites located at a wide variety of elevations from around 0 to 200 meters above sea level.
+Elevation correlates with the ecological diversity of plants and animals in Korea [@sohn2019effect; @park2020temporal] and thus elevation is a suitable proxy for the environmental context of human behavior.
 As we saw for artifact density, there is an increase in the upper bound of site elevation in more recent times.
@@
 Sites containing stemmed points reach a maximum elevation of 175 m at 28 ka.
-We acknowledged that site elevation might not be the best proxy for the environmental context due to mountainous geography of Korea and possible disturbance during the site formation processes.
-However recent studies show that elevation makes huge impact on the diversity and ecology of plants and animals in Korea [@sohn2019effect; @park2020temporal] and thus enable us to explore the elevation as a environmental proxy.
+While the elevation difference is well within the total range of elevation throughout the Korean Peninsula (which exceeds 500 m in the more mountainous eastern regions), the slightly higher elevation sites occupied by stemmed point users probably provided better views of the valleys in search of large game.
 
@@
 The environmental context of this technological innovation was a gradual decrease in temperature into the LGM.
-And the population increase before the appearance of stemmed points.
+Analysis of radiocarbon ages suggest a population increase before the appearance of stemmed points.
 
@@
 We examined the artifact volumetric density and retouch frequency to understand the difference in mobility strategies and site occupation patterns associated with the introduction of stemmed points.
-The analyses are based on the premise that highly mobile groups carried lightweight toolkits with more retouched and easily replaceable composite tools, and left smaller amounts of artifacts at sites as a result of short-term occupations.
+The analyses are based on the premise, established on global ethnographic data, that highly mobile groups carried lightweight toolkits with more retouched and easily replaceable composite tools, and left smaller amounts of artifacts at sites as a result of short-term occupations.
 These curated technologies reflect an investment of time and effort in stone artifact manufacturing and maintenance to minimize travel costs of carrying the tools for high mobility or special purpose foray groups.
@@
 These expedient technologies are more typical for groups with greater residential stability who can make, use, abandon, and remake their tools frequently in the same location [@binford1979organization; @torrence1989time; @kuhn1994formal; @meignen2006middle; @kuhn2014mousterian; @vaquero2018searching; @clark2019landscapes].
-As mentioned above, the sites we addressed here are all open-air localities with limited material remains to infer the clear view of subsistence strategies but our data still supports the general concept of the curated and expedient technology by showing a strong pattern of artifacts from less dense assemblages having higher retouch frequencies, while more dense assemblages have lower retouch frequencies.
+As mentioned above, the sites we addressed here are all open-air localities with limited material remains to infer the clear view of subsistence strategies, nevertheless, our data are highly consistent with these concepts of curated and expedient technologies, showing a strong pattern of artifacts from less dense assemblages having higher retouch frequencies, and more dense assemblages having lower retouch frequencies.
+In general, the new technology of stemmed points was more associated with expedient technological strategies.
 Our results show that foragers with stemmed points tended to stay in the same site for longer periods with higher artifact densities and fewer retouched pieces, compared to assemblages without stemmed points.
-In general, the new technology of stemmed points was more associated with expedient technological strategies.
 
 If stemmed points were primarily a hunting tool, as prior work has assumed, we might expect foragers to have been using stemmed points in highly mobile groups and stemmed points mostly found in small assemblages at task-specific sites [@chang2013human; @lee2019were].
-However, our results show the opposite, namely that this new technology was more often associated with expedient technological strategies.
+However, our results show the opposite, namely that this new technology was more often associated with expedient technological strategies, implying that the tools may have been multipurpose.
 Furthermore, our results show that different sites represent different occupation patterns during the time of technological transition, making it difficult to characterize this period with a single land use strategy.
@@
 Both Seong and Lee mention that core tools made of quartz or quartz vein never disappeared and they point out the appearance of stemmed points and blade technology as an addition, rather than replacement.
-Our results, covering 49-24 ka and corresponding to the first three types of Seong's divisions, support the continuation of core tools abundantly throughout the Late Pleistocene.
+Our results, covering 49-24 ka and corresponding to the first three types of Seong's divisions, show that the divisions are subtle, with the continuation of core tools abundantly throughout the Late Pleistocene.
 In addition, our data show that side scrapers remained as major tools while other tools occupied only a small portion of the assemblages.
@@
 Combining the stone projectile with a wooden shaft further materials enhances functionality of the tool by improving penetration by increasing weight and sharpness [@browne1940projectile].
-We expected that the functional advantage of the new composite tool would have allowed foragers to be more effective in less productive landscapes.
+We predicted that the functional advantage of the new composite tool would have allowed foragers to be more effective in less productive or more marginal landscapes.
 To explore the environmental and demographic contexts related to the technological transition, we examined the distribution of site elevations, and simulated MAT during the Late Pleistocene.
 The results show that groups with stemmed points generally reached higher elevations, where the MAT was generally lower.
-This supports our hypothesis that stemmed points were associated with the occupation of more harsh habitats.
-Through MIS 3 the temperature gradually decreased until the LGM, suggesting that stemmed points may have been part of a suite of adaptations to cooler temperatures.
+This supports our hypothesis that stemmed points were associated with the occupation of more challenging habitats.
+Furthermore, through MIS 3 the temperature gradually decreased until the LGM, suggesting that stemmed points may have been part of a suite of adaptations to cooler temperatures at any elevation.
 We see a similar pattern of new technologies enabling expansion into marginal areas at a later time in north Paleo-Honshu Island, Japan, where the appearance of blade and projectile points at about 25 ka were associated with foragers moving into cold grassland areas [@Morisaki_Izuho_Terry_Sato_2015].
@@
 Bae claims that stemmed points are the result of the continuous influx of modern humans from a Siberian migration [@bae2010origin].
-Our results are partly consistent with this scenario, showing a gradual increase in population before the appearance of stemmed points.
+Our results are partly consistent with this scenario, showing a gradual increase in population before the appearance of stemmed points, however with the evidence currently available we cannot discern if this is endogenous population growth or external migration.
 
@@
 The temperature gradually decreased during the technological transition and after the appearance of stemmed points.
-Overall, the result indicates that the use of stemmed points was associated with occupation of marginal or extreme environments.
+Overall, the result indicates that the use of stemmed points was associated with occupation of marginal environments.
 In addition, the groups with stemmed points had expedient technologies which reflect residential and less mobile land use patterns.
@@
 The absence of detailed stratigraphic data limit the chronological resolution for the assemblages in our sample.
-Furthermore, we currently lack functional data on individual artifact assemblages to support our conclusions about site use.
-With future work we may be better able to distinguish a high density of stone artifacts as results from a long-term/small group occupation versus a short-term/large group occupation.
-Further research should focus on detailed assemblage analysis at each site to better understand foragers mobility and site use strategies.
+Furthermore, we currently lack functional data on individual artifact assemblages to support our conclusions about site use to evaluate whether stemmed points were primarily hunting projectiles, or multipurpose, as we have proposed here.
+With future work we may be better able to distinguish a high density of stone artifacts resulting from a long-term/small group occupation versus a short-term/large group occupation.
+We acknowledge that our coarse resolution of environmental data is not fully satisfying to patternize the subsistence strategies of forager groups.
+Therefore we aim to increase the spatial resolution of the data from the sites in our future work.
+Further research should also aim to identify the mechanisms behind the population increase that we have reported here, was it local population growth, or a result of new groups arriving from outside of Korea?
 

### 2021-10-12 analysis/paper/008-summed-probability-distributions.R

@@ -4,7 +4,7 @@ library(ggrepel)
 library(here)
 
 dates <-
-  read.csv(here("analysis/data/raw_data/Radiocarbon dates.csv"))
+  read.csv(here("analysis/data/raw_data/Radiocarbon dates_revision.csv"))
 
 
 dates_clean <-

### 2021-12-09 analysis/paper/paper.Rmd
added affliation and edit 7D to 9D

@@
 author:
-  - Gayoung Park
-  - Ben Marwick
+  - Gayoung Park (University of Washington)
+  - Ben Marwick (University of Washington)
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 The first appearance of stemmed points occurred in the middle of the decreasing MAT trend in MIS 3, at 40-35 ka.
-Figure 7D shows a negative relationship between temperature and elevation.
+Figure 9D shows a negative relationship between temperature and elevation.
 For example, Gihwari Cave site, with one of the lowest MAT distributions, is located in the highest elevation around 300m above sea level.

### 2021-12-09 analysis/paper/paper.Rmd
Deleted the word "supplementary"

@@
 
-The entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is included in the Supplementary Online Materials at https://osf.io/jxwy5/ to enable re-use of materials and improve reproducibility and transparency [@Marwick2017].
+The entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is shared through open repository at https://osf.io/jxwy5/ to enable re-use of materials and improve reproducibility and transparency [@Marwick2017].
 Also in this version-controlled compendium [@marwick2018packaging] are the raw data for all the visualizations and tests reported here.

### 2021-12-09 analysis/paper/paper.Rmd
add affiliation, DOI to SOM, start using renv

--- 2021-10-12 analysis/paper/paper.Rmd
@@
 author:
-  - Gayoung Park
-  - Ben Marwick
+  - Gayoung Park:
+      email: gayoungp@uw.edu
+      institute: [UofW]
+      correspondence: true
+  - Ben Marwick:
+      email: bmarwick@uw.edu
+      institute: [UofW]
+      correspondence: false
+institute:
+  - UofW: Department of Anthropology, University of Washington, Seattle, USA
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
       reference_docx: "../templates/template.docx" # Insert path for the DOCX file
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
 bibliography: references.bib
@@
 
-The entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is included in the Supplementary Online Materials at https://osf.io/jxwy5/ to enable re-use of materials and improve reproducibility and transparency [@Marwick2017].
+The entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is included in the Supplementary Online Materials at https://doi.org/10.17605/OSF.IO/JXWY5 to enable re-use of materials and improve reproducibility and transparency [@Marwick2017].
 Also in this version-controlled compendium [@marwick2018packaging] are the raw data for all the visualizations and tests reported here.

### 2021-12-09 analysis/paper/paper.Rmd
minor edits to respond to QSR manuscript checking

@@
 
-The entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is included in the Supplementary Online Materials at https://doi.org/10.17605/OSF.IO/JXWY5 to enable re-use of materials and improve reproducibility and transparency [@Marwick2017].
+The entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is openly available at https://doi.org/10.17605/OSF.IO/JXWY5 to enable re-use of materials and improve reproducibility and transparency [@Marwick2017].
 Also in this version-controlled compendium [@marwick2018packaging] are the raw data for all the visualizations and tests reported here.
@@
 The first appearance of stemmed points occurred in the middle of the decreasing MAT trend in MIS 3, at 40-35 ka.
-Figure 7D shows a negative relationship between temperature and elevation.
+Figure \@ref(fig:climate-model)D shows a negative relationship between temperature and elevation.
 For example, Gihwari Cave site, with one of the lowest MAT distributions, is located in the highest elevation around 300m above sea level.

### 2023-05-09 analysis/paper/paper.Rmd
updated text

@@
 42-35 ka) in Korea is often assumed to have transformed forager lifeways because it was thought to have indicated more specialized hunting practices [@seong2008tanged; @chang2013human].
-Stemmed points are projectile points made from an elongated flake or blade with slight retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft.
+Stemmed points are projectile points made from an elongated flake or blade with slight retouch on the distal end to shape an acute tip and on the proximal end to make a stem, which connects to a shaft.
 These types of artifacts are often called tanged points in the western hemisphere [@serwatka20162d; @ivanovaite2020all], but we use ‘stemmed point’ here to distinguish from Bronze Age stone projectile points that have long been called ‘tanged points’ in Korea.
